### 2015-03-10 oil_pres_new.tex
moved directory from google drive

%oil_pres_new.tex

\documentclass[12pt]{article}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{natbib} %for citet and citep
\usepackage{syntonly}
\usepackage{esdiff} %for writing partial derivatives
\usepackage{url} %for inserting urls
\usepackage{placeins} % for limitting floats
\usepackage{comment} %for excluding figures
% \excludecomment{figure} % for printing without figures
% \syntaxonly % for quickly checking document
%set document settings

\doublespacing % from package setspacs

% table font size
\let\oldtabular\tabular
\renewcommand{\tabular}{\scriptsize\oldtabular}

\title{The Effect of Oil Price on Field Production: Evidence from the Norwegian Continental Shelf}
\author{Johannes Mauritzen\\
		Department of Business and Management Science\\
        NHH Norwegian School of Economics\\
        Helleveien 30, 5045\\
        Bergen, Norway\\
        johannes.mauritzen@nhh.no\\
        \url{jmaurit.github.io}\\
		}
\date{\today}

\begin{document}
 \begin{spacing}{1} %sets spacing to single for title page
	\maketitle

\begin{abstract}
I use detailed field-level data on Norwegian off-shore oil field production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production.  I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 2 to 4\% for a 10 dollar per barrel increase in the real price of oil.  Most of this effect appears to come in the planning phase of a field's development.\\
Keywords: Oil production,Oil Prices, Norway, Semiparametric.
\end{abstract}

\thanks{I would like to thank Klaus Mohn, Jonas Andersson, Sturla Kvamsdal, Harrison Fell, R\"ognvaldur Hannesson and Henrik Horn for valuable discussion and comments.}
% JEL Codes: Q4, L71
 \end{spacing}

\section{Introduction}

For most of the last century, crude oil has been the world’s single most important and valuable fuel source.\footnote{Oil is the world's largest single energy source, consisting of approximately 33\% of total energy consumption in 2013 as well as the most valuable in terms of market price per unit of energy \citep{british_petroleum_statistical_2013}} Naturally, questions of how the oil price affects the world economy as well as how oil production reacts to the oil price have been fundamental topics in economics.

However a significant gap exists in the literature.  While numerous studies have taken up the issue of how the price of oil affects searching for new fields as well as total oil production at both the regional and global level, few studies exist of the effects of oil prices on oil production at the field level.  The studies that do use field- and well-level data, such as \citet{rao_taxation_2010}, often use data from on-shore installations.  However an increasing amount of the world's oil production comes from hard-to-reach off-shore fields.  Production from challenging off-shore environments is substantially different in character than from on-shore installations.

Moreover, studies within the economics literature fail to take into account how oil prices have varying effects on the different stages of field development and production.  The planning, build-out and depletion phase of a field are conceptually unique and how oil companies operating off-shore respond to oil prices can be expected to vary substantially.

The effect of oil prices on producing fields is an important topic in understanding the mechanisms of how total oil supply reacts in response to price. The response of oil production in existing fields is especially important now as many of the major oil- producing areas, like the Norwegian Continental Shelf are mature and an increasing share of investments will be directed at capacity enhancement in producing fields rather than new field developments. How production from these fields will respond to changes in oil prices has major implications for the oil industry, the state finances of oil-producing countries and long-run oil price formation.

The question of the effect of oil prices on production and the more general question of optimal oil extraction has spawned a large theoretical literature dating back to the seminal work of \citet{hotelling_economics_1931}. \citet{krautkraemer_nonrenewable_1998} provides a good overview. At a basic level a central idea of much of this theoretical work is that with a non-renewable resource, production is a decision that involves a significant opportunity cost: more production in the current period means less production in future periods.  Within this frameworks, prices and expectations of prices become important variables in the production decision. A simple Hotelling model suggests that a producer would immediately change their production in response to a change in oil price in order to dynamically optimize the total extraction.

But in practice, the question is not as clear cut.  Production in the Norwegian Continental Shelf - as well as most other offshore production areas - has extremely high fixed and operating costs.  Keeping spare capacity available in order to adjust to changing oil prices might simply be too expensive.  Instead, producers may find it more beneficial to use storage and financial instruments in order to hedge short-term price movements. Higher oil prices may however still lead a producer to invest in increased capacity.  Since lag times are significant in the off-shore sector, we would then expect to see a multi-year lagged effect of prices on production.

However even with the question of lagged production and investment, some ambiguity exists.  \citet{mohn_efforts_2008} suggests and finds evidence for the idea that in periods of high oil prices off-shore producers will invest more in risky wild-cat drilling in search of new fields, but concentrate investments in lower-risk ventures, like expanding production in existing fields, when prices are low.  If this effect were to dominate, then it may even be plausible that production in existing fields reacts \emph{negatively} to increases in price.

%Implications for theories and explanations for understanding of the price mechanism and %the role of speculation.  Hamilton block quote.

For such a prominent subject, the lack of research on the role of price in oil field production is striking. Two main factors likely contribute to the limited literature - the availability of data and the non-linear time profile of field production.  Large private oil companies, notably the “super majors” and state-owned oil companies have historically accounted for the vast majority of oil production and reserves. \footnote{See for example the economist article titled Supermajordämmerung from August 3rd, 2013: \url{http://www.economist.com/news/briefing/21582522-day-huge-integrated-international-oil-company-drawing}} These entities tend to consider field-level data as either company or state secrets.

However, over the last few years, a movement towards making the petroleum and other extractive industries more transparent has taken form.  The Norwegian government has been on the forefront of this movement and increasingly committed itself to transparency in the petroleum sector. \footnote{see http://www.regjeringen.no/en/sub/eiti---extractive-industries-tranparency/about-eiti.html?id=633586} Over the last several years, detailed data on most aspects of the country\’s oil industry has become openly available.

In this article, I use historical production data from all 77 currently or formerly oil-producing fields on the Norwegian continental shelf in order to estimate the effect that prices have on oil production.

By looking only at the effect of price on fields that currently or previously have produced oil I am limiting the scope of this article.  The effect of oil prices on total production over an extended period of time is due not just to reactions in production in existing fields but also increased searching for new fields.  In fact, an implication of this work is that much of the total production response from higher oil prices is likely from increased searching as well as production from previously un-economic fields.

The main finding in this article is that oil production at the field level has no significant reaction to concurrent changes in the oil price, where concurrent is broadly defined as within the first three years.  A slight effect can be detected at a lag of between 4 and 8 years, with a magnitude of about a 2 to 4\% increase in yearly production for a 10 dollar increase in the price of oil.  This effect is somewhat greater and with less of a lag in large fields compared to small fields.  Price appears to have the most significant affect during the planning stage - before production begins in a field.  In the depletion phase of production, price is found to have little to no significant effect.

The main methodological problem is the non-linear production profile of oil fields.  Once full-scale extraction is started in an oil field, pressure in wells will start declining. Compensating measures such as gas and water injection have been successful, but their impact is temporary. In turn production rates will therefore inevitably decline.

More so, oil field production is correlated across fields - that is, increases and decreases in production in fields are not randomly distributed across time.  Instead, as figure \ref{top10_production} shows with the production profile of the 10 largest Norwegian oil fields, production tends to be correlated across fields.  The result is a total production curve that is bell-shaped over time as in figure \ref{oil_decline}.  Since the oil price series is autocorrelated and non-stationary, failure to properly account for the production profile will lead to spurious estimation of the price terms in a regression.

The direction of this bias can be gleaned in figure \ref{oil_decline}.  High oil prices were present at periods of relatively low production in the late 1970s and early 1980s as well as over the last 10 years, however real prices reached some of their lowest levels at the same time as the top of production around the year 2000. These oil price dynamics almost certainly affected investment decisions in the industry and in turn total production (\citep{osmundsen_is_2007}, \citep{aune_financial_2010}). However the inverse contemporaneous relationship at the field level is entirely coincidental, but will heavily bias the estimation of the effect of price on production if field production profiles are not properly accounted for.

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/top10_production_print.png}
	\caption{The production profile of the 10 largest oil fields on the Norwegian Continental Shelf.  Production tends to be correlated across fields}
	\label{top10_production}	
	\end{figure}

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/oil_decline_print.png}
	\caption{The production profile for the entire Norwegian Continental Shelf is bell-shaped, reflecting the correlated production profiles of the fields.  With oil prices that are autocorrelated, }
	\label{oil_decline}
\end{figure}

As a solution I use a semi-parametric model within the Generalized Additive Model frameworks of \cite{hastie_generalized_1990}.  Here I use a two-dimensional smoothed spline function to account for the general non-parametric shape of the production profile while allowing price to enter the equation linearly.  The coefficient of price can then be interpreted as the average effect of price on production over the entire production profile.

\FloatBarrier
\section{The effect of oil price on production: theory, simulations and empirics}

For modeling aggregate oil production, shape-fitting models, notably \citet{hubbert_energy_1962} and more recently advocated by \citet{deffeyes_hubberts_2001}, have had some success in estimating the timing of peak production at the regional and national level, but they tend to seriously underestimate the total recoverable resource of oil-producing regions and the models can be shown to be fundamentally misspecified \citep{boyce_prediction_2013}.   Simulation type studies where aggregate oil production is modeled through an often complex combination of physical and economic processes also exist in the geo-engineering literature, but their usefulness tends to be weighed down by their complexity as they require quite detailed data and specific assumptions about functional form that can be difficult to justify \citet{brandt_review_2010}.

More important to this paper are empirical estimates of the effect of price on production. Several econometric papers seek to answer the question of how aggregated oil supply is affected by oil prices.  \citet{farzin_impact_2001} attempts to estimate an elasticity for the effect on added reserves of increased oil prices and finds a small though statistically significant effect.  \citet{ramcharran_oil_2002} estimates a supply function for the total supply of oil from several OPEC countries based on data from 1973 to 1997.  The author finds a negative price elasticity for several of the countries, and interprets this as evidence of producers targeting revenue.  However since the author does not take into account the production profile of oil fields and the spurious correlation that can arise with autocorrelated prices, these estimates come under considerable doubt.

the effect of oil-price uncertainty on drilling and exploration has also been explored.  \citet{kellogg_effect_2014} finds that oil exploration firms in Texas do approximately respond as real-option theory would predict when it comes to the timing of drilling.  A model and test using data from North Sea producers on the UK continental shelf by \citet{hurn_geology_1994}, on the other hand, fails to find evidence that the variance in the oil price affects the timing of oil field development.  I do not attempt to directly model uncertainty, however given that the investments needed to increase oil production in an existing field are to a certain extent irreversible and that oil prices are highly volatile, the results can and probably should be interpreted with the real options framework in mind.

Only a few papers utilize field-level data.  \citet{black_is_1998} tests the relevance of “nesting” a structural empirical model of profit maximization that takes into account oil prices into a typical geo-engineering model of oil field production.  They find strong evidence that taking into account profit maximization, and implicitly price, substantially improves the fit compared to a purely geo-engineering type model.   The limitation of their methodology is that they are only able to test whether including economic factors like price affects the fit of the model but are not able to give an estimate of the effect.  Methodologically, the paper also relies heavily on assumptions about the functional form of both the geo-engineering aspects of the oil producer as well as their profit-maximization.  By taking a more flexible, semi-parametric approach to estimating the effects of oil field production profile, this paper avoids problems with overly restrictive assumptions.

\citet{rao_taxation_2010} uses data on land based oil wells in California to estimate the effects of tax changes and price controls.  The author finds that short-term tax changes caused small, but significant retiming of production from oil wells.  These findings are over-all consistent with the findings of this paper, though the author finds concurrent effects of taxation on production while I do not find any concurrent effect of prices on production.  This difference is most likely due to the significant differences in cost and complexity between operating offshore and onshore.  How producers react to a short-term tax change as opposed to a change in price is likely also different.  More so, the author does not consider differences between the different stages of production.

Studies using detailed Norwegian data on offshore activity also exist, though the focus has mainly been on exploration and drilling.  \citet{mohn_exploration_2008} finds that long-term changes in the oil-price has a strong effect on exploratory drilling though little effect is measured from short-term changes in the oil price.  \citet{osmundsen_exploration_2010} analyses drilling productivity over time on the Norwegian Continental Shelf while \citet{mohn_efforts_2008} finds that higher oil prices leads to higher reserves and as well as that oil prices affect producer risk-preferences - with higher prices leading to lower success rates but larger discovery size.

\section{Oil production on the Norwegian Continental Shelf}

The first commercial oil well in Norwegian continental waters was discovered in December of 1969 in what became the Ekofisk oil field, the largest Norwegian oil field by estimated recoverable reserves.  As figure \ref{north_sea_reserves} shows, most of the largest fields in the North Sea were found relatively early on while more recent finds have tended to be smaller - a pattern typical of oil producing areas called creaming.  A major exception to this trend was the recent find of the Johan Sverdrup field which is estimated to have approximately 300 million Standard Cubic Meters (SM3) of recoverable oil. \footnote{The Johan Sverdrup field is estimated to begin producing oil in 2019 and so is not present in the data set used for the analysis.}

\begin{figure}
\includegraphics[width=1.2\textwidth]{figures/north_sea_reserves_print.png}
\caption{Oil fields in the Norwegian territorial North Sea.  The largest oil fields tended to be discovered earliest, while newer finds tend to be smaller.  An exception is the large Johan Sverdrup field, which is expected to begin producing in 2019.}
\label{north_sea_reserves}
\end{figure}

Exploration in the Norwegian Sea was opened in the early 1980’s and the first commercial field started production in 1981.  While several mid-sized fields have been discovered, the Norwegian Sea has generally disappointed in terms of commercial oil finds and most finds have been relatively small (\ref{norwegian_sea_reserves}).

\begin{figure}
\includegraphics[width=1.2\textwidth]{figures/norwegian_sea_reserves_print.png}
\caption{Oil fields in the Norwegian Sea.  Production from the Norwegian Sea has generally been dissapointing compared with expectations when the area was opened to exploration in the early 1980's.}
\label{norwegian_sea_reserves}
\end{figure}

Norwegian waters in the Barents Sea have also been open to exploration since the 1980s, however up until recently only a few, small finds were made and none came into commercial production.  However recently several significant oil and gas finds have been made in the Barents Sea - notably the oil field Goliat and the large gas find Sn\o hvit, which are both currently under development but not yet producing.  The agreement between Norway and Russia in June 2011 settling a long-running dispute over the maritime delimitation has also given a boost to new exploration in the region.

Profits from oil and gas production in Norway are subject to a resource tax of 50 \% on top of the ordinary income tax of 28 \%, thus income from petroleum production is taxed at a total marginal tax rate of 78 \%.  The central government also receives revenues through ownership stakes in companies, notably Statoil, where the state is the majority stakeholder.  The over-all tax rate has been fairly constant through the history of Norwegian oil production, however several important changes related to the tax code have occurred.  In 1991 a C02 tax was introduced, and in the year 2012 the tax was doubled.  But this was mainly levied on the use and import of petroleum products.  In the offshore sector it was levied on the burning of oil and gas and thus the main effect was on the practice of flaring natural gas that could not be transported and sold commercially.

More importantly to the offshore sector were accounting changes that were implemented in 2002 and 2005 that were meant to encourage new entrants by allowing losses to be carried forward for tax purposes and by introducing a rebate on the tax-value of losses associated with searching and drilling.  In general though, these rules mainly affected searching and discoveries of new fields rather than production from existing fields and I do not control for the tax changes in my model.  More information on taxation and revenues from the offshore sector can be found at the website of the Norwegian Ministry of Finance. \footnote{\url{http://www.regjeringen.no/en/dep/fin/Selected-topics/taxes-and-duties/bedriftsbeskatning/Taxation-of-petroleum-activities.html?id=417318}}

Rights to explore and eventually produce on the Norwegian Continental Shelf are based on a system where the government announces geographic blocks that will be opened to oil exploration and production subject to production licenses.  Production licenses are initially granted for between 4-6 years subject to requirements that firms are actively searching in their awarded blocks.  If oil or gas deposits are proven then the production license can be extended for up to 30 years.  In general, the frameworks are fair, predictable and stable for companies who find commercially extractable oil deposits and regulatory interference is unlikely to be the cause of any observed changes in production from existing oil fields.  For more information see the website of the Norwegian Petroleum Directorate. \footnote{\url{http://www.npd.no/en/Topics/Production-licences/Theme-articles/Production-licence--licence-to-explore-discover-and-produce-/}}

\section{Norwegian oil field production data}
Production data of Norwegian oil fields is obtained from the website of the Norwegian Petroleum Directorate. \footnote{http://factpages.npd.no/}  Production data is available at a monthly frequency for all fields, though I choose to aggregate up to yearly values both to smooth over seasonality as well as short-term volatility of output due to factors such as weather or technical issues that are not relevant for this article.

In addition to data on field-level production, I also make use of data on estimated total recoverable reserves. The use of this variable is complicated as it is an estimate subject to a large amount of uncertainty, especially in relatively young fields.  However, the methodologies used to estimate the total recoverable resource of a field are constantly evolving and it is a fair assumption that any consistent bias of the estimates are observed in older fields and corrected for in estimates for newer fields.  I can then assume that existing errors are random and will not significantly bias the estimates.

Moreover, the estimate is likely endogenous, in the sense that it is also effected by prices.  However since I use the variable as a control variable and not for the purpose of estimating a parameter with a causal interpretation, this should not significantly affect the validity of the results.

I use yearly data from the US Energy Information Agency on the real price of Brent-traded oil in 2010 dollars.The Brent benchmark oil price is likely the best oil price measure for Norwegian production as it is based on light sweet crude oil sourced from the North Sea.

An argument can be made that expectations of future oil prices can be equally if not even more important for production decisions as the current oil price.  Forecasts for future oil prices are available from, among others, the International Energy Agency, but these have tended to be notoriously inaccurate and it is unlikely oil companies use these projections for their investment decisions.

On the other hand, given the size and liquidity of oil spot markets, it is a fair assumption that the current oil prices do a good job of incorporating much of the available information about crude oil markets and that future price movements are generally difficult to predict \citep{hamilton_understanding_2008}.  An active futures market does exist, but several studies have found that current oil prices are in general better than prices on futures contracts at predicting future oil prices \citep{alquist_what_2010, chinn_predictive_2005}.  \citet{mohn_investment_2008} as well as \citet{pesaran_econometric_1990} and \citet{farzin_impact_2001} find evidence for adaptive expectations, where expectations of future prices is based on a weighted average of current and past prices.  I take account of this by including several years of price lags in my regression equations.

A cleaned data set and the full code for the analysis are available upon request. I use the R statistical programming package for all the analysis in this article \citep{r_core_team_r:_2013}.  I use the R packages ggplot2 and ggmap for plotting \citep{wickham_ggplot2:_2009, kahle_ggmap:_2013}, plyr for data manipulation and cleaning \citep{wickham_split-apply-combine_2011}, texreg for table formatting \citep{leifeld_texreg:_2013} and mgcv for implementation of the Generalized Additive Models \citep{wood_fast_2011}.

### 2015-03-10 oil_prices_new_pres.tex
moved directory from google drive

% oil_mc_pres.tex
%Last updated February 2015

\documentclass{beamer}
\usepackage{url} 
\usepackage{amsmath}
\usepackage{esdiff} %for writing partial derivatives

\usetheme{default}

\title{Estimating the Effect of Price on Oil Production: Evidence from the Norwegian Continental Shelf}
\author{Johannes Mauritzen}
\institute % (optional)
{
 Department of Business and Management Science\\
 NHH Norwegian School of Economics
}
%\logo{\includegraphics[height=1.5cm]{/Users/johannesmauritzen/Google Drive/Logos/NHH.jpg}}

\date{March 5th, 2015}

\begin{document}

%--- the titlepage frame -------------------------%
\begin{frame}[plain]
  \titlepage
\end{frame}

\begin{frame}
	\begin{figure}
		\includegraphics[width=1\textwidth]{figures/oil_decline_print.png}
		% \caption{The production profile for the entire Norwegian Continental Shelf is bell-shaped, reflecting the correlated production profiles of the fields.  With oil prices that are autocorrelated, }
		\label{oil_decline}
	\end{figure}
\end{frame}

\begin{frame}[plain]
	\begin{itemize}
		\item Effect of Price on Drilling / Reserve Replacement
			\begin{itemize} \item Mohn and Osmundsen (2008), Mohn (2008), Ringlund (2008) \end{itemize}
		 \item Aggregate Production

			\begin{itemize}
				\item Curve-fitting/Simulation (geo-engineering)
				\item Econometric

				\begin{itemize}
					\item Kaufman (1990), Kaufman and Cleveland (2001)
					\item Ramcharran (2002)
	 			\end{itemize}		
	 	\end{itemize}
	 	\item Field-level
	 			\begin{itemize}
	 				\item Rao (2012)
	 			\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}[plain]
	Generalized Additive Models
	\begin{itemize}
	\item  Hastie and Tibshirani (1990) 
	\item  Wood (2006)
	\end {itemize}
\end{frame}

\begin{frame}[plain]
	Main Results
	\begin{itemize}
		\item No significant contemporary effect of oil price on field production (within 3 years)
		\item Slight lagged effect found after 4-8 years, magnitude of around 2\%
		\item Most of this effect seems to come in the planning stage of an oil field
		\item Little to no effect - contemporary or lagged - in depleting fields
	\end{itemize}
\end{frame}

\begin{frame}[plain]
	\begin{figure}
	\includegraphics[width=.8\textwidth]{figures/north_sea_reserves.png}	
	\label{north_sea_reserves}
	\end{figure}
\end{frame}

\begin{frame}[plain]
	\begin{figure}
	\includegraphics[width=.8\textwidth]{figures/norwegian_sea_reserves.png}

	\label{norwegian_sea_reserves}
	\end{figure}
\end{frame}

\begin{frame}[plain]

\begin{figure}
	\includegraphics[width=.8\textwidth]{figures/oil_decline.png}

	\label{oil_decline}
\end{figure}

\end{frame}

\begin{frame}[plain]
	\begin{figure}
	\includegraphics[width=.8\textwidth]{figures/top10_production.png}

	\label{top10_production}	
	\end{figure}
\end{frame}

\begin{frame}
	\begin{figure}
		\includegraphics[width=1\textwidth]{figures/statfjord_dem_print.png}
		\label{statfjord_dem}
	\end{figure}
\end{frame}

\begin{frame}[plain]
	\begin{equation}
		\begin{split}
		 Log(Production_{i,t}) & = \alpha_0 + Poly(time\_to\_peak_{i,t}) \\ 
		 & +  Poly(peak\_to\_end_{i,t}) \\
		 & + \gamma total\_recoverable\_oil_i \\
		 & + \beta oil\_price_{t,t-1, t-2,.., t-n} + \epsilon
		\end{split}
	\label{glm_eqn}
	\end{equation}
\end{frame}

\begin{frame}[plain]
	\begin{figure}
		\includegraphics[width=.8\textwidth]{figures/glm_dirty_box.png}
		\label{glm_dirty_box}
	\end{figure}
\end{frame}

\begin{frame}[plain]
	\begin{equation}
	\begin{split}
		Log(Production_{i,t})&=f(time\_to\_peak_{i,t}, total\_recoverable\_oil_i) \\
		& \quad + f(peak\_to\_end_{i,t}, total\_recoverable\_oil_i) \\
		& \quad + \beta oil\_price_{t,t-1, t-2,.., t-n} + \epsilon
	\end{split}
	\label{gam_price_eqn}
	\end{equation}
\end{frame}

\begin{frame}[plain]
	\begin{equation}
	Production_{t}=f(time) + \epsilon
		\label{simp_eqn}
	\end{equation}
\end{frame}

\begin{frame}[plain]
	\begin{figure}
		\includegraphics[width=.8\textwidth]{figures/statfjord_gam.png}

		\label{statfjord_gam}
	\end{figure}
\end{frame}

\begin{frame}[plain]
	\begin{equation}
	\begin{split}
		Log(Production_{i,t})&=f(time\_to\_peak_{i,t}, total\_recoverable\_oil_i) \\
		& \quad + f(peak\_to\_end_{i,t}, total\_recoverable\_oil_i) \\
		& \quad + \beta oil\_price_{t,t-1, t-2,.., t-n} + \epsilon
	\end{split}
	\label{gam_price_eqn}
	\end{equation}
\end{frame}

\begin{frame}[plain]
	Thin Plate (Regression) Splines (Duchon 1977)
	\begin{equation}
	y_i = g(x_1, x_2)
	\end{equation}

	\begin{equation}
	\min \|\boldsymbol{y-f}\|^2 + \lambda J_{md}(f)
	\end{equation}

	\begin{equation}
	J_{22}{f}= \diffp[2]{f}{x_1}^2 + \diffp[2]{f}{{x_1}{x_2}} + \diffp[2]{f}{x_2}^2dx_1 dx_2
	\end{equation}
\end{frame}

\begin{frame}[plain]	
	\begin{figure}
	Build-out Phase
	\includegraphics[width=1\textwidth]{figures/gam_prepeak_pres.png}
	% \caption{The estimated coefficients on the price terms in a model of fields in the build-out phase.  For large fields, positive coefficients are estimated at the 6th, 7th and 8th lag, but the estimates are imprecise and not statistically significant.  Significant positive coefficients are estimated for small fields at the fifth and eighth lag.  In a pooled model, significant positive coefficients are found at the fourth through eight lags.}
	\label{gam_prepeak_print}
	\end{figure}
\end{frame}

\begin{frame}[plain]
	\begin{figure}
	Depletion Phase
	\includegraphics[width=1\textwidth]{figures/gam_postpeak_print.png}
	% \caption{The estimated coefficients on the price terms in a model of oil field production in depleting fields.  No significant coefficients are estimated except for on the 8th lag of small fields.  This effect is not, however, robust to changes in specification.}
	\label{gam_postpeak_pres}
	\end{figure}
\end{frame}

\begin{frame}[plain]
	\begin{figure}
		\includegraphics[width=.8\textwidth]{figures/field_time_to_peak_pres.png}		
		\label{field_time_to_peak}
	\end{figure}
\end{frame}

\begin{frame}
\begin{figure}
	\includegraphics[width=1\textwidth]{figures/oil_price_series.png}
	\label{oil_price_series}	
\end{figure}
\end{frame}

\begin{frame}[plain, fragile]
Monte-Carlo Simulation Sudo-code
	\begin{verbatim}
	>Generate X fields with size
	 from exponential-normal distribution
	>Generate random starting year for each field
	>Generate logistic cumulative production profile
	\end{verbatim}
\end{frame}

\begin{frame}[plain, fragile]
	\begin{verbatim}
		In loop:
			>Create production profiles from derivative of
			logistic function, price component
			 and stochastic component
			>Regress ``fake'' data with GLM and GAM model
			>Store point estimates
	\end{verbatim}
\end{frame}

\begin{frame}
	\begin{equation}
		\begin{split}
		cumProd&=\frac{size}{1+exp(\frac{-prodTime_t}{3})} \\
		log(production) &= f`(time) + beta*log(price) + epsilon
		\end{split}
	\end{equation}
\end{frame}

\begin{frame}[plain, fragile]
	\begin{equation}
		\begin{split}
		prod_t&=poly(prod\_time_t) + field\_size_i \\
		prod_t&=s(prod\_time,field\_size_i) + \beta price_t
		\end{split}
	\end{equation}
\end{frame}

\begin{frame}[plain]
	\begin{figure}
		\includegraphics[width=1\textwidth]{figures/simulated_production.png}
		\caption{Simulated production of 77 oil fields}
		\label{simulated_production}	
	\end{figure}
\end{frame}

\begin{frame}[plain]
	\begin{figure}
		\includegraphics[width=1\textwidth]{figures/lin_model_price_mc.png}
		\caption{Estimated coefficients on price from linear model from Monte Carlo Experiment}
		\label{lin_model_price_mc}	
	\end{figure}
\end{frame}

\begin{frame}[plain]
	\begin{figure}
		\includegraphics[width=1\textwidth]{figures/gam_model_price_mc.png}
		\caption{Estimated coefficients on price from GAM model from Monte Carlo Experiment}
		\label{gam_model_price_mc.png}	
	\end{figure}
\end{frame}

\begin{frame}[plain]
	\begin{figure}
		\includegraphics[width=.8\textwidth]{figures/tot_forecast.png}

		\label{tot_forecast}
	\end{figure}
\end{frame}

\begin{frame}[plain]
\begin{figure}
		\includegraphics[width=.8\textwidth]{figures/NPD_oil_forecast.png}

		\label{NPD_oil_forecast}
	\end{figure}
\end{frame}

\begin{frame}[plain]
	\begin{itemize}
	\item[] Criticisms
		\begin{itemize}
			\item Ignores significant technical changes in the oil industry
		\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}[plain]
	\begin{itemize}
	\item[] Criticisms
		\begin{itemize}
			\item Ignores significant technical changes in the oil industry
			\item Estimate of field size itself is affected by price
		\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}[plain]
	\begin{itemize}
	\item[] Criticisms
		\begin{itemize}
			\item Ignores significant technical changes in the oil industry
			\item Field size itself is affected by price
			\item Time at which production peaks, as measured from the start of production, is likely affected by price as well
		\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}[plain]
	\begin{itemize}
	\item[] Criticisms
		\begin{itemize}
			\item Ignores significant technical changes in the oil industry
			\item Field size itself is affected by price
			\item Time at which production peaks, as measured from the start of production, is likely affected by price as well
			\item Costs in industry are correlated with oil price
		\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}
	\begin{itemize}
		\item[] johannes.mauritzen@nhh.no 
		\item[] jmaurit.github.io\#oil\_prices
	\end{itemize}
\end{frame}

\end{document}

### 2015-03-10 oil_prices_old.tex
moved directory from google drive

\documentclass[12pt]{article}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{natbib} %for citet and citep
\usepackage{syntonly}
\usepackage{esdiff} %for writing partial derivatives
\usepackage{url} %for inserting urls
\usepackage{placeins} % for limitting floats
\usepackage{comment} %for excluding figures
% \excludecomment{figure} % for printing without figures
% \syntaxonly % for quickly checking document
%set document settings

 \doublespacing % from package setspacs

% table font size
\let\oldtabular\tabular
\renewcommand{\tabular}{\scriptsize\oldtabular}

\title{The Effect of Oil Price on Field Production: Evidence from the Norwegian Continental Shelf}
\author{Johannes Mauritzen\\
		Department of Business and Management Science\\
        NHH Norwegian School of Economics\\
        Helleveien 30, 5045\\
        Bergen, Norway\\
        johannes.mauritzen@nhh.no\\
        \url{jmaurit.github.io}\\
		}
\date{\today}

\begin{document}
 \begin{spacing}{1} %sets spacing to single for title page
	\maketitle

\begin{abstract}
I use detailed field-level data on Norwegian off-shore oil field production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production.  I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 2 to 4\% for a 10 dollar per barrel increase in the real price of oil.  Most of this effect appears to come in the planning phase of a field's development.
\end{abstract}

\thanks{I would like to thank Klaus Mohn, Jonas Andersson, Sturla Kvamsdal, Harrison Fell, R\"ognvaldur Hannesson and Henrik Horn for valuable discussion and comments.}
% JEL Codes: Q4, L71
 \end{spacing}

\section{Introduction}

For most of the last century, crude oil has been the world’s single most important and valuable fuel source.\footnote{Oil is the world's largest single energy source, consisting of approximately 33\% of total energy consumption in 2013 as well as the most valuable in terms of market price per unit of energy \citep{british_petroleum_statistical_2013}} Naturally, questions of how the oil price affects the world economy as well as how oil production reacts to the oil price have been fundamental topics in economics.

However a significant gap exists in the literature.  While numerous studies have taken up the issue of how the price of oil affects searching for new fields as well as total oil production at both the regional and global level, few studies exist of the effects of oil prices on oil production at the field level.  The studies that do use field- and well-level data, such as \citet{rao_taxation_2010}, often use data from on-shore installations.  However an increasing amount of the world's oil production comes from hard-to-reach off-shore fields.  Production from challenging off-shore environments is substantially different in character than from on-shore installations.

Moreover, studies within the economics literature fail to take into account how oil prices have varying effects on the different stages of field development and production.  The planning, build-out and depletion phase of a field are conceptually unique and how oil companies operating off-shore respond to oil prices can be expected to vary substantially.

The effect of oil prices on producing fields is an important topic in understanding the mechanisms of how total oil supply reacts in response to price. The response of oil production in existing fields is especially important now as many of the major oil- producing areas, like the Norwegian Continental Shelf are mature and an increasing share of investments will be directed at capacity enhancement in producing fields rather than new field developments. How production from these fields will respond to changes in oil prices has major implications for the oil industry, the state finances of oil-producing countries and long-run oil price formation.

The question of the effect of oil prices on production and the more general question of optimal oil extraction has spawned a large theoretical literature dating back to the seminal work of \citet{hotelling_economics_1931}. \citet{krautkraemer_nonrenewable_1998} provides a good overview. At a basic level a central idea of much of this theoretical work is that with a non-renewable resource, production is a decision that involves a significant opportunity cost: more production in the current period means less production in future periods.  Within this frameworks, prices and expectations of prices become important variables in the production decision. A simple Hotelling model suggests that a producer would immediately change their production in response to a change in oil price in order to dynamically optimize the total extraction.

But in practice, the question is not as clear cut.  Production in the Norwegian Continental Shelf - as well as most other offshore production areas - has extremely high fixed and operating costs.  Keeping spare capacity available in order to adjust to changing oil prices might simply be too expensive.  Instead, producers may find it more beneficial to use storage and financial instruments in order to hedge short-term price movements. Higher oil prices may however still lead a producer to invest in increased capacity.  Since lag times are significant in the off-shore sector, we would then expect to see a multi-year lagged effect of prices on production.

However even with the question of lagged production and investment, some ambiguity exists.  \citet{mohn_efforts_2008} suggests and finds evidence for the idea that in periods of high oil prices off-shore producers will invest more in risky wild-cat drilling in search of new fields, but concentrate investments in lower-risk ventures, like expanding production in existing fields, when prices are low.  If this effect were to dominate, then it may even be plausible that production in existing fields reacts \emph{negatively} to increases in price.

%Implications for theories and explanations for understanding of the price mechanism and %the role of speculation.  Hamilton block quote.

For such a prominent subject, the lack of research on the role of price in oil field production is striking. Two main factors likely contribute to the limited literature - the availability of data and the non-linear time profile of field production.  Large private oil companies, notably the “super majors” and state-owned oil companies have historically accounted for the vast majority of oil production and reserves. \footnote{See for example the economist article titled Supermajordämmerung from August 3rd, 2013: \url{http://www.economist.com/news/briefing/21582522-day-huge-integrated-international-oil-company-drawing}} These entities tend to consider field-level data as either company or state secrets.

However, over the last few years, a movement towards making the petroleum and other extractive industries more transparent has taken form.  The Norwegian government has been on the forefront of this movement and increasingly committed itself to transparency in the petroleum sector. \footnote{see http://www.regjeringen.no/en/sub/eiti---extractive-industries-tranparency/about-eiti.html?id=633586} Over the last several years, detailed data on most aspects of the country\’s oil industry has become openly available.

In this article, I use historical production data from all 77 currently or formerly oil-producing fields on the Norwegian continental shelf in order to estimate the effect that prices have on oil production.

By looking only at the effect of price on fields that currently or previously have produced oil I am limiting the scope of this article.  The effect of oil prices on total production over an extended period of time is due not just to reactions in production in existing fields but also increased searching for new fields.  In fact, an implication of this work is that much of the total production response from higher oil prices is likely from increased searching as well as production from previously un-economic fields.

The main finding in this article is that oil production at the field level has no significant reaction to concurrent changes in the oil price, where concurrent is broadly defined as within the first three years.  A slight effect can be detected at a lag of between 4 and 8 years, with a magnitude of about a 2 to 4\% increase in yearly production for a 10 dollar increase in the price of oil.  This effect is somewhat greater and with less of a lag in large fields compared to small fields.  Price appears to have the most significant affect during the planning stage - before production begins in a field.  In the depletion phase of production, price is found to have little to no significant effect.

The main methodological problem is the non-linear production profile of oil fields.  Once full-scale extraction is started in an oil field, pressure in wells will start declining. Compensating measures such as gas and water injection have been successful, but their impact is temporary. In turn production rates will therefore inevitably decline.

More so, oil field production is correlated across fields - that is, increases and decreases in production in fields are not randomly distributed across time.  Instead, as figure \ref{top10_production} shows with the production profile of the 10 largest Norwegian oil fields, production tends to be correlated across fields.  The result is a total production curve that is bell-shaped over time as in figure \ref{oil_decline}.  Since the oil price series is autocorrelated and non-stationary, failure to properly account for the production profile will lead to spurious estimation of the price terms in a regression.

The direction of this bias can be gleaned in figure \ref{oil_decline}.  High oil prices were present at periods of relatively low production in the late 1970s and early 1980s as well as over the last 10 years, however real prices reached some of their lowest levels at the same time as the top of production around the year 2000. These oil price dynamics almost certainly affected investment decisions in the industry and in turn total production (\citep{osmundsen_is_2007}, \citep{aune_financial_2010}). However the inverse contemporaneous relationship at the field level is entirely coincidental, but will heavily bias the estimation of the effect of price on production if field production profiles are not properly accounted for.

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/top10_production_print.png}
	\caption{The production profile of the 10 largest oil fields on the Norwegian Continental Shelf.  Production tends to be correlated across fields}
	\label{top10_production}	
	\end{figure}

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/oil_decline_print.png}
	\caption{The production profile for the entire Norwegian Continental Shelf is bell-shaped, reflecting the correlated production profiles of the fields.  With oil prices that are autocorrelated, }
	\label{oil_decline}
\end{figure}

As a solution I use a semi-parametric model within the Generalized Additive Model frameworks of \cite{hastie_generalized_1990}.  Here I use a two-dimensional smoothed spline function to account for the general non-parametric shape of the production profile while allowing price to enter the equation linearly.  The coefficient of price can then be interpreted as the average effect of price on production over the entire production profile.

\FloatBarrier
\section{The effect of oil price on production: theory, simulations and empirics}

For modeling aggregate oil production, shape-fitting models, notably \citet{hubbert_energy_1962} and more recently advocated by \citet{deffeyes_hubberts_2001}, have had some success in estimating the timing of peak production at the regional and national level, but they tend to seriously underestimate the total recoverable resource of oil-producing regions and the models can be shown to be fundamentally misspecified \citep{boyce_prediction_2013}.   Simulation type studies where aggregate oil production is modeled through an often complex combination of physical and economic processes also exist in the geo-engineering literature, but their usefulness tends to be weighed down by their complexity as they require quite detailed data and specific assumptions about functional form that can be difficult to justify \citet{brandt_review_2010}.

More important to this paper are empirical estimates of the effect of price on production. Several econometric papers seek to answer the question of how aggregated oil supply is affected by oil prices.  \citet{farzin_impact_2001} attempts to estimate an elasticity for the effect on added reserves of increased oil prices and finds a small though statistically significant effect.  \citet{ramcharran_oil_2002} estimates a supply function for the total supply of oil from several OPEC countries based on data from 1973 to 1997.  The author finds a negative price elasticity for several of the countries, and interprets this as evidence of producers targeting revenue.  However since the author does not take into account the production profile of oil fields and the spurious correlation that can arise with autocorrelated prices, these estimates come under considerable doubt.

the effect of oil-price uncertainty on drilling and exploration has also been explored.  \citet{kellogg_effect_2014} finds that oil exploration firms in Texas do approximately respond as real-option theory would predict when it comes to the timing of drilling.  A model and test using data from North Sea producers on the UK continental shelf by \citet{hurn_geology_1994}, on the other hand, fails to find evidence that the variance in the oil price affects the timing of oil field development.  I do not attempt to directly model uncertainty, however given that the investments needed to increase oil production in an existing field are to a certain extent irreversible and that oil prices are highly volatile, the results can and probably should be interpreted with the real options framework in mind.

Only a few papers utilize field-level data.  \citet{black_is_1998} tests the relevance of “nesting” a structural empirical model of profit maximization that takes into account oil prices into a typical geo-engineering model of oil field production.  They find strong evidence that taking into account profit maximization, and implicitly price, substantially improves the fit compared to a purely geo-engineering type model.   The limitation of their methodology is that they are only able to test whether including economic factors like price affects the fit of the model but are not able to give an estimate of the effect.  Methodologically, the paper also relies heavily on assumptions about the functional form of both the geo-engineering aspects of the oil producer as well as their profit-maximization.  By taking a more flexible, semi-parametric approach to estimating the effects of oil field production profile, this paper avoids problems with overly restrictive assumptions.

\citet{rao_taxation_2010} uses data on land based oil wells in California to estimate the effects of tax changes and price controls.  The author finds that short-term tax changes caused small, but significant retiming of production from oil wells.  These findings are over-all consistent with the findings of this paper, though the author finds concurrent effects of taxation on production while I do not find any concurrent effect of prices on production.  This difference is most likely due to the significant differences in cost and complexity between operating offshore and onshore.  How producers react to a short-term tax change as opposed to a change in price is likely also different.  More so, the author does not consider differences between the different stages of production.

Studies using detailed Norwegian data on offshore activity also exist, though the focus has mainly been on exploration and drilling.  \citet{mohn_exploration_2008} finds that long-term changes in the oil-price has a strong effect on exploratory drilling though little effect is measured from short-term changes in the oil price.  \citet{osmundsen_exploration_2010} analyses drilling productivity over time on the Norwegian Continental Shelf while \citet{mohn_efforts_2008} finds that higher oil prices leads to higher reserves and as well as that oil prices affect producer risk-preferences - with higher prices leading to lower success rates but larger discovery size.

\section{Oil production on the Norwegian Continental Shelf}

The first commercial oil well in Norwegian continental waters was discovered in December of 1969 in what became the Ekofisk oil field, the largest Norwegian oil field by estimated recoverable reserves.  As figure \ref{north_sea_reserves} shows, most of the largest fields in the North Sea were found relatively early on while more recent finds have tended to be smaller - a pattern typical of oil producing areas called creaming.  A major exception to this trend was the recent find of the Johan Sverdrup field which is estimated to have approximately 300 million Standard Cubic Meters (SM3) of recoverable oil. \footnote{The Johan Sverdrup field is estimated to begin producing oil in 2019 and so is not present in the data set used for the analysis.}

\begin{figure}
\includegraphics[width=1.2\textwidth]{figures/north_sea_reserves_print.png}
\caption{Oil fields in the Norwegian territorial North Sea.  The largest oil fields tended to be discovered earliest, while newer finds tend to be smaller.  An exception is the large Johan Sverdrup field, which is expected to begin producing in 2019.}
\label{north_sea_reserves}
\end{figure}

Exploration in the Norwegian Sea was opened in the early 1980’s and the first commercial field started production in 1981.  While several mid-sized fields have been discovered, the Norwegian Sea has generally disappointed in terms of commercial oil finds and most finds have been relatively small (\ref{norwegian_sea_reserves}).

\begin{figure}
\includegraphics[width=1.2\textwidth]{figures/norwegian_sea_reserves_print.png}
\caption{Oil fields in the Norwegian Sea.  Production from the Norwegian Sea has generally been dissapointing compared with expectations when the area was opened to exploration in the early 1980's.}
\label{norwegian_sea_reserves}
\end{figure}

Norwegian waters in the Barents Sea have also been open to exploration since the 1980s, however up until recently only a few, small finds were made and none came into commercial production.  However recently several significant oil and gas finds have been made in the Barents Sea - notably the oil field Goliat and the large gas find Sn\o hvit, which are both currently under development but not yet producing.  The agreement between Norway and Russia in June 2011 settling a long-running dispute over the maritime delimitation has also given a boost to new exploration in the region.

Profits from oil and gas production in Norway are subject to a resource tax of 50 \% on top of the ordinary income tax of 28 \%, thus income from petroleum production is taxed at a total marginal tax rate of 78 \%.  The central government also receives revenues through ownership stakes in companies, notably Statoil, where the state is the majority stakeholder.  The over-all tax rate has been fairly constant through the history of Norwegian oil production, however several important changes related to the tax code have occurred.  In 1991 a C02 tax was introduced, and in the year 2012 the tax was doubled.  But this was mainly levied on the use and import of petroleum products.  In the offshore sector it was levied on the burning of oil and gas and thus the main effect was on the practice of flaring natural gas that could not be transported and sold commercially.

More importantly to the offshore sector were accounting changes that were implemented in 2002 and 2005 that were meant to encourage new entrants by allowing losses to be carried forward for tax purposes and by introducing a rebate on the tax-value of losses associated with searching and drilling.  In general though, these rules mainly affected searching and discoveries of new fields rather than production from existing fields and I do not control for the tax changes in my model.  More information on taxation and revenues from the offshore sector can be found at the website of the Norwegian Ministry of Finance. \footnote{\url{http://www.regjeringen.no/en/dep/fin/Selected-topics/taxes-and-duties/bedriftsbeskatning/Taxation-of-petroleum-activities.html?id=417318}}

Rights to explore and eventually produce on the Norwegian Continental Shelf are based on a system where the government announces geographic blocks that will be opened to oil exploration and production subject to production licenses.  Production licenses are initially granted for between 4-6 years subject to requirements that firms are actively searching in their awarded blocks.  If oil or gas deposits are proven then the production license can be extended for up to 30 years.  In general, the frameworks are fair, predictable and stable for companies who find commercially extractable oil deposits and regulatory interference is unlikely to be the cause of any observed changes in production from existing oil fields.  For more information see the website of the Norwegian Petroleum Directorate. \footnote{\url{http://www.npd.no/en/Topics/Production-licences/Theme-articles/Production-licence--licence-to-explore-discover-and-produce-/}}

\section{Norwegian oil field production data}
Production data of Norwegian oil fields is obtained from the website of the Norwegian Petroleum Directorate. \footnote{http://factpages.npd.no/}  Production data is available at a monthly frequency for all fields, though I choose to aggregate up to yearly values both to smooth over seasonality as well as short-term volatility of output due to factors such as weather or technical issues that are not relevant for this article.

In addition to data on field-level production, I also make use of data on estimated total recoverable reserves. The use of this variable is complicated as it is an estimate subject to a large amount of uncertainty, especially in relatively young fields.  However, the methodologies used to estimate the total recoverable resource of a field are constantly evolving and it is a fair assumption that any consistent bias of the estimates are observed in older fields and corrected for in estimates for newer fields.  I can then assume that existing errors are random and will not significantly bias the estimates.

Moreover, the estimate is likely endogenous, in the sense that it is also effected by prices.  However since I use the variable as a control variable and not for the purpose of estimating a parameter with a causal interpretation, this should not significantly affect the validity of the results.

I use yearly data from the US Energy Information Agency on the real price of Brent-traded oil in 2010 dollars.The Brent benchmark oil price is likely the best oil price measure for Norwegian production as it is based on light sweet crude oil sourced from the North Sea.

An argument can be made that expectations of future oil prices can be equally if not even more important for production decisions as the current oil price.  Forecasts for future oil prices are available from, among others, the International Energy Agency, but these have tended to be notoriously inaccurate and it is unlikely oil companies use these projections for their investment decisions.

On the other hand, given the size and liquidity of oil spot markets, it is a fair assumption that the current oil prices do a good job of incorporating much of the available information about crude oil markets and that future price movements are generally difficult to predict \citep{hamilton_understanding_2008}.  An active futures market does exist, but several studies have found that current oil prices are in general better than prices on futures contracts at predicting future oil prices \citep{alquist_what_2010, chinn_predictive_2005}.  \citet{mohn_investment_2008} as well as \citet{pesaran_econometric_1990} and \citet{farzin_impact_2001} find evidence for adaptive expectations, where expectations of future prices is based on a weighted average of current and past prices.  I take account of this by including several years of price lags in my regression equations.

A cleaned data set and the full code for the analysis are available upon request. I use the R statistical programming package for all the analysis in this article \citep{r_core_team_r:_2013}.  I use the R packages ggplot2 and ggmap for plotting \citep{wickham_ggplot2:_2009, kahle_ggmap:_2013}, plyr for data manipulation and cleaning \citep{wickham_split-apply-combine_2011}, texreg for table formatting \citep{leifeld_texreg:_2013} and mgcv for implementation of the Generalized Additive Models \citep{wood_fast_2011}.

\section{A generalized additive model of oil field production}
Parametric linear models have the sizeable advantages of simplicity and interpretability and are therefore usually a good starting point for an analysis.  However, when attempting to model the effect of price on oil field production, a standard linear model is unable to sufficiently control for the production profile and therefore heavily biases the estimate of the effect of price.  As an example, consider a generalized linear model written as in equation \ref{glm_eqn}.

	\begin{equation}
	\begin{split}	
	 Log(Production_{i,t})  &= \alpha_0 + \alpha_1 timetopeak_{i,t} + \alpha_2 timetopeak_{i,t}^2 \\
	  \quad & + \alpha_3 timetopeak_{i,t}^3  + \alpha_4 peaktoend_{i,t} + \alpha_5 peaktoend_{i,t}^2 \\
	  \quad & + \alpha_6 peaktoend_{i,t}^3 + \gamma totalrecoverableoil_i \\
	  \quad & + \beta_1 oilprice + \beta_2 oilpricel1 + ...+ \beta_6 oilpricel5 + \epsilon	
	\label{glm_eqn}
	\end{split}
	\end{equation}

Here the left hand side variable is yearly production in year $t$ for field $i$.  To try to account for the time profile of production, I split up model-time into a time-to-peak and peak-to-end variable as demonstrated for data for the Statfjord field in figure \ref{statfjord_dem} and then represent each as a cubic function.  Yearly production is assumed to be proportional to the total size of the field as represented by the estimate of the total recoverable resource.  Finally, a term for the oil price as well as five lagged terms are added in order to capture the effects of price.

\begin{figure}
\includegraphics[width=1\textwidth]{figures/statfjord_dem_print.png}
\caption{The production profile can be modeled in two phases - time-to-peak, representing the build-out phase of the oil field and time-from-peak, representing the period of depletion. The is illustrated with data from the Statfjord field.}
\label{statfjord_dem}
\end{figure}

Figure \ref{glm_dirty_box} shows the estimates of the coefficients on the oil price and its lags \footnote{The dots on the figure represent 1000 simulations of the estimated coefficient based on the estimated standard error and point estimate from the model}.  The lags are not estimated to have an effect significantly different from zero. However a literal interpretation of the coefficient on the concurrent oil price term would indicate that a 10-dollar increase in the oil price would lead to a lowering of field production of around 4-5\%.

\begin{figure}
\includegraphics[width=1\textwidth]{figures/glm_dirty_box_print.png}
\caption{A generalized linear model generates a negative estimate of the coefficient on price in a regression on production.  This result is spurious, caused by a failure to properly control for the production profiles of the fields.}
\label{glm_dirty_box}
\end{figure}

This estimate is heavily biased downwards due to the spurious correlation between the field production profiles and the autocorrelated and non-stationary time series of oil prices.  A more formal analysis of the bias can be found in the online appendix, \url{jmaurit.github.io#oil_price_sim}, where I simulated data of oil field production in a monte carlo experiment.  However, the basic problem is that the parametric representation is not flexible enough to control for the production profiles of the fields.   Instead, a more flexible estimation of the production profile is needed.

Instead of attempting to estimate the shape of the production profiles of the fields by estimating parameters on linear terms I estimate a non-parametric function for the production profile.  As an illustration, consider the production profile of a single field.  The simplest possible model would then have the form of equation \ref{simp_eqn}.

\begin{equation}
Production_{t}=f(time) + \epsilon
	\label{simp_eqn}
\end{equation}

Again considering the production profile for the Statfjord field, a smoothed function might look like the black line in figure \ref{statfjord_gam}.

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/statfjord_gam_print.png}
	\caption{An illustration of fitting a simple non-parametric function to the production profile of a single field.}
	\label{statfjord_gam}
\end{figure}

In principle any number of well-behaved smoothers could be used to estimate the function, for example a Loess or a kernel smoother.  In practice regression splines are most commonly used.  With a regression spline the data points for the function to be estimated are broken into bins.  For each bin of data a local linear regression is estimated.  For functions of one variable, a cubic parameterization is often used.  These regressions are then essentially tied together at what are called “knots” and the smoothness of the overall function is controlled by a penalty function that consists of the second derivative of the estimated function.

In my regression, I find it helpful to represent the production profile of fields as a two-dimensional function.  This can not be accomplished with a standard cubic regression spline, but I can instead use a Thin Plate (Regression) Spline \citep{wood_thin_2003}.  Though the full details of the implementation are beyond the scope of this paper, for the basic idea consider equation \ref{thin_plate_1}.

	\begin{equation}
	y_i = g(x_1, x_2)
	\label{thin_plate_1}
	\end{equation}

Following \citet{wood_generalized_2006}, $g$ is the function of $x_1$ and $x_2$ that is to be estimated by $f$, which in turn is estimated by minimizing equation \ref{thin_plate_2}.  Here $\boldsymbol{y}$ represents a vector of $y_i$’s and $\boldsymbol{f} = (f(\boldsymbol{x_1}),f(\boldsymbol{x_2}))^t$.

	\begin{equation}
\min \|\boldsymbol{y-f}\|^2 + \lambda J_{22}(f)
\label{thin_plate_2}
	\end{equation}

$J_{22}$ represents the penalty function for the smoothness of the function which can be written as in \ref{thin_plate_3}.  The $22$ represents the fact that it is a penalty function of two variables with smoothness measured by the second derivative.

	\begin{equation}
	J_{22}{f}= \diffp[2]{f}{x_1}^2 + \diffp[2]{f}{{x_1}{x_2}} + \diffp[2]{f}{x_2}^2dx_1 dx_2
\label{thin_plate_3}
	\end{equation}

In short, a function of $x_1$ and $x_2$ is found that is minimizing errors in the sense of minimizing Euclidean distance subject to a penalty function of “wiggiliness”.  The actual implementation is somewhat more involved in order to increase the computational efficiency of the estimation.  For further details I again refer to \citet{wood_thin_2003}.

The advantage to using splines over other smoothing methods is that it can be represented in a linear form.  Thus estimation of the model can be done using standard and efficient matrix algebra algorithms. For further details I refer to the discussion in \citet{hastie_generalized_1990} and \citet{wood_generalized_2006}.  The latter is a particularly useful reference for implementing generalized additive models in R.

I do not want to estimate smoothed curves individually for each field.  While this would provide a good overall fit to the full data set, not enough variation in the data would be left to estimate the effect of price.  Instead I want to estimate a general shape of the production profile for all fields and then use the remaining variation in the data to estimate the effect of price.  My model can be written as in equations \ref{gam_price_eqn}.

\begin{equation}
\begin{split}

	Log(Production_{i,t}) & = f(time\_to\_peak_{i,t}, total\_recoverable\_oil_i) \\
 	 \quad & + f(peak\_to\_end_{i,t}, total\_recoverable\_oil_i) \\
	 \quad & + \beta_1 oil\_price + \beta_2 oil\_price\_l1 + ... +  \epsilon
\label{gam_price_eqn}
\end{split}
\end{equation}

In this model I am estimating the parameters and functions from all fields $i$. As in the parametric model presented earlier, the left-hand-side variable is yearly oil production for field $i$. Also like the parametric model I split the production-time component element in two: up to and after the peak in production. 
While a shape for the entire production profile could be estimated with one smoothed function, splitting it up allows for more flexibility and better overall fit of the model as estimated by deviance score and the related estimated degrees of freedom of the model.  More so, production up to a peak and the subsequent decline represents two distinct processes.  The first is driven by the physical investment and build-out of the field, while the latter is dominated by depletion of the field and subsequent drop in pressure. It makes sense to model these two processes separately, and later I will also attempt to estimate separate price effects for the two phases.

I also allow the smoothed functions to vary with the total size of the field as measured by the estimated total recoverable oil since the shape of the production profile tends to vary substantially by field size.  Inspection of a selection of fields, such as shown in figure \ref{field_inspection}, shows that smaller fields tend to reach their peak quickly while larger fields take more time.

The estimated field-size variable is however an imperfect measure.  First, estimates for newer fields will likely have higher measurement errors than established fields that are well explored and where a significant amount of the oil has already been extracted.  The total reserves estimates are also likely correlated with the price terms as estimates of recoverable reserves likely change depending on the level of the oil price.  Nonetheless, since I am only using the variable to help to control for the general shape of the production profiles of fields and not as an exogenous regressor, the inclusion of the variable should not materially affect the estimation.

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/field_inspection_print.png}
	\caption{A selection of production profiles of oil fields.  The shape can vary substantially by field size as well as other factors.}
	\label{field_inspection}
\end{figure}

Even with a smoothed function that is allowed to vary by field-size, a substantially better fit, as well as more nuanced analysis could be obtained by splitting the estimation into small and large fields, this time measured by the maximum production of the field - a measure of field size that is less likely to be highly correlated with price.  The improvement in fit can be seen by inspecting the fitted values of the models as in figure \ref{bench_vs_split}. The split estimation provides a particularly better fit for smaller and mid-sized fields.

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/bench_vs_split_print.png}
	\caption{A benchmarking model shows that a substantially better fit can be achieved by splitting the analysis into small and large fields, where a somewhat arbitrary maximum yearly production of 8 million SMU is used as the separating value.}
	\label{bench_vs_split}
\end{figure}

Plots of the estimated smoothed functions in the model are not directly informative about the effects of price and so I do not include them in the main text.  However they can be helpful in gaining a intuition behind the estimation procedure so I have included them in appendix 2 along with other model diagnostics and robustness checks.

\section{The Effect of Oil Price on Field Production}

The variables of interest in this article is the oil price and its lags, which I include as nine linear parametric terms in the model.  The idea of including both a concurrent oil price term as well as eight lags is that a change in price could conceivably have two effects on oil production in a field.  First, the field operator could be operating on the basis of some short-term extraction rule - choosing to pump out less at times of lower prices in order to pump out more at periods of high prices.

Alternatively, a change in price of oil can be seen as a lifting of a production constraint.  A higher oil price means that added investments in production become attractive in order to either increase the total amount extracted from a field or to shift production forward.  However investments in the offshore sector can be complex and lengthy, and any production lift would be expected to happen with a lag.  As mentioned earlier, including several lags also allows for the possibility of adaptive expectations of future oil prices.The results for the price terms are shown in figure \ref{gam_price_8}.

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/gam_price_8_print.png}
	\caption{The coefficient estimates on the price terms where two additional lags are added to the model.  Significant positive results are found on the 8th lag for both large and small fields, providing evidence for adaptive price expectations by producers.}
	\label{gam_price_8}
\end{figure}

The estimated coefficients for the concurrent and first three lags of the oil price are not estimated to be significantly different than zero.  For large fields a modest effect at the fourth, sixth and seventh lags are estimated, though only at a 10\% significance level, \footnote{The lines in the boxplot represent conservative 95\% confidence intervals and are calculated using a bayesian-inspired simulation from the posterior distribution of the coefficients.  P-values (stars) shown in the table in the appendix were calculated by standard asymptotic methods and show significant results for the 4th and 6th lags at the 5\% level.} while for small fields an effect is estimated only at the sixth lags. The magnitude of these effects is in the neighborhood of a 2\% increase in production for a 10 dollar increase in the oil price.  A larger and more significant effect is found for both small and large fields at the 8th lag, with a magnitude implying an approximately 3 to 4\% increase in production for a 10 dollar increase in the oil price.

Despite the clear implications of most simplified Hotelling models, the lack of significant results for the concurrent price and the modest lagged effect is not surprising. In general, operating an oil production rig and related offshore infrastructure is an extremely expensive venture with high fixed costs.  In the challenging conditions of the North Sea, the expenses are multiplied.  Thus any short-term benefit of strategically altering production in relation to movements in the oil price are dominated by the large costs of having excess pumping capacity.  In other words, oil producers have a strong incentive to pump as much oil out at any given time given the existing production capacity.

Instead, the results point to a mechanism where oil producers react to higher oil prices by increasing investment in those fields, leading to higher oil production only with a considerable lag.   This story is in line with a trend of increased total extraction estimates from the Norwegian continental shelf as a whole as well as from existing fields over the last 15 years of strongly rising oil prices \footnote{\url{http://npd.no/Templates/OD/Article.aspx?id=4731&epslanguage=en}}.

I prefer the un-pooled models I have presented above because of the improvement in model fit and because operators of small and large fields may be expected to have substantially different reactions to changes in price as well as different time profiles of production. Nonetheless, it can be instructive to show results from a pooled model where a fixed effect and interaction variable is introduced to distinguish the effects of small and large fields. The model can be written as equation \ref{pooled_eqn}.  The terms in the model are mostly the same as in \ref{gam_price_eqn} but with an added dummy term for small fields - those with a maximum yearly production of 8 million SM3 or below, $smallfield$, and interaction terms.

\begin{equation}
\begin{split}
	Log(Production_{i,t})&=f(timetopeak_{i,t}, totalrecoverableoil_i) \\
	 \quad & + f(peaktoend_{i,t}, totalrecoverableoil_i) \\
	 \quad & + \beta_1 oilprice + \beta_2 oilpricel1 + \dots + oilpricel8\\
	 \quad & + \delta smallfield \\
 	 \quad & + \gamma_1 oilprice*smallfield + \gamma_2 oilprice*smallfield + \dots \\
 	 \quad & + oilpricel8*smallfield +  \epsilon
\label{pooled_eqn}
\end{split}
\end{equation}

The full results for the estimates of all the parametric terms can be found in table \ref{table:pooled} in the appendix.  The estimated coefficients on the price terms are shown in figure \ref{gam_price_pooled}. In the regression none of the interaction effects were significantly different than zero, nor were the concurrent price term or the first three lags.  The fourth and sixth through eighth lag were however significant and positive, again at a magnitude of approximately 2\% increase in production for a \$10 increase in the oil price.

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/gam_price_pooled_print.png}
	\caption{The estimated coefficients on the price terms for a pooled model of oil field production.  Significant estimates are found on the fourth, sixth, seventh and eighth price lags}
	\label{gam_price_pooled}
\end{figure}

In the above discussion, I have been implicitly giving the coefficients a causal interpretation, which deserves some discussion.  My main identifying assumption is that production at the field level can not cause significant changes in the oil price.  A stronger but still relevant assumption that may be necessary since field production is correlated across fields is that total production from the Norwegian continental shelf does not affect oil price.

Both of these assumptions are likely satisfied.  Oil is a globally traded commodity and total Norwegian production accounts only for a small fraction of total production.  In 2013 Norwegian production made up only 2.3 \% of the world total. \footnote{\url{http://www.eia.gov/countries/country-data.cfm?fips=NO#pet}}  A drastic change in production, on par with the halt in production that occurred in Libya in 2011, \footnote{Before the Libyan revolution of 2011, Norwegian and Libyan yearly oil production were of a similar magnitude, 2.1 versus 1.8 million barrels of oil in 2010 \url{http://www.eia.gov/countries/country-data.cfm?fips=LY#pet}} would have been required to have had any significant effect on world oil supply and in turn prices.

One of the main implications of the interpretation that higher oil prices leads to increased extraction through increased investment is that while the effect on production will be lagged, higher oil prices will have a more immediate effect on investments.  I test for this with the model as written in equation \ref{gam_invest_eqn}

\begin{equation}
\begin{split}
	Log(Investment_{i,t})&=f(timetopeak_{i,t}, totalrecoverableoil_i) \\
	 \quad &+ f(peaktoend_{i,t}, totalrecoverableoil_i) \\
	 \quad &+ \alpha oilproduction_{i,t} \\
	 \quad & + \beta_1 oilprice + \beta_2 oilpricel1 + ... + oil_pricel8 \\
	 \quad &+ \delta smallfield \\
	 \quad &+ \gamma_1 oilprice*smallfield + \gamma_2 oilprice*smallfield + ... 
	+  \epsilon \\
\label{gam_invest_eqn}
\end{split}
\end{equation}

In the equation investment in each field $i$ at time $t$ is a function of the state of field development, as modeled by a non-parametric function of time to and from the peak as well as the total size of the field - as in the model of oil field production.  In addition I include a term for oil production in field $i$ at time $t$.  The coefficients of interest are again those on the oil price and its lags which I interact with a dummy variable for large and small fields, $smallfield$.

The results for the estimation of the estimated coefficients on the oil price and its lags is shown in figure \ref{gam_price_invest_box} while full results are also shown in table \ref{table:investment} in appendix 1.  The coefficient on the concurrent oil price as well as the third and fourth lags are all significantly positive, with coefficients that can be interpreted to mean that a 10 dollar increase in the oil price leads to between a 5 to 10 \% increase in oil field investment in the concurrent and second subsequent year and between a 10 to 20\% increase in the third subsequent year.

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/invest_pooled_print.png}
	\caption{Estimated coefficients for the price terms in a model of oil field investments.  A rise in the oil price leads to an increase in investment in the concurrent year as well as in the second and third lag. Negative coefficients are found for the sixth through 8th lags.}
	\label{gam_price_invest_box}
\end{figure}

The results of the regression on investment fits nicely with the previous results on oil production which showed a significant effect in the 4th through 6th lags.  A story consistent with both results is that higher oil prices induce increased investment in production capacity, though it takes time to get the extra capacity in place and the effects on production to be felt.

The estimated coefficients on the 6th, 7th and 8th lagged coefficients are shown to be significantly negative. An explanation is that a higher oil price induces producers to speed up a build out of a field, leading to more investment earlier, but relatively less in later years.  An implication of this theory is that the effect of price will primarily be in the build-out phase of a fields production profile.  This is explored in the following section.

\section{The effect of price in the phases of oil production}

Treating the entire production life of an oil field as essentially the same process is an oversimplification of the industry and field dynamics.  A field's life can be split into roughly three periods.  Before physical investment and production can begin, an often times lengthy planning stage occurs.  In this stage, the extent and details of a fields build-out are decided and must receive approval from the Norwegian Petroleum Directorate as well as other government agencies.

Once plans have been approved and finalized, the build-out phase of a field can begin.  The field operator has an incentive to build out the field as quickly as possible as large amounts of capital have at that point been allocated and it is important to get oil production and in turn income flowing as quickly as possible.  Once the build-out is complete and production has reached its peak, geophysical forces that come from depletion and subsequent drop in field pressure dominate.  Here technological change can also play a role - with improvements over time - like the introduction of gas and water injection - allowing for increased recovery percentage of the available reservoir.

In the preceding sections I have partially taken account of the different dynamics of the phases of field production by modeling post- and pre-peak production as seperate smoothed terms.  However, the price variable was estimated for the entire production profile.  The advantage of this approach is that I am able to use the full data-set and estimate a total average effect.  However the effect of price can reasonably be expected to differ in the different phases of production.

In this section I model the pre-peak and post-peak periods of production separately - including the price terms. The vast majority of producing fields in Norway are well past their peak production and because larger fields tended to be found earlier, an even larger share of total oil production comes from depleting fields.  In terms of policy implications, the most important question is what the effect of price has on post-peak production.

To ensure that no fields are included in the post-peak group that have not been fully built-out yet and which have not reached their peak production, I exclude all fields that began oil production prior to 2008.  While some large fields may take slightly more than five years to fully build out, none such fields began production in that time frame.

 The estimated coefficients on the price term for post-peak production is shown in figure \ref{gam_postpeak_print}.  The full results are shown in table \ref{table:postpeak} in the appendix.  The results mostly show no significant effect of price on production through the 7th lag, though positive coefficients for large fields are estimated for all but the 5th lag.

 A significant positive coefficient is estimated on the first lag for small fields, however this result is not robust to specification.  Using the maximum production instead of estimated total oil reserves in the smoothed function gives a somewhat better fit for some small fields (see figure \ref{chart:Max_prod_vs_TRO_print} in Appendex 2).  In this regression the coefficient is estiamted to be smaller and is no longer significant (see table \ref{table:postpeak}).  When small and large fields are pooled in one regression, this coefficient is again not estimated to be significantly different from zero.  This estimated coefficient should be met with some skepticism.

 On the other hand, the significant coefficient on the 8th lag for small fields is robust to specification. From an industry perspective, why price should have a 8-year-lagged effect on small fields but not large fields may not be immediately clear, but the length of the lag combined with the relatively short life of small fields suggests that the effect of price occurs well before the depletion phase begins.

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/gam_postpeak_print.png}
	\caption{The estimated coefficients on the price terms in a model of oil field production in depleting fields.  Significant coefficients are only estimated on the concurrent price and eight lag, however the significance of the estimated coefficient on the eighth lag is not robust to specification}
	\label{gam_postpeak_print}
\end{figure}

The effect that price has on production from a declining fields appears to be, at best, slight.  The point estimates on the price terms for large fields are of a magnitude of around 1\% - 2\% for a 10 dollar increase in the price of oil, but as noted, these estimates are not significantly different from zero.  In general, no clear relationship appears in the data between production and prices.

However, to say that price has no effect on production from depleting fields may be too strong. One mechanism for prices to effect production in depleting oil fields that this regression may not directly capture is through technological change.  Higher prices may encourage more research and development and in turn lead to innovations that increase the total recovery percentage of a field.  For example, over the course of the history of Norwegian oil production, innovations such as horizontal drilling and four-dimensional geological visualization have led to increased estimates of the total recovery rate of Norwegian oil fields.  However, this mechanism is likely too diffuse - lacking any consistent relationship between price, innovation and subsequent increases in production - to be captured by the regressions in this analysis.

While price appears to have little affect on production in depleting fields, it may have an effect in the planning and build-out phase of production. I run a regression model of field production up to peak production, representing the build-out phase of a field.  Because of the double-peak profile of the Ekofisk field, I only include data from this field up to the initial peak in 1976.  The full results can be found in table \ref{table:prepeak} in Appendix I.  I show the estimated coefficient on the price terms in table \ref{gam_prepeak_print} below.

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/gam_prepeak_print.png}
	\caption{The estimated coefficients on the price terms in a model of fields in the build-out phase.  For large fields, positive coefficients are estimated at the 6th, 7th and 8th lag, but the estimates are imprecise and not statistically significant.  Significant positive coefficients are estimated for small fields at the fifth and eighth lag.  In a pooled model, significant positive coefficients are found at the fourth through eighth lags.}
	\label{gam_prepeak_print}
\end{figure}

What is immediately noticeably is the imprecision of the price estimates in the model of the large fields.  This is primarily due to a lack of data points.  The estimates are based on only 76 data points from 11 fields.  Positive coefficients are estimated at the 6th, 7th and 8th lag, though none of these are statistically significant at the 95\% level.  The coefficient on the lagged price terms for small fields, however, are estimated to be positive and significant at the 5th and 8th lags.

I also estimate a pooled model, with both small and large fields.  This model provides a worse overall fit than the split model, but can nonetheless be useful given the few data points available for large fields.  The coefficients on the 1st as well as 3rd to through 8th lags are all estimated to be significant and positive at a magnitude that corresponds to approximately 2-8 \% increase in oil production for a 10 dollar increase in the cost of a barrel of oil.

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/field_time_to_peak_print.png}
	\caption{A histogram of the number of year before a field is fully built out and a field reaches its peak production.  The vast majority of fields, especially small fields, are built out quickly - within one or two years.}
	\label{gam_buildout_hist}
\end{figure}

Figure \ref{gam_buildout_hist} shows a histogram of the time it takes to fully build out an oil field and reach peak production.  We see that the vast majority of fields are built out quickly, reaching their peak production within one or two years.  Since the effect of price on the build-out phase is estimated with a lag of between 5 and 8 year, this indicates that changes in price likely has most effect well before production in a field even begins.  In turn, this suggests that changes in price that occur during the planning stage of a fields development are the most important in determining the production of the field.  This result fits well with the industry structure.  As mentioned, before physical investment in a field can begin and production started, detailed plans need to be approved by several government agencies and capital must be allocated.  Changing the level of investments in reaction to a change in prices after plans have been finalized is likely to lead to both an expensive delay as well as other costs that heavily outweigh any benefits of increased future production.

\section{Conclusion}

The main results of this research is to show that production in existing Norwegian offshore fields has no significant concurrent reaction to higher oil prices while a slight effect is estimated  with a lag of between 4 and 8 years.  Oil producers do not appear to be behaving strategically in relation to short-term production - increasing or reducing production in response to changes in oil price.  Instead they are likely using storage or financial instruments to hedge short-term price movements. Changes in oil prices can rather be seen as a relaxing of a production constraint, justifying increased investment that leads to either a higher total extraction rate or an inter-temporal shifting of production. Furthermore, it appears that the most of the effect of price comes at the planning phase of a field's production life.  Little to no effect of price on production is found in the build-out or depletion phase.

The modest estimated effect of prices on production adds weight to the argument of \citet{hamilton_oil_2012} that most of the increased supply of oil that comes from higher prices is from expanding the geographic and technological boundaries of oil production.  For example exploration of deep-water oil deposits off the coast of Brazil and extraction of oil sands in western Canada.

% \begin{spacing}{1}

\bibliographystyle{plainnat}
\bibliography{oil_prices}

% \end{spacing}
% \end{document} for shorter version with no appendix
\FloatBarrier

\appendix

\begin{table}
\begin{center}
\begin{tabular}{l c c c c c c }
\hline
	&\multicolumn{2}{|c|}{6 lags} & \multicolumn{2}{c}{8 lags} &\multicolumn{2}{c}{4-8 lags} \\
                                     & Small & Large & Small  & Large & Small & Large\\
\hline
(Intercept)                          & $-1.46^{***}$ & $-0.15$       & $-1.58^{***}$ & $-0.25$       & $-1.54^{***}$ & $-1.54^{***}$ \\
                                     & $(0.37)$      & $(0.50)$      & $(0.38)$      & $(0.51)$      & $(0.37)$      & $(0.37)$      \\
oilpricereal                       & $0.00$        & $-0.02$       & $0.00$        & $-0.01$       &               &               \\
                                     & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      &               &               \\
oilpricereall1                    & $0.01$        & $-0.01$       & $0.01$        & $0.01$        &               &               \\
                                     & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      &               &               \\
oilpricereall2                    & $-0.02$       & $0.01$        & $-0.01$       & $0.00$        &               &               \\
                                     & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      &               &               \\
oilpricereall3                    & $-0.02$       & $0.01$        & $-0.03^{*}$   & $0.00$        &               &               \\
                                     & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      &               &               \\
oilpricereall4                    & $0.00$        & $0.03^{*}$    & $0.00$        & $0.03^{*}$    & $-0.01$       & $-0.01$       \\
                                     & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
oilpricereall5                    & $-0.01$       & $0.01$        & $-0.01$       & $0.01$        & $-0.02$       & $-0.02$       \\
                                     & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
oilpricereall6                    & $0.02^{*}$    & $0.03^{*}$    & $0.02$        & $0.01$        & $0.02^{*}$    & $0.02^{*}$    \\
                                     & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
EDF: s(timetopeak,recoverableoil) & $28.92^{***}$ & $26.89^{***}$ & $28.91^{***}$ & $26.89^{***}$ & $28.91^{***}$ & $28.91^{***}$ \\
                                     & $(29.00)$     & $(27.00)$     & $(29.00)$     & $(27.00)$     & $(29.00)$     & $(29.00)$     \\
EDF: s(peaktoend,recoverableoil)  & $17.75^{***}$ & $10.71^{***}$ & $17.58^{***}$ & $11.41^{***}$ & $17.47^{***}$ & $17.47^{***}$ \\
                                     & $(28.00)$     & $(28.00)$     & $(28.00)$     & $(28.00)$     & $(28.00)$     & $(28.00)$     \\
oilpricereall7                    &               &               & $-0.01$       & $0.02^{*}$    & $-0.01$       & $-0.01$       \\
                                     &               &               & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
oilpricereall8                    &               &               & $0.04^{***}$  & $0.03^{**}$   & $0.03^{***}$  & $0.03^{***}$  \\
                                     &               &               & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
\hline
AIC                                  &               & 1211.33       &               & 1186.97       &               &               \\
BIC                                  &               & 1377.97       &               & 1363.28       &               &               \\
Log Likelihood                       &               & -559.07       &               & -544.18       &               &               \\
Deviance                             & 12319.62      & 301090.57     & 12104.52      & 268975.69     & 12347.62      & 12347.62      \\
Deviance explained                   & 0.84          & 0.96          & 0.85          & 0.96          & 0.84          & 0.84          \\
Dispersion                           & 15.17         & 1378.61       & 14.93         & 1247.02       & 15.16         & 15.16         \\
R$^2$                                & 0.82          & 0.95          & 0.83          & 0.95          & 0.82          & 0.82          \\
GCV score                            & 16.19         & 1666.44       & 15.98         & 1526.28       & 16.13         & 16.13         \\
Num. obs.                            & 865           & 264           & 865           & 264           & 865           & 865           \\
Num. smooth terms                    & 2             & 2             & 2             & 2             & 2             & 2             \\
\hline
\multicolumn{7}{l}{\scriptsize{\textsuperscript{***}$p<0.001$, 
  \textsuperscript{**}$p<0.01$, 
  \textsuperscript{*}$p<0.05$}}
\end{tabular}
\caption{Full results for model of field production with price, seperate estimation of small and large fields.}
\label{table:unpooled}
\end{center}
\end{table}

%table two pooled
\begin{table}
\begin{center}
\begin{tabular}{l c c }
\hline
                                     & Pooled Model & Insignificant terms dropped \\
\hline
(Intercept)                          & $-1465.35^{***}$ & $-1500.51^{***}$ \\
                                     & $(65.21)$        & $(66.28)$        \\
oilpricereal                       & $-0.01$          &                  \\
                                     & $(0.01)$         &                  \\
largefieldsmall                     & $-2.32^{***}$    & $-1.19^{***}$    \\
                                     & $(0.27)$         & $(0.10)$         \\
oilpricereall1                    & $0.02^{*}$       &                  \\
                                     & $(0.01)$         &                  \\
oilpricereall2                    & $0.01$           &                  \\
                                     & $(0.01)$         &                  \\
oilpricereall3                    & $0.00$           &                  \\
                                     & $(0.01)$         &                  \\
oilpricereall4                    & $0.03^{***}$     & $0.03^{***}$     \\
                                     & $(0.01)$         & $(0.01)$         \\
oilpricereall5                    & $0.01$           & $0.01$           \\
                                     & $(0.01)$         & $(0.01)$         \\
oilpricereall6                    & $0.02^{***}$     & $0.03^{***}$     \\
                                     & $(0.01)$         & $(0.01)$         \\
oilpricereall7                    & $0.02^{**}$      & $0.02^{***}$     \\
                                     & $(0.01)$         & $(0.01)$         \\
oilpricereall8                    & $0.03^{***}$     & $0.03^{***}$     \\
                                     & $(0.00)$         & $(0.00)$         \\
oilpricereal:largefieldsmall      & $-0.03$          &                  \\
                                     & $(0.06)$         &                  \\
largefieldsmall:oilpricereall1   & $0.03$           &                  \\
                                     & $(0.07)$         &                  \\
largefieldsmall:oilpricereall2   & $-0.05$          &                  \\
                                     & $(0.08)$         &                  \\
largefieldsmall:oilpricereall3   & $0.03$           &                  \\
                                     & $(0.07)$         &                  \\
largefieldsmall:oilpricereall4   & $0.03$           &                  \\
                                     & $(0.07)$         &                  \\
largefieldsmall:oilpricereall5   & $0.04$           &                  \\
                                     & $(0.07)$         &                  \\
largefieldsmall:oilpricereall6   & $0.18^{*}$       &                  \\
                                     & $(0.07)$         &                  \\
largefieldsmall:oilpricereall7   & $0.06$           &                  \\
                                     & $(0.07)$         &                  \\
largefieldsmall:oilpricereall8   & $0.07$           &                  \\
                                     & $(0.07)$         &                  \\
EDF: s(timetopeak,recoverableoil) & $28.99^{***}$    & $28.99^{***}$    \\
                                     & $(29.00)$        & $(29.00)$        \\
EDF: s(peaktoend,recoverableoil)  & $27.98^{***}$    & $26.99^{***}$    \\
                                     & $(28.00)$        & $(27.00)$        \\
\hline
AIC                                  &                  &                  \\
BIC                                  &                  &                  \\
Log Likelihood                       &                  &                  \\
Deviance                             & 371079.07        & 396494.49        \\
Deviance explained                   & 0.96             & 0.96             \\
Dispersion                           & 352.06           & 371.24           \\
R$^2$                                & 0.96             & 0.95             \\
GCV score                            & 377.77           & 393.13           \\
Num. obs.                            & 1129             & 1129             \\
Num. smooth terms                    & 2                & 2                \\
\hline
\multicolumn{3}{l}{\scriptsize{\textsuperscript{***}$p<0.001$, 
  \textsuperscript{**}$p<0.01$, 
  \textsuperscript{*}$p<0.05$}}
\end{tabular}
\caption{Full results for pooled model of field production with price terms.}
\label{table:pooled}
\end{center}
\end{table}

%investment tables

\begin{table}
\begin{center}
\begin{tabular}{l c c c }
\hline
                                     & Pooled & Small Fields & Large Fields \\
\hline
(Intercept)                          & $-199.84^{***}$ & $7.03^{***}$   & $7.64^{***}$    \\
                                     & $(44.00)$       & $(0.36)$       & $(0.24)$        \\
yearprod                            & $-0.06^{***}$   & $-0.26^{***}$  & $-0.06^{***}$   \\
                                     & $(0.01)$        & $(0.02)$       & $(0.01)$        \\
oilpricereal                       & $0.08^{***}$    & $-0.03^{**}$   & $0.04$          \\
                                     & $(0.01)$        & $(0.01)$       & $(0.03)$        \\
largefieldsmall                     & $1.80^{***}$    &                &                 \\
                                     & $(0.52)$        &                &                 \\
oilpricereall1                    & $-0.02$         & $0.00$         & $0.00$          \\
                                     & $(0.01)$        & $(0.01)$       & $(0.03)$        \\
oilpricereall2                    & $0.06^{***}$    & $0.05^{***}$   & $0.09^{***}$    \\
                                     & $(0.01)$        & $(0.01)$       & $(0.03)$        \\
oilpricereall3                    & $0.16^{***}$    & $0.25^{***}$   & $0.19^{***}$    \\
                                     & $(0.01)$        & $(0.03)$       & $(0.03)$        \\
oilpricereall4                    & $0.00$          & $-0.08^{**}$   & $-0.03$         \\
                                     & $(0.01)$        & $(0.03)$       & $(0.03)$        \\
oilpricereall5                    & $-0.02$         & $-0.09^{**}$   & $-0.07^{*}$     \\
                                     & $(0.02)$        & $(0.03)$       & $(0.03)$        \\
oilpricereall6                    & $-0.09^{***}$   & $-0.10^{**}$   & $-0.04$         \\
                                     & $(0.02)$        & $(0.04)$       & $(0.04)$        \\
oilpricereall7                    & $-0.11^{***}$   & $-0.26^{***}$  & $-0.09^{*}$     \\
                                     & $(0.02)$        & $(0.04)$       & $(0.04)$        \\
oilpricereall8                    & $-0.10^{***}$   & $-0.19^{***}$  & $-0.12^{***}$   \\
                                     & $(0.01)$        & $(0.03)$       & $(0.03)$        \\
oilpricereal:smallfield      & $-0.11^{*}$     &                &                 \\
                                     & $(0.05)$        &                &                 \\
largefieldsmall:oilpricereall1   & $0.02$          &                &                 \\
                                     & $(0.05)$        &                &                 \\
largefieldsmall:oilpricereall2   & $0.01$          &                &                 \\
                                     & $(0.05)$        &                &                 \\
largefieldsmall:oilpricereall3   & $0.03$          &                &                 \\
                                     & $(0.06)$        &                &                 \\
EDF: s(timetopeak,recoverableoil) & $28.99^{***}$   & $26.92^{***}$  & $28.79^{***}$   \\
                                     & $(29.00)$       & $(28.11)$      & $(28.97)$       \\
EDF: s(peaktoend,recoverableoil)  & $16.90^{***}$   & $4.83^{*}$     & $5.39^{**}$     \\
                                     & $(18.82)$       & $(6.40)$       & $(7.44)$        \\
\hline
AIC                                  &                 &                & 4747.87         \\
BIC                                  &                 &                & 4912.98         \\
Log Likelihood                       &                 &                & -2327.76        \\
Deviance                             & 257371493567.88 & 10630461533.61 & 198564735228.80 \\
Deviance explained                   & 0.82            & 0.81           & 0.84            \\
Dispersion                           & 243467395.17    & 13087671.29    & 907398412.98    \\
R$^2$                                & 0.80            & 0.79           & 0.81            \\
GCV score                            & 257721863.13    & 13776495.42    & 1094706852.34   \\
Num. obs.                            & 1117            & 853            & 264             \\
Num. smooth terms                    & 2               & 2              & 2               \\
\hline
\multicolumn{4}{l}{\scriptsize{\textsuperscript{***}$p<0.001$, 
  \textsuperscript{**}$p<0.01$, 
  \textsuperscript{*}$p<0.05$}}
\end{tabular}
\caption{Full results for model of field investment with price terms.}
\label{table:investment}
\end{center}
\end{table}

% Table post-peak

\begin{table}
\begin{center}
\begin{tabular}{l c c c c }
\hline
&\multicolumn{2}{c}{Recoverable Oil} & \multicolumn{2}{c}{Max Production} \\
                                    & Large & Small & Large & Small \\
\hline
(Intercept)                         & $1.54^{***}$  & $-0.31^{*}$   & $2.25^{***}$  & $-1.61^{***}$ \\
                                    & $(0.06)$      & $(0.14)$      & $(0.08)$      & $(0.25)$      \\
oilpricereal                      & $0.01$        & $0.03^{***}$  & $-0.02^{*}$   & $0.02^{*}$    \\
                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
oilpricereall1                   & $0.02$        & $-0.01$       & $-0.01$       & $0.00$        \\
                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
oilpricereall2                   & $0.00$        & $-0.01$       & $-0.02$       & $-0.01$       \\
                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
oilpricereall3                   & $0.02$        & $-0.03^{**}$  & $0.00$        & $-0.02^{*}$   \\
                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
oilpricereall4                   & $0.03$        & $0.00$        & $0.01$        & $0.00$        \\
                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
oilpricereall5                   & $-0.01$       & $-0.03^{*}$   & $-0.01$       & $-0.02$       \\
                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
oilpricereall6                   & $0.01$        & $0.01$        & $0.00$        & $0.00$        \\
                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
oilpricereall7                   & $0.01$        & $-0.03$       & $0.02$        & $-0.01$       \\
                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
oilpricereall8                   & $0.01$        & $0.06^{***}$  & $-0.01$       & $0.04^{***}$  \\
                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
EDF: s(peaktoend,recoverableoil) & $27.63^{***}$ & $23.05^{***}$ &               &               \\
                                    & $(29.00)$     & $(29.00)$     &               &               \\
EDF: s(peaktoend,maxprod)        &               &               & $23.82^{***}$ & $28.82^{***}$ \\
                                    &               &               & $(29.00)$     & $(29.00)$     \\
\hline
AIC                                 & 571.09        & 1543.57       & 490.28        & 1421.72       \\
BIC                                 & 690.61        & 1696.16       & 598.01        & 1600.20       \\
Log Likelihood                      & -246.92       & -737.74       & -210.31       & -671.04       \\
Deviance                            & 49634.76      & 6868.08       & 31676.52      & 5599.00       \\
Deviance explained                  & 0.99          & 0.87          & 0.99          & 0.90          \\
Dispersion                          & 395.91        & 11.08         & 245.22        & 9.12          \\
R$^2$                               & 0.98          & 0.86          & 0.99          & 0.88          \\
GCV score                           & 514.74        & 11.67         & 309.43        & 9.69          \\
Num. obs.                           & 163           & 653           & 163           & 653           \\
Num. smooth terms                   & 1             & 1             & 1             & 1             \\
\hline
\multicolumn{5}{l}{\scriptsize{\textsuperscript{***}$p<0.001$, 
  \textsuperscript{**}$p<0.01$, 
  \textsuperscript{*}$p<0.05$}}
\end{tabular}
\caption{Full results for model of field production with price terms in depletion phase.  Using both estimate of total recoverable oil and max production to control for field size}
\label{table:postpeak}
\end{center}
\end{table}

% prepeak

\begin{table}
\begin{center}
\begin{tabular}{l c c c }
\hline
                                     & Small Fields & Large Fields & Pooled \\
\hline
(Intercept)                          & $-0.43$       & $1.26$        & $-0.91$       \\
                                     & $(0.29)$      & $(2.28)$      & $(0.68)$      \\
oilpricereal                       & $0.01$        & $-0.04$       & $0.02$        \\
                                     & $(0.03)$      & $(0.07)$      & $(0.04)$      \\
oilpricereall1                    & $0.03$        & $-0.06$       & $0.10^{**}$   \\
                                     & $(0.04)$      & $(0.09)$      & $(0.03)$      \\
oilpricereall2                    & $-0.01$       & $-0.11$       & $0.05^{*}$    \\
                                     & $(0.03)$      & $(0.11)$      & $(0.02)$      \\
oilpricereall3                    & $-0.07^{*}$   & $-0.11$       & $-0.02$       \\
                                     & $(0.03)$      & $(0.13)$      & $(0.02)$      \\
oilpricereall4                    & $0.06$        & $-0.02$       & $0.07^{***}$  \\
                                     & $(0.03)$      & $(0.12)$      & $(0.02)$      \\
oilpricereall5                    & $0.09^{**}$   & $0.01$        & $0.06^{***}$  \\
                                     & $(0.03)$      & $(0.10)$      & $(0.02)$      \\
oilpricereall6                    & $-0.01$       & $0.07$        & $0.04^{*}$    \\
                                     & $(0.03)$      & $(0.07)$      & $(0.02)$      \\
oilpricereall7                    & $0.01$        & $0.10$        & $0.05^{**}$   \\
                                     & $(0.03)$      & $(0.06)$      & $(0.02)$      \\
oilpricereall8                    & $0.07^{**}$   & $0.06$        & $0.05^{**}$   \\
                                     & $(0.02)$      & $(0.04)$      & $(0.02)$      \\
EDF: s(time_to_peak,recoverable_oil) & $16.91^{***}$ & $28.37^{***}$ & $20.51^{***}$ \\
                                     & $(29.00)$     & $(29.00)$     & $(29.00)$     \\
\hline
AIC                                  & 551.59        & 330.20        & 1534.58       \\
BIC                                  & 638.09        & 421.97        & 1644.24       \\
Log Likelihood                       & -247.89       & -125.73       & -735.78       \\
Deviance                             & 3949.57       & 34833.67      & 300487.73     \\
Deviance explained                   & 0.79          & 0.99          & 0.92          \\
Dispersion                           & 28.81         & 925.75        & 1434.35       \\
R$^2$                                & 0.74          & 0.98          & 0.91          \\
GCV score                            & 34.46         & 1869.82       & 1643.21       \\
Num. obs.                            & 164           & 76            & 240           \\
Num. smooth terms                    & 1             & 1             & 1             \\
\hline
\multicolumn{4}{l}{\scriptsize{\textsuperscript{***}$p<0.001$, 
  \textsuperscript{**}$p<0.01$, 
  \textsuperscript{*}$p<0.05$}}
\end{tabular}
\caption{Full results for model of field production with price terms in depletion phase.}
\label{table:prepeak}
\end{center}
\end{table}

\begin{figure}
	\includegraphics[width=.8\textwidth]{figures/gam_under_2d_plot_1.png}
	\caption{Plot of estimated function of the production profile of small fields, time-to-peak.}
	\label{chart:gam_under_2d_plot_1}
\end{figure}

\begin{figure}
	\includegraphics[width=.8\textwidth]{figures/gam_under_2d_plot_2.png}
	\caption{Plot of estimated function of the production profile of small fields, peak-to-end.}
	\label{chart:gam_under_2d_plot_2}
\end{figure}

\begin{figure}
	\includegraphics[width=.8\textwidth]{figures/gam_over_2d_plot_1.png}
	\caption{Plot of estimated function of the production profile of large fields, time-to-peak.}
	\label{chart:gam_over_2d_plot_1}
\end{figure}

\begin{figure}
	\includegraphics[width=.8\textwidth]{figures/gam_over_2d_plot_2.png}
	\caption{Plot of estimated function of the production profile of large fields, peak-to-end.}
	\label{chart:gam_over_2d_plot_2}
\end{figure}

\begin{figure}
	\includegraphics[width=.8\textwidth]{figures/qq_gam_price_over_2d.png}
	\caption{QQplot of model residuals against theoretical quantiles of normal distribution.  A Gaussian distribution appears to be adequate.}
	\label{chart:qqplot}
\end{figure}

%Fitted vs. residuals
\begin{figure}
	\includegraphics[width=.8\textwidth]{figures/under_fitted.png}
	\caption{Fitted verses actual production for small fields.}
	\label{chart:under_fitted}
\end{figure}

\begin{figure}
	\includegraphics[width=.8\textwidth]{figures/over_fitted.png}
	\caption{Fitted verses actual production for larger fields.}
	\label{chart:under_fitted}
\end{figure}

% Check to see if overfitting of model - change gamma parameter to 1.4
\begin{table}
\begin{center}
\begin{tabular}{l c c }
\hline
                                     & Small Fields & Large Fields \\
\hline
(Intercept)                          & $-0.24^{***}$ & $1.88^{***}$  \\
                                     & $(0.07)$      & $(0.11)$      \\
oilpricereal                       & $0.02$        & $-0.04$       \\
                                     & $(0.01)$      & $(0.02)$      \\
oilpricereall1                    & $-0.01$       & $-0.02$       \\
                                     & $(0.01)$      & $(0.03)$      \\
oilpricereall2                    & $-0.01$       & $0.00$        \\
                                     & $(0.01)$      & $(0.03)$      \\
oilpricereall3                    & $-0.02$       & $0.00$        \\
                                     & $(0.01)$      & $(0.03)$      \\
oilpricereall4                    & $0.02$        & $0.02$        \\
                                     & $(0.01)$      & $(0.03)$      \\
oilpricereall5                    & $0.01$        & $-0.01$       \\
                                     & $(0.02)$      & $(0.03)$      \\
oilpricereall6                    & $-0.04^{*}$   & $0.09^{**}$   \\
                                     & $(0.02)$      & $(0.03)$      \\
EDF: s(time_to_peak,recoverable_oil) & $23.93^{***}$ & $28.58^{***}$ \\
                                     & $(29.00)$     & $(29.00)$     \\
EDF: s(peak_to_end,recoverable_oil)  & $25.91^{***}$ & $6.10^{***}$  \\
                                     & $(28.00)$     & $(28.00)$     \\
\hline
AIC                                  & 59890.49      & 441209.87     \\
BIC                                  & 60170.73      & 441366.08     \\
Log Likelihood                       & -29886.40     & -220561.25    \\
Deviance                             & 4653.55       & 12307.60      \\
Deviance explained                   & 0.80          & 0.83          \\
Dispersion                           & 5.75          & 55.61         \\
R$^2$                                & 0.77          & 0.89          \\
GCV score                            & 6.16          & 66.34         \\
Num. obs.                            & 865           & 264           \\
Num. smooth terms                    & 2             & 2             \\
\hline
\multicolumn{3}{l}{\scriptsize{\textsuperscript{***}$p<0.001$, 
  \textsuperscript{**}$p<0.01$, 
  \textsuperscript{*}$p<0.05$}}
\end{tabular}
\caption{Results when changing the $\gamma$ parameter that controls smoothness to 1.4 in case of overfitting, though this does not significantly change results}
\label{table:gamma_check}
\end{center}
\end{table}

% max production vs. total fields
\begin{figure}
	\includegraphics[width=.8\textwidth]{figures/Max_prod_vs_TRO_print.png}
	\caption{Comparison of fits between using max production and estimate of total reserves to control for field size.}
	\label{chart:Max_prod_vs_TRO_print}
\end{figure}

\end{document}

### 2015-03-10 oil_prices_rr.tex
moved directory from google drive

--- 2015-03-10 oil_prices_old.tex
@@
-\documentclass[12pt]{article}
+% revised paper for Energy Journal
+
+\documentclass[11pt]{article}
+\usepackage[margin=1in]{geometry}
 \usepackage{setspace}
@@
 
- \doublespacing % from package setspacs
+%\doublespacing % from package setspacs
 
@@
 I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 2 to 4\% for a 10 dollar per barrel increase in the real price of oil.
-Most of this effect appears to come in the planning phase of a field's development.
+Most of this effect appears to come in the planning phase of a field's development.\\
+Keywords: Oil production,Oil Prices, Norway, Semiparametric.
 \end{abstract}
@@
 This estimate is heavily biased downwards due to the spurious correlation between the field production profiles and the autocorrelated and non-stationary time series of oil prices.
-A more formal analysis of the bias can be found in the online appendix, \url{jmaurit.github.io#oil_price_sim}, where I simulated data of oil field production in a monte carlo experiment.
-However, the basic problem is that the parametric representation is not flexible enough to control for the production profiles of the fields.
+A more formal analysis of the bias can be found in Appendix B, where I simulate data of oil field production in a Monte Carlo experiment.
+However, the basic idea is that the parametric representation is not flexible enough to control for the production profiles of the fields.
 Instead, a more flexible estimation of the production profile is needed.
@@
 
-% \begin{spacing}{1}
+ \begin{spacing}{1}
 
@@
 \appendix
+
+\section{Appendix: Tables of full results and robustness checks}
 
@@
 \end{tabular}
-\caption{Results when changing the $\gamma$ parameter that controls smoothness to 1.4 in case of overfitting, though this does not significantly change results}
+\caption{Results when changing the $\gamma$ parameter that controls smoothness to 1.4 in case of overfitting, though this does not signficantly change results}
 \label{table:gamma_check}
@@
 
+\FloatBarrier
+
+\section{Appendix: Monte Carlo Experiment}
+
+The data set used in this paper has a complex structure.
+The dependent variable - field level oil production - has a non-linear autocorrelated structure.
+The production profiles are also correlated with each other - leading to a bell-shaped total production curve.
+In addition the exogenous variable of interest - oil prices - are autocorrelated and non-stationary.
+In such a setting, standard rules for inference may not hold.
+
+In this appendix I instead present a Monte Carlo simulation experiment.
+Generating artificial data that is similar in character to the data used, I first show how failing to properly control for the production profile can lead to a biased estimation of the effect of oil prices.
+I then show how using a generalized additive model can lead to better estimates.
+
+I use the R statistical programming language \citet{r_core_team_r:_2013} for the analysis.
+For illustration, I include code snippets in this document.
+The full code for the analysis is available upon request.
+The figures shown are created using the r package ggplot2 \citet{wickham_ggplot2:_2009}.
+
+\subsection{Generating Data}
+
+The first step is generate a set of fields from a size distribution.
+In the simulation, I generate 77 fields from a exponential normal distribution with mean 2.3 and standard deviation of 1.5.
+This is similar to the distribution of the actual 77 productive Norwegian fields.
+\begin{verbatim}
+field_size<-round(exp(rnorm(77, mean=2.3, sd=1.5)), digits=1)
+\end{verbatim}
+
+Using the generated field sizes as an input, I then create a function to generate starting times.
+The goal here is to mimic what is called creaming in the exploration industry - the tendency of finding and starting production on larger fields first.
+For fields that are smaller than 10 million SM3, I generate a starting time from a uniform distribution with endpoints 1975 and 2008.
+For larger fields, I generate a starting time from a normal distribution with standard deviation of 10 and a mean as in formula 1:
+
+\begin{equation}
+mean_start(size) = trunc(1973 + (max + 300)/size)
+\end{equation}
+
+\begin{verbatim}
+gen_year<-function(size, maxsize){
+	#let small fields be distributed uniformly from 1975 to 2013
+	if(size<10){
+		year<-trunc(runif(1, 1975, 2008))	
+	} else{	#while large fields should be more common earlier on
+		range<-FALSE
+		while(range==FALSE){
+			year<-trunc(rnorm(1,mean=(1973+(maxsize+300)/(size+300)), sd=10))
+			ifelse(year>=1970 & year<=2013, range<-TRUE, range<-FALSE)
+			}	
+		}
+return(year)	
+}
+\end{verbatim}
+
+Generating the starting times in this way was essentially ad-hoc.
+However the method is sufficient since the purpose of this exercise is not to model the underlying physical and economic processes, but rather to generate data that shares key similarities with the actual data.
+
+In the simulation I use the actual Brent oil price series shown in figure \ref{oil_price_series}.
+This series can be shown to be non-stationary.
+In order to get reliable inference in a simpler time-series setting the solution might be to take the first difference of the series.
+This is not possible in this setting since it is the actual levels of oil prices that are important.
+Intuitively, identification comes from comparison of fields of similar size and at similar points in their production lifetime but at periods at differing oil prices.
+This variation would be discarded by purely looking at first differences.
+
+\begin{figure}
+	\includegraphics[width=1\textwidth]{figures/oil_price_series.png}
+	\caption{The Real Price of Brent Traded Oil from 1960 to 2013 in 2010 U.S. Dollars.}
+	\label{oil_price_series}	
+\end{figure}
+
+So far, all the simulated data has been outside the Monte Carlo function.
+The data is treated as being fixed.Within the Monte-Carlo loop I create a function that takes as input a fields size, start date as well as the price series and generates a production profile.
+The code for this function is below.
+
+\begin{verbatim}
+gen_production<-function(field, prices){
+	size<-field[1]
+	start<-field[2]
+	name<-field[3]
+
+	names(prices)<-c("year", "price")
+	t<-trunc(sqrt(size)) + 3  #generate production time
+	prod_time<- -t:(t)
+	year=start:(start+2*t-1)
+
+	#use a cumulative logistic function to represent shape of function
+	cum_production = 1/(1+exp(-(prod_time)/3))*size
+
+	#take difference to create production per year
+	prod_shape = diff(cum_production)
+	production<-data.frame(year, prod_shape)
+
+	#Formula: log(production) = f`(time) + beta*log(price) + epsilon
+
+	production$prod<-production$prod_shape*
+	exp(beta0*production$price*rlnorm(length(production$price), meanlog=0, sdlog=.05))
+
+	return(production)
+}
+\end{verbatim}
+
+First, the production time of the field is created as a deterministic function of field size.
+I then model cumulative production from the field over time as a logistic function as in equation \ref{cumProd_equation}.
+This series is then first-differenced to create the general production profile.
+
+\begin{equation}
+% 1/(1+exp(-(prod_time)/3))*size
+cumProd=\frac{size}{1+exp(\frac{-prodTime_t}{3})}
+\label{cumProd_equation}
+
+\end{equation}
+
+I multiply the production profile series with an exponential price term and a coefficient representing the effect of price as well as a random component with mean 0 and standard deviation .05 drawn from a log-normal distribution.
+Taking the logarithm of both sides, the production series for a given field can be written as equation \ref{production_equation}.
+Importantly, the term $\beta$ represents the true effect of price on production.
+
+\begin{equation}
+log(production_t)=log(prodShape_t) + \beta price_t + \epsilon_t
+\label{production_equation}
+\end{equation}
+
+Within the Monte-Carlo loop, production data for each field is generated as above.
+Thus the variation in each run of the Monte Carlo simulation is the random term $\epsilon_t$ from equation \ref{production_equation}.
+
+An example of the data generated can be seen in figure \ref{simulated_production}.
+By inspection, it appears reasonably similar to the data on actual oil production on the Norwegian Continental Shelf.
+
+\begin{figure}
+	\includegraphics[width=1\textwidth]{figures/simulated_production.png}
+	\caption{Simulated production of 77 oil fields}
+	\label{simulated_production}	
+\end{figure}
+
+\subsection{Estimation of Coefficient}
+
+Within the Monte Carlo function, a regression is run using the gam() function from the package mgcv \citet{wood_generalized_2006}.
+This will handle both estimation of linear models or generalized additive models, depending on what formula is passed to the Monte Carlo function.
+
+\begin{verbatim}
+gam_sim<-gam(formula,
+	family=gaussian(link=log), weights=size, data=sim_fields)
+summary(gam_sim)
+\end{verbatim}
+
+Outside of the Monte Carlo function, First I specify a linear regression formula with a 3rd degree polynomial terms to control for the production profile as below.
+
+\begin{verbatim}
+formula_0=formula(prod~time_to_peak + time_to_peak_sq + time_to_peak_cu + 
+	peak_to_end + peak_to_end_sq + peak_to_end_cu +size + price)
+\end{verbatim}
+
+I then replicate the data 1000 times, letting the $\beta$ term equal zero, that is having no effect of prices on oil production.
+The formula used is shown below, where $gam_mc()$ is the Monte Carlo function.
+
+\begin{verbatim}
+gamm_mc_0<-replicate(1000, gam_mc(beta=0, formula=formula_0, use_true_prices=TRUE))
+\end{verbatim}
+
+The above line returns 1000 estimates of $\beta$ from the linear regression specification.  Figure \ref{lin_model_price_mc} shows the results in the form of a histogram.
+
+\begin{figure}
+	\includegraphics[width=1\textwidth]{figures/lin_model_price_mc.png}
+	\caption{Estimated coefficients on price from linear model from Monte Carlo Experiment}
+	\label{lin_model_price_mc}	
+\end{figure}
+
+As the figure shows, a negative bias in the coefficient is apparent.  The distribution appears double-peaked, with most of the estimates of $\beta$ begin between -.05 and -.01.
+
+Now consider a specification using a smooth term for the production profile as below.
+
+\begin{verbatim}
+formula_1= formula(prod~s(prod_time,size) + price)
+\end{verbatim}
+
+Now, the production profile is modeled as a smooth function of production time and size with an additive linear term for price.
+Unlike the specification in the body of the article, I do not specify separate smooth terms for pre- and post-peak.
+This is simply because I know that the logistic function I used to create the shape of the production profile is symmetric.
+
+The results of running the Monte Carlo experiment with the specification with the smoothed term are shown in figure \ref{gam_model_price_mc.png}.
+Here the estimated coefficient on the price term is centered close to zero, correctly reflecting the true $\beta$.
+
+\begin{figure}
+	\includegraphics[width=1\textwidth]{figures/gam_model_price_mc.png}
+	\caption{Estimated coefficients on price from GAM model from Monte Carlo Experiment}
+	\label{gam_model_price_mc.png}	
+\end{figure}
+
 \end{document}

### 2015-03-10 modeling.py

@@ -1,0 +1,51 @@
+#modeling.py
+
+#first run prep.py
+ipython
+
+
+
+%run prep.py #opens data and manipulates, loads standard python libraries
+
+from rpy2.robjects import FloatVector
+from rpy2.robjects import Vector
+from rpy2.robjects import StrVector
+
+from rpy2.robjects.packages import importr
+stats = importr('stats')
+base = importr('base')
+
+ctl = FloatVector([4.17,5.58,5.18,6.11,4.50,4.61,5.17,4.53,5.33,5.14])
+trt = FloatVector([4.81,4.17,4.41,3.59,5.87,3.83,6.03,4.89,4.32,4.69])
+group = base.gl(2, 10, 20, labels = ["Ctl","Trt"])
+weight = ctl + trt
+
+robjects.globalenv["weight"] = weight
+robjects.globalenv["group"] = group
+lm_D9 = stats.lm("weight ~ group")
+print(stats.anova(lm_D9))
+
+# omitting the intercept
+lm_D90 = stats.lm("weight ~ group - 1")
+print(base.summary(lm_D90))
+
+coef=lm_D9.rx2('coefficients')
+
+#Now try with mgcv
+mgcv = importr('mgcv')
+
+#variables
+robjects.globalenv["year_prod"]  = FloatVector(np.array(fields["year_prod"]))
+robjects.globalenv["producing_time"]  = FloatVector(np.array(fields["producing_time"]))
+robjects.globalenv["recoverable_oil"] = FloatVector(np.array(fields["recoverable_oil"]))
+
+#prod_gam_under<-gam(year_prod~s(producing_time, recoverable_oil) + s(peak_to_end, recoverable_oil) +  s(year), 
+#	family=gaussian(link=log), weights=recoverable_oil, data=fields_p[fields_p$max_prod<=split,])
+
+stat_mod=stats.lm("year_prod ~ producing_time + recoverable_oil")
+gam_mod=mgcv.gam("year_prod ~ s(producing_time)", family=gaussian(link=log))
+
+
+
+
+

### 2015-03-10 prep.py

@@ -1,0 +1,79 @@
+#prep.py
+import numpy as np
+import pandas as pd
+import statsmodels.api as sm
+import statsmodels.formula.api as smf
+import matplotlib.pyplot as plt
+
+fields=pd.read_csv("/Users/johannesmauritzen/Google Drive/Research/rOil/data/oil_fields.csv")
+
+del fields["Unnamed: 0"]
+
+#interpret at 10$ change in oil price
+fields["oil_price_real"]=fields["oil_price_real"]/10
+fields["oil_price_real_l1"]=fields["oil_price_real_l1"]/10
+fields["oil_price_real_l2"]=fields["oil_price_real_l2"]/10
+fields["oil_price_real_l3"]=fields["oil_price_real_l3"]/10
+fields["oil_price_real_l4"]=fields["oil_price_real_l4"]/10
+fields["oil_price_real_l5"]=fields["oil_price_real_l5"]/10
+fields["oil_price_real_l6"]=fields["oil_price_real_l6"]/10
+fields["oil_price_real_l7"]=fields["oil_price_real_l7"]/10
+fields["oil_price_real_l8"]=fields["oil_price_real_l8"]/10
+
+fields["diff_oil_price"]=fields["diff_oil_price"]/10
+
+#get rid of fields where tot. production is zero
+fields=fields[fields["tot.prod"]>0]
+
+#datetime format
+fields["producing_from"]=pd.to_datetime(fields["producing_from"], format="%Y-%m-%d")
+fields["producing_to"]=pd.to_datetime(fields["producing_to"], format="%Y-%m-%d")
+fields["year"]=pd.to_datetime(fields["year"], format="%Y")
+
+#drop duplicates
+fields=fields.drop_duplicates(subset=["name", "year"])
+
+#set name and year as multi-index
+
+#finds max year of data frame
+def max_year(df):
+	#test
+	#df=fields
+	#test
+	peak_prod=df.year_prod.max()
+	peak_year=df.year[df.year_prod==peak_prod]
+	return peak_year
+
+fields=fields.reset_index()
+
+maxyear=pd.DataFrame(fields.groupby("name").apply(max_year))
+maxyear=maxyear.reset_index()
+del maxyear["level_1"]
+maxyear.columns=["name", "maxyear"]
+#fields=fields.set_index(["name", "year"])
+
+fields=fields.merge(maxyear, on="name", how="left")
+
+#get rid of NAs 
+fields=fields[fields["year_prod"].notnull()]
+
+#change year to int:
[…]

### 2015-03-10 rr_analysis.py

@@ -1,0 +1,188 @@
+#rr_analysis.py
+
+#attempts using python
+
+#use this?
+#http://patsy.readthedocs.org/en/latest/spline-regression.html
+
+#with mixed models in statsmodels
+
+#attempt using rp2
+import numpy as np
+import pandas as pd
+import statsmodels.api as sm
+import statsmodels.formula.api as smf
+import matplotlib.pyplot as plt
+import random
+
+import rpy2.robjects as robjects
+
+from rpy2.robjects.packages import importr
+
+mgcv = importr("mgcv")
+
+
+# R code:
+# ctl <- c(4.17,5.58,5.18,6.11,4.50,4.61,5.17,4.53,5.33,5.14)
+# trt <- c(4.81,4.17,4.41,3.59,5.87,3.83,6.03,4.89,4.32,4.69)
+# group <- gl(2, 10, 20, labels = c("Ctl","Trt"))
+# weight <- c(ctl, trt)
+
+# anova(lm.D9 <- lm(weight ~ group))
+
+# summary(lm.D90 <- lm(weight ~ group - 1))# omitting intercept
+
+
+
+#rpy2*********
+x = robjects.Vector([4.17,5.58,5.18,6.11,4.50,4.61,5.17,4.53,5.33,5.14])
+y = robjects.Vector([random.randrange(1,10) for i in range(10)])
+
+fmla = robjects.Formula('y ~ s(x) -1')
+env = fmla.environment
+env['x'] = x
+env['y'] = y
+
+mgcv = importr('mgcv')
+fit = mgcv.gam(fmla)
+
+# omitting the intercept
+lm_D90 = r.lm("ctl ~ trt")
+print(r.summary(lm_D90))
+
+#now with mgcv
+fmla = robjects.Formula('y ~ x')
+mgcv.gam()
+
+
+
+
+#Using Stan
+#http://jofrhwld.github.io/blog/2014/07/29/nonlinear_stan.html
+
+import pystan
+import numpy as np
+import pandas as pd
+intercept = 0.785
+deltas = np.array([0, 0.1,0.097, 0.096,0.089]).reshape((5,1))
+durations = np.array([50,60,70,80,90]).reshape((5,1))
+cumsum=deltas.cumsum().reshape((5,1))
+time_mu=cumsum + intercept
+
+durations=np.concatenate((deltas, durations, cumsum, time_mu), axis=1)
+
+test_data = pd.read_table("http://jofrhwld.github.io/data/I_jean.txt")
+
+model_code = """
+  data{
[…]

### 2015-03-10 rr_analysis.r

@@ -1,0 +1,196 @@
+#rr_analysis.r
+
+library(ggplot2)
+library(plyr)
+library(reshape2)
+library(mgcv)
+library(lubridate) 
+library(grid)
+library(boot)
+library(arm)
+library(gamm4)
+library(lme4)
+
+#clear workspace
+rm(list = ls())
+
+#Use time dummies - first three years, first four years
+
+#Use multilevel (random effects) model.  
+
+#Combine many graphs in a single figure. 
+
+#Showing how a parametric model heavily biases estimation - Put in an online appendix.  
+
+#Instead of splitting up into small and large, use mixed effects to take care of heterogeneity.  
+
+#go straight to preferred set up, splitting up in post, build-out and draw-down.  
+
+#Waiting time model - Time to start modeled as a function of time (bayesian model)
+
+#prep.py
+
+fields<-read.csv("/Users/johannesmauritzen/Google Drive/Research/rOil/data/fields_new.csv")
+
+
+
+large_fields<-fields[fields$recoverable_oil>5,]
+large_fields<-large_fields[large_fields$name!="EKOFISK",]
+large_fields<-large_fields[!is.na(large_fields$year_prod),]
+
+#simple GAM
+gam_full<-gam(year_prod~s(producing_time, bs="cr") + s(recoverable_oil) + 
+     s(year, bs="cr") +
+oil_price_real + oil_price_real_l1 + oil_price_real_l2 + 
+oil_price_real_l3 + oil_price_real_l4 + oil_price_real_l5 + oil_price_real_l6 +
+oil_price_real_l7 + oil_price_real_l8, 
+     family=gaussian(link=log), weights=recoverable_oil, data=large_fields)
+summary(gam_full)
+
+gam_2<-gam(year_prod ~ s(producing_time, bs="cr") + recoverable_oil + 
+     s(year, bs="cr") +oil_price_real_l2, 
+     family=gaussian(link=log), weights=recoverable_oil, data=large_fields)
+summary(gam_2)
+
+gam_2_alt<-gam(year_prod ~ s(producing_time, bs="cr") + recoverable_oil + 
+     year + I(year^2) +oil_price_real_l2, 
+     family=gaussian(link=log), weights=recoverable_oil, data=large_fields)
+summary(gam_2_alt)
+
[…]

### 2015-03-10 rr_analysis_1.r

@@ -1,0 +1,10 @@
+#rr_analysis_1.r
+
+#last updated 9. march 2015
+
+fields<-read.csv("/Users/johannesmauritzen/Google Drive/Research/rOil/data/fields_new.csv")
+
+#remove ekofisk
+#remove small fields
+
+#compare amount of oil production

### 2015-03-10 rr_clean.py

@@ -1,0 +1,137 @@
+#rr_clean.py
+#last updated 6. march 2015
+
+import pandas as pd
+import matplotlib.pyplot as plt
+import sys
+import seaborn as sn
+import Quandl
+import numpy as np
+
+#download, merge and clean dataset
+
+brentprices=pd.read_excel("http://www.eia.gov/dnav/pet/hist_xls/RBRTEm.xls", sheetname="Data 1")
+brentprices.columns=["date", "oil_price_nom"]
+brentprices=brentprices[2:]
+brentprices["date"]=pd.to_datetime(brentprices["date"])
+brentprices=brentprices.set_index("date")
+brentprices["oil_price_nom"]=[float(i) for i in brentprices["oil_price_nom"]]
+brentprices["year"]=brentprices.index.year
+brent_yearly=brentprices.groupby("year").mean()
+
+#from bp via Quandl
+
+oil_long = Quandl.get("BP/CRUDE_OIL_PRICES", authtoken="WHFoD2EKU2J8H8yGn_bS")
+oil_long.columns=["oil_price_nom", "oil_price_real"]
+oil_long["year"]=oil_long.index.year
+
+oil_nom=oil_long.reset_index()
+oil_nom=oil_nom[["year", "oil_price_nom"]]
+oil_nom=oil_nom[oil_nom.year<1987].append(brent_yearly.reset_index())
+oil_nom=oil_nom.set_index("year")
+
+#real - deflate
+#investment deflator
+deflator=pd.read_table('http://research.stlouisfed.org/fred2/data/USAGDPDEFAISMEI.txt', skiprows=16, sep=" ")
+deflator.iloc[-5:,2]=deflator.iloc[-5:,1]
+deflator=deflator[[0,2]]
+deflator.columns=["year", "deflator_2010"]
+deflator["year"]=pd.to_datetime(deflator["year"])
+deflator=deflator.set_index("year")
+deflator.index=deflator.index.year
+
+oil_price=oil_nom.merge(deflator, how="inner", left_index=True, right_index=True)
+oil_price["oil_price_real"]=oil_price["oil_price_nom"]*(100/oil_price["deflator_2010"])
+oil_price=oil_price.reset_index()
+oil_price.columns=["year", "oil_price_nom", "deflator_2010", "oil_price_real"]
+
+#from npd
[…]

### 2015-03-10 rr_descriptives.py

@@ -1,0 +1,67 @@
+#rr_descriptives.py
+
+#Descriptives for revise and resubmit
+
+import pandas as pd
+import matplotlib.pyplot as plt
+import matplotlib as mpl
+import numpy as np
+import sys
+import seaborn as sns
+import matplotlib.pyplot as plt
+import random
+import statsmodels.api as sm
+
+#first on individual field data
+field_data=pd.read_csv('/Users/johannesmauritzen/research/oil_prices/data/field_data.csv')
+prod_data=pd.read_csv('/Users/johannesmauritzen/research/oil_prices/data/prod_data.csv')
+
+#remove fields with zero oil
+
+field_data=field_data[field_data["in_place_oil_mill_sm3"]!=0]
+
+#histogram of size, in terms of in place oil
+plt.hist(field_data["in_place_oil_mill_sm3"], 50, facecolor='green', alpha=0.5)
+plt.hist(field_data["recoverable_oil"], 50, facecolor='red', alpha=0.5)
+plt.show()
+
+#show scatter between recoverable and in_place_oil
+fig, ax = plt.subplots()
+ax.scatter(x=field_data["in_place_oil_mill_sm3"], y=field_data["recoverable_oil"])
+plt.show()
+
+# Fit and summarize OLS model
+X=field_data["in_place_oil_mill_sm3"]
+X=sm.add_constant(X)
+Y=field_data["recoverable_oil"]
+mod = sm.OLS(Y,X)
+results = mod.fit()
+fitted=np.column_stack((results.fittedvalues, mod.exog[:,1]))
+x_fit=np.linspace(0, 1200, 100)
+a,b=results.params
+y_fit = a+x_fit*b
+
+
+
+
+fig, ax = plt.subplots()
+ax.scatter(x=field_data["in_place_oil_mill_sm3"], y=field_data["recoverable_oil"])
+ax.plot(x_fit, y_fit)
+ax.annotate('dy/dx~0.52', xy=(400,300), xytext=(300,300))
+fig.set_size_inches(5,4)
+fig.savefig('/Users/johannesmauritzen/research/oil_prices/figures/in_place_vs_recov.png',dpi=100)
+
+plt.show()
+
+print res.summary()
+
+
+sns.lmplot("in_place_oil_mill_sm3", "recoverable_oil", field_data, 
+	scatter_kws={"marker": ".", "color": "slategray"},
[…]

### 2015-03-10 spline_intro.py

@@ -1,0 +1,98 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <headingcell level=1>
+
+# Splines
+
+# <markdowncell>
+
+# <codecell>
+ipython --pylab
+import numpy as np
+import pandas as pd
+import statsmodels.api as sm
+import statsmodels.formula.api as smf
+import matplotlib.pyplot as plt
+
+size=[1.42,1.58,1.78,1.99,1.99,1.99,2.13,2.13,2.13, 2.32,2.32,2.32,2.32,2.32,2.43,2.43,2.78,2.98,2.98]
+wear=np.array([4.0,4.2,2.5,2.6,2.8,2.4,3.2,2.4,2.6,4.8,2.9, 3.8,3.0,2.7,3.1,3.3,3.0,2.8,1.7])
+x=size-np.min(size)
+x=x/np.max(x)
+
+
+%matplotlib inline
+plt.scatter(x, wear)
+
+
+
+# Now I want to create the class for making a spline
+
+
+class simple_spline(object):
+    
+    def __init__(self, data_x, data_y, knots):
+        self.x = data_x
+        self.k = knots
+        self.y = data_y
+        
+    def basis_poly(self, x, z):
+        return ((z-0.5)**2-1/12)*((x-0.5)**2-1/12)/4 - ((abs(x-z)-0.5)**4-(abs(x-z)-0.5)**2/2+7/240)/24
+    
+    def spline_model(self):
+        q=len(self.k)+2
+        n=len(self.x)
+        X=np.ones((n,q))
+        X[:, 1] = self.x
+        
+        temp_X = []
+        for i in self.x:
+            for j in self.k:
+                temp_X.append(self.basis_poly(i, j))
+        temp_X=np.matrix(temp_X).reshape((n, q-2))
+        X[:,2:q] = temp_X
+        return(X)
+        
+        
+
+k=[i/5.0 for i in range(1,5)]
+spline1=simple_spline(x,wear,k)
+
+X_spline = spline1.spline_model()
+results = sm.OLS(wear, X_spline).fit()
+results.summary()
+
+xp = [i/100.0 for i in range(1,100)]
+Xp=simple_spline(xp, wear, k).spline_model()
+plot()
+
+y_hat = np.dot(Xp,results.params)
+
+
+
+# for penalized least squares
+
+def basis_poly(x, z):
[…]

### 2015-03-11 industry_costs.py
added cost measure

@@ -1,0 +1,50 @@
+#costs.py
+
+import pandas as pd
+import matplotlib.pyplot as plt
+import sys
+import seaborn as sn
+import Quandl
+import numpy as np
+import statsmodels as sm
+
+#producer price index:ssb, norway 2000 dollars
+ppi=pd.read_csv("http://data.ssb.no/api/v0/dataset/27263.csv?lang=en")
+ppi_oandg=ppi[ppi["industry/commodity group"]=="SNN06 Extraction of oil and natural gas"]
+ppi_oandg=ppi_oandg[ppi_oandg["market"]=="1 Domestic and export market, total"]
+ppi_oandg=ppi_oandg[ppi_oandg["contents"]=="Producer price index (2000=100)"]
+ppi_oandg=ppi_oandg[["time","Producer price index. Oil/ gas extraction, manufacturing, mining and electricity by market, industry/commodity group, time and contents"]]
+ppi_oandg.columns=["period", "ppi_oandg"]
+
+year=[]
+month=[]
+for i in ppi_oandg["period"]:
+	year.append(int(i[:4]))
+	month.append(int(i[5:]))
+
+ppi_oandg["year"]=year
+ppi_oandg["month"]=month
+
+#convert to float
+ppi_oandg["ppi_oandg"]=[float(i) for i in ppi_oandg["ppi_oandg"]]
+ppi_yearly=ppi_oandg[["year", "ppi_oandg"]].groupby("year").mean()
+
+#merge with prices
+ppi_yearly=oil_price.merge(ppi_yearly.reset_index(), on="year", how="outer")
+
+#info from eia
+#Costs of Crude Oil and Natural Gas Wells Drilled
+#All (Real*) Thousand dollars per well
+#http://www.eia.gov/dnav/pet/pet_crd_wellcost_s1_a.htm
+
+well_costs=pd.read_csv("research/oil_prices/data/well_costs.csv", sep=";", decimal=",")
+well_costs["cost_index"]=well_costs["cost"]/754.6*100
+
+cost_index=well_costs.merge(ppi_yearly.reset_index(), on="year", how="outer")
+cost_index["ppi_oandg"][cost_index["ppi_oandg"].isnull()]=cost_index["cost_index"][cost_index["ppi_oandg"].isnull()]
+
+cost_index=cost_index[["year", "cost_index"]]
+cost_index.to_csv("research/oil_prices/data/cost_index.csv")
+
[…]

### 2015-03-11 rr_analysis.py

@@ -14,12 +14,7 @@ import statsmodels.api as sm
 import statsmodels.formula.api as smf
 import matplotlib.pyplot as plt
 import random
-
-import rpy2.robjects as robjects
-
-from rpy2.robjects.packages import importr
-
-mgcv = importr("mgcv")
+import math
 
 
 # R code:
@@ -32,27 +27,12 @@ mgcv = importr("mgcv")
 
 # summary(lm.D90 <- lm(weight ~ group - 1))# omitting intercept
 
-
+#try https://github.com/thehackerwithin/Python2010/wiki/PyBCSession08
 
 #rpy2*********
-x = robjects.Vector([4.17,5.58,5.18,6.11,4.50,4.61,5.17,4.53,5.33,5.14])
-y = robjects.Vector([random.randrange(1,10) for i in range(10)])
-
-fmla = robjects.Formula('y ~ s(x) -1')
-env = fmla.environment
-env['x'] = x
-env['y'] = y
 
 mgcv = importr('mgcv')
 fit = mgcv.gam(fmla)
-
-# omitting the intercept
-lm_D90 = r.lm("ctl ~ trt")
-print(r.summary(lm_D90))
-
-#now with mgcv
-fmla = robjects.Formula('y ~ x')
-mgcv.gam()
 
 
 

### 2015-03-11 rr_analysis_1.r

@@ -1,10 +1,33 @@
 #rr_analysis_1.r
 
-#last updated 9. march 2015
+#last update march 11th, 2015
+#First new modeling
+rm(list = ls())
 
-fields<-read.csv("/Users/johannesmauritzen/Google Drive/Research/rOil/data/fields_new.csv")
+library(ggplot2)
+library(dplyr)
+library(reshape2)
+library(mgcv)
+library(lubridate) 
+library(grid)
+library(boot)
+library(arm)
+library(gamm4)
+library(lme4)
 
-#remove ekofisk
-#remove small fields
+prod_data<-read.csv('/Users/johannesmauritzen/research/oil_prices/data/prod_data.csv')
 
-#compare amount of oil production
+head(prod_data)
+
+#all fields
+
+gam_full<-gam(oil_prod_mill_sm3~s(producing_time, bs="cr") + s(recoverable_oil) + 
+     s(year, bs="cr") + price + price_l1 + price_l2 + price_l3 + price_l4 + 
+     price_l5 + price_l6 + price_l7 + price_l8, 
+     family=gaussian(link=log), weights=recoverable_oil, data=prod_data)
+summary(gam_full)
+
+
+
+
+

### 2015-03-11 rr_clean.py

@@ -7,6 +7,7 @@ import sys
 import seaborn as sn
 import Quandl
 import numpy as np
+import statsmodels as sm
 
 #download, merge and clean dataset
 
@@ -133,5 +134,17 @@ oil_price.columns=["year", "deflator", "price", "price_l1",
 prod_data=oil_prod.merge(oil_price, on="year")
 prod_data=prod_data.merge(yearly_investments, on=["name", "year"])
 prod_data=prod_data.merge(field_data, on="name")
+
+#add data on cost
+
+cost_index=pd.read_csv("research/oil_prices/data/cost_index.csv")
+
+#add production year data
+def prod_year(group):
+	return(pd.DataFrame([i for i in range(len(group["name"]))]))
+
+prod_data["prod_year"]=prod_data.groupby("name").apply(prod_year)
+prod_year.reset_index(inplace=True)
+prod_data["prod_year"]=prod_year.iloc[:,2]
 
 prod_data.to_csv('/Users/johannesmauritzen/research/oil_prices/data/prod_data.csv')

### 2015-03-12 industry_costs.py
update work on modeling

@@ -29,22 +29,33 @@ ppi_oandg["month"]=month
 ppi_oandg["ppi_oandg"]=[float(i) for i in ppi_oandg["ppi_oandg"]]
 ppi_yearly=ppi_oandg[["year", "ppi_oandg"]].groupby("year").mean()
 
+oil_price=pd.read_csv("research/oil_prices/data/oil_price.csv")
 #merge with prices
 ppi_yearly=oil_price.merge(ppi_yearly.reset_index(), on="year", how="outer")
-
+ppi_yearly["year"]=ppi_yearly["year"].apply(int)
 #info from eia
 #Costs of Crude Oil and Natural Gas Wells Drilled
 #All (Real*) Thousand dollars per well
 #http://www.eia.gov/dnav/pet/pet_crd_wellcost_s1_a.htm
 
 well_costs=pd.read_csv("research/oil_prices/data/well_costs.csv", sep=";", decimal=",")
-well_costs["cost_index"]=well_costs["cost"]/754.6*100
+well_costs["well_costs"]=well_costs["cost"]/754.6*100
 
-cost_index=well_costs.merge(ppi_yearly.reset_index(), on="year", how="outer")
-cost_index["ppi_oandg"][cost_index["ppi_oandg"].isnull()]=cost_index["cost_index"][cost_index["ppi_oandg"].isnull()]
+cost_index=well_costs.merge(ppi_yearly, on="year", how="outer")
+cost_index=cost_index.sort(columns="year")
+cost_index["cost_index"]=cost_index["ppi_oandg"]
+cost_index["cost_index"][cost_index["ppi_oandg"].isnull()]=cost_index["well_costs"][cost_index["ppi_oandg"].isnull()]
 
-cost_index=cost_index[["year", "cost_index"]]
+#make plot
+fig, (ax1, ax2) = plt.subplots(2, sharex=True, sharey=False)
+fig.set_size_inches(8,6)
+ax1.plot(cost_index["year"], cost_index["cost_index"])
+ax1.set_xlabel("Real Cost Index, Oil and Gas Extraction, 2000=100")
+ax2.plot(cost_index["year"], cost_index["oil_price_real"])
+ax2.set_xlabel("Real Price of Crude Oil, 2010 Dollars")
+plt.savefig("research/oil_prices/figures/oil_cost_and_price.png", dpi=160)
+
+
[…]

### 2015-03-12 rr_analysis_1.r

@@ -1,6 +1,6 @@
 #rr_analysis_1.r
 
-#last update march 11th, 2015
+#last update march 12th, 2015
 #First new modeling
 rm(list = ls())
 
@@ -14,6 +14,7 @@ library(boot)
 library(arm)
 library(gamm4)
 library(lme4)
+library(nlme)
 
 prod_data<-read.csv('/Users/johannesmauritzen/research/oil_prices/data/prod_data.csv')
 
@@ -21,13 +22,61 @@ head(prod_data)
 
 #all fields
 
-gam_full<-gam(oil_prod_mill_sm3~s(producing_time, bs="cr") + s(recoverable_oil) + 
-     s(year, bs="cr") + price + price_l1 + price_l2 + price_l3 + price_l4 + 
+large_fields<-prod_data[prod_data["recoverable_oil"]>3,]
+large_fields<-prod_data[prod_data["name"]!="EKOFISK",]
+
+
+
+gam_full<-gam(oil_prod_mill_sm3~s(prod_year) + 
+	in_place_oil_mill_sm3 + cost_index +
+    year + I(year^2) + price + price_l1 + price_l2 + price_l3 + price_l4 + 
      price_l5 + price_l6 + price_l7 + price_l8, 
-     family=gaussian(link=log), weights=recoverable_oil, data=prod_data)
+     family=gaussian(link=log), weights=in_place_oil_mill_sm3, data=large_fields, 
+     select=TRUE)
 summary(gam_full)
 
+gam_full2<-gam(oil_prod_mill_sm3~s(prod_year, bs="cr") + 
+	in_place_oil_mill_sm3 + cost_index +
+    s(year, bs="cr", k=4) + price + price_l1 + price_l2 + price_l3 + price_l4 + 
+     price_l5 + price_l6 + price_l7 + price_l8, 
+     family=gaussian(link=log), weights=recoverable_oil, 
+     data=large_fields)
+summary(gam_full2)
 
+gam_full3<-gam(oil_prod_mill_sm3~s(prod_year, in_place_oil_mill_sm3) + 
+	 cost_index +
+     year + I(year^2) + price + price_l1 + price_l2 + price_l3 + price_l4 + 
+     price_l5 + price_l6 + price_l7 + price_l8, 
+     family=gaussian(link=log), weights=in_place_oil_mill_sm3, data=large_fields)
+summary(gam_full3)
 
+#With random effects
+
[…]

### 2015-03-12 rr_clean.py

@@ -45,7 +45,7 @@ oil_price=oil_nom.merge(deflator, how="inner", left_index=True, right_index=True
 oil_price["oil_price_real"]=oil_price["oil_price_nom"]*(100/oil_price["deflator_2010"])
 oil_price=oil_price.reset_index()
 oil_price.columns=["year", "oil_price_nom", "deflator_2010", "oil_price_real"]
-
+oil_price.to_csv("research/oil_prices/data/oil_price.csv", index=False)
 #from npd
 overview=pd.read_csv("http://factpages.npd.no/ReportServer?/FactPages/TableView/field&rs:Command=Render&rc:Toolbar=false&rc:Parameters=f&rs:Format=CSV&Top100=false&IpAddress=158.37.94.112&CultureCode=en")
 month_prod=pd.read_csv("http://factpages.npd.no/ReportServer?/FactPages/TableView/field_production_monthly&rs:Command=Render&rc:Toolbar=false&rc:Parameters=f&rs:Format=CSV&Top100=false&IpAddress=158.37.94.56&CultureCode=nb-no")
@@ -136,15 +136,16 @@ prod_data=prod_data.merge(yearly_investments, on=["name", "year"])
 prod_data=prod_data.merge(field_data, on="name")
 
 #add data on cost
-
+#prod_data=pd.read_csv('/Users/johannesmauritzen/research/oil_prices/data/prod_data.csv')
+#from data cleaned from industry_cost.py
 cost_index=pd.read_csv("research/oil_prices/data/cost_index.csv")
+prod_data=prod_data.merge(cost_index[["year", "cost_index"]], on="year", how="left")
 
+prod_data=prod_data.sort(columns=["name", "year"])
 #add production year data
-def prod_year(group):
-	return(pd.DataFrame([i for i in range(len(group["name"]))]))
+def create_prod_year(year):
+	return(pd.DataFrame({"prod_year":[i for i in range(len(year))]}))
 
-prod_data["prod_year"]=prod_data.groupby("name").apply(prod_year)
-prod_year.reset_index(inplace=True)
-prod_data["prod_year"]=prod_year.iloc[:,2]
+prod_data['prod_year']=prod_data.groupby("name")["year"].transform(create_prod_year)
 
-prod_data.to_csv('/Users/johannesmauritzen/research/oil_prices/data/prod_data.csv')
+prod_data.to_csv('/Users/johannesmauritzen/research/oil_prices/data/prod_data.csv', index=False)

### 2015-03-13 oil_prices_rr.tex
some modeling work

@@
 
-Here the left hand side variable is yearly production in year $t$ for field $i$.
-To try to account for the time profile of production, I split up model-time into a time-to-peak and peak-to-end variable as demonstrated for data for the Statfjord field in figure \ref{statfjord_dem} and then represent each as a cubic function.
-Yearly production is assumed to be proportional to the total size of the field as represented by the estimate of the total recoverable resource.
-Finally, a term for the oil price as well as five lagged terms are added in order to capture the effects of price.
-
-\begin{figure}
-\includegraphics[width=1\textwidth]{figures/statfjord_dem_print.png}
-\caption{The production profile can be modeled in two phases - time-to-peak, representing the build-out phase of the oil field and time-from-peak, representing the period of depletion.
-The is illustrated with data from the Statfjord field.}
-\label{statfjord_dem}
-\end{figure}
-
-Figure \ref{glm_dirty_box} shows the estimates of the coefficients on the oil price and its lags \footnote{The dots on the figure represent 1000 simulations of the estimated coefficient based on the estimated standard error and point estimate from the model}.
-The lags are not estimated to have an effect significantly different from zero.
-However a literal interpretation of the coefficient on the concurrent oil price term would indicate that a 10-dollar increase in the oil price would lead to a lowering of field production of around 4-5\%.
-
-\begin{figure}
-\includegraphics[width=1\textwidth]{figures/glm_dirty_box_print.png}
-\caption{A generalized linear model generates a negative estimate of the coefficient on price in a regression on production.
-This result is spurious, caused by a failure to properly control for the production profiles of the fields.}
-\label{glm_dirty_box}
-\end{figure}
-
 This estimate is heavily biased downwards due to the spurious correlation between the field production profiles and the autocorrelated and non-stationary time series of oil prices.
@@
 The simplest possible model would then have the form of equation \ref{simp_eqn}.
-
-\begin{equation}
-Production_{t}=f(time) + \epsilon
-	\label{simp_eqn}
-\end{equation}
-
-Again considering the production profile for the Statfjord field, a smoothed function might look like the black line in figure \ref{statfjord_gam}.
-
-\begin{figure}
-	\includegraphics[width=1\textwidth]{figures/statfjord_gam_print.png}
-	\caption{An illustration of fitting a simple non-parametric function to the production profile of a single field.}
-	\label{statfjord_gam}
-\end{figure}
 
@@
 
-In my regression, I find it helpful to represent the production profile of fields as a two-dimensional function.
-This can not be accomplished with a standard cubic regression spline, but I can instead use a Thin Plate (Regression) Spline \citep{wood_thin_2003}.
-Though the full details of the implementation are beyond the scope of this paper, for the basic idea consider equation \ref{thin_plate_1}.
-
-	\begin{equation}
-	y_i = g(x_1, x_2)
-	\label{thin_plate_1}
-	\end{equation}
-
-Following \citet{wood_generalized_2006}, $g$ is the function of $x_1$ and $x_2$ that is to be estimated by $f$, which in turn is estimated by minimizing equation \ref{thin_plate_2}.
-Here $\boldsymbol{y}$ represents a vector of $y_i$’s and $\boldsymbol{f} = (f(\boldsymbol{x_1}),f(\boldsymbol{x_2}))^t$.
-
-	\begin{equation}
-\min \|\boldsymbol{y-f}\|^2 + \lambda J_{22}(f)
-\label{thin_plate_2}
-	\end{equation}
-
-$J_{22}$ represents the penalty function for the smoothness of the function which can be written as in \ref{thin_plate_3}.
-The $22$ represents the fact that it is a penalty function of two variables with smoothness measured by the second derivative.
-
-	\begin{equation}
-	J_{22}{f}= \diffp[2]{f}{x_1}^2 + \diffp[2]{f}{{x_1}{x_2}} + \diffp[2]{f}{x_2}^2dx_1 dx_2
-\label{thin_plate_3}
-	\end{equation}
-
-In short, a function of $x_1$ and $x_2$ is found that is minimizing errors in the sense of minimizing Euclidean distance subject to a penalty function of “wiggiliness”.
-The actual implementation is somewhat more involved in order to increase the computational efficiency of the estimation.
-For further details I again refer to \citet{wood_thin_2003}.
-
 The advantage to using splines over other smoothing methods is that it can be represented in a linear form.
@@
 Also like the parametric model I split the production-time component element in two: up to and after the peak in production.
-While a shape for the entire production profile could be estimated with one smoothed function, splitting it up allows for more flexibility and better overall fit of the model as estimated by deviance score and the related estimated degrees of freedom of the model.
-More so, production up to a peak and the subsequent decline represents two distinct processes.
-The first is driven by the physical investment and build-out of the field, while the latter is dominated by depletion of the field and subsequent drop in pressure.
-It makes sense to model these two processes separately, and later I will also attempt to estimate separate price effects for the two phases.
-
-I also allow the smoothed functions to vary with the total size of the field as measured by the estimated total recoverable oil since the shape of the production profile tends to vary substantially by field size.
-Inspection of a selection of fields, such as shown in figure \ref{field_inspection}, shows that smaller fields tend to reach their peak quickly while larger fields take more time.
-
-The estimated field-size variable is however an imperfect measure.
-First, estimates for newer fields will likely have higher measurement errors than established fields that are well explored and where a significant amount of the oil has already been extracted.
-The total reserves estimates are also likely correlated with the price terms as estimates of recoverable reserves likely change depending on the level of the oil price.
-Nonetheless, since I am only using the variable to help to control for the general shape of the production profiles of fields and not as an exogenous regressor, the inclusion of the variable should not materially affect the estimation.
-
-\begin{figure}
-	\includegraphics[width=1\textwidth]{figures/field_inspection_print.png}
-	\caption{A selection of production profiles of oil fields.  The shape can vary substantially by field size as well as other factors.}
-	\label{field_inspection}
-\end{figure}
-
-Even with a smoothed function that is allowed to vary by field-size, a substantially better fit, as well as more nuanced analysis could be obtained by splitting the estimation into small and large fields, this time measured by the maximum production of the field - a measure of field size that is less likely to be highly correlated with price.
-The improvement in fit can be seen by inspecting the fitted values of the models as in figure \ref{bench_vs_split}.
-The split estimation provides a particularly better fit for smaller and mid-sized fields.
-
-\begin{figure}
-	\includegraphics[width=1\textwidth]{figures/bench_vs_split_print.png}
-	\caption{A benchmarking model shows that a substantially better fit can be achieved by splitting the analysis into small and large fields, where a somewhat arbitrary maximum yearly production of 8 million SMU is used as the separating value.}
-	\label{bench_vs_split}
-\end{figure}
-
-Plots of the estimated smoothed functions in the model are not directly informative about the effects of price and so I do not include them in the main text.
-However they can be helpful in gaining a intuition behind the estimation procedure so I have included them in appendix 2 along with other model diagnostics and robustness checks.
 
@@
 
-\begin{figure}
-	\includegraphics[width=1\textwidth]{figures/gam_price_8_print.png}
-	\caption{The coefficient estimates on the price terms where two additional lags are added to the model.
-Significant positive results are found on the 8th lag for both large and small fields, providing evidence for adaptive price expectations by producers.}
-	\label{gam_price_8}
-\end{figure}
-
-The estimated coefficients for the concurrent and first three lags of the oil price are not estimated to be significantly different than zero.
-For large fields a modest effect at the fourth, sixth and seventh lags are estimated, though only at a 10\% significance level, \footnote{The lines in the boxplot represent conservative 95\% confidence intervals and are calculated using a bayesian-inspired simulation from the posterior distribution of the coefficients.
-P-values (stars) shown in the table in the appendix were calculated by standard asymptotic methods and show significant results for the 4th and 6th lags at the 5\% level.} while for small fields an effect is estimated only at the sixth lags.
-The magnitude of these effects is in the neighborhood of a 2\% increase in production for a 10 dollar increase in the oil price.
-A larger and more significant effect is found for both small and large fields at the 8th lag, with a magnitude implying an approximately 3 to 4\% increase in production for a 10 dollar increase in the oil price.
-
-Despite the clear implications of most simplified Hotelling models, the lack of significant results for the concurrent price and the modest lagged effect is not surprising.
-In general, operating an oil production rig and related offshore infrastructure is an extremely expensive venture with high fixed costs.
-In the challenging conditions of the North Sea, the expenses are multiplied.
-Thus any short-term benefit of strategically altering production in relation to movements in the oil price are dominated by the large costs of having excess pumping capacity.
-In other words, oil producers have a strong incentive to pump as much oil out at any given time given the existing production capacity.
-
-Instead, the results point to a mechanism where oil producers react to higher oil prices by increasing investment in those fields, leading to higher oil production only with a considerable lag.
-This story is in line with a trend of increased total extraction estimates from the Norwegian continental shelf as a whole as well as from existing fields over the last 15 years of strongly rising oil prices \footnote{\url{http://npd.no/Templates/OD/Article.aspx?id=4731&epslanguage=en}}.
-
-I prefer the un-pooled models I have presented above because of the improvement in model fit and because operators of small and large fields may be expected to have substantially different reactions to changes in price as well as different time profiles of production.
-Nonetheless, it can be instructive to show results from a pooled model where a fixed effect and interaction variable is introduced to distinguish the effects of small and large fields.
-The model can be written as equation \ref{pooled_eqn}.
-The terms in the model are mostly the same as in \ref{gam_price_eqn} but with an added dummy term for small fields - those with a maximum yearly production of 8 million SM3 or below, $smallfield$, and interaction terms.
+\begin{table}
+\begin{center}
+\begin{tabular}{l c c c }
+\hline
+                                 & w/out Rand. Effects & w Rand. Effects & Lags 1-5 \\
+\hline
+(Intercept)                      & $0.335$        & $-4.016^{***}$ & $-3.971^{***}$ \\
+                                 & $(0.265)$      & $(0.297)$      & $(0.212)$      \\
+I(in\_place\_oil\_mill\_sm3/100) & $0.337^{***}$  & $-0.014$       & $-0.015$       \\
+                                 & $(0.009)$      & $(0.022)$      & $(0.022)$      \\
+I(cost\_index/10)                & $-0.018^{**}$  & $-0.030^{***}$ & $-0.030^{***}$ \\
+                                 & $(0.006)$      & $(0.006)$      & $(0.005)$      \\
+I(price/10)                      & $-0.006$       & $0.014$        &                \\
+                                 & $(0.017)$      & $(0.016)$      &                \\
+I(price\_l1/10)                  & $0.004$        & $0.021$        & $0.029^{*}$    \\
+                                 & $(0.017)$      & $(0.017)$      & $(0.014)$      \\
+I(price\_l2/10)                  & $0.037^{*}$    & $0.047^{**}$   & $0.048^{***}$  \\
+                                 & $(0.017)$      & $(0.016)$      & $(0.014)$      \\
+I(price\_l3)                     & $0.003$        & $0.003$        & $0.002$        \\
+                                 & $(0.002)$      & $(0.001)$      & $(0.001)$      \\
+I(price\_l4)                     & $0.000$        & $0.004^{**}$   & $0.004^{***}$  \\
+                                 & $(0.002)$      & $(0.001)$      & $(0.001)$      \\
+I(price\_l5/10)                  & $-0.020$       & $0.012$        & $0.019$        \\
+                                 & $(0.013)$      & $(0.012)$      & $(0.010)$      \\
+I(price\_l6/10)                  & $-0.010$       & $0.008$        &                \\
+                                 & $(0.012)$      & $(0.011)$      &                \\
+I(price\_l7/10)                  & $-0.008$       & $0.007$        &                \\
+                                 & $(0.012)$      & $(0.012)$      &                \\
+I(price\_l8/10)                  & $-0.044^{***}$ & $-0.011$       &                \\
+                                 & $(0.011)$      & $(0.011)$      &                \\
+EDF: s(prod\_year)               & $6.381^{***}$  &                &                \\
+                                 & $(7.249)$      &                &                \\
+EDF: s(year)                     & $2.768^{***}$  & $3.000^{***}$  & $3.000^{***}$  \\
+                                 & $(2.956)$      & $(3.000)$      & $(3.000)$      \\
+EDF: s(prod\_year,name)          &                & $33.002^{***}$ & $33.014^{***}$ \\
+                                 &                & $(72.000)$     & $(72.000)$     \\
+\hline
+AIC                              &                &                &                \\
+BIC                              &                &                &                \\
+Log Likelihood                   &                &                &                \\
+Deviance                         & 2306267.814    & 1976585.232    & 1980468.734    \\
+Deviance explained               & 0.848          & 0.870          & 0.870          \\
+Dispersion                       & 1785.243       & 1562.520       & 1560.671       \\
+R$^2$                            & 0.841          & 0.865          & 0.866          \\
+GCV score                        & 1814.469       & 1621.811       & 1614.802       \\
+Num. obs.                        & 1088           & 1088           & 1088           \\
+Num. smooth terms                & 2              & 2              & 2              \\
+\hline
+\multicolumn{4}{l}{\scriptsize{$^{***}p<0.001$, $^{**}p<0.01$, $^*p<0.05$}}
+\end{tabular}
+\caption{Statistical models}
+\label{GAM_model_table}
+\end{center}
+\end{table}
 
@@
 
-The full results for the estimates of all the parametric terms can be found in table \ref{table:pooled} in the appendix.
-The estimated coefficients on the price terms are shown in figure \ref{gam_price_pooled}.
-In the regression none of the interaction effects were significantly different than zero, nor were the concurrent price term or the first three lags.
-The fourth and sixth through eighth lag were however significant and positive, again at a magnitude of approximately 2\% increase in production for a \$10 increase in the oil price.
-
-\begin{figure}
-	\includegraphics[width=1\textwidth]{figures/gam_price_pooled_print.png}
-	\caption{The estimated coefficients on the price terms for a pooled model of oil field production.  Significant estimates are found on the fourth, sixth, seventh and eighth price lags}
-	\label{gam_price_pooled}
-\end{figure}
-
-In the above discussion, I have been implicitly giving the coefficients a causal interpretation, which deserves some discussion.
-My main identifying assumption is that production at the field level can not cause significant changes in the oil price.
-A stronger but still relevant assumption that may be necessary since field production is correlated across fields is that total production from the Norwegian continental shelf does not affect oil price.
-
-Both of these assumptions are likely satisfied.
-Oil is a globally traded commodity and total Norwegian production accounts only for a small fraction of total production.
-In 2013 Norwegian production made up only 2.3 \% of the world total.
-\footnote{\url{http://www.eia.gov/countries/country-data.cfm?fips=NO#pet}}  A drastic change in production, on par with the halt in production that occurred in Libya in 2011, \footnote{Before the Libyan revolution of 2011, Norwegian and Libyan yearly oil production were of a similar magnitude, 2.1 versus 1.8 million barrels of oil in 2010 \url{http://www.eia.gov/countries/country-data.cfm?fips=LY#pet}} would have been required to have had any significant effect on world oil supply and in turn prices.
-
-One of the main implications of the interpretation that higher oil prices leads to increased extraction through increased investment is that while the effect on production will be lagged, higher oil prices will have a more immediate effect on investments.
-I test for this with the model as written in equation \ref{gam_invest_eqn}
-
-\begin{equation}
-\begin{split}
-	Log(Investment_{i,t})&=f(timetopeak_{i,t}, totalrecoverableoil_i) \\
-	 \quad &+ f(peaktoend_{i,t}, totalrecoverableoil_i) \\
-	 \quad &+ \alpha oilproduction_{i,t} \\
-	 \quad & + \beta_1 oilprice + \beta_2 oilpricel1 + ... + oil_pricel8 \\
-	 \quad &+ \delta smallfield \\
-	 \quad &+ \gamma_1 oilprice*smallfield + \gamma_2 oilprice*smallfield + ... 
-	+  \epsilon \\
-\label{gam_invest_eqn}
-\end{split}
-\end{equation}
-
-In the equation investment in each field $i$ at time $t$ is a function of the state of field development, as modeled by a non-parametric function of time to and from the peak as well as the total size of the field - as in the model of oil field production.
-In addition I include a term for oil production in field $i$ at time $t$.
-The coefficients of interest are again those on the oil price and its lags which I interact with a dummy variable for large and small fields, $smallfield$.
-
-The results for the estimation of the estimated coefficients on the oil price and its lags is shown in figure \ref{gam_price_invest_box} while full results are also shown in table \ref{table:investment} in appendix 1.
-The coefficient on the concurrent oil price as well as the third and fourth lags are all significantly positive, with coefficients that can be interpreted to mean that a 10 dollar increase in the oil price leads to between a 5 to 10 \% increase in oil field investment in the concurrent and second subsequent year and between a 10 to 20\% increase in the third subsequent year.
-
-\begin{figure}
-	\includegraphics[width=1\textwidth]{figures/invest_pooled_print.png}
-	\caption{Estimated coefficients for the price terms in a model of oil field investments.
-A rise in the oil price leads to an increase in investment in the concurrent year as well as in the second and third lag.
-Negative coefficients are found for the sixth through 8th lags.}
-	\label{gam_price_invest_box}
-\end{figure}
-
-The results of the regression on investment fits nicely with the previous results on oil production which showed a significant effect in the 4th through 6th lags.
-A story consistent with both results is that higher oil prices induce increased investment in production capacity, though it takes time to get the extra capacity in place and the effects on production to be felt.
-
-The estimated coefficients on the 6th, 7th and 8th lagged coefficients are shown to be significantly negative.
-An explanation is that a higher oil price induces producers to speed up a build out of a field, leading to more investment earlier, but relatively less in later years.
-An implication of this theory is that the effect of price will primarily be in the build-out phase of a fields production profile.
-This is explored in the following section.
-
 \section{The effect of price in the phases of oil production}
@@
 Changing the level of investments in reaction to a change in prices after plans have been finalized is likely to lead to both an expensive delay as well as other costs that heavily outweigh any benefits of increased future production.
-
 \section{Conclusion}

### 2015-03-13 rr_analysis_1.r

@@ -15,68 +15,116 @@ library(arm)
 library(gamm4)
 library(lme4)
 library(nlme)
+library(texreg)
+
+field_data<-read.csv('/Users/johannesmauritzen/research/oil_prices/data/field_data.csv')
+
+#How much of the total oil reserves are we excluding
+tot_rec_oil<-sum(field_data["recoverable_oil"])
+small_rec_oil<-sum(field_data["recoverable_oil"][field_data["recoverable_oil"]<=3])
+28.3/4779.2
+#.5% of total recoverable oil.  
 
 prod_data<-read.csv('/Users/johannesmauritzen/research/oil_prices/data/prod_data.csv')
-
-head(prod_data)
-
-#all fields
 
 large_fields<-prod_data[prod_data["recoverable_oil"]>3,]
 large_fields<-prod_data[prod_data["name"]!="EKOFISK",]
 
 
-
-gam_full<-gam(oil_prod_mill_sm3~s(prod_year) + 
-	in_place_oil_mill_sm3 + cost_index +
-    year + I(year^2) + price + price_l1 + price_l2 + price_l3 + price_l4 + 
-     price_l5 + price_l6 + price_l7 + price_l8, 
-     family=gaussian(link=log), weights=in_place_oil_mill_sm3, data=large_fields, 
-     select=TRUE)
-summary(gam_full)
-
-gam_full2<-gam(oil_prod_mill_sm3~s(prod_year, bs="cr") + 
-	in_place_oil_mill_sm3 + cost_index +
-    s(year, bs="cr", k=4) + price + price_l1 + price_l2 + price_l3 + price_l4 + 
-     price_l5 + price_l6 + price_l7 + price_l8, 
-     family=gaussian(link=log), weights=recoverable_oil, 
-     data=large_fields)
-summary(gam_full2)
-
-gam_full3<-gam(oil_prod_mill_sm3~s(prod_year, in_place_oil_mill_sm3) + 
-	 cost_index +
-     year + I(year^2) + price + price_l1 + price_l2 + price_l3 + price_l4 + 
-     price_l5 + price_l6 + price_l7 + price_l8, 
-     family=gaussian(link=log), weights=in_place_oil_mill_sm3, data=large_fields)
-summary(gam_full3)
+gam_mod<-gam(oil_prod_mill_sm3~s(prod_year, bs="cr") + 
[…]

### 2015-03-13 rr_analysis_alt_models.r

@@ -1,0 +1,54 @@
+#rr_analysis.r
+
+library(ggplot2)
+library(plyr)
+library(reshape2)
+library(mgcv)
+library(lubridate) 
+library(grid)
+library(boot)
+library(arm)
+library(gamm4)
+library(lme4)
+
+#clear workspace
+rm(list = ls())
+
+#Use time dummies - first three years, first four years
+
+#Use multilevel (random effects) model.  
+
+#Combine many graphs in a single figure. 
+
+#Showing how a parametric model heavily biases estimation - Put in an online appendix.  
+
+#Instead of splitting up into small and large, use mixed effects to take care of heterogeneity.  
+
+#go straight to preferred set up, splitting up in post, build-out and draw-down.  
+
+#Waiting time model - Time to start modeled as a function of time (bayesian model)
+
+#prep.py
+
+#other models not used
+
+large_fields<-prod_data[prod_data["recoverable_oil"]>3,]
+large_fields<-prod_data[prod_data["name"]!="EKOFISK",]
+
+
+
+gam_full<-gam(oil_prod_mill_sm3~s(prod_year) + 
+     in_place_oil_mill_sm3 + cost_index +
+    year + I(year^2) + price + price_l1 + price_l2 + price_l3 + price_l4 + 
+     price_l5 + price_l6 + price_l7 + price_l8, 
+     family=gaussian(link=log), weights=in_place_oil_mill_sm3, data=large_fields, 
+     select=TRUE)
+summary(gam_full)
+
+gam_full3<-gam(oil_prod_mill_sm3~s(prod_year, in_place_oil_mill_sm3) + 
+      cost_index +
+     year + I(year^2) + price + price_l1 + price_l2 + price_l3 + price_l4 + 
+     price_l5 + price_l6 + price_l7 + price_l8, 
+     family=gaussian(link=log), weights=in_place_oil_mill_sm3, data=large_fields)
+summary(gam_full3)
+

### 2015-03-17 rr_descriptives.py
updated monte carlo experiment

@@ -12,6 +12,10 @@ import matplotlib.pyplot as plt
 import random
 import statsmodels.api as sm
 
+
+brentprices=pd.read_excel("http://www.eia.gov/dnav/pet/hist_xls/RBRTEm.xls", sheetname="Data 1")
+
+
 #first on individual field data
 field_data=pd.read_csv('/Users/johannesmauritzen/research/oil_prices/data/field_data.csv')
 prod_data=pd.read_csv('/Users/johannesmauritzen/research/oil_prices/data/prod_data.csv')

### 2015-03-17 rr_oil_bootstrap.r

@@ -1,0 +1,32 @@
+#oil_bootstram
+library(boot)
+library(arm)
+#with model
+#prod_gam_under<-gam(year_prod~s(time_to_peak, recoverable_oil) +  s(year), 
+#	family=Gamma(link=log), weights=recoverable_oil, data=fields_p[fields_p$max_prod<=split,])
+#prod_gam_over<-gam(year_prod~s(time_to_peak, recoverable_oil) + s(year), 
+#	family=Gamma(link=log), weights=recoverable_oil, data=fields_p[fields_p$max_prod>split,])
+
+prod_gam_over_bootstrap <- function(field_data){
+prod_gam_over<-gam(year_prod~s(time_to_peak, recoverable_oil) + s(year), 
+	family=Gamma(link=log), weights=recoverable_oil, data=field_data[field_data$max_prod>split,])
+	return(prod_gam_over$fitted.values)
+}
+
+prod_gam_under_bootstrap <- function(field_data){
+prod_gam_under<-gam(year_prod~s(time_to_peak, recoverable_oil) +  s(year), 
+	family=Gamma(link=log), weights=recoverable_oil, data=field_data[field_data$max_prod<=split,])
+	return(prod_gam_under$fitted.values)
+
+}
+#With the function fc defined, we can use the boot command, 
+#providing our dataset name, our function, and the number of bootstrap samples to be drawn.
+
+prod_gam_over_boot <- boot(fields_p, prod_gam_over_bootstrap , R=10)
+
+
+
+#better idea - generate one forecast point at a time
+# for each point where time to peak, peak to end, and recoverable_oil are approximate
+#
+

### 2015-03-17 rr_simulation.r

@@ -1,0 +1,163 @@
+library(ggplot2) #for plotting
+library(mgcv) #for gam modeling
+library(lubridate) #for extracting "year" component
+library(dplyr) 
+library(reshape2)
+library(zoo)
+library(sandwich)
+
+#make function that inputs simulated field_size and outputs simulated
+#first production year
+#should be so that large fields are found more often early on.  
+
+gen_year<-function(size, maxsize){
+	#test
+	#size=15
+	#maxsize=1267
+	#
+
+	#let small fields be distributed uniformly from 1975 to 2013
+	if(size<10){
+		year<-trunc(runif(1, 1975, 2008))	
+	} else{	#while large fields should be more common earlier on
+		range<-FALSE
+		while(range==FALSE){
+			year<-trunc(rnorm(1,mean=(1973+(maxsize+300)/(size+300)), sd=10))
+			ifelse(year>=1970 & year<=2013, range<-TRUE, range<-FALSE)
+			#print(paste("in while loop", year))
+			}	
+		}
+return(year)	
+}
+
+
+#Simulation of data****************************************************************
+
+gen_production_shape<-function(field){
+	#
+	#field<-c(20, 1990, 1)
+	#
+	size<-as.numeric(field[1])
+	start<-as.numeric(field[2])
+	name<-field[3]
+
+	t<-trunc(sqrt(size)) + 3
+	time<- -t:(t)
+	year<-start:(start+2*t-1)
+
+	#use a cumulative logistic function to represent shape of function
+	cum_production <- 1/(1+exp(-(time)/3))*size
+
+	#take difference to create production per year
+	prod_shape <- diff(cum_production)
+	prod_years <- 1:length(prod_shape)
+	production<-data.frame(year, prod_shape, name, size, prod_years)
+	return(production)
+}
+
+
+
+#GAM simualtion *********************************
+add_uncertainty<-function(beta0=0, fields){ 
+	fields$prod<-fields$prod_shape*
+		(exp(beta0*(fields$price))*
+			rlnorm(length(fields$price), meanlog=0, sdlog=.05))
+	return(fields)
+}
+
+run_gam<-function(formula, fields, beta=0){
+	#
+	#fields <- sim_fields_unc
+	#formula<-formula(prod ~ s(prod_time) + size + prices)
+	#
[…]

### 2015-03-19 oil_prices_rr.tex
work on mc

@@
 
-%\doublespacing % from package setspacs
+\doublespacing % from package setspacs
 
@@
 
-\begin{figure}
-	\includegraphics[width=1\textwidth]{figures/top10_production_print.png}
-	\caption{The production profile of the 10 largest oil fields on the Norwegian Continental Shelf.  Production tends to be correlated across fields}
-	\label{top10_production}	
-	\end{figure}
+As a solution I use a semi-parametric model within the Generalized Additive Model frameworks of \cite{hastie_generalized_1990}.
+Here I use a cubic spline function to account for the general non-parametric shape of the production profile while allowing price to enter the equation linearly while also attempting to control for the effects of varying production costs and the effects of technological change and other time-varying effects.
 
-\begin{figure}
-	\includegraphics[width=1\textwidth]{figures/oil_decline_print.png}
-	\caption{The production profile for the entire Norwegian Continental Shelf is bell-shaped, reflecting the correlated production profiles of the fields.  With oil prices that are autocorrelated, }
-	\label{oil_decline}
-\end{figure}
-
-As a solution I use a semi-parametric model within the Generalized Additive Model frameworks of \cite{hastie_generalized_1990}.
-Here I use a two-dimensional smoothed spline function to account for the general non-parametric shape of the production profile while allowing price to enter the equation linearly.
-The coefficient of price can then be interpreted as the average effect of price on production over the entire production profile.
+The results suggest that little to no concurrent effect of oil prices on production exists.  However modest lagged effects of between 2 to 5 years exists.  These results are in line with the intuition
 
@@
 
-\begin{figure}
-\includegraphics[width=1.2\textwidth]{figures/north_sea_reserves_print.png}
-\caption{Oil fields in the Norwegian territorial North Sea.
-The largest oil fields tended to be discovered earliest, while newer finds tend to be smaller.
-An exception is the large Johan Sverdrup field, which is expected to begin producing in 2019.}
-\label{north_sea_reserves}
-\end{figure}
-
 Exploration in the Norwegian Sea was opened in the early 1980’s and the first commercial field started production in 1981.
 While several mid-sized fields have been discovered, the Norwegian Sea has generally disappointed in terms of commercial oil finds and most finds have been relatively small (\ref{norwegian_sea_reserves}).
-
-\begin{figure}
-\includegraphics[width=1.2\textwidth]{figures/norwegian_sea_reserves_print.png}
-\caption{Oil fields in the Norwegian Sea.  Production from the Norwegian Sea has generally been dissapointing compared with expectations when the area was opened to exploration in the early 1980's.}
-\label{norwegian_sea_reserves}
-\end{figure}
 
@@
 
-A cleaned data set and the full code for the analysis are available upon request.
-I use the R statistical programming package for all the analysis in this article \citep{r_core_team_r:_2013}.
-I use the R packages ggplot2 and ggmap for plotting \citep{wickham_ggplot2:_2009, kahle_ggmap:_2013}, plyr for data manipulation and cleaning \citep{wickham_split-apply-combine_2011}, texreg for table formatting \citep{leifeld_texreg:_2013} and mgcv for implementation of the Generalized Additive Models \citep{wood_fast_2011}.
+A cleaned data set and the full code for the analysis are available at my website \url{jmaurit.github.io#oil_prices}.
 
@@
 
+\begin{figure}
+	\includegraphics[width=1\textwidth]{figures/mc_plot.png}
+	\caption{Results from a Monte-Carlo experiment.
+I generate data mimicking oil price production from a set of fields, as displayed in the top frame.
+Using GLM and GAM models, and regenerating the error component in the data, I replicate estimates of the effect of price on production, which is set at $\beta = 0$ and $\beta=0.05$.
+The density of the results for the GLM model are shown in the lower left panel, where a clear and large bias is shown relative to the true $\beta$ which are represented by the vertical lines.
+A GAM model however provides estimates close to the true $\beta$}
+	\label{mc_results}
+\end{figure}
+
 \section{The effect of price in the phases of oil production}
@@
 
-\begin{figure}
-	\includegraphics[width=1\textwidth]{figures/gam_postpeak_print.png}
-	\caption{The estimated coefficients on the price terms in a model of oil field production in depleting fields.
-Significant coefficients are only estimated on the concurrent price and eight lag, however the significance of the estimated coefficient on the eighth lag is not robust to specification}
-	\label{gam_postpeak_print}
-\end{figure}
-
 The effect that price has on production from a declining fields appears to be, at best, slight.
@@
 I show the estimated coefficient on the price terms in table \ref{gam_prepeak_print} below.
-
-\begin{figure}
-	\includegraphics[width=1\textwidth]{figures/gam_prepeak_print.png}
-	\caption{The estimated coefficients on the price terms in a model of fields in the build-out phase.
-For large fields, positive coefficients are estimated at the 6th, 7th and 8th lag, but the estimates are imprecise and not statistically significant.
-Significant positive coefficients are estimated for small fields at the fifth and eighth lag.
-In a pooled model, significant positive coefficients are found at the fourth through eighth lags.}
-	\label{gam_prepeak_print}
-\end{figure}
 
@@
 
+\section{software}
+I use the R statistical programming package for all the analysis in this article \citep{r_core_team_r:_2013}.
+I use the R package ggplot2 for plotting \citep{wickham_ggplot2:_2009}, plyr for data manipulation and cleaning \citep{wickham_split-apply-combine_2011}, texreg for table formatting \citep{leifeld_texreg:_2013} and mgcv for implementation of the Generalized Additive Models \citep{wood_fast_2011}.
+
  \begin{spacing}{1}
@@
 \FloatBarrier
-
-\appendix
-
-\section{Appendix: Tables of full results and robustness checks}
-
-\begin{table}
-\begin{center}
-\begin{tabular}{l c c c c c c }
-\hline
-	&\multicolumn{2}{|c|}{6 lags} & \multicolumn{2}{c}{8 lags} &\multicolumn{2}{c}{4-8 lags} \\
-                                     & Small & Large & Small  & Large & Small & Large\\
-\hline
-(Intercept)                          & $-1.46^{***}$ & $-0.15$       & $-1.58^{***}$ & $-0.25$       & $-1.54^{***}$ & $-1.54^{***}$ \\
-                                     & $(0.37)$      & $(0.50)$      & $(0.38)$      & $(0.51)$      & $(0.37)$      & $(0.37)$      \\
-oilpricereal                       & $0.00$        & $-0.02$       & $0.00$        & $-0.01$       &               &               \\
-                                     & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      &               &               \\
-oilpricereall1                    & $0.01$        & $-0.01$       & $0.01$        & $0.01$        &               &               \\
-                                     & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      &               &               \\
-oilpricereall2                    & $-0.02$       & $0.01$        & $-0.01$       & $0.00$        &               &               \\
-                                     & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      &               &               \\
-oilpricereall3                    & $-0.02$       & $0.01$        & $-0.03^{*}$   & $0.00$        &               &               \\
-                                     & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      &               &               \\
-oilpricereall4                    & $0.00$        & $0.03^{*}$    & $0.00$        & $0.03^{*}$    & $-0.01$       & $-0.01$       \\
-                                     & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
-oilpricereall5                    & $-0.01$       & $0.01$        & $-0.01$       & $0.01$        & $-0.02$       & $-0.02$       \\
-                                     & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
-oilpricereall6                    & $0.02^{*}$    & $0.03^{*}$    & $0.02$        & $0.01$        & $0.02^{*}$    & $0.02^{*}$    \\
-                                     & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
-EDF: s(timetopeak,recoverableoil) & $28.92^{***}$ & $26.89^{***}$ & $28.91^{***}$ & $26.89^{***}$ & $28.91^{***}$ & $28.91^{***}$ \\
-                                     & $(29.00)$     & $(27.00)$     & $(29.00)$     & $(27.00)$     & $(29.00)$     & $(29.00)$     \\
-EDF: s(peaktoend,recoverableoil)  & $17.75^{***}$ & $10.71^{***}$ & $17.58^{***}$ & $11.41^{***}$ & $17.47^{***}$ & $17.47^{***}$ \\
-                                     & $(28.00)$     & $(28.00)$     & $(28.00)$     & $(28.00)$     & $(28.00)$     & $(28.00)$     \\
-oilpricereall7                    &               &               & $-0.01$       & $0.02^{*}$    & $-0.01$       & $-0.01$       \\
-                                     &               &               & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
-oilpricereall8                    &               &               & $0.04^{***}$  & $0.03^{**}$   & $0.03^{***}$  & $0.03^{***}$  \\
-                                     &               &               & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
-\hline
-AIC                                  &               & 1211.33       &               & 1186.97       &               &               \\
-BIC                                  &               & 1377.97       &               & 1363.28       &               &               \\
-Log Likelihood                       &               & -559.07       &               & -544.18       &               &               \\
-Deviance                             & 12319.62      & 301090.57     & 12104.52      & 268975.69     & 12347.62      & 12347.62      \\
-Deviance explained                   & 0.84          & 0.96          & 0.85          & 0.96          & 0.84          & 0.84          \\
-Dispersion                           & 15.17         & 1378.61       & 14.93         & 1247.02       & 15.16         & 15.16         \\
-R$^2$                                & 0.82          & 0.95          & 0.83          & 0.95          & 0.82          & 0.82          \\
-GCV score                            & 16.19         & 1666.44       & 15.98         & 1526.28       & 16.13         & 16.13         \\
-Num. obs.                            & 865           & 264           & 865           & 264           & 865           & 865           \\
-Num. smooth terms                    & 2             & 2             & 2             & 2             & 2             & 2             \\
-\hline
-\multicolumn{7}{l}{\scriptsize{\textsuperscript{***}$p<0.001$, 
-  \textsuperscript{**}$p<0.01$, 
-  \textsuperscript{*}$p<0.05$}}
-\end{tabular}
-\caption{Full results for model of field production with price, seperate estimation of small and large fields.}
-\label{table:unpooled}
-\end{center}
-\end{table}
-
-%table two pooled
-\begin{table}
-\begin{center}
-\begin{tabular}{l c c }
-\hline
-                                     & Pooled Model & Insignificant terms dropped \\
-\hline
-(Intercept)                          & $-1465.35^{***}$ & $-1500.51^{***}$ \\
-                                     & $(65.21)$        & $(66.28)$        \\
-oilpricereal                       & $-0.01$          &                  \\
-                                     & $(0.01)$         &                  \\
-largefieldsmall                     & $-2.32^{***}$    & $-1.19^{***}$    \\
-                                     & $(0.27)$         & $(0.10)$         \\
-oilpricereall1                    & $0.02^{*}$       &                  \\
-                                     & $(0.01)$         &                  \\
-oilpricereall2                    & $0.01$           &                  \\
-                                     & $(0.01)$         &                  \\
-oilpricereall3                    & $0.00$           &                  \\
-                                     & $(0.01)$         &                  \\
-oilpricereall4                    & $0.03^{***}$     & $0.03^{***}$     \\
-                                     & $(0.01)$         & $(0.01)$         \\
-oilpricereall5                    & $0.01$           & $0.01$           \\
-                                     & $(0.01)$         & $(0.01)$         \\
-oilpricereall6                    & $0.02^{***}$     & $0.03^{***}$     \\
-                                     & $(0.01)$         & $(0.01)$         \\
-oilpricereall7                    & $0.02^{**}$      & $0.02^{***}$     \\
-                                     & $(0.01)$         & $(0.01)$         \\
-oilpricereall8                    & $0.03^{***}$     & $0.03^{***}$     \\
-                                     & $(0.00)$         & $(0.00)$         \\
-oilpricereal:largefieldsmall      & $-0.03$          &                  \\
-                                     & $(0.06)$         &                  \\
-largefieldsmall:oilpricereall1   & $0.03$           &                  \\
-                                     & $(0.07)$         &                  \\
-largefieldsmall:oilpricereall2   & $-0.05$          &                  \\
-                                     & $(0.08)$         &                  \\
-largefieldsmall:oilpricereall3   & $0.03$           &                  \\
-                                     & $(0.07)$         &                  \\
-largefieldsmall:oilpricereall4   & $0.03$           &                  \\
-                                     & $(0.07)$         &                  \\
-largefieldsmall:oilpricereall5   & $0.04$           &                  \\
-                                     & $(0.07)$         &                  \\
-largefieldsmall:oilpricereall6   & $0.18^{*}$       &                  \\
-                                     & $(0.07)$         &                  \\
-largefieldsmall:oilpricereall7   & $0.06$           &                  \\
-                                     & $(0.07)$         &                  \\
-largefieldsmall:oilpricereall8   & $0.07$           &                  \\
-                                     & $(0.07)$         &                  \\
-EDF: s(timetopeak,recoverableoil) & $28.99^{***}$    & $28.99^{***}$    \\
-                                     & $(29.00)$        & $(29.00)$        \\
-EDF: s(peaktoend,recoverableoil)  & $27.98^{***}$    & $26.99^{***}$    \\
-                                     & $(28.00)$        & $(27.00)$        \\
-\hline
-AIC                                  &                  &                  \\
-BIC                                  &                  &                  \\
-Log Likelihood                       &                  &                  \\
-Deviance                             & 371079.07        & 396494.49        \\
-Deviance explained                   & 0.96             & 0.96             \\
-Dispersion                           & 352.06           & 371.24           \\
-R$^2$                                & 0.96             & 0.95             \\
-GCV score                            & 377.77           & 393.13           \\
-Num. obs.                            & 1129             & 1129             \\
-Num. smooth terms                    & 2                & 2                \\
-\hline
-\multicolumn{3}{l}{\scriptsize{\textsuperscript{***}$p<0.001$, 
-  \textsuperscript{**}$p<0.01$, 
-  \textsuperscript{*}$p<0.05$}}
-\end{tabular}
-\caption{Full results for pooled model of field production with price terms.}
-\label{table:pooled}
-\end{center}
-\end{table}
-
-%investment tables
-
-\begin{table}
-\begin{center}
-\begin{tabular}{l c c c }
-\hline
-                                     & Pooled & Small Fields & Large Fields \\
-\hline
-(Intercept)                          & $-199.84^{***}$ & $7.03^{***}$   & $7.64^{***}$    \\
-                                     & $(44.00)$       & $(0.36)$       & $(0.24)$        \\
-yearprod                            & $-0.06^{***}$   & $-0.26^{***}$  & $-0.06^{***}$   \\
-                                     & $(0.01)$        & $(0.02)$       & $(0.01)$        \\
-oilpricereal                       & $0.08^{***}$    & $-0.03^{**}$   & $0.04$          \\
-                                     & $(0.01)$        & $(0.01)$       & $(0.03)$        \\
-largefieldsmall                     & $1.80^{***}$    &                &                 \\
-                                     & $(0.52)$        &                &                 \\
-oilpricereall1                    & $-0.02$         & $0.00$         & $0.00$          \\
-                                     & $(0.01)$        & $(0.01)$       & $(0.03)$        \\
-oilpricereall2                    & $0.06^{***}$    & $0.05^{***}$   & $0.09^{***}$    \\
-                                     & $(0.01)$        & $(0.01)$       & $(0.03)$        \\
-oilpricereall3                    & $0.16^{***}$    & $0.25^{***}$   & $0.19^{***}$    \\
-                                     & $(0.01)$        & $(0.03)$       & $(0.03)$        \\
-oilpricereall4                    & $0.00$          & $-0.08^{**}$   & $-0.03$         \\
-                                     & $(0.01)$        & $(0.03)$       & $(0.03)$        \\
-oilpricereall5                    & $-0.02$         & $-0.09^{**}$   & $-0.07^{*}$     \\
-                                     & $(0.02)$        & $(0.03)$       & $(0.03)$        \\
-oilpricereall6                    & $-0.09^{***}$   & $-0.10^{**}$   & $-0.04$         \\
-                                     & $(0.02)$        & $(0.04)$       & $(0.04)$        \\
-oilpricereall7                    & $-0.11^{***}$   & $-0.26^{***}$  & $-0.09^{*}$     \\
-                                     & $(0.02)$        & $(0.04)$       & $(0.04)$        \\
-oilpricereall8                    & $-0.10^{***}$   & $-0.19^{***}$  & $-0.12^{***}$   \\
-                                     & $(0.01)$        & $(0.03)$       & $(0.03)$        \\
-oilpricereal:smallfield      & $-0.11^{*}$     &                &                 \\
-                                     & $(0.05)$        &                &                 \\
-largefieldsmall:oilpricereall1   & $0.02$          &                &                 \\
-                                     & $(0.05)$        &                &                 \\
-largefieldsmall:oilpricereall2   & $0.01$          &                &                 \\
-                                     & $(0.05)$        &                &                 \\
-largefieldsmall:oilpricereall3   & $0.03$          &                &                 \\
-                                     & $(0.06)$        &                &                 \\
-EDF: s(timetopeak,recoverableoil) & $28.99^{***}$   & $26.92^{***}$  & $28.79^{***}$   \\
-                                     & $(29.00)$       & $(28.11)$      & $(28.97)$       \\
-EDF: s(peaktoend,recoverableoil)  & $16.90^{***}$   & $4.83^{*}$     & $5.39^{**}$     \\
-                                     & $(18.82)$       & $(6.40)$       & $(7.44)$        \\
-\hline
-AIC                                  &                 &                & 4747.87         \\
-BIC                                  &                 &                & 4912.98         \\
-Log Likelihood                       &                 &                & -2327.76        \\
-Deviance                             & 257371493567.88 & 10630461533.61 & 198564735228.80 \\
-Deviance explained                   & 0.82            & 0.81           & 0.84            \\
-Dispersion                           & 243467395.17    & 13087671.29    & 907398412.98    \\
-R$^2$                                & 0.80            & 0.79           & 0.81            \\
-GCV score                            & 257721863.13    & 13776495.42    & 1094706852.34   \\
-Num. obs.                            & 1117            & 853            & 264             \\
-Num. smooth terms                    & 2               & 2              & 2               \\
-\hline
-\multicolumn{4}{l}{\scriptsize{\textsuperscript{***}$p<0.001$, 
-  \textsuperscript{**}$p<0.01$, 
-  \textsuperscript{*}$p<0.05$}}
-\end{tabular}
-\caption{Full results for model of field investment with price terms.}
-\label{table:investment}
-\end{center}
-\end{table}
-
-% Table post-peak
-
-\begin{table}
-\begin{center}
-\begin{tabular}{l c c c c }
-\hline
-&\multicolumn{2}{c}{Recoverable Oil} & \multicolumn{2}{c}{Max Production} \\
-                                    & Large & Small & Large & Small \\
-\hline
-(Intercept)                         & $1.54^{***}$  & $-0.31^{*}$   & $2.25^{***}$  & $-1.61^{***}$ \\
-                                    & $(0.06)$      & $(0.14)$      & $(0.08)$      & $(0.25)$      \\
-oilpricereal                      & $0.01$        & $0.03^{***}$  & $-0.02^{*}$   & $0.02^{*}$    \\
-                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
-oilpricereall1                   & $0.02$        & $-0.01$       & $-0.01$       & $0.00$        \\
-                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
-oilpricereall2                   & $0.00$        & $-0.01$       & $-0.02$       & $-0.01$       \\
-                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
-oilpricereall3                   & $0.02$        & $-0.03^{**}$  & $0.00$        & $-0.02^{*}$   \\
-                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
-oilpricereall4                   & $0.03$        & $0.00$        & $0.01$        & $0.00$        \\
-                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
-oilpricereall5                   & $-0.01$       & $-0.03^{*}$   & $-0.01$       & $-0.02$       \\
-                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
-oilpricereall6                   & $0.01$        & $0.01$        & $0.00$        & $0.00$        \\
-                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
-oilpricereall7                   & $0.01$        & $-0.03$       & $0.02$        & $-0.01$       \\
-                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
-oilpricereall8                   & $0.01$        & $0.06^{***}$  & $-0.01$       & $0.04^{***}$  \\
-                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
-EDF: s(peaktoend,recoverableoil) & $27.63^{***}$ & $23.05^{***}$ &               &               \\
-                                    & $(29.00)$     & $(29.00)$     &               &               \\
-EDF: s(peaktoend,maxprod)        &               &               & $23.82^{***}$ & $28.82^{***}$ \\
-                                    &               &               & $(29.00)$     & $(29.00)$     \\
-\hline
-AIC                                 & 571.09        & 1543.57       & 490.28        & 1421.72       \\
-BIC                                 & 690.61        & 1696.16       & 598.01        & 1600.20       \\
-Log Likelihood                      & -246.92       & -737.74       & -210.31       & -671.04       \\
-Deviance                            & 49634.76      & 6868.08       & 31676.52      & 5599.00       \\
-Deviance explained                  & 0.99          & 0.87          & 0.99          & 0.90          \\
-Dispersion                          & 395.91        & 11.08         & 245.22        & 9.12          \\
-R$^2$                               & 0.98          & 0.86          & 0.99          & 0.88          \\
-GCV score                           & 514.74        & 11.67         & 309.43        & 9.69          \\
-Num. obs.                           & 163           & 653           & 163           & 653           \\
-Num. smooth terms                   & 1             & 1             & 1             & 1             \\
-\hline
-\multicolumn{5}{l}{\scriptsize{\textsuperscript{***}$p<0.001$, 
-  \textsuperscript{**}$p<0.01$, 
-  \textsuperscript{*}$p<0.05$}}
-\end{tabular}
-\caption{Full results for model of field production with price terms in depletion phase.  Using both estimate of total recoverable oil and max production to control for field size}
-\label{table:postpeak}
-\end{center}
-\end{table}
-
-% prepeak
-
-\begin{table}
-\begin{center}
-\begin{tabular}{l c c c }
-\hline
-                                     & Small Fields & Large Fields & Pooled \\
-\hline
-(Intercept)                          & $-0.43$       & $1.26$        & $-0.91$       \\
-                                     & $(0.29)$      & $(2.28)$      & $(0.68)$      \\
-oilpricereal                       & $0.01$        & $-0.04$       & $0.02$        \\
-                                     & $(0.03)$      & $(0.07)$      & $(0.04)$      \\
-oilpricereall1                    & $0.03$        & $-0.06$       & $0.10^{**}$   \\
-                                     & $(0.04)$      & $(0.09)$      & $(0.03)$      \\
-oilpricereall2                    & $-0.01$       & $-0.11$       & $0.05^{*}$    \\
-                                     & $(0.03)$      & $(0.11)$      & $(0.02)$      \\
-oilpricereall3                    & $-0.07^{*}$   & $-0.11$       & $-0.02$       \\
-                                     & $(0.03)$      & $(0.13)$      & $(0.02)$      \\
-oilpricereall4                    & $0.06$        & $-0.02$       & $0.07^{***}$  \\
-                                     & $(0.03)$      & $(0.12)$      & $(0.02)$      \\
-oilpricereall5                    & $0.09^{**}$   & $0.01$        & $0.06^{***}$  \\
-                                     & $(0.03)$      & $(0.10)$      & $(0.02)$      \\
-oilpricereall6                    & $-0.01$       & $0.07$        & $0.04^{*}$    \\
-                                     & $(0.03)$      & $(0.07)$      & $(0.02)$      \\
-oilpricereall7                    & $0.01$        & $0.10$        & $0.05^{**}$   \\
-                                     & $(0.03)$      & $(0.06)$      & $(0.02)$      \\
-oilpricereall8                    & $0.07^{**}$   & $0.06$        & $0.05^{**}$   \\
-                                     & $(0.02)$      & $(0.04)$      & $(0.02)$      \\
-EDF: s(time_to_peak,recoverable_oil) & $16.91^{***}$ & $28.37^{***}$ & $20.51^{***}$ \\
-                                     & $(29.00)$     & $(29.00)$     & $(29.00)$     \\
-\hline
-AIC                                  & 551.59        & 330.20        & 1534.58       \\
-BIC                                  & 638.09        & 421.97        & 1644.24       \\
-Log Likelihood                       & -247.89       & -125.73       & -735.78       \\
-Deviance                             & 3949.57       & 34833.67      & 300487.73     \\
-Deviance explained                   & 0.79          & 0.99          & 0.92          \\
-Dispersion                           & 28.81         & 925.75        & 1434.35       \\
-R$^2$                                & 0.74          & 0.98          & 0.91          \\
-GCV score                            & 34.46         & 1869.82       & 1643.21       \\
-Num. obs.                            & 164           & 76            & 240           \\
-Num. smooth terms                    & 1             & 1             & 1             \\
-\hline
-\multicolumn{4}{l}{\scriptsize{\textsuperscript{***}$p<0.001$, 
-  \textsuperscript{**}$p<0.01$, 
-  \textsuperscript{*}$p<0.05$}}
-\end{tabular}
-\caption{Full results for model of field production with price terms in depletion phase.}
-\label{table:prepeak}
-\end{center}
-\end{table}
-
-\begin{figure}
-	\includegraphics[width=.8\textwidth]{figures/gam_under_2d_plot_1.png}
-	\caption{Plot of estimated function of the production profile of small fields, time-to-peak.}
-	\label{chart:gam_under_2d_plot_1}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=.8\textwidth]{figures/gam_under_2d_plot_2.png}
-	\caption{Plot of estimated function of the production profile of small fields, peak-to-end.}
-	\label{chart:gam_under_2d_plot_2}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=.8\textwidth]{figures/gam_over_2d_plot_1.png}
-	\caption{Plot of estimated function of the production profile of large fields, time-to-peak.}
-	\label{chart:gam_over_2d_plot_1}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=.8\textwidth]{figures/gam_over_2d_plot_2.png}
-	\caption{Plot of estimated function of the production profile of large fields, peak-to-end.}
-	\label{chart:gam_over_2d_plot_2}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=.8\textwidth]{figures/qq_gam_price_over_2d.png}
-	\caption{QQplot of model residuals against theoretical quantiles of normal distribution.  A Gaussian distribution appears to be adequate.}
-	\label{chart:qqplot}
-\end{figure}
-
-%Fitted vs. residuals
-\begin{figure}
-	\includegraphics[width=.8\textwidth]{figures/under_fitted.png}
-	\caption{Fitted verses actual production for small fields.}
-	\label{chart:under_fitted}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=.8\textwidth]{figures/over_fitted.png}
-	\caption{Fitted verses actual production for larger fields.}
-	\label{chart:under_fitted}
-\end{figure}
-
-% Check to see if overfitting of model - change gamma parameter to 1.4
-\begin{table}
-\begin{center}
-\begin{tabular}{l c c }
-\hline
-                                     & Small Fields & Large Fields \\
-\hline
-(Intercept)                          & $-0.24^{***}$ & $1.88^{***}$  \\
-                                     & $(0.07)$      & $(0.11)$      \\
-oilpricereal                       & $0.02$        & $-0.04$       \\
-                                     & $(0.01)$      & $(0.02)$      \\
-oilpricereall1                    & $-0.01$       & $-0.02$       \\
-                                     & $(0.01)$      & $(0.03)$      \\
-oilpricereall2                    & $-0.01$       & $0.00$        \\
-                                     & $(0.01)$      & $(0.03)$      \\
-oilpricereall3                    & $-0.02$       & $0.00$        \\
-                                     & $(0.01)$      & $(0.03)$      \\
-oilpricereall4                    & $0.02$        & $0.02$        \\
-                                     & $(0.01)$      & $(0.03)$      \\
-oilpricereall5                    & $0.01$        & $-0.01$       \\
-                                     & $(0.02)$      & $(0.03)$      \\
-oilpricereall6                    & $-0.04^{*}$   & $0.09^{**}$   \\
-                                     & $(0.02)$      & $(0.03)$      \\
-EDF: s(time_to_peak,recoverable_oil) & $23.93^{***}$ & $28.58^{***}$ \\
-                                     & $(29.00)$     & $(29.00)$     \\
-EDF: s(peak_to_end,recoverable_oil)  & $25.91^{***}$ & $6.10^{***}$  \\
-                                     & $(28.00)$     & $(28.00)$     \\
-\hline
-AIC                                  & 59890.49      & 441209.87     \\
-BIC                                  & 60170.73      & 441366.08     \\
-Log Likelihood                       & -29886.40     & -220561.25    \\
-Deviance                             & 4653.55       & 12307.60      \\
-Deviance explained                   & 0.80          & 0.83          \\
-Dispersion                           & 5.75          & 55.61         \\
-R$^2$                                & 0.77          & 0.89          \\
-GCV score                            & 6.16          & 66.34         \\
-Num. obs.                            & 865           & 264           \\
-Num. smooth terms                    & 2             & 2             \\
-\hline
-\multicolumn{3}{l}{\scriptsize{\textsuperscript{***}$p<0.001$, 
-  \textsuperscript{**}$p<0.01$, 
-  \textsuperscript{*}$p<0.05$}}
-\end{tabular}
-\caption{Results when changing the $\gamma$ parameter that controls smoothness to 1.4 in case of overfitting, though this does not signficantly change results}
-\label{table:gamma_check}
-\end{center}
-\end{table}
-
-% max production vs. total fields
-\begin{figure}
-	\includegraphics[width=.8\textwidth]{figures/Max_prod_vs_TRO_print.png}
-	\caption{Comparison of fits between using max production and estimate of total reserves to control for field size.}
-	\label{chart:Max_prod_vs_TRO_print}
-\end{figure}
-
-\FloatBarrier
-
-\section{Appendix: Monte Carlo Experiment}
-
-The data set used in this paper has a complex structure.
-The dependent variable - field level oil production - has a non-linear autocorrelated structure.
-The production profiles are also correlated with each other - leading to a bell-shaped total production curve.
-In addition the exogenous variable of interest - oil prices - are autocorrelated and non-stationary.
-In such a setting, standard rules for inference may not hold.
-
-In this appendix I instead present a Monte Carlo simulation experiment.
-Generating artificial data that is similar in character to the data used, I first show how failing to properly control for the production profile can lead to a biased estimation of the effect of oil prices.
-I then show how using a generalized additive model can lead to better estimates.
-
-I use the R statistical programming language \citet{r_core_team_r:_2013} for the analysis.
-For illustration, I include code snippets in this document.
-The full code for the analysis is available upon request.
-The figures shown are created using the r package ggplot2 \citet{wickham_ggplot2:_2009}.
-
-\subsection{Generating Data}
-
-The first step is generate a set of fields from a size distribution.
-In the simulation, I generate 77 fields from a exponential normal distribution with mean 2.3 and standard deviation of 1.5.
-This is similar to the distribution of the actual 77 productive Norwegian fields.
-\begin{verbatim}
-field_size<-round(exp(rnorm(77, mean=2.3, sd=1.5)), digits=1)
-\end{verbatim}
-
-Using the generated field sizes as an input, I then create a function to generate starting times.
-The goal here is to mimic what is called creaming in the exploration industry - the tendency of finding and starting production on larger fields first.
-For fields that are smaller than 10 million SM3, I generate a starting time from a uniform distribution with endpoints 1975 and 2008.
-For larger fields, I generate a starting time from a normal distribution with standard deviation of 10 and a mean as in formula 1:
-
-\begin{equation}
-mean_start(size) = trunc(1973 + (max + 300)/size)
-\end{equation}
-
-\begin{verbatim}
-gen_year<-function(size, maxsize){
-	#let small fields be distributed uniformly from 1975 to 2013
-	if(size<10){
-		year<-trunc(runif(1, 1975, 2008))	
-	} else{	#while large fields should be more common earlier on
-		range<-FALSE
-		while(range==FALSE){
-			year<-trunc(rnorm(1,mean=(1973+(maxsize+300)/(size+300)), sd=10))
-			ifelse(year>=1970 & year<=2013, range<-TRUE, range<-FALSE)
-			}	
-		}
-return(year)	
-}
-\end{verbatim}
-
-Generating the starting times in this way was essentially ad-hoc.
-However the method is sufficient since the purpose of this exercise is not to model the underlying physical and economic processes, but rather to generate data that shares key similarities with the actual data.
-
-In the simulation I use the actual Brent oil price series shown in figure \ref{oil_price_series}.
-This series can be shown to be non-stationary.
-In order to get reliable inference in a simpler time-series setting the solution might be to take the first difference of the series.
-This is not possible in this setting since it is the actual levels of oil prices that are important.
-Intuitively, identification comes from comparison of fields of similar size and at similar points in their production lifetime but at periods at differing oil prices.
-This variation would be discarded by purely looking at first differences.
-
-\begin{figure}
-	\includegraphics[width=1\textwidth]{figures/oil_price_series.png}
-	\caption{The Real Price of Brent Traded Oil from 1960 to 2013 in 2010 U.S. Dollars.}
-	\label{oil_price_series}	
-\end{figure}
-
-So far, all the simulated data has been outside the Monte Carlo function.
-The data is treated as being fixed.Within the Monte-Carlo loop I create a function that takes as input a fields size, start date as well as the price series and generates a production profile.
-The code for this function is below.
-
-\begin{verbatim}
-gen_production<-function(field, prices){
-	size<-field[1]
-	start<-field[2]
-	name<-field[3]
-
-	names(prices)<-c("year", "price")
-	t<-trunc(sqrt(size)) + 3  #generate production time
-	prod_time<- -t:(t)
-	year=start:(start+2*t-1)
-
-	#use a cumulative logistic function to represent shape of function
-	cum_production = 1/(1+exp(-(prod_time)/3))*size
-
-	#take difference to create production per year
-	prod_shape = diff(cum_production)
-	production<-data.frame(year, prod_shape)
-
-	#Formula: log(production) = f`(time) + beta*log(price) + epsilon
-
-	production$prod<-production$prod_shape*
-	exp(beta0*production$price*rlnorm(length(production$price), meanlog=0, sdlog=.05))
-
-	return(production)
-}
-\end{verbatim}
-
-First, the production time of the field is created as a deterministic function of field size.
-I then model cumulative production from the field over time as a logistic function as in equation \ref{cumProd_equation}.
-This series is then first-differenced to create the general production profile.
-
-\begin{equation}
-% 1/(1+exp(-(prod_time)/3))*size
-cumProd=\frac{size}{1+exp(\frac{-prodTime_t}{3})}
-\label{cumProd_equation}
-
-\end{equation}
-
-I multiply the production profile series with an exponential price term and a coefficient representing the effect of price as well as a random component with mean 0 and standard deviation .05 drawn from a log-normal distribution.
-Taking the logarithm of both sides, the production series for a given field can be written as equation \ref{production_equation}.
-Importantly, the term $\beta$ represents the true effect of price on production.
-
-\begin{equation}
-log(production_t)=log(prodShape_t) + \beta price_t + \epsilon_t
-\label{production_equation}
-\end{equation}
-
-Within the Monte-Carlo loop, production data for each field is generated as above.
-Thus the variation in each run of the Monte Carlo simulation is the random term $\epsilon_t$ from equation \ref{production_equation}.
-
-An example of the data generated can be seen in figure \ref{simulated_production}.
-By inspection, it appears reasonably similar to the data on actual oil production on the Norwegian Continental Shelf.
-
-\begin{figure}
-	\includegraphics[width=1\textwidth]{figures/simulated_production.png}
-	\caption{Simulated production of 77 oil fields}
-	\label{simulated_production}	
-\end{figure}
-
-\subsection{Estimation of Coefficient}
-
-Within the Monte Carlo function, a regression is run using the gam() function from the package mgcv \citet{wood_generalized_2006}.
-This will handle both estimation of linear models or generalized additive models, depending on what formula is passed to the Monte Carlo function.
-
-\begin{verbatim}
-gam_sim<-gam(formula,
-	family=gaussian(link=log), weights=size, data=sim_fields)
-summary(gam_sim)
-\end{verbatim}
-
-Outside of the Monte Carlo function, First I specify a linear regression formula with a 3rd degree polynomial terms to control for the production profile as below.
-
-\begin{verbatim}
-formula_0=formula(prod~time_to_peak + time_to_peak_sq + time_to_peak_cu + 
-	peak_to_end + peak_to_end_sq + peak_to_end_cu +size + price)
-\end{verbatim}
-
-I then replicate the data 1000 times, letting the $\beta$ term equal zero, that is having no effect of prices on oil production.
-The formula used is shown below, where $gam_mc()$ is the Monte Carlo function.
-
-\begin{verbatim}
-gamm_mc_0<-replicate(1000, gam_mc(beta=0, formula=formula_0, use_true_prices=TRUE))
-\end{verbatim}
-
-The above line returns 1000 estimates of $\beta$ from the linear regression specification.  Figure \ref{lin_model_price_mc} shows the results in the form of a histogram.
-
-\begin{figure}
-	\includegraphics[width=1\textwidth]{figures/lin_model_price_mc.png}
-	\caption{Estimated coefficients on price from linear model from Monte Carlo Experiment}
-	\label{lin_model_price_mc}	
-\end{figure}
-
-As the figure shows, a negative bias in the coefficient is apparent.  The distribution appears double-peaked, with most of the estimates of $\beta$ begin between -.05 and -.01.
-
-Now consider a specification using a smooth term for the production profile as below.
-
-\begin{verbatim}
-formula_1= formula(prod~s(prod_time,size) + price)
-\end{verbatim}
-
-Now, the production profile is modeled as a smooth function of production time and size with an additive linear term for price.
-Unlike the specification in the body of the article, I do not specify separate smooth terms for pre- and post-peak.
-This is simply because I know that the logistic function I used to create the shape of the production profile is symmetric.
-
-The results of running the Monte Carlo experiment with the specification with the smoothed term are shown in figure \ref{gam_model_price_mc.png}.
-Here the estimated coefficient on the price term is centered close to zero, correctly reflecting the true $\beta$.
-
-\begin{figure}
-	\includegraphics[width=1\textwidth]{figures/gam_model_price_mc.png}
-	\caption{Estimated coefficients on price from GAM model from Monte Carlo Experiment}
-	\label{gam_model_price_mc.png}	
-\end{figure}
-
 \end{document}

### 2015-03-19 rr_descriptives.py

@@ -13,7 +13,6 @@ import random
 import statsmodels.api as sm
 
 
-brentprices=pd.read_excel("http://www.eia.gov/dnav/pet/hist_xls/RBRTEm.xls", sheetname="Data 1")
 
 
 #first on individual field data

### 2015-03-19 rr_simulation.r

@@ -5,12 +5,14 @@ library(dplyr)
 library(reshape2)
 library(zoo)
 library(sandwich)
+library(grid)
 
 #make function that inputs simulated field_size and outputs simulated
 #first production year
 #should be so that large fields are found more often early on.  
 
 gen_year<-function(size, maxsize){
+	#Generates initial year as a function of size/maxsize
 	#test
 	#size=15
 	#maxsize=1267
@@ -34,6 +36,7 @@ return(year)
 #Simulation of data****************************************************************
 
 gen_production_shape<-function(field){
+	#Given size of field, generates production shape based on
 	#
 	#field<-c(20, 1990, 1)
 	#
@@ -58,33 +61,58 @@ gen_production_shape<-function(field){
 
 
 #GAM simualtion *********************************
-add_uncertainty<-function(beta0=0, fields){ 
+add_uncertainty<-function(beta, fields){ 
 	fields$prod<-fields$prod_shape*
-		(exp(beta0*(fields$price))*
-			rlnorm(length(fields$price), meanlog=0, sdlog=.05))
+		exp(beta*(fields$prices))*
+			rlnorm(length(fields$prices), meanlog=0, sdlog=.10)
 	return(fields)
 }
 
-run_gam<-function(formula, fields, beta=0){
-	#
-	#fields <- sim_fields_unc
-	#formula<-formula(prod ~ s(prod_time) + size + prices)
-	#
+run_gam<-function(formula, fields){
+	#runs GAM model and returns coefficient on price
+	# formula<-formula(prod~s(prod_years, bs="cr") + prices + size)
+
+	# beta<-0.05
+	# fields<-add_uncertainty(fields=sim_fields, beta=0)
+
 	gam_sim<-gam(formula,
 		family=gaussian(link=log), weights=size, data=fields, 
 		na.action='na.omit')
+	coefficients(gam_sim)
 	return(coefficients(gam_sim)["prices"])
 }
 
+run_glm<-function(formula, fields){
+	#test
+	#formula <- formula_glm
+	#fields <- sim_fields_unc
+	#test
+	
+	glm_sim<-lm(formula, data=fields)
+	return(coefficients(glm_sim)["prices"])
+}
 
[…]

### 2015-03-21 mc_sim.rmd
created appendix

---
title: "Monte Carlo Simulations"
output:
  html_document:
    toc: true
    theme: united
---

#Appendix A: Monte Carlo Simulations

In this appendix, I use Monte Carlo simulations to motivate and verify the use of Generalized Additive Models in order to estimate the effect of price on production.  For the sake of the reproducibility, the full r-code and results are imbedded and explained in the document.

First, load some r packages that we will need

```{r}
library(ggplot2) #for plotting
library(reshape2) #for some data manipulation
library(mgcv) #for GAM modeling
```

##Functions
Below is the code for a function that generates a starting year for an oil field conditional on the size of the oil field.  I could use the actual field sizes and starting years from the Norwegian Continental Shelf, but generating them randomly inserts a generality to the Monte Carlo results.

The code takes as input the size of the field, in terms of recoverable oil, and then if the field is smaller than 10 returns a year uniformly distributed between 1970 and 2013.  If the size is larger than 10, then the starting year is taken from a constrained normal distribution with a mean at 1975.  This is meant to reflect the common industry pattern of "creaming" - where large finds tend to be found earlier and which is apparent in the actual field-time distribution on the Norwegian Continental Shelf.

```{r}
gen_year<-function(size){
	#Generates initial year as a function of size/maxsize
	#let small fields be distributed uniformly from 1975 to 2013
	if(size<10){
		year<-trunc(runif(1, 1970, 2013))	
	} else{	
	#while large fields should be more common earlier - from a normal dist.
	#with mean at 1975
	#constrained to be between 1970 and 2013
		range<-FALSE
		while(range==FALSE){
			year<-trunc(rnorm(1,mean=(1975), sd=10))
			ifelse(year>=1970 & year<=2013, range<-TRUE, range<-FALSE)
			}
		}
return(year)
}	
```

The following function takes a variable field with three components: size, starting year, and id. It then generates a time grid of production proportional to the square root of the size.  It then generates a smooth cumulative time path of production based on the logistic function:

$cumProd=\frac{size}{1+exp(\frac{-prodTime_t}{3})}$

The cumulative production series is then first-differenced to created the production profile of the field, and a data frame with the production shape and other field characteristics are returned.

```{r}
gen_production_shape<-function(field){
	#Given size of field, generates production shape based on
	#
	#field<-c(20, 1990, 1)
	#
	size<-as.numeric(field[1])
	start<-as.numeric(field[2])
	name<-field[3]

	t<-trunc(sqrt(size)) + 3
	time<- -t:(t)
	year<-start:(start+2*t-1)

	#use a cumulative logistic function to represent shape of function
	cum_production <- 1/(1+exp(-(time)/3))*size

	#take difference to create production per year
	prod_shape <- diff(cum_production)
	prod_years <- 1:length(prod_shape)
	production<-data.frame(year, prod_shape, name, size, prod_years)
	return(production)
}
```

The next function takes input of a data frame with the smooth production profile and oil prices, as well as the "true" effect of oil price on production beta.  The oil price effect is then added to the production series as well as uncertainty in the form of a draws from a log-normal distribution.	The final production data can then be described as:

$log(production) = f`(time) + beta*log(price) + epsilon$

where f is the logistic cumulative distribution function.

```{r}
add_uncertainty<-function(beta, fields){ 
	fields$prod<-fields$prod_shape*
		exp(beta*(fields$prices))*
			rlnorm(length(fields$prices), meanlog=0, sdlog=.10)
	return(fields)
}
```

The below function takes a regression formula and a data frame of field production data and runs a GAM regression, returning the coefficient on $prices$ variable.
```{r}
run_gam<-function(formula, fields){
	#runs GAM model and returns coefficient on price

	gam_sim<-gam(formula,
		family=gaussian(link=log), weights=size, data=fields, 
		na.action='na.omit')
	coefficients(gam_sim)
	return(coefficients(gam_sim)["prices"])
}
```

The below function takes a formula and a data frame of field production data and runs a GLM model, returning the estimated coefficient on the $prices$ variable. 
```{r}
run_glm<-function(formula, fields){	
	glm_sim<-lm(formula, data=fields)
	return(coefficients(glm_sim)["prices"])
}
```

The below function encapsulates the two core operations involved in each step of the Monte Carlo experiment. The effect of price and uncertainty are added to each field via the "add_uncertainty()" function.  Then either a GAM or GLM regression is fun on the data.

```{r}
mc_run<-function(formula, use_gam=TRUE, fields, beta){
	#adds uncertainty then runs regression
	#
	fields_unc<-add_uncertainty(fields=fields, beta=beta)
	if(use_gam){
		return(run_gam(formula, fields_unc))
	}
	else{
		return(run_glm(formula, fields_unc))
	}
}
```

##Generate fields

```{r}
field_size<-round(exp(rnorm(40,mean=4, sd=1.5)), digits=1)
init_year<-trunc(sapply(field_size, gen_year))
fields<-cbind(field_size, init_year,1:length(init_year))
```

Then, we can load the oil price data

```{r}
oil_price<-read.csv("/Users/johannesmauritzen/research/oil_prices/data/oil_price.csv")
prices<-oil_price[c("year", "oil_price_real")]
names(prices)<-c("year", "prices")
#change units to $10 to aid interpretation
prices$prices<-prices$prices/10
```

Now we create the simulated smoothed production profiles of the fields
```{r}
sim_fields<-apply(fields, 1, gen_production_shape)
sim_fields<-Reduce(rbind, sim_fields)
sim_fields<-merge(sim_fields, prices, by="year")
sim_fields<-sim_fields[order(sim_fields$name, sim_fields$year),]
````

Below we do a single run of generating production uncertainty, while setting the effect of prices to zero to visualize what the data looks like.  It looks reasonable.     
```{r ggplot, fig.width=8, fig.height=5, message=FALSE}
sim_fields_unc<-add_uncertainty(fields=sim_fields, beta=0)
ggplot(sim_fields_unc)+
 geom_line(aes(x=year, y=prod, color=factor(name)),alpha=.3, size=1) +
 guides(color=FALSE) +
 scale_color_grey(start=0, end=.3) +
 theme_bw() +
 labs(x="", y="Simulated Production from Fields")
 ```
###GLM Monte Carlo Experiment
Lets start with a monte-carlo regression using standard GLM, using a 4th-degree polynomial in order to try to control for the production profile of the fields.  
```{r}
formula_glm <- formula(I(log(prod)) ~ prod_years + I(prod_years^2) + I(prod_years^3) + 
	I(prod_years^4) + size + prices + year + I(year^2) + I(year^3))
```

We then do a 100 runs of the Monte Carlo experiment, setting the "true" effect of price, $beta=0$ and $beta=0.05$.  The latter being interpreted as a 10 dollar increase in the oil price leading to 5% more production.

```{r}
glm_mc_beta0<-replicate(100, mc_run(beta=0, formula=formula_glm, fields=sim_fields, use_gam=FALSE))
glm_mc_beta05<-replicate(100, mc_run(beta=0.05, formula=formula_glm, fields=sim_fields, use_gam=FALSE))
```

We transform the data in order to make it easier to visualize
```{r}
glm_mc_data<-data.frame(beta0=glm_mc_beta0, beta05=glm_mc_beta05)
glm_mc_data<-melt(glm_mc_data)
colnames(glm_mc_data)<-c("beta", "coefficient")
levels(glm_mc_data$beta)<-c("0", "0.05")
```

Below we plot the results in the form of emperical density functions of the results.  The vertical black lines represent the true values of beta.  Clearly, the GLM estimation has a significant negative bias.
```{r, fig.width=8, fig.height=5, message=FALSE}
ggplot(glm_mc_data, aes(x=coefficient, fill=beta)) +
geom_density(alpha=.5, size=0) +
geom_vline(aes(xintercept=c(0, 0.05))) +
scale_fill_grey() +
theme_bw() +
guides(fill=FALSE) +
xlab("beta_hat, estimated by GLM") +
ylab("Density")
```
### GAM Monte Carlo Experiment
Below we set the formula as our preffered GAM model with field-level random effects and smooth terms for both the production years and for calendar time, the latter restricted to have a 4-knot basis.

```{r}
formula_1<- formula(prod~s(prod_years, name, bs="re") + prices + size + s(year, bs="cr", k=4))
```

We then run the Monte Carlo experiment 100 times, again setting $\beta=0$ and $\beta=0.05$.

```{r}
gam_mc_beta0<-replicate(100, mc_run(beta=0, formula=formula_1, fields=sim_fields, use_gam=TRUE))
gam_mc_beta05<-replicate(100, mc_run(beta=.05, formula=formula_1, fields=sim_fields, use_gam=TRUE))
```

Converting the data to a format for plotting
```{r}
gam_mc_data<-data.frame(beta0=gam_mc_beta0, beta05=gam_mc_beta05)
gam_mc_data<-melt(gam_mc_data)
colnames(gam_mc_data)<-c("beta", "coefficient")
levels(gam_mc_data$beta)<-c("0", "0.05")
levels(gam_mc_data)
```

Plotting the results, we see that the GAM model does a substantially better job of getting close to the "true" value.  
```{r, fig.width=8, fig.height=5, message=FALSE}
ggplot(gam_mc_data, aes(x=coefficient, fill=beta)) +
geom_density(alpha=.5, size=0) +
geom_vline(aes(xintercept=c(0, 0.05))) +
xlab("beta_hat, estimated by GAM") +
ylab("") +
scale_fill_grey() +
theme_bw()
```

### 2015-03-21 oil_prices_rr.tex
created appendix

@@
 \section{A generalized additive model of oil field production}
-Parametric linear models have the sizeable advantages of simplicity and interpretability and are therefore usually a good starting point for an analysis.
-However, when attempting to model the effect of price on oil field production, a standard linear model is unable to sufficiently control for the production profile and therefore heavily biases the estimate of the effect of price.
+Parametric linear models have the sizeable advantages of simplicity and interpretability and are therefore usually a good starting point for an analysis.However, when attempting to model the effect of price on oil field production, a standard linear model is unable to sufficiently control for the production profile and therefore heavily biases the estimate of the effect of price.
 As an example, consider a generalized linear model written as in equation \ref{glm_eqn}.

### 2015-03-21 rr_simulation.r

@@ -6,40 +6,33 @@ library(reshape2)
 library(zoo)
 library(sandwich)
 library(grid)
+library(rmarkdown)
 
+#render rmd file
+rmarkdown::render("/Users/johannesmauritzen/research/oil_prices/mc_sim.rmd")
 #make function that inputs simulated field_size and outputs simulated
+
 #first production year
 #should be so that large fields are found more often early on.  
 
-gen_year<-function(size, maxsize){
-	#Generates initial year as a function of size/maxsize
-	#test
-	#size=15
-	#maxsize=1267
-	#
-
-	#let small fields be distributed uniformly from 1975 to 2013
-	if(size<10){
-		year<-trunc(runif(1, 1975, 2008))	
-	} else{	#while large fields should be more common earlier on
-		range<-FALSE
-		while(range==FALSE){
-			year<-trunc(rnorm(1,mean=(1973+(maxsize+300)/(size+300)), sd=10))
-			ifelse(year>=1970 & year<=2013, range<-TRUE, range<-FALSE)
-			#print(paste("in while loop", year))
-			}	
+gen_year<-function(size){
+	#while large fields should be more common earlier - from a normal dist.
+	#with mean at 1970
+	#constrained to be between 1970 and 2013
+	range<-FALSE
+	while(range==FALSE){
+		year<-trunc(rnorm(1,mean=(1970), sd=10))
+		ifelse(year>=1970 & year<=1990, range<-TRUE, range<-FALSE)
 		}
-return(year)	
+		
+return(year)
 }
 
 
 #Simulation of data****************************************************************
 
 gen_production_shape<-function(field){
-	#Given size of field, generates production shape based on
-	#
-	#field<-c(20, 1990, 1)
-	#
+	#Given size of field, generates production shape
 	size<-as.numeric(field[1])
 	start<-as.numeric(field[2])
 	name<-field[3]
@@ -64,17 +57,12 @@ gen_production_shape<-function(field){
 add_uncertainty<-function(beta, fields){ 
 	fields$prod<-fields$prod_shape*
 		exp(beta*(fields$prices))*
-			rlnorm(length(fields$prices), meanlog=0, sdlog=.10)
[…]

### 2015-03-23 mc_sim.rmd
redid analysis with 2d smooth

@@
 
-First, load some r packages that we will need
+First, we load some r packages that we will need:
 
@@
 
-The code takes as input the size of the field, in terms of recoverable oil, and then if the field is smaller than 10 returns a year uniformly distributed between 1970 and 2013.
-If the size is larger than 10, then the starting year is taken from a constrained normal distribution with a mean at 1975.
-This is meant to reflect the common industry pattern of "creaming" - where large finds tend to be found earlier and which is apparent in the actual field-time distribution on the Norwegian Continental Shelf.
+The function generates a starting year for field production from a half-normal distribution with a mean at 1970, but constrained to be between 1970 to 2000.
+This is meant to reflect the common industry pattern that most significant finds are found early in a regions activity.
 
@@
 gen_year<-function(size){
-	#Generates initial year as a function of size/maxsize
-	#let small fields be distributed uniformly from 1975 to 2013
-	if(size<10){
-		year<-trunc(runif(1, 1970, 2013))	
-	} else{	
-	#while large fields should be more common earlier - from a normal dist.
-	#with mean at 1975
-	#constrained to be between 1970 and 2013
+	#half-normal distribution with mean at 1970
+	#constrained to be between 1970 and 2000
 	range<-FALSE
 	while(range==FALSE){
-			year<-trunc(rnorm(1,mean=(1975), sd=10))
-			ifelse(year>=1970 & year<=2013, range<-TRUE, range<-FALSE)
+		year<-trunc(rnorm(1,mean=(1970), sd=10))
+		ifelse(year>=1970 & year<=2000, range<-TRUE, range<-FALSE)
 		}
-		}
+
 return(year)
@@
 The effect of price and uncertainty are added to each field via the "add_uncertainty()" function.
-Then either a GAM or GLM regression is fun on the data.
+Then either a GAM or GLM regression is run on the data.
 
@@
 
-Below we plot the results in the form of emperical density functions of the results.
+Below we plot the results in the form of empirical density functions of the results.
 The vertical black lines represent the true values of beta.
@@
 ### GAM Monte Carlo Experiment
-Below we set the formula as our preffered GAM model with field-level random effects and smooth terms for both the production years and for calendar time, the latter restricted to have a 4-knot basis.
+Below we set the formula as a slightly simplified version of our preffered GAM model.
+I have removed the cost index as this was not taken into account when generating the data.
+For simplicity I also do not take into account lagged price terms as the main point of this monte-carlo experiment is to show how the production profile can be effectively controlled for.
 
 ```{r}
-formula_1<- formula(prod~s(prod_years, name, bs="re") + prices + size + s(year, bs="cr", k=4))
+formula_1<- formula(prod~s(prod_years, size) + prices + year + I(year^2))
 ```

### 2015-03-23 rr_analysis_1.r

@@ -33,41 +33,80 @@ large_fields<-prod_data[prod_data["name"]!="EKOFISK",]
 
 gam_mod<-gam(oil_prod_mill_sm3~s(prod_year, bs="cr") + 
 	I(in_place_oil_mill_sm3/100) + I(cost_index/10) +
-    s(year, bs="cr", k=4) + I(price/10) + I(price_l1/10) + I(price_l2/10) +
+    year + I(year^2) + I(price/10) + I(price_l1/10) + I(price_l2/10) +
     I(price_l3) + I(price_l4) + I(price_l5/10) + I(price_l6/10) + 
     I(price_l7/10) + I(price_l8/10), 
     family=gaussian(link=log), weights=in_place_oil_mill_sm3, data=large_fields)
 summary(gam_mod)
 
+
+
+gam_mod_2d<-gam(oil_prod_mill_sm3~s(prod_year, in_place_oil_mill_sm3) + 
+  + I(cost_index/10) +
+    year + I(year^2) + (year^3) + I(price/10) + I(price_l1/10) + I(price_l2/10) +
+    I(price_l3/10) + I(price_l4/10) + I(price_l5/10) + I(price_l6/10) + 
+    I(price_l7/10) + I(price_l8/10), 
+    family=gaussian(link=log), weights=in_place_oil_mill_sm3, data=large_fields)
+summary(gam_mod_2d)
+
+coef<-gam_mod_2d$coefficients[5:13]
+se<-summary(gam_mod_2d)$p.table[5:13,2]
+ymin<-coef-2*se
+ymax<-coef+2*se
+gam_2d_coef<-data.frame(coef=coef, se=se, lag=0:8, ymin=ymin, ymax=ymax)
+
+2d_coef_plot<-ggplot(gam_2d_coef) +
+geom_pointrange(aes(x=lag, y=coef, ymin = ymin, ymax = ymax)) +
+geom_hline(aes(y=0)) +
+theme_bw() +
+labs(x="Lags", y="Estimated Coefficient on Oil Price")
+
+
 #With random effects
 
 gam_mod_re<-gam(oil_prod_mill_sm3~s(prod_year, name, bs="re") + 
 	I(in_place_oil_mill_sm3/100) + I(cost_index/10) +
-     s(year, bs="cr", k=4) + I(price/10) + I(price_l1/10) + I(price_l2/10) + 
-     I(price_l3) + I(price_l4) + I(price_l5/10) + I(price_l6/10) + 
[…]

### 2015-03-23 rr_clean.py

@@ -149,3 +149,31 @@ def create_prod_year(year):
 prod_data['prod_year']=prod_data.groupby("name")["year"].transform(create_prod_year)
 
 prod_data.to_csv('/Users/johannesmauritzen/research/oil_prices/data/prod_data.csv', index=False)
+
+
+#add build-out dummy
+prod_data=pd.read_csv('/Users/johannesmauritzen/research/oil_prices/data/prod_data.csv')
+
+def build_out_phase(field):
+	"""
+	returns dummy variables with input of fields
+	"""
+	#test
+	#field=prod_data[prod_data.name=="STATFJORD"]
+	#
+	prod_max=field.oil_prod_mill_sm3.max()
+	year_max=field.year[field.oil_prod_mill_sm3==prod_max].max()
+	field["build_out"]=0
+	field["build_out"][field.year<year_max]=1
+	return(field.build_out)
+
+build_out=prod_data.groupby("name").apply(build_out_phase).reset_index()
+
+prod_data["build_out"]=build_out["build_out"]
+
+prod_data.to_csv('/Users/johannesmauritzen/research/oil_prices/data/prod_data.csv', index=False)
+
+
+
+
+

### 2015-03-23 rr_simulation.r

@@ -57,7 +57,7 @@ gen_production_shape<-function(field){
 add_uncertainty<-function(beta, fields){ 
 	fields$prod<-fields$prod_shape*
 		exp(beta*(fields$prices))*
-			rlnorm(length(fields$prices), meanlog=0, sdlog=.03)
+			rlnorm(length(fields$prices), meanlog=0, sdlog=.05)
 	return(fields)
 }
 
@@ -123,7 +123,7 @@ simulated_production<-ggplot(sim_fields_unc)+
 formula_1<- formula(prod~s(prod_years, name, bs="re") + prices + size + s(year, bs="cr", k=4))
 formula_2<- formula(prod~s(prod_years, bs="cr") + prices + size + s(year, bs="cr", k=4))
 formula_3<- formula(prod~s(prod_years, bs="cr") + prices + size)
-formula_4<- formula(prod~s(prod_years, size) + prices + s(year, bs="cr", k=4))
+formula_4<- formula(prod~s(prod_years, size) + prices + year + I(year^2))
 
 gam_mc_beta0<-replicate(100, mc_run(beta=0, formula=formula_4, fields=sim_fields, use_gam=TRUE))
 gam_mc_beta05<-replicate(100, mc_run(beta=.05, formula=formula_4, fields=sim_fields, use_gam=TRUE))

### 2015-03-31 mc_sim.rmd
update on paper

@@
 
-The function generates a starting year for field production from a half-normal distribution with a mean at 1970, but constrained to be between 1970 to 2000.
+The function generates a starting year for field production from a half-normal distribution with a mean at 1970, but constrained to be between 1970 to 1990.
 This is meant to reflect the common industry pattern that most significant finds are found early in a regions activity.
@@
 ```{r}
-gen_year<-function(size){
+gen_year<-function(sizexz){
 	#half-normal distribution with mean at 1970
-	#constrained to be between 1970 and 2000
+	#constrained to be between 1970 and 1990
 	range<-FALSE
@@
 		exp(beta*(fields$prices))*
-			rlnorm(length(fields$prices), meanlog=0, sdlog=.10)
+			rlnorm(length(fields$prices), meanlog=0, sdlog=.05)
 	return(fields)
@@
 The below function takes a regression formula and a data frame of field production data and runs a GAM regression, returning the coefficient on $prices$ variable.
+
 ```{r}
@@
 The below function takes a formula and a data frame of field production data and runs a GLM model, returning the estimated coefficient on the $prices$ variable.
+
 ```{r}
@@
 Now we create the simulated smoothed production profiles of the fields
+
 ```{r}
@@
 Below we do a single run of generating production uncertainty, while setting the effect of prices to zero to visualize what the data looks like.  It looks reasonable.
+
 ```{r ggplot, fig.width=8, fig.height=5, message=FALSE}
@@
  ```
-###GLM Monte Carlo Experiment
+
+##GLM Monte Carlo Experiment
+
 Lets start with a monte-carlo regression using standard GLM, using a 4th-degree polynomial in order to try to control for the production profile of the fields.
+
 ```{r}
@@
 We transform the data in order to make it easier to visualize
+
 ```{r}
@@
 Clearly, the GLM estimation has a significant negative bias.
+
 ```{r, fig.width=8, fig.height=5, message=FALSE}
@@
 ```
-### GAM Monte Carlo Experiment
+
+## GAM Monte Carlo Experiment
+
 Below we set the formula as a slightly simplified version of our preffered GAM model.
@@
 Converting the data to a format for plotting
+
 ```{r}
@@
 Plotting the results, we see that the GAM model does a substantially better job of getting close to the "true" value.
+
 ```{r, fig.width=8, fig.height=5, message=FALSE}

### 2015-03-31 oil_prices_old.tex
update on paper

@@
 This estimate is heavily biased downwards due to the spurious correlation between the field production profiles and the autocorrelated and non-stationary time series of oil prices.
-A more formal analysis of the bias can be found in the online appendix, \url{jmaurit.github.io#oil_price_sim}, where I simulated data of oil field production in a monte carlo experiment.
+A more formal analysis of the bias can be found in the on-line appendix, \url{jmaurit.github.io#oil_price_sim}, where I simulated data of oil field production in a monte carlo experiment.
 However, the basic problem is that the parametric representation is not flexible enough to control for the production profiles of the fields.

### 2015-03-31 oil_prices_rr.tex
update on paper

--- 2015-03-19 oil_prices_rr.tex
@@
-% revised paper for Energy Journal
-
 \documentclass[11pt]{article}
@@
 \doublespacing % from package setspacs
-
-% table font size
-\let\oldtabular\tabular
-\renewcommand{\tabular}{\scriptsize\oldtabular}
 
@@
 I use detailed field-level data on Norwegian off-shore oil field production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production.
-I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 2 to 4\% for a 10 dollar per barrel increase in the real price of oil.
-Most of this effect appears to come in the planning phase of a field's development.\\
-Keywords: Oil production,Oil Prices, Norway, Semiparametric.
+I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 3 to 5\% for a 10 dollar per barrel increase in the real price of oil.
+Most of this effect appears to come in the build-out or planning phase of a field's development.\\
+Keywords: Oil production,Oil Prices, Norway, Generalized Additive Model.
 \end{abstract}
@@
 For most of the last century, crude oil has been the world’s single most important and valuable fuel source.\footnote{Oil is the world's largest single energy source, consisting of approximately 33\% of total energy consumption in 2013 as well as the most valuable in terms of market price per unit of energy \citep{british_petroleum_statistical_2013}} Naturally, questions of how the oil price affects the world economy as well as how oil production reacts to the oil price have been fundamental topics in economics.
+Recent volatility in the price of oil has made this an especially relevant topic.
 
@@
 
-The main finding in this article is that oil production at the field level has no significant reaction to concurrent changes in the oil price, where concurrent is broadly defined as within the first three years.
-A slight effect can be detected at a lag of between 4 and 8 years, with a magnitude of about a 2 to 4\% increase in yearly production for a 10 dollar increase in the price of oil.
-This effect is somewhat greater and with less of a lag in large fields compared to small fields.
-Price appears to have the most significant affect during the planning stage - before production begins in a field.
-In the depletion phase of production, price is found to have little to no significant effect.
+The main finding in this article is that oil production at the field level has no significant reaction to concurrent changes in the oil price.
+A slight effect can be detected at a lag of between 1 and 4 years, with a magnitude of about a 2 to 5\% increase in yearly production for a 10 dollar increase in the price of oil.
+
+% Price appears to have the most significant affect during the planning and build-out stage.  In the depletion phase of production, price is found to have little to no significant effect.
 
@@
 More so, oil field production is correlated across fields - that is, increases and decreases in production in fields are not randomly distributed across time.
-Instead, as figure \ref{top10_production} shows with the production profile of the 10 largest Norwegian oil fields, production tends to be correlated across fields.
-The result is a total production curve that is bell-shaped over time as in figure \ref{oil_decline}.
+Instead, as the top panel of figure \ref{oil_decline} shows with the production profile of the 10 largest Norwegian oil fields, production tends to be correlated across fields.
+The result is a total production curve that is bell-shaped over time as in the lower panel of figure \ref{oil_decline}.
 Since the oil price series is autocorrelated and non-stationary, failure to properly account for the production profile will lead to spurious estimation of the price terms in a regression.
+
+\begin{figure}
+	\includegraphics[width=1\textwidth]{figures/oil_decline.png}
+	\caption{The panel shows the production from the largest 10 fields, which are temporally correlated with each other.  The total production over time is bell-shaped, as shown by the bottom panel.}
+	\label{oil_decline}
+\end{figure}
 
@@
 As a solution I use a semi-parametric model within the Generalized Additive Model frameworks of \cite{hastie_generalized_1990}.
-Here I use a cubic spline function to account for the general non-parametric shape of the production profile while allowing price to enter the equation linearly while also attempting to control for the effects of varying production costs and the effects of technological change and other time-varying effects.
+Here I use both a cubic spline and a two-dimensional thin-plate spline to account for the general non-parametric shape of the production profile and allowing price to enter the equation linearly while also attempting to control for the effects of varying production costs and the effects of technological change and other time-varying effects.
 
-The results suggest that little to no concurrent effect of oil prices on production exists.  However modest lagged effects of between 2 to 5 years exists.  These results are in line with the intuition
+The results suggest that little to no concurrent effect of oil prices on production exists.
+However modest lagged effects of between 1 to 4 years exist.These results are in line with an understanding of the industry as being highly capital intensive with high operating costs.
+Maintaining spare capacity in order to adjust to changing oil prices is likely prohibitively inexpensive.
+Instead, production is increased through investment, which may take up to several years to show an effect.
 
@@
 For modeling aggregate oil production, shape-fitting models, notably \citet{hubbert_energy_1962} and more recently advocated by \citet{deffeyes_hubberts_2001}, have had some success in estimating the timing of peak production at the regional and national level, but they tend to seriously underestimate the total recoverable resource of oil-producing regions and the models can be shown to be fundamentally misspecified \citep{boyce_prediction_2013}.
-Simulation type studies where aggregate oil production is modeled through an often complex combination of physical and economic processes also exist in the geo-engineering literature, but their usefulness tends to be weighed down by their complexity as they require quite detailed data and specific assumptions about functional form that can be difficult to justify \citet{brandt_review_2010}.
+Simulation type studies where aggregate oil production is modeled through an often complex combination of physical and economic processes also exist in the geo-engineering literature, but their usefulness tends to be weighed down by their complexity as they require quite detailed data and specific assumptions about functional form that can be difficult to justify \citep{brandt_review_2010}.
 
@@
 A model and test using data from North Sea producers on the UK continental shelf by \citet{hurn_geology_1994}, on the other hand, fails to find evidence that the variance in the oil price affects the timing of oil field development.
+
 I do not attempt to directly model uncertainty, however given that the investments needed to increase oil production in an existing field are to a certain extent irreversible and that oil prices are highly volatile, the results can and probably should be interpreted with the real options framework in mind.
@@
 \citet{black_is_1998} tests the relevance of “nesting” a structural empirical model of profit maximization that takes into account oil prices into a typical geo-engineering model of oil field production.
-They find strong evidence that taking into account profit maximization, and implicitly price, substantially improves the fit compared to a purely geo-engineering type model.
+They find strong evidence that taking into account profit maximization - and implicitly price - substantially improves the fit compared to a purely geo-engineering type model.
 The limitation of their methodology is that they are only able to test whether including economic factors like price affects the fit of the model but are not able to give an estimate of the effect.
@@
 
-\citet{rao_taxation_2010} uses data on land based oil wells in California to estimate the effects of tax changes and price controls.
+\citet{rao_taxation_2010} uses data on land-based oil wells in California to estimate the effects of tax changes and price controls.
 The author finds that short-term tax changes caused small, but significant retiming of production from oil wells.
 These findings are over-all consistent with the findings of this paper, though the author finds concurrent effects of taxation on production while I do not find any concurrent effect of prices on production.
-This difference is most likely due to the significant differences in cost and complexity between operating offshore and onshore.
+The significant differences in cost and complexity between operating offshore and onshore likely account for the different results.
 How producers react to a short-term tax change as opposed to a change in price is likely also different.
-More so, the author does not consider differences between the different stages of production.
 
@@
 \citet{mohn_exploration_2008} finds that long-term changes in the oil-price has a strong effect on exploratory drilling though little effect is measured from short-term changes in the oil price.
-\citet{osmundsen_exploration_2010} analyses drilling productivity over time on the Norwegian Continental Shelf while \citet{mohn_efforts_2008} finds that higher oil prices leads to higher reserves and as well as that oil prices affect producer risk-preferences - with higher prices leading to lower success rates but larger discovery size.
+\citet{osmundsen_exploration_2010} analyses drilling productivity over time on the Norwegian Continental Shelf while \citet{mohn_efforts_2008} finds that higher oil prices leads to higher reserves as well as that oil prices affect producer risk-preferences - with higher prices leading to lower success rates but larger discovery size.
 
@@
 The first commercial oil well in Norwegian continental waters was discovered in December of 1969 in what became the Ekofisk oil field, the largest Norwegian oil field by estimated recoverable reserves.
-As figure \ref{north_sea_reserves} shows, most of the largest fields in the North Sea were found relatively early on while more recent finds have tended to be smaller - a pattern typical of oil producing areas called creaming.
+Most of the largest fields in the North Sea were found relatively early on while more recent finds have tended to be smaller - a pattern typical of oil producing areas called creaming.
 A major exception to this trend was the recent find of the Johan Sverdrup field which is estimated to have approximately 300 million Standard Cubic Meters (SM3) of recoverable oil.
@@
 Exploration in the Norwegian Sea was opened in the early 1980’s and the first commercial field started production in 1981.
-While several mid-sized fields have been discovered, the Norwegian Sea has generally disappointed in terms of commercial oil finds and most finds have been relatively small (\ref{norwegian_sea_reserves}).
+While several mid-sized fields have been discovered, the Norwegian Sea has generally disappointed in terms of commercial oil finds and most finds have been relatively small.
 
 Norwegian waters in the Barents Sea have also been open to exploration since the 1980s, however up until recently only a few, small finds were made and none came into commercial production.
-However recently several significant oil and gas finds have been made in the Barents Sea - notably the oil field Goliat and the large gas find Sn\o hvit, which are both currently under development but not yet producing.
+Recently several significant oil and gas finds have been made in the Barents Sea - notably the oil field Goliat and the large gas find Sn\o hvit, which are both currently under development but not yet producing.
 The agreement between Norway and Russia in June 2011 settling a long-running dispute over the maritime delimitation has also given a boost to new exploration in the region.
 
-Profits from oil and gas production in Norway are subject to a resource tax of 50 \% on top of the ordinary income tax of 28 \%, thus income from petroleum production is taxed at a total marginal tax rate of 78 \%.
+Profits from oil and gas production in Norway are subject to a resource tax of 51 \% on top of the ordinary income tax of 27 \%, thus income from petroleum production is taxed at a total marginal tax rate of 78 \%.
 The central government also receives revenues through ownership stakes in companies, notably Statoil, where the state is the majority stakeholder.
+
 The over-all tax rate has been fairly constant through the history of Norwegian oil production, however several important changes related to the tax code have occurred.
@@
 
-More importantly to the offshore sector were accounting changes that were implemented in 2002 and 2005 that were meant to encourage new entrants by allowing losses to be carried forward for tax purposes and by introducing a rebate on the tax-value of losses associated with searching and drilling.
+More important to the offshore sector were accounting changes that were implemented in 2002 and 2005, which were meant to encourage new entrants by allowing losses to be carried forward for tax purposes and by introducing a rebate on the tax-value of losses associated with searching and drilling.
 In general though, these rules mainly affected searching and discoveries of new fields rather than production from existing fields and I do not control for the tax changes in my model.
@@
 
-\section{Norwegian oil field production data}
-Production data of Norwegian oil fields is obtained from the website of the Norwegian Petroleum Directorate.
-\footnote{http://factpages.npd.no/}  Production data is available at a monthly frequency for all fields, though I choose to aggregate up to yearly values both to smooth over seasonality as well as short-term volatility of output due to factors such as weather or technical issues that are not relevant for this article.
+\section{Data}
+Production data of Norwegian oil fields is obtained from the website of the Norwegian Petroleum Directorate.\footnote{http://factpages.npd.no/} The data I use includes all current or formerly oil-producing fields, from 1971 through 2014.
+Production data is available at a monthly frequency, though I choose to aggregate up to yearly values both to smooth over seasonality as well as short-term volatility of output due to factors such as weather or technical issues that are not relevant for this article.
 
-In addition to data on field-level production, I also make use of data on estimated total recoverable reserves.
-The use of this variable is complicated as it is an estimate subject to a large amount of uncertainty, especially in relatively young fields.
-However, the methodologies used to estimate the total recoverable resource of a field are constantly evolving and it is a fair assumption that any consistent bias of the estimates are observed in older fields and corrected for in estimates for newer fields.
-I can then assume that existing errors are random and will not significantly bias the estimates.
+In addition to data on field-level production, I also have access to data on estimates of total recoverable reserves as well as estimates of total in-place oil for each field.
+The use of total recoverable reserves is complicated since it is dependent on the price of oil and thus endogenous in the analysis.
+Therefor, as a proxy for the size of the fields, I use estimates of in-place oil from the Norwegian Petroleum Directorate, which is unlikely to be correlated with prices.
+The left panels of figure \ref{data_descriptives} show the histogram of field sizes in my data set.
 
-Moreover, the estimate is likely endogenous, in the sense that it is also effected by prices.
-However since I use the variable as a control variable and not for the purpose of estimating a parameter with a causal interpretation, this should not significantly affect the validity of the results.
+\begin{figure}
+	\includegraphics[width=1\textwidth]{figures/data_descriptives.png}
+	\caption{Data descriptives. The left panels show histograms of estimates of recoverable oil and in-place oil by field. The right panels shows the production cost index and real brent oil prices.}
+	\label{data_descriptives}
+\end{figure}
 
-I use yearly data from the US Energy Information Agency on the real price of Brent-traded oil in 2010 dollars.The Brent benchmark oil price is likely the best oil price measure for Norwegian production as it is based on light sweet crude oil sourced from the North Sea.
+To simplify the analysis, I eliminate the smallest fields from my analysis - those with less than 3 million Standard Cubic Meters (SM3) of recoverable reserves.
+These fields tend to have substantially different production profiles than the larger fields.
+However, these fields only account for around .5 \% of total oil production so this exclusion is unlikely to have any serious practical implications.
+
+I also eliminate the large Ekofisk field from my analysis.
+I do this because it has a unique double-humped production profile.
+Ekofisk was the first field to be found and the first to start production on the Norwegian Continental Shelf.
+It reached an initial peak in production in the late 1970's and began depleting.
+However, the introduction of water and gas injection technology led to a boost in production and a new peak in production in the early 2000's.
+All subsequent major fields used gas and water injection and have a single-hump profile.
+
+I use yearly data from the US Energy Information Agency on the real price of Brent-traded oil in 2010 dollars.
+The Brent benchmark oil price is likely the best oil price measure for Norwegian production as it is based on light sweet crude oil sourced from the North Sea.
+
+Production costs are known to vary substantially with the price of oil.
+To control for this, I include the producer price index for Norwegian oil and gas extraction industry, obtained from Statistics Norway.
+\footnote{\url{https://www.ssb.no/en/priser-og-prisindekser/statistikker/ppi}} This series only goes back to 2000, however, so I also include an index of oil and gas well costs from the US energy information agency for the years 1960 to 2000.
+\footnote{\url{http://www.eia.gov/dnav/pet/pet_crd_wellcost_s1_a.htm}} While this measure is unlikely to be a perfect substitute for costs in Norway, oil and gas extraction is a global industry, where costs are highly correlated across regions.
+The combined production cost index and the real Brent oil price series are shown in the right panels in figure \ref{data_descriptives}.
 
@@
 On the other hand, given the size and liquidity of oil spot markets, it is a fair assumption that the current oil prices do a good job of incorporating much of the available information about crude oil markets and that future price movements are generally difficult to predict \citep{hamilton_understanding_2008}.
+
 An active futures market does exist, but several studies have found that current oil prices are in general better than prices on futures contracts at predicting future oil prices \citep{alquist_what_2010, chinn_predictive_2005}.
@@
 
-A cleaned data set and the full code for the analysis are available at my website \url{jmaurit.github.io#oil_prices}.
+A cleaned data set and the full code for the analysis are available at \url{jmaurit.github.io/#oil_prices}.
 
@@
 However, when attempting to model the effect of price on oil field production, a standard linear model is unable to sufficiently control for the production profile and therefore heavily biases the estimate of the effect of price.
-As an example, consider a generalized linear model written as in equation \ref{glm_eqn}.
-
-	\begin{equation}
-	\begin{split}	
-	 Log(Production_{i,t})  &= \alpha_0 + \alpha_1 timetopeak_{i,t} + \alpha_2 timetopeak_{i,t}^2 \\
-	  \quad & + \alpha_3 timetopeak_{i,t}^3  + \alpha_4 peaktoend_{i,t} + \alpha_5 peaktoend_{i,t}^2 \\
-	  \quad & + \alpha_6 peaktoend_{i,t}^3 + \gamma totalrecoverableoil_i \\
-	  \quad & + \beta_1 oilprice + \beta_2 oilpricel1 + ...+ \beta_6 oilpricel5 + \epsilon	
-	\label{glm_eqn}
-	\end{split}
-	\end{equation}
-
-This estimate is heavily biased downwards due to the spurious correlation between the field production profiles and the autocorrelated and non-stationary time series of oil prices.
-A more formal analysis of the bias can be found in Appendix B, where I simulate data of oil field production in a Monte Carlo experiment.
-However, the basic idea is that the parametric representation is not flexible enough to control for the production profiles of the fields.
-Instead, a more flexible estimation of the production profile is needed.
 
 Instead of attempting to estimate the shape of the production profiles of the fields by estimating parameters on linear terms I estimate a non-parametric function for the production profile.
-As an illustration, consider the production profile of a single field.
-The simplest possible model would then have the form of equation \ref{simp_eqn}.
+At a conceptual level, this model is decomposing the variation in the data into a smoothed component representing the technical and physical forces that produce the typical bell-shaped production profile, and the remaining component, that I estimate the effect of price with.
 
-In principle any number of well-behaved smoothers could be used to estimate the function, for example a Loess or a kernel smoother.
-In practice regression splines are most commonly used.
-With a regression spline the data points for the function to be estimated are broken into bins.
-For each bin of data a local linear regression is estimated.
-For functions of one variable, a cubic parameterization is often used.
-These regressions are then essentially tied together at what are called “knots” and the smoothness of the overall function is controlled by a penalty function that consists of the second derivative of the estimated function.
-
-The advantage to using splines over other smoothing methods is that it can be represented in a linear form.
-Thus estimation of the model can be done using standard and efficient matrix algebra algorithms.
-For further details I refer to the discussion in \citet{hastie_generalized_1990} and \citet{wood_generalized_2006}.
-The latter is a particularly useful reference for implementing generalized additive models in R.
-
-I do not want to estimate smoothed curves individually for each field.
+I do not, however, want to estimate smoothed curves individually for each field.
 While this would provide a good overall fit to the full data set, not enough variation in the data would be left to estimate the effect of price.
 Instead I want to estimate a general shape of the production profile for all fields and then use the remaining variation in the data to estimate the effect of price.
-My model can be written as in equations \ref{gam_price_eqn}.
+
+A simple model can be written as in equation \ref{gam_price_eqn}.
 
@@
 
-	Log(Production_{i,t}) & = f(time\_to\_peak_{i,t}, total\_recoverable\_oil_i) \\
- 	 \quad & + f(peak\_to\_end_{i,t}, total\_recoverable\_oil_i) \\
-	 \quad & + \beta_1 oil\_price + \beta_2 oil\_price\_l1 + ... +  \epsilon
+	Log(production_{i,t}) & = f(production\_time_{i,t}) + \beta_1 in\_place\_oil_{i,t} \\
+	 \quad & + \beta_2 oil\_price\_lags_t + \beta_3 cost\_index_{t} +  
+	 f(year_t) + \epsilon_{i,t}
 \label{gam_price_eqn}
@@
 In this model I am estimating the parameters and functions from all fields $i$.
-As in the parametric model presented earlier, the left-hand-side variable is yearly oil production for field $i$.
-Also like the parametric model I split the production-time component element in two: up to and after the peak in production.
+The left-hand-side variable, $Log(production_{i,t})$ is the yearly oil production for field $i$ at production time $t$ in millions of standard cubic meters (SM3).
+On the right hand side is a smoothed function of the production profile over time $f(production\_time_{i,t})$, $in\_place\_oil_{i,t}$ is the estimated in place oil of the field, and $cost\_index_{t}$ represents the producer price index for oil and gas extraction.
+
+$f(year_t)$ represents a function of the calendar year.
+The main purpose of including this term is to control for the effects of technological change.
+I could use a smoothed function here as well, however in order to avoid over fitting the model, I use a simple quadratic representation.
+
+Finally, $oil\_price\_lags_t$ represents the variables of interest - a vector of concurrent and up to 8 lagged terms for the oil price.
+I also experimented with using quadratic terms for the oil price and its lags, however these could not be shown to be significant.
+
+The smoothed function is estimated using a cubic regression spline...
+
+One objection to this model may be that a fixed effect controlling for the general production profile of oil fields does not sufficiently address idiosyncratic differences between the fields.
+To address this I can add the term, $\beta_r Z$, where $Z$ represents field random effects that takes into account uncertainty based on random variation between the fields.
+
+Another substantial objection may be that the smoothed function of production time may not be sufficient in controlling for the variation in production profiles shapes across fields.
+In particular, production profiles differ substantially by the size of the field.
+A solution is to allow the shape to vary by production.
+Thus the smoothed can be written as a function of both production time and in place oil: $f(production\_time_{i,t}, in\_place\_oil_{i,t})$.
+
+Estimating a smoothed function of two variables requires a somewhat more involved method.
+In particular I use Thin-plate regression spline \citep{wood_thin_2003}.
+Consider equation \ref{thin_plate_1}.
+
+	\begin{equation}
+	y_i = g(x_1, x_2)
+	\label{thin_plate_1}
+	\end{equation}
+
+Following the notation of \citet{wood_generalized_2006}, $g$ is the function of $x_1$ and $x_2$ that is to be estimated by $f$, which in turn is estimated by minimizing equation \ref{thin_plate_2}.
+Here $\boldsymbol{y}$ represents a vector of $y_i$’s and $\boldsymbol{f} = (f(\boldsymbol{x_1}),f(\boldsymbol{x_2}))^t$.
+
+	\begin{equation}
+\min \|\boldsymbol{y-f}\|^2 + \lambda J_{22}(f)
+\label{thin_plate_2}
+	\end{equation}
+
+$J_{22}$ represents the penalty function for the smoothness of the function which can be written as in \ref{thin_plate_3}.
+The $22$ represents the fact that it is a penalty function of two variables with smoothness measured by the second derivative.
+
+	\begin{equation}
+	J_{22}{f}= \diffp[2]{f}{x_1}^2 + \diffp[2]{f}{{x_1}{x_2}} + \diffp[2]{f}{x_2}^2dx_1 dx_2
+\label{thin_plate_3}
+	\end{equation}
+
+In short, a function of $x_1$ and $x_2$ is found that is minimizing errors in the sense of minimizing Euclidean distance subject to a penalty function of “wiggiliness”.
+The actual implementation is somewhat more involved in order to increase the computational efficiency of the estimation.
+For further details I again refer to \citet{wood_thin_2003}.
 
@@
 \hline
-                                 & w/out Rand. Effects & w Rand. Effects & Lags 1-5 \\
+                                             & w/out Rand. Effects & w Rand. Effects & 2-d Smooth \\
 \hline
-(Intercept)                      & $0.335$        & $-4.016^{***}$ & $-3.971^{***}$ \\
-                                 & $(0.265)$      & $(0.297)$      & $(0.212)$      \\
-I(in\_place\_oil\_mill\_sm3/100) & $0.337^{***}$  & $-0.014$       & $-0.015$       \\
-                                 & $(0.009)$      & $(0.022)$      & $(0.022)$      \\
-I(cost\_index/10)                & $-0.018^{**}$  & $-0.030^{***}$ & $-0.030^{***}$ \\
-                                 & $(0.006)$      & $(0.006)$      & $(0.005)$      \\
-I(price/10)                      & $-0.006$       & $0.014$        &                \\
-                                 & $(0.017)$      & $(0.016)$      &                \\
-I(price\_l1/10)                  & $0.004$        & $0.021$        & $0.029^{*}$    \\
-                                 & $(0.017)$      & $(0.017)$      & $(0.014)$      \\
-I(price\_l2/10)                  & $0.037^{*}$    & $0.047^{**}$   & $0.048^{***}$  \\
-                                 & $(0.017)$      & $(0.016)$      & $(0.014)$      \\
-I(price\_l3)                     & $0.003$        & $0.003$        & $0.002$        \\
-                                 & $(0.002)$      & $(0.001)$      & $(0.001)$      \\
-I(price\_l4)                     & $0.000$        & $0.004^{**}$   & $0.004^{***}$  \\
-                                 & $(0.002)$      & $(0.001)$      & $(0.001)$      \\
-I(price\_l5/10)                  & $-0.020$       & $0.012$        & $0.019$        \\
-                                 & $(0.013)$      & $(0.012)$      & $(0.010)$      \\
-I(price\_l6/10)                  & $-0.010$       & $0.008$        &                \\
-                                 & $(0.012)$      & $(0.011)$      &                \\
-I(price\_l7/10)                  & $-0.008$       & $0.007$        &                \\
-                                 & $(0.012)$      & $(0.012)$      &                \\
-I(price\_l8/10)                  & $-0.044^{***}$ & $-0.011$       &                \\
-                                 & $(0.011)$      & $(0.011)$      &                \\
-EDF: s(prod\_year)               & $6.381^{***}$  &                &                \\
-                                 & $(7.249)$      &                &                \\
-EDF: s(year)                     & $2.768^{***}$  & $3.000^{***}$  & $3.000^{***}$  \\
-                                 & $(2.956)$      & $(3.000)$      & $(3.000)$      \\
-EDF: s(prod\_year,name)          &                & $33.002^{***}$ & $33.014^{***}$ \\
-                                 &                & $(72.000)$     & $(72.000)$     \\
+(Intercept)                                  & $-28746.236^{***}$ & $-70463.727^{***}$ & $-58741.031^{***}$ \\
+                                             & $(3386.956)$       & $(2942.538)$       & $(2959.556)$       \\
+I(in\_place\_oil\_mill\_sm3/100)             & $0.333^{***}$      & $-0.028$           &                    \\
+                                             & $(0.009)$          & $(0.023)$          &                    \\
+I(cost\_index/10)                            & $-0.005$           & $-0.019^{***}$     & $-0.000$           \\
+                                             & $(0.006)$          & $(0.006)$          & $(0.004)$          \\
+year                                         & $28.793^{***}$     & $70.938^{***}$     & $58.352^{***}$     \\
+                                             & $(3.393)$          & $(2.951)$          & $(2.955)$          \\
+I(year^2)                                    & $-0.007^{***}$     & $-0.018^{***}$     & $-0.015^{***}$     \\
+                                             & $(0.001)$          & $(0.001)$          & $(0.001)$          \\
+I(price/10)                                  & $0.030$            & $0.026$            & $0.008$            \\
+                                             & $(0.015)$          & $(0.015)$          & $(0.010)$          \\
+I(price\_l1/10)                              & $0.050^{**}$       & $0.034^{*}$        & $0.020$            \\
+                                             & $(0.016)$          & $(0.015)$          & $(0.010)$          \\
+I(price\_l2/10)                              & $0.070^{***}$      & $0.058^{***}$      & $0.055^{***}$      \\
+                                             & $(0.017)$          & $(0.015)$          & $(0.010)$          \\
+I(price\_l3)                                 & $0.004^{*}$        &                    &                    \\
+                                             & $(0.002)$          &                    &                    \\
+I(price\_l4)                                 & $0.003$            &                    &                    \\
+                                             & $(0.002)$          &                    &                    \\
+I(price\_l5/10)                              & $-0.008$           & $0.010$            & $-0.003$           \\
+                                             & $(0.013)$          & $(0.012)$          & $(0.008)$          \\
+I(price\_l6/10)                              & $-0.001$           & $0.007$            & $-0.003$           \\
+                                             & $(0.012)$          & $(0.011)$          & $(0.008)$          \\
+I(price\_l7/10)                              & $0.009$            & $0.008$            & $0.002$            \\
+                                             & $(0.012)$          & $(0.011)$          & $(0.008)$          \\
+I(price\_l8/10)                              & $-0.004$           & $-0.001$           & $-0.021^{***}$     \\
+                                             & $(0.010)$          & $(0.009)$          & $(0.006)$          \\
+EDF: s(prod\_year)                           & $6.505^{***}$      &                    &                    \\
+                                             & $(7.326)$          &                    &                    \\
+I(price\_l3/10)                              &                    & $0.024$            & $0.031^{**}$       \\
+                                             &                    & $(0.015)$          & $(0.010)$          \\
+I(price\_l4/10)                              &                    & $0.046^{***}$      & $0.026^{**}$       \\
+                                             &                    & $(0.013)$          & $(0.009)$          \\
+EDF: s(prod\_year,name)                      &                    & $33.287^{***}$     &                    \\
+                                             &                    & $(72.000)$         &                    \\
+EDF: s(prod\_year,in\_place\_oil\_mill\_sm3) &                    &                    & $28.195^{***}$     \\
+                                             &                    &                    & $(28.469)$         \\
 \hline
@@
 Log Likelihood                               &                    &                    &                    \\
-Deviance                         & 2306267.814    & 1976585.232    & 1980468.734    \\
-Deviance explained               & 0.848          & 0.870          & 0.870          \\
-Dispersion                       & 1785.243       & 1562.520       & 1560.671       \\
-R$^2$                            & 0.841          & 0.865          & 0.866          \\
-GCV score                        & 1814.469       & 1621.811       & 1614.802       \\
+Deviance                                     & 2391636.349        & 1980765.291        & 914018.219         \\
+Deviance explained                           & 0.843              & 0.870              & 0.940              \\
+Dispersion                                   & 1850.403           & 1564.940           & 718.678            \\
+R$^2$                                        & 0.835              & 0.865              & 0.937              \\
+GCV score                                    & 1879.759           & 1623.406           & 741.957            \\
 Num. obs.                                    & 1088               & 1088               & 1088               \\
-Num. smooth terms                & 2              & 2              & 2              \\
+Num. smooth terms                            & 1                  & 1                  & 1                  \\
 \hline
@@
 
-\begin{equation}
-\begin{split}
-	Log(Production_{i,t})&=f(timetopeak_{i,t}, totalrecoverableoil_i) \\
-	 \quad & + f(peaktoend_{i,t}, totalrecoverableoil_i) \\
-	 \quad & + \beta_1 oilprice + \beta_2 oilpricel1 + \dots + oilpricel8\\
-	 \quad & + \delta smallfield \\
- 	 \quad & + \gamma_1 oilprice*smallfield + \gamma_2 oilprice*smallfield + \dots \\
- 	 \quad & + oilpricel8*smallfield +  \epsilon
-\label{pooled_eqn}
-\end{split}
-\end{equation}
+\begin{figure}
+	\includegraphics[width=1\textwidth]{figures/price_coefficents.png}
+	\caption{Estimated coefficients from two GAM models: a model with random field effects and single smoothed function of production time, and another with a two dimensional smoothed function of production time and field size.
+The points represent the point estimates, while the lines represent approximate 95\% confidence intervals.
+Significant positive coefficients are estimated on lags of between 1-4.
+However, little to no effect is measured on the concurrent price term.}
+	\label{price\_coefficients}
+\end{figure}
 
@@
 
-\section{The effect of price in the phases of oil production}
-
-Treating the entire production life of an oil field as essentially the same process is an oversimplification of the industry and field dynamics.
-A field's life can be split into roughly three periods.
-Before physical investment and production can begin, an often times lengthy planning stage occurs.
-In this stage, the extent and details of a fields build-out are decided and must receive approval from the Norwegian Petroleum Directorate as well as other government agencies.
-
-Once plans have been approved and finalized, the build-out phase of a field can begin.
-The field operator has an incentive to build out the field as quickly as possible as large amounts of capital have at that point been allocated and it is important to get oil production and in turn income flowing as quickly as possible.
-Once the build-out is complete and production has reached its peak, geophysical forces that come from depletion and subsequent drop in field pressure dominate.
-Here technological change can also play a role - with improvements over time - like the introduction of gas and water injection - allowing for increased recovery percentage of the available reservoir.
-
-In the preceding sections I have partially taken account of the different dynamics of the phases of field production by modeling post- and pre-peak production as seperate smoothed terms.
-However, the price variable was estimated for the entire production profile.
-The advantage of this approach is that I am able to use the full data-set and estimate a total average effect.
-However the effect of price can reasonably be expected to differ in the different phases of production.
-
-In this section I model the pre-peak and post-peak periods of production separately - including the price terms.
-The vast majority of producing fields in Norway are well past their peak production and because larger fields tended to be found earlier, an even larger share of total oil production comes from depleting fields.
-In terms of policy implications, the most important question is what the effect of price has on post-peak production.
-
-To ensure that no fields are included in the post-peak group that have not been fully built-out yet and which have not reached their peak production, I exclude all fields that began oil production prior to 2008.
-While some large fields may take slightly more than five years to fully build out, none such fields began production in that time frame.
-
- The estimated coefficients on the price term for post-peak production is shown in figure \ref{gam_postpeak_print}.
-The full results are shown in table \ref{table:postpeak} in the appendix.
-The results mostly show no significant effect of price on production through the 7th lag, though positive coefficients for large fields are estimated for all but the 5th lag.
-
- A significant positive coefficient is estimated on the first lag for small fields, however this result is not robust to specification.
-Using the maximum production instead of estimated total oil reserves in the smoothed function gives a somewhat better fit for some small fields (see figure \ref{chart:Max_prod_vs_TRO_print} in Appendex 2).
-In this regression the coefficient is estiamted to be smaller and is no longer significant (see table \ref{table:postpeak}).
-When small and large fields are pooled in one regression, this coefficient is again not estimated to be significantly different from zero.
-This estimated coefficient should be met with some skepticism.
-
- On the other hand, the significant coefficient on the 8th lag for small fields is robust to specification.
-From an industry perspective, why price should have a 8-year-lagged effect on small fields but not large fields may not be immediately clear, but the length of the lag combined with the relatively short life of small fields suggests that the effect of price occurs well before the depletion phase begins.
-
-The effect that price has on production from a declining fields appears to be, at best, slight.
-The point estimates on the price terms for large fields are of a magnitude of around 1\% - 2\% for a 10 dollar increase in the price of oil, but as noted, these estimates are not significantly different from zero.
-In general, no clear relationship appears in the data between production and prices.
-
-However, to say that price has no effect on production from depleting fields may be too strong.
-One mechanism for prices to effect production in depleting oil fields that this regression may not directly capture is through technological change.
-Higher prices may encourage more research and development and in turn lead to innovations that increase the total recovery percentage of a field.
-For example, over the course of the history of Norwegian oil production, innovations such as horizontal drilling and four-dimensional geological visualization have led to increased estimates of the total recovery rate of Norwegian oil fields.
-However, this mechanism is likely too diffuse - lacking any consistent relationship between price, innovation and subsequent increases in production - to be captured by the regressions in this analysis.
-
-While price appears to have little affect on production in depleting fields, it may have an effect in the planning and build-out phase of production.
-I run a regression model of field production up to peak production, representing the build-out phase of a field.
-Because of the double-peak profile of the Ekofisk field, I only include data from this field up to the initial peak in 1976.
-The full results can be found in table \ref{table:prepeak} in Appendix I.
-I show the estimated coefficient on the price terms in table \ref{gam_prepeak_print} below.
-
-What is immediately noticeably is the imprecision of the price estimates in the model of the large fields.
-This is primarily due to a lack of data points.
-The estimates are based on only 76 data points from 11 fields.
-Positive coefficients are estimated at the 6th, 7th and 8th lag, though none of these are statistically significant at the 95\% level.
-The coefficient on the lagged price terms for small fields, however, are estimated to be positive and significant at the 5th and 8th lags.
-
-I also estimate a pooled model, with both small and large fields.
-This model provides a worse overall fit than the split model, but can nonetheless be useful given the few data points available for large fields.
-The coefficients on the 1st as well as 3rd to through 8th lags are all estimated to be significant and positive at a magnitude that corresponds to approximately 2-8 \% increase in oil production for a 10 dollar increase in the cost of a barrel of oil.
-
-\begin{figure}
-	\includegraphics[width=1\textwidth]{figures/field_time_to_peak_print.png}
-	\caption{A histogram of the number of year before a field is fully built out and a field reaches its peak production.
-The vast majority of fields, especially small fields, are built out quickly - within one or two years.}
-	\label{gam_buildout_hist}
-\end{figure}
-
-Figure \ref{gam_buildout_hist} shows a histogram of the time it takes to fully build out an oil field and reach peak production.
-We see that the vast majority of fields are built out quickly, reaching their peak production within one or two years.
-Since the effect of price on the build-out phase is estimated with a lag of between 5 and 8 year, this indicates that changes in price likely has most effect well before production in a field even begins.
-In turn, this suggests that changes in price that occur during the planning stage of a fields development are the most important in determining the production of the field.
-This result fits well with the industry structure.
-As mentioned, before physical investment in a field can begin and production started, detailed plans need to be approved by several government agencies and capital must be allocated.
-Changing the level of investments in reaction to a change in prices after plans have been finalized is likely to lead to both an expensive delay as well as other costs that heavily outweigh any benefits of increased future production.
 \section{Conclusion}
@@
 
-\section{software}
+\section{Software}
 I use the R statistical programming package for all the analysis in this article \citep{r_core_team_r:_2013}.
 I use the R package ggplot2 for plotting \citep{wickham_ggplot2:_2009}, plyr for data manipulation and cleaning \citep{wickham_split-apply-combine_2011}, texreg for table formatting \citep{leifeld_texreg:_2013} and mgcv for implementation of the Generalized Additive Models \citep{wood_fast_2011}.
-
- \begin{spacing}{1}
 
@@
 
-% \end{spacing}
-% \end{document} for shorter version with no appendix
 \FloatBarrier

### 2015-03-31 rr_analysis_1.r

@@ -53,13 +53,14 @@ coef<-gam_mod_2d$coefficients[5:13]
 se<-summary(gam_mod_2d)$p.table[5:13,2]
 ymin<-coef-2*se
 ymax<-coef+2*se
-gam_2d_coef<-data.frame(coef=coef, se=se, lag=0:8, ymin=ymin, ymax=ymax)
+gam_2d_coef<-data.frame(coef=coef, se=se, lag=0:8, ymin=ymin, ymax=ymax, type="2d Smooth Model")
 
-2d_coef_plot<-ggplot(gam_2d_coef) +
+coef_plot_2d<-ggplot(gam_2d_coef) +
 geom_pointrange(aes(x=lag, y=coef, ymin = ymin, ymax = ymax)) +
 geom_hline(aes(y=0)) +
 theme_bw() +
-labs(x="Lags", y="Estimated Coefficient on Oil Price")
+labs(x="Lags", y="Estimated Coefficient on Oil Price", 
+    title="2-dim. Smooth Model")
 
 
 #With random effects
@@ -76,13 +77,29 @@ coef<-gam_mod_re$coefficients[6:14]
 se<-summary(gam_mod_re)$p.table[6:14,2]
 ymin<-coef-2*se
 ymax<-coef+2*se
-gam_re<-data.frame(coef=coef, se=se, lag=0:8, ymin=ymin, ymax=ymax)
+gam_re<-data.frame(coef=coef, se=se, lag=0:8, ymin=ymin, ymax=ymax, type="Random Effects Model")
+
+model_coef<-rbind(gam_re, gam_2d_coef)
 
 re_coef_plot<-ggplot(gam_re) +
 geom_pointrange(aes(x=lag, y=coef, ymin = ymin, ymax = ymax)) +
 geom_hline(aes(y=0)) +
 theme_bw() +
-labs(x="Lags", y="Estimated Coefficient on Oil Price", main="Random Effects Model")
+labs(x="Lags", y="Estimated Coefficient on Oil Price", 
+   title="Random Effects Model")
+
+coef_plot<-ggplot(model_coef) +
+geom_pointrange(aes(x=lag, y=coef, ymin = ymin, ymax = ymax)) +
+geom_hline(aes(y=0)) +
+facet_wrap(~type, nrow=2) +
+theme_bw() +
+labs(x="Lags", y="Estimated Coefficient on Oil Price")
+
+png("/Users/johannesmauritzen/research/oil_prices/figures/price_coefficents.png",
[…]

### 2015-03-31 rr_descriptives.py

@@ -14,7 +14,6 @@ import statsmodels.api as sm
 
 
 
-
 #first on individual field data
 field_data=pd.read_csv('/Users/johannesmauritzen/research/oil_prices/data/field_data.csv')
 prod_data=pd.read_csv('/Users/johannesmauritzen/research/oil_prices/data/prod_data.csv')

### 2015-03-31 rr_descriptives.r

@@ -1,0 +1,100 @@
+#rr_descriptives.r
+#last updated March 24th
+
+library(ggplot2)
+library(dplyr)
+library(reshape2)
+library(mgcv)
+library(lubridate) 
+library(grid)
+library(arm)
+library(texreg)
+
+field_data<-read.csv('/Users/johannesmauritzen/research/oil_prices/data/field_data.csv')
+prod_data<-read.csv('/Users/johannesmauritzen/research/oil_prices/data/prod_data.csv')
+
+#How much of the total oil reserves are we excluding
+tot_rec_oil<-sum(field_data["recoverable_oil"])
+small_rec_oil<-sum(field_data["recoverable_oil"][field_data["recoverable_oil"]<=3])
+28.3/4779.2
+#.5% of total recoverable oil.  
+
+
+top10<-as.character(head(field_data$name[order(field_data$recoverable_oil, decreasing=TRUE)],10))
+
+#function to to show field-level of data
+fields_lim<-subset(fields, name %in% top10)
+fields_lim<-fields_lim[,c("name", "year", "year_prod")]
+fields_lim_long<-melt(fields_lim, id.vars=c("name", "year", "year_prod"))
+
+top10_production<-ggplot(fields_lim_long, aes(x=year, y=year_prod, color=name)) +
+geom_line() +
+scale_color_grey() +
+theme_bw() +
+labs(x="", y="Yearly Oil Production, Mill SM3", color="Field") +
+guides(color=FALSE)
+
+
+#total production
+prod_grouped<-group_by(prod_data, year)
+tot_year_prod<-summarise(prod_grouped, tot_year_prod=sum(oil_prod_mill_sm3, na.rm=TRUE))
+
+#plot total production with oil price
+oil_decline<-ggplot(tot_year_prod) +
+geom_line(aes(x=year, y=tot_year_prod)) +
+xlab("") + ylab("Total Norwegian Production of Oil, Mill SM3") +
+theme_bw()
+
+
+png("/Users/johannesmauritzen/research/oil_prices/figures/oil_decline.png", 
+	width = 25, height = 20, units = "cm", res=100, pointsize=12)
+grid.newpage()
+	pushViewport(viewport(layout = grid.layout(2, 1)))
+	vplayout <- function(x, y)
+	  viewport(layout.pos.row = x, layout.pos.col = y)
+	print(top10_production, vp = vplayout(1, 1))
[…]

### 2015-03-31 rr_simulation.r

@@ -128,7 +128,7 @@ formula_4<- formula(prod~s(prod_years, size) + prices + year + I(year^2))
 gam_mc_beta0<-replicate(100, mc_run(beta=0, formula=formula_4, fields=sim_fields, use_gam=TRUE))
 gam_mc_beta05<-replicate(100, mc_run(beta=.05, formula=formula_4, fields=sim_fields, use_gam=TRUE))
 gam_mc_data<-data.frame(beta0=gam_mc_beta0, beta05=gam_mc_beta05)
-#write.csv(gamm_mc_data, "/Users/johannesmauritzen/research/oil_prices/mc_data.csv")
+write.csv(gam_mc_data, "/Users/johannesmauritzen/research/oil_prices/mc_data.csv")
 #gam_mc_data<-read.csv("/Users/johannesmauritzen/research/oil_prices/mc_data.csv")
 #gam_mc_data$X<-NULL
 gam_mc_data<-melt(gam_mc_data)

### 2015-04-01 oil_prices_rr.tex
updated methods section

@@
 
-Finally, $oil\_price\_lags_t$ represents the variables of interest - a vector of concurrent and up to 8 lagged terms for the oil price.
+Finally, $oil\_price\_lags_t$ represents the variables of interest - a vector of concurrent and lagged terms for the oil price.
+Initially I included eight lagged terms of the oil price, however in the preferred specification I drop lags 5 through 8 as they where not individually significant, nor as a group did they improve the fit of the model - as tested by an F-test.
 I also experimented with using quadratic terms for the oil price and its lags, however these could not be shown to be significant.
 
-The smoothed function is estimated using a cubic regression spline...
+The smoothed function is estimated using a cubic regression spline.
+Here a cubic polynomial function is used to fit the shape in sections, separated at points known as knots, but continuous up to the second derivative.
+The major advantage of this method, is that it can be represented in linear form $\boldsymbol{X \beta} $, and thus normal matrix algebra techniques can be used.
+
+The model is then fit by minimizing,
+
+\begin{equation}
+
+||\boldsymbol{y} - \boldsymbol{X\beta} ||^2 + \lambda \int_{0}^{1} [f´´(x)]^2dx
+
+\end{equation}
+
+The latter term, an estimate of the second derivative of the function, acts a penalty for the ``wiggiliness'' of the function.
+The total smoothness can then be controlled by adjusting the $\lambda$ term.
+However, instead of setting this arbitrarily, cross-validation is used.
+Intuitively, each data point is left out and the smooth term that provides the best predicted fit for each data point is chosen.
+For further details on the cubic regression spline, I refer to \citet{wood_generalized_2006}.
 
 One objection to this model may be that a fixed effect controlling for the general production profile of oil fields does not sufficiently address idiosyncratic differences between the fields.
-To address this I can add the term, $\beta_r Z$, where $Z$ represents field random effects that takes into account uncertainty based on random variation between the fields.
+To address this I can add field random effects that takes into account uncertainty based on random variation between the fields.
+The smoothed function can then be written as $f(production\_time_{i,t})*Z$.
+Where $Z~N(0, \sigma)$ and $\sigma$ is to be estimated.
 
-Another substantial objection may be that the smoothed function of production time may not be sufficient in controlling for the variation in production profiles shapes across fields.
+Another potential issue may be that the smoothed function of production time may not be sufficient in controlling for the variation in production profiles shapes across fields.
 In particular, production profiles differ substantially by the size of the field.
@@
 For further details I again refer to \citet{wood_thin_2003}.
+
+In the formulation with a two-dimensional smoothed function, it is not possible to estimate a random effects component for computational reasons.
 

### 2015-04-01 rr_analysis_1.r

@@ -27,26 +27,26 @@ small_rec_oil<-sum(field_data["recoverable_oil"][field_data["recoverable_oil"]<=
 
 prod_data<-read.csv('/Users/johannesmauritzen/research/oil_prices/data/prod_data.csv')
 
-large_fields<-prod_data[prod_data["recoverable_oil"]>3,]
+large_fields<-prod_data[prod_data["recoverable_oil"]>5,]
 large_fields<-prod_data[prod_data["name"]!="EKOFISK",]
 
 
-gam_mod<-gam(oil_prod_mill_sm3~s(prod_year, bs="cr") + 
+gam_mod<-gam(oil_prod_mill_sm3~s(prod_year, bs="cr") +
 	I(in_place_oil_mill_sm3/100) + I(cost_index/10) +
     year + I(year^2) + I(price/10) + I(price_l1/10) + I(price_l2/10) +
-    I(price_l3) + I(price_l4) + I(price_l5/10) + I(price_l6/10) + 
+    I(price_l3/10) + I(price_l4/10) + I(price_l5/10) + I(price_l6/10) + 
     I(price_l7/10) + I(price_l8/10), 
-    family=gaussian(link=log), weights=in_place_oil_mill_sm3, data=large_fields)
+    family=gaussian(link=log), data=large_fields)
 summary(gam_mod)
 
 
-
-gam_mod_2d<-gam(oil_prod_mill_sm3~s(prod_year, in_place_oil_mill_sm3) + 
-  + I(cost_index/10) +
-    year + I(year^2) + (year^3) + I(price/10) + I(price_l1/10) + I(price_l2/10) +
+gam_mod_2d<-gam(oil_prod_mill_sm3~s(prod_year, in_place_oil_mill_sm3) +
+    s(name, bs="re") +
+    I(cost_index/10) +
+    year + I(year^2) + I(price/10) + I(price_l1/10) + I(price_l2/10) +
     I(price_l3/10) + I(price_l4/10) + I(price_l5/10) + I(price_l6/10) + 
     I(price_l7/10) + I(price_l8/10), 
-    family=gaussian(link=log), weights=in_place_oil_mill_sm3, data=large_fields)
+    family=gaussian(link=log), I(in_place_oil_mill_sm3/100), data=large_fields)
 summary(gam_mod_2d)
 
 coef<-gam_mod_2d$coefficients[5:13]
[…]

### 2015-04-02 oil_prices_rr.tex
updated analysis

@@
 Estimating a smoothed function of two variables requires a somewhat more involved method.
-In particular I use Thin-plate regression spline \citep{wood_thin_2003}.
+In particular, a Thin-plate regression spline \citep{wood_thin_2003} is used.
 Consider equation \ref{thin_plate_1}.
@@
 
-In the formulation with a two-dimensional smoothed function, it is not possible to estimate a random effects component for computational reasons.
+Unfortunately, In the formulation with a two-dimensional smoothed function, it is not possible to estimate a random effects component for computational reasons.
 
@@
 \hline
-(Intercept)                                  & $-28746.236^{***}$ & $-70463.727^{***}$ & $-58741.031^{***}$ \\
-                                             & $(3386.956)$       & $(2942.538)$       & $(2959.556)$       \\
-I(in\_place\_oil\_mill\_sm3/100)             & $0.333^{***}$      & $-0.028$           &                    \\
-                                             & $(0.009)$          & $(0.023)$          &                    \\
-I(cost\_index/10)                            & $-0.005$           & $-0.019^{***}$     & $-0.000$           \\
-                                             & $(0.006)$          & $(0.006)$          & $(0.004)$          \\
-year                                         & $28.793^{***}$     & $70.938^{***}$     & $58.352^{***}$     \\
-                                             & $(3.393)$          & $(2.951)$          & $(2.955)$          \\
-I(year^2)                                    & $-0.007^{***}$     & $-0.018^{***}$     & $-0.015^{***}$     \\
-                                             & $(0.001)$          & $(0.001)$          & $(0.001)$          \\
-I(price/10)                                  & $0.030$            & $0.026$            & $0.008$            \\
-                                             & $(0.015)$          & $(0.015)$          & $(0.010)$          \\
-I(price\_l1/10)                              & $0.050^{**}$       & $0.034^{*}$        & $0.020$            \\
-                                             & $(0.016)$          & $(0.015)$          & $(0.010)$          \\
-I(price\_l2/10)                              & $0.070^{***}$      & $0.058^{***}$      & $0.055^{***}$      \\
-                                             & $(0.017)$          & $(0.015)$          & $(0.010)$          \\
-I(price\_l3)                                 & $0.004^{*}$        &                    &                    \\
-                                             & $(0.002)$          &                    &                    \\
-I(price\_l4)                                 & $0.003$            &                    &                    \\
-                                             & $(0.002)$          &                    &                    \\
-I(price\_l5/10)                              & $-0.008$           & $0.010$            & $-0.003$           \\
-                                             & $(0.013)$          & $(0.012)$          & $(0.008)$          \\
-I(price\_l6/10)                              & $-0.001$           & $0.007$            & $-0.003$           \\
-                                             & $(0.012)$          & $(0.011)$          & $(0.008)$          \\
-I(price\_l7/10)                              & $0.009$            & $0.008$            & $0.002$            \\
-                                             & $(0.012)$          & $(0.011)$          & $(0.008)$          \\
-I(price\_l8/10)                              & $-0.004$           & $-0.001$           & $-0.021^{***}$     \\
-                                             & $(0.010)$          & $(0.009)$          & $(0.006)$          \\
-EDF: s(prod\_year)                           & $6.505^{***}$      &                    &                    \\
-                                             & $(7.326)$          &                    &                    \\
-I(price\_l3/10)                              &                    & $0.024$            & $0.031^{**}$       \\
-                                             &                    & $(0.015)$          & $(0.010)$          \\
-I(price\_l4/10)                              &                    & $0.046^{***}$      & $0.026^{**}$       \\
-                                             &                    & $(0.013)$          & $(0.009)$          \\
-EDF: s(prod\_year,name)                      &                    & $33.287^{***}$     &                    \\
+(Intercept)                                  & $0.335^{*}$    & $-4.017^{***}$ & $-574.287^{***}$ \\
+                                             & $(0.146)$      & $(0.183)$      & $(51.125)$       \\
+I(in\_place\_oil\_mill\_sm3/100)             & $0.337^{***}$  & $-0.014$       &                  \\
+                                             & $(0.005)$      & $(0.014)$      &                  \\
+I(cost\_index/10)                            & $-0.018^{***}$ & $-0.030^{***}$ & $-0.014^{***}$   \\
+                                             & $(0.003)$      & $(0.003)$      & $(0.002)$        \\
+I(price/10)                                  & $-0.006$       & $0.014$        & $0.007$          \\
+                                             & $(0.009)$      & $(0.010)$      & $(0.006)$        \\
+I(price\_l1/10)                              & $0.004$        & $0.021^{*}$    & $0.015^{*}$      \\
+                                             & $(0.010)$      & $(0.010)$      & $(0.006)$        \\
+I(price\_l2/10)                              & $0.037^{***}$  & $0.047^{***}$  & $0.048^{***}$    \\
+                                             & $(0.009)$      & $(0.010)$      & $(0.006)$        \\
+I(price\_l3/10)                              & $0.030^{***}$  & $0.025^{**}$   & $0.033^{***}$    \\
+                                             & $(0.009)$      & $(0.009)$      & $(0.006)$        \\
+I(price\_l4/10)                              & $0.003$        & $0.043^{***}$  & $0.028^{***}$    \\
+                                             & $(0.008)$      & $(0.008)$      & $(0.005)$        \\
+I(price\_l5/10)                              & $-0.020^{**}$  & $0.012$        & $-0.001$         \\
+                                             & $(0.007)$      & $(0.007)$      & $(0.005)$        \\
+I(price\_l6/10)                              & $-0.010$       & $0.008$        & $-0.001$         \\
+                                             & $(0.007)$      & $(0.007)$      & $(0.004)$        \\
+I(price\_l7/10)                              & $-0.008$       & $0.007$        & $0.004$          \\
+                                             & $(0.007)$      & $(0.007)$      & $(0.004)$        \\
+I(price\_l8/10)                              & $-0.044^{***}$ & $-0.011$       & $-0.027^{***}$   \\
+                                             & $(0.006)$      & $(0.007)$      & $(0.004)$        \\
+EDF: s(prod\_year)                           & $6.381^{***}$  &                &                  \\
+                                             & $(7.249)$      &                &                  \\
+EDF: s(year)                                 & $2.768^{***}$  & $3.000^{***}$  & $3.000^{***}$    \\
+                                             & $(2.956)$      & $(3.000)$      & $(3.000)$        \\
+EDF: s(prod\_year,name)                      &                & $33.002^{***}$ &                  \\
                                              &                & $(72.000)$     &                  \\
-EDF: s(prod\_year,in\_place\_oil\_mill\_sm3) &                    &                    & $28.195^{***}$     \\
-                                             &                    &                    & $(28.469)$         \\
+EDF: s(prod\_year,in\_place\_oil\_mill\_sm3) &                &                & $27.445^{***}$   \\
+                                             &                &                & $(27.895)$       \\
 \hline
@@
 Log Likelihood                               &                &                &                  \\
-Deviance                                     & 2391636.349        & 1980765.291        & 914018.219         \\
-Deviance explained                           & 0.843              & 0.870              & 0.940              \\
-Dispersion                                   & 1850.403           & 1564.940           & 718.678            \\
-R$^2$                                        & 0.835              & 0.865              & 0.937              \\
-GCV score                                    & 1879.759           & 1623.406           & 741.957            \\
+Deviance                                     & 23062.678      & 19765.833      & 9057.204         \\
+Deviance explained                           & 0.848          & 0.870          & 0.940            \\
+Dispersion                                   & 5.410          & 5.965          & 2.325            \\
+R$^2$                                        & 0.841          & 0.865          & 0.938            \\
+GCV score                                    & 18.145         & 16.218         & 7.355            \\
 Num. obs.                                    & 1088           & 1088           & 1088             \\
-Num. smooth terms                            & 1                  & 1                  & 1                  \\
+Num. smooth terms                            & 2              & 2              & 2                \\
 \hline

### 2015-04-02 rr_analysis_1.r

@@ -12,7 +12,6 @@ library(lubridate)
 library(grid)
 library(boot)
 library(arm)
-library(gamm4)
 library(lme4)
 library(nlme)
 library(texreg)
@@ -33,21 +32,46 @@ large_fields<-prod_data[prod_data["name"]!="EKOFISK",]
 
 gam_mod<-gam(oil_prod_mill_sm3~s(prod_year, bs="cr") +
 	I(in_place_oil_mill_sm3/100) + I(cost_index/10) +
-    year + I(year^2) + I(price/10) + I(price_l1/10) + I(price_l2/10) +
+    s(year, bs="cr", k=4) + I(price/10) + I(price_l1/10) + I(price_l2/10) +
     I(price_l3/10) + I(price_l4/10) + I(price_l5/10) + I(price_l6/10) + 
     I(price_l7/10) + I(price_l8/10), 
-    family=gaussian(link=log), data=large_fields)
+    family=gaussian(link=log), weights=I(in_place_oil_mill_sm3/100), data=large_fields)
 summary(gam_mod)
+#,weights=I(in_place_oil_mill_sm3/100)
+
+#terms
+
+pdat <- with(large_fields,
+             data.frame(prod_year = round(seq(min(prod_year), max(prod_year), length = 36)),
+                        year = round(seq(min(year), max(year), length = 36)),
+                        oil_prod_mill_sm3 = rep(mean(oil_prod_mill_sm3), 36),
+                        in_place_oil_mill_sm3 = rep(mean(in_place_oil_mill_sm3), 36),
+                        cost_index = rep(mean(cost_index), 36),
+                        price = rep(mean(price), 36),
+                        price_l1 = rep(mean(price_l1), 36),
+                        price_l2= rep(mean(price_l2), 36),
+                        price_l3= rep(mean(price_l3), 36),
+                        price_l4 = rep(mean(price_l4), 36),
+                        price_l5= rep(mean(price_l5), 36),
+                        price_l6 = rep(mean(price_l6), 36),
+                        price_l7 = rep(mean(price_l7), 36),
+                        price_l8= rep(mean(price_l8), 36)
+                        ))
[…]

### 2015-04-05 oil_prices_rr.tex
small update@

@@
 In this model I am estimating the parameters and functions from all fields $i$.
-The left-hand-side variable, $Log(production_{i,t})$ is the yearly oil production for field $i$ at production time $t$ in millions of standard cubic meters (SM3).
-On the right hand side is a smoothed function of the production profile over time $f(production\_time_{i,t})$, $in\_place\_oil_{i,t}$ is the estimated in place oil of the field, and $cost\_index_{t}$ represents the producer price index for oil and gas extraction.
+The left-hand-side variable, $Log(production_{i,t})$ is the log of yearly oil production for field $i$ at production time $t$ in millions of standard cubic meters (SM3).
+On the right hand side is a smoothed function of the production profile over time $f(production\_time_{i,t})$; $in\_place\_oil_{i,t}$ is the estimated in place oil of the field, and $cost\_index_{t}$ represents the producer price index for oil and gas extraction.
 
-$f(year_t)$ represents a function of the calendar year.
-The main purpose of including this term is to control for the effects of technological change.
-I could use a smoothed function here as well, however in order to avoid over fitting the model, I use a simple quadratic representation.
+$f(year_t)$ represents a function of the calendar year.  The main purpose of including this term is to control for the effects of technological change.
 
@@
 
-One objection to this model may be that a fixed effect controlling for the general production profile of oil fields does not sufficiently address idiosyncratic differences between the fields.
-To address this I can add field random effects that takes into account uncertainty based on random variation between the fields.
+The advantage of this simple model that use smoothed functions of single-variable cubic regression splines is that the smoothed terms can be interpreted directly, and thus we can do a visual check of their appropriateness.
+The top panel of figure \ref{smooths} shows the estimated smoothed function of the field production profile, where all other co-variates are held at their average value.
+The function appears reasonable, showing the expected pattern of an initial build-out period followed by an extended depletion phase.
+
+In the second panel, the smoothed function of calendar time $year_t$ is shown.
+Here a cubic regression spline is also used, however the number of knots is restricted to four in order to avoid over-fitting - capturing gradual trends over time like technological change, but not the variation  caused by volatile oil prices.
+The function increases strongly in the early period, likely reflecting the effects of learning and technological change in a new industry in Norwegian waters, while flattening out in the latter years.
+
+Extended differences in the oil price will likely have some long-term effects on technology and learning, which in turn may be picked up in the smoothed function here, rather than in the parametric price terms.
+I discuss this further in the conclusion.
+
+\begin{figure}
+	\includegraphics[width=1\textwidth]{figures/smooths.png}
+	\caption{The estimated smoothed functions over production time and calendar time.}
+	\label{smooths}
+\end{figure}
+
+One objection to the simple model is that a fixed effect controlling for the general production profile of oil fields does not sufficiently address idiosyncratic differences between the fields.
+To address this I can add field random effects that takes into account uncertainty based on random variation between the fields in the estimation.
 The smoothed function can then be written as $f(production\_time_{i,t})*Z$.
-Where $Z~N(0, \sigma)$ and $\sigma$ is to be estimated.
+Where $Z~N(0, \sigma)$ and $\sigma$ is the variance term to be estimated.
 
@@
 
-Unfortunately, In the formulation with a two-dimensional smoothed function, it is not possible to estimate a random effects component for computational reasons.
+Unfortunately, estimating a random effects component is not possible with a two-dimensional smoothed function for computational reasons.
+In the presentation of results, I therefor present models side-by-side that use both one-dimensional smooth functions with random effects of the field production profile and two-dimensional smooth functions.
 
@@
 
-The variables of interest in this article is the oil price and its lags, which I include as nine linear parametric terms in the model.
+Table \ref{GAM_model_table} shows the parametric coefficient terms and other summary statistics of the regression results.
+The first column shows results for the simple model with a single-dimensional smoothed function of the production profile.
+The second column shows results with field-level random effects, while the third columns shows results when a two-dimensional smoothed function is used to control for production profiles across field size.
+The latter two models are my preferred models, providing a substantially better fit, as measured by both deviance explained and R^2.
+Monte Carlo Experiments, which I describe further below, also indicates that the latter two models provide consistently less-biased parametric estimates of the price terms.
+
+The variables of interest in this article is the oil price and its lags, which I include as eight linear parametric terms in the model.
 The idea of including both a concurrent oil price term as well as eight lags is that a change in price could conceivably have two effects on oil production in a field.
@@
 However investments in the offshore sector can be complex and lengthy, and any production lift would be expected to happen with a lag.
-As mentioned earlier, including several lags also allows for the possibility of adaptive expectations of future oil prices.The results for the price terms are shown in figure \ref{gam_price_8}.
+As mentioned earlier, including several lags also allows for the possibility of adaptive expectations of future oil prices.
 
@@
 \hline
-AIC                                          &                &                &                  \\
-BIC                                          &                &                &                  \\
-Log Likelihood                               &                &                &                  \\
 Deviance                                     & 23062.678      & 19765.833      & 9057.204         \\
@@
 \end{tabular}
-\caption{Statistical models}
+\caption{Model results. Standard errors in parenthesis.}
 \label{GAM_model_table}
@@
 \end{table}
+
+For convenience the estimated coefficients on the price terms from the two preferred models are shown in figure \ref{price_coefficients}.
+In general the results from the two models agree with each other, indicating little to no effect in the concurrent price term, but with significant effects estimated on the 1st to 4th lags, with a peak at the 2nd lag.
+F-tests of restricted models without the concurrent and 5-8th lag confirm that these terms do not significantly improve the fit of the models.
 
@@
 However, little to no effect is measured on the concurrent price term.}
-	\label{price\_coefficients}
+	\label{price_coefficients}
 \end{figure}
+
+For computationally intensive Generalized Additive Models, few if any analytic results are available to confirm that the estimates are unbiased.
+Instead, I rely on Monte Carlo experiments to test for biased results.
+Here, I create artificial field data using a cumulative logistic function with a linear price term and random variation.
+I then estimate the effect of price repeatedly, generating a new random component before each new estimation.
+Details of the Monte Carlo experiment can be found in the online appendix \url{https://github.com/jmaurit/oil_prices/blob/master/mc_sim.html}.
+
+The main results are shown in figure \ref{mc_results}.
+The top panel shows one run of the artificially generated data.
+I compare estimations of a concurrent price term, $beta_hat$ made with a Generalized Linear Model and Generalized Additive Model with a 2-dimensional smooth function.
+I set the true parameter $beta$ to both $0$ and $0.05$, which are shown as vertical lines in the figures.
+The figure shows that while a bias exists in the estimation of the GLM model, especially for when the true value is $0.05$, the Generalized Additive Model generally provides unbiased results.
 
@@
 
-The main results of this research is to show that production in existing Norwegian offshore fields has no significant concurrent reaction to higher oil prices while a slight effect is estimated  with a lag of between 4 and 8 years.
+The main results of this research is to show that production in existing Norwegian offshore fields has no significant concurrent reaction to higher oil prices while a slight effect is estimated  with a lag of between 1 and 4 years.
 Oil producers do not appear to be behaving strategically in relation to short-term production - increasing or reducing production in response to changes in oil price.
@@
 Changes in oil prices can rather be seen as a relaxing of a production constraint, justifying increased investment that leads to either a higher total extraction rate or an inter-temporal shifting of production.
-Furthermore, it appears that the most of the effect of price comes at the planning phase of a field's production life.
-Little to no effect of price on production is found in the build-out or depletion phase.
 

### 2015-04-05 rr_analysis_1.r

@@ -41,25 +41,79 @@ summary(gam_mod)
 
 #terms
 
-pdat <- with(large_fields,
-             data.frame(prod_year = round(seq(min(prod_year), max(prod_year), length = 36)),
-                        year = round(seq(min(year), max(year), length = 36)),
-                        oil_prod_mill_sm3 = rep(mean(oil_prod_mill_sm3), 36),
-                        in_place_oil_mill_sm3 = rep(mean(in_place_oil_mill_sm3), 36),
-                        cost_index = rep(mean(cost_index), 36),
-                        price = rep(mean(price), 36),
-                        price_l1 = rep(mean(price_l1), 36),
-                        price_l2= rep(mean(price_l2), 36),
-                        price_l3= rep(mean(price_l3), 36),
-                        price_l4 = rep(mean(price_l4), 36),
-                        price_l5= rep(mean(price_l5), 36),
-                        price_l6 = rep(mean(price_l6), 36),
-                        price_l7 = rep(mean(price_l7), 36),
-                        price_l8= rep(mean(price_l8), 36)
+pdat1 <- with(large_fields,
+             data.frame(prod_year = round(seq(min(prod_year), max(prod_year), length = 200)),
+                        year = rep(mean(year), 200),
+                        oil_prod_mill_sm3 = rep(mean(oil_prod_mill_sm3), 200),
+                        in_place_oil_mill_sm3 = rep(mean(in_place_oil_mill_sm3), 200),
+                        cost_index = rep(mean(cost_index), 200),
+                        price = rep(mean(price), 200),
+                        price_l1 = rep(mean(price_l1), 200),
+                        price_l2= rep(mean(price_l2), 200),
+                        price_l3= rep(mean(price_l3), 200),
+                        price_l4 = rep(mean(price_l4), 200),
+                        price_l5= rep(mean(price_l5), 200),
+                        price_l6 = rep(mean(price_l6), 200),
+                        price_l7 = rep(mean(price_l7), 200),
+                        price_l8= rep(mean(price_l8), 200)
                         ))
-pred2 <- predict(gam_mod, pdat, type = "terms", se.fit = TRUE)
-pred2$fit
-pred2$se
+
[…]

### 2015-04-06 oil_prices_rr.tex
added pdf of appendix

@@
 
-Moreover, studies within the economics literature fail to take into account how oil prices have varying effects on the different stages of field development and production.
+% Moreover, studies within the economics literature fail to take into account how oil prices have varying effects on the different stages of field development and production.
 The planning, build-out and depletion phase of a field are conceptually unique and how oil companies operating off-shore respond to oil prices can be expected to vary substantially.
@@
 \citet{mohn_efforts_2008} suggests and finds evidence for the idea that in periods of high oil prices off-shore producers will invest more in risky wild-cat drilling in search of new fields, but concentrate investments in lower-risk ventures, like expanding production in existing fields, when prices are low.
-If this effect were to dominate, then it may even be plausible that production in existing fields reacts \emph{negatively} to increases in price.
+If this effect were to dominate, then it may even be plausible that a fall in prices leads to an \emph{increase} in production.
 
@@
 
-In this article, I use historical production data from all 77 currently or formerly oil-producing fields on the Norwegian continental shelf in order to estimate the effect that prices have on oil production.
+In this article, I use historical production data from all 77 currently or formerly oil-producing fields on the Norwegian Continental Shelf in order to estimate the effect that prices have on oil production.
 
@@
 
-\FloatBarrier
 \section{The effect of oil price on production: theory, simulations and empirics}
@@
 
-	Log(production_{i,t}) & = f(production\_time_{i,t}) + \beta_1 in\_place\_oil_{i,t} \\
-	 \quad & + \beta_2 oil\_price\_lags_t + \beta_3 cost\_index_{t} +  
-	 f(year_t) + \epsilon_{i,t}
+	Log(production_{i,t}) & = f(production\_time_{i,t}) + f(year_t) + \beta_1 in\_place\_oil_{i,t}\\
+	 \quad & + \beta_3 cost\_index_{t} + \mathbf{\beta_2 oil\_price\_lags_t} +  
+	  \epsilon_{i,t}
 \label{gam_price_eqn}
@@
 
-||\boldsymbol{y} - \boldsymbol{X\beta} ||^2 + \lambda \int_{0}^{1} [f´´(x)]^2dx
+\| \mathbf{y} - \mathbf{X\beta} \| ^2 + \lambda \int_{0}^{1} [f^{\prime \prime}(x)]^2dx
 
@@
 To address this I can add field random effects that takes into account uncertainty based on random variation between the fields in the estimation.
-The smoothed function can then be written as $f(production\_time_{i,t})*Z$.
-Where $Z~N(0, \sigma)$ and $\sigma$ is the variance term to be estimated.
+The smoothed function can then be written as $f(production\_time_{i,t}*Z)$.
+Where $Z\~N(0, \sigma)$ and $\sigma$ is the variance term to be estimated.
 
@@
 The second column shows results with field-level random effects, while the third columns shows results when a two-dimensional smoothed function is used to control for production profiles across field size.
-The latter two models are my preferred models, providing a substantially better fit, as measured by both deviance explained and R^2.
+The latter two models are my preferred models, providing a substantially better fit, as measured by both deviance explained and $R^2$.
 Monte Carlo Experiments, which I describe further below, also indicates that the latter two models provide consistently less-biased parametric estimates of the price terms.
@@
                                              & $(0.146)$      & $(0.183)$      & $(51.125)$       \\
-I(in\_place\_oil\_mill\_sm3/100)             & $0.337^{***}$  & $-0.014$       &                  \\
+in\_place\_oil\_mill\_100sm3             & $0.337^{***}$  & $-0.014$       &                  \\
                                              & $(0.005)$      & $(0.014)$      &                  \\
@@
                                              & $(0.003)$      & $(0.003)$      & $(0.002)$        \\
-I(price/10)                                  & $-0.006$       & $0.014$        & $0.007$          \\
+price/10                                  & $-0.006$       & $0.014$        & $0.007$          \\
                                              & $(0.009)$      & $(0.010)$      & $(0.006)$        \\
-I(price\_l1/10)                              & $0.004$        & $0.021^{*}$    & $0.015^{*}$      \\
+price\_l1/10                              & $0.004$        & $0.021^{*}$    & $0.015^{*}$      \\
                                              & $(0.010)$      & $(0.010)$      & $(0.006)$        \\
-I(price\_l2/10)                              & $0.037^{***}$  & $0.047^{***}$  & $0.048^{***}$    \\
-                                             & $(0.009)$      & $(0.010)$      & $(0.006)$        \\
-I(price\_l3/10)                              & $0.030^{***}$  & $0.025^{**}$   & $0.033^{***}$    \\
+price\_l2/10                              & $0.037^{***}$  & $0.047^{***}$  & $0.048^{***}$    \\
+                                             & $(0.009)$      & $(0.010$      & $(0.006)$        \\
+price\_l3/10                              & $0.030^{***}$  & $0.025^{**}$   & $0.033^{***}$    \\
                                              & $(0.009)$      & $(0.009)$      & $(0.006)$        \\
-I(price\_l4/10)                              & $0.003$        & $0.043^{***}$  & $0.028^{***}$    \\
+price\_l4/10                              & $0.003$        & $0.043^{***}$  & $0.028^{***}$    \\
                                              & $(0.008)$      & $(0.008)$      & $(0.005)$        \\
-I(price\_l5/10)                              & $-0.020^{**}$  & $0.012$        & $-0.001$         \\
+price\_l5/10                              & $-0.020^{**}$  & $0.012$        & $-0.001$         \\
                                              & $(0.007)$      & $(0.007)$      & $(0.005)$        \\
-I(price\_l6/10)                              & $-0.010$       & $0.008$        & $-0.001$         \\
+price\_l6/10                              & $-0.010$       & $0.008$        & $-0.001$         \\
                                              & $(0.007)$      & $(0.007)$      & $(0.004)$        \\
-I(price\_l7/10)                              & $-0.008$       & $0.007$        & $0.004$          \\
+price\_l7/10                              & $-0.008$       & $0.007$        & $0.004$          \\
                                              & $(0.007)$      & $(0.007)$      & $(0.004)$        \\
-I(price\_l8/10)                              & $-0.044^{***}$ & $-0.011$       & $-0.027^{***}$   \\
+price\_l8/10                              & $-0.044^{***}$ & $-0.011$       & $-0.027^{***}$   \\
                                              & $(0.006)$      & $(0.007)$      & $(0.004)$        \\
@@
                                              & $(2.956)$      & $(3.000)$      & $(3.000)$        \\
-EDF: s(prod\_year,name)                      &                & $33.002^{***}$ &                  \\
+EDF: s(prod\_year,re\_field)                      &                & $33.002^{***}$ &                  \\
                                              &                & $(72.000)$     &                  \\
@@
 For convenience the estimated coefficients on the price terms from the two preferred models are shown in figure \ref{price_coefficients}.
-In general the results from the two models agree with each other, indicating little to no effect in the concurrent price term, but with significant effects estimated on the 1st to 4th lags, with a peak at the 2nd lag.
-F-tests of restricted models without the concurrent and 5-8th lag confirm that these terms do not significantly improve the fit of the models.
+In general the results from the two models agree with each other, indicating little to no effect in the concurrent price term, but with significant coefficients estimated on the 1st to 4th lags, with a peak at the 2nd lag.
+F-tests with comparisons to restricted models without the concurrent and 5-8th lag confirm that these terms do not significantly improve the fit of the models.
 
@@
 For computationally intensive Generalized Additive Models, few if any analytic results are available to confirm that the estimates are unbiased.
-Instead, I rely on Monte Carlo experiments to test for biased results.
+Instead, I rely on Monte Carlo experiments to test for biasedness.
 Here, I create artificial field data using a cumulative logistic function with a linear price term and random variation.
-I then estimate the effect of price repeatedly, generating a new random component before each new estimation.
-Details of the Monte Carlo experiment can be found in the online appendix \url{https://github.com/jmaurit/oil_prices/blob/master/mc_sim.html}.
+I then estimate the effect of price repeatedly, generating the random component before each new estimation.
+Details of the Monte Carlo experiment can be found in the on-line appendix \url{https://github.com/jmaurit/oil_prices/blob/master/mc_sim.html}.
 
@@
 
-\section{Conclusion}
+\section{Conclusion and Discussion}
 
@@
 
-The modest estimated effect of prices on production adds weight to the argument of \citet{hamilton_oil_2012} that most of the increased supply of oil that comes from higher prices is from expanding the geographic and technological boundaries of oil production.
+One important caveat is that I do not claim to capture all possible mechanisms for price to influence production.
+Mechanisms that are likely gradual and smooth in nature will be absorbed by the smooth functions in the regression.
+In particular, price-induced technological change will likely be a smooth process with significant lag times.
+Referring back to figure \ref{smooths}Perhaps some evidence of this can be seen in the estimated smooth function of the calendar year variable, which shows a slight (and statistically insignificant) upwards slope after approximately 2005, corresponding to a preceding period of rising oil prices.
+
+The findings in this paper can provide some guidance on the effects of the recent steep fall in the oil price on Norwegian production, as well as production from other off-shore production areas.
+The fall in total oil production that has continued since the peak in production in 2000 has recently been arrested as several new finds began production after years of increased searching and investment.
+As mentioned, a further significant boost is expected in 2019 when production from the large Sverdrup field begins production.
+The results of this paper indicate that the recent fall in prices is unlikely to have major ramifications for production in existing fields.
+Anecdotal evidence however suggests that searching and exploration are being curtailed.
+Anecdotal evidence also suggests that lower oil prices may be leading to earlier shut-down decisions of depleting fields - a hypothesis that could be explored with future research.
+
+Finally, relating back to the literature on total oil supply, the modest estimated effect of prices on production adds weight to the argument of \citet{hamilton_oil_2012} that most of the increased supply of oil that comes from higher prices is from expanding the geographic and technological boundaries of oil production.
 For example exploration of deep-water oil deposits off the coast of Brazil and extraction of oil sands in western Canada.
 
+\FloatBarrier
 \section{Software}
@@
 
-\FloatBarrier
 \end{document}

### 2015-04-09 reply_referee2.txt
wrote response to referees

Reply to Referee 2

Thank you for the careful and constructive review of my paper.  I have made a major revision, including a complete overhaul of the methodology, and a rewriting of most of the sections based on your comments, as well as those of the other referees.  I have also substantially shortened and at some places simplified the analysis in order to comply with Energy Journal length limits, as well as to focus on the main results.  I give a short summary of the major changes below, and then follow with a point-by-point response to your comments.

Methodologically, I have tried to reduce the complexity of the modeling by excluding the smallest of the fields - chosen as under 3 mill sm3 in recoverable reserves since these show significantly different production profiles than larger fields.  This should not have a major impact on the implications of the paper since they account for only .5 % of total recoverable oil reserves. I also exclude the Ekofisk field, as this has a unique double-hump profile, related to its early history (see p.9).

I was then able to in some ways simplify and extend the methodology. Significantly, I now use a single-variable smoothed function to control for the profile. This can be interpreted directly, and thus adds intuition to the analysis.  Perhaps more importantly, i was then able to add random effects terms - modeling the idiosyncratic variance between the fields (p. 14).  I compare these results with a model that includes a two-dimensional smoothed function (p.14-15) (but where adding a random effects term is not possible) and the main results generally agree.

To deal with the potential that the peak of production could be endogenous to price, I now also include only a single smoothed function for the full production profile, rather than splitting the function into two pieces before and after the peak.  I have further extended the model to try to deal with the issue of technological change, as well as the possibility that costs are correlated with price, which I discuss below.

The results are on the whole similar, though differ in detail. I still find no significant evidence of a contemporary effect of oil prices on production, however now I define contemporary as the same year.  In lags 1-4, modest, but significant effects are found, with the highest effect found after 2 years.

I have chosen to limit my results to the total average effect of prices on production.  Partly this is out of purely practical consideration of keeping the exposition within Energy Journal limits. But in addition, the results for different phases of production are considerably less robust, mainly because I am only using about half the data for each analysis, but also because I now am using a single smoothed function to model the production profile. Estimations of the effect of prices on investment probably deserves a seperate paper and I have also cut this section. The analysis certainly does lose some detail with these omissions, but I believe the trade-off of better clearity and robustness, and appropriate length is worth it.

Response to referee 2 comments:

"1) My greatest objection is that the paper lacks a discussion of why the author chooses to analyze a regression function as in e.g. Eq. (6) in the paper. The author has to revise this specification...To include investment and/or production costs is crucial as an increase in (marginal) costs will lead to declined production for a given oil price.

I think this is good advice, and I have redone the specifications. Significantly, I have added a new variable to account for costs.  From the year 2000 to the present this is the producer price index for oil and gas extraction from Statistics Norway. This takes into account the skyrocketing costs for rigs, steel, etc, that you note.  Unfortunately, the series does not extend farther back, so between 1960 and 2000 I use an index of US oil and gas industry costs as a proxy, with the rationale that the industry is global and costs tend to be highly correlated across regions. (p.9-10)

"Here I will add that I do not understand the sentence on p.31/32:”..a trend of increased total extraction estimates from the Norwegian continental shelf as a whole as well as from existing fields over the last 15 years of strongly rising oil prices”."

I agree, this was poorly worded - I have rewritten the entire section.

"Failure to include the demand side might be misleading, as price is a function of both demand and supply..."

I think you are right that I needed to do a better job considering endogeneity of demand in the paper.  Since prices are established in equilibrium, a bias could result if demand also was correlated with changes in field production other than through the direct price-mechanism. For example, higher GDP could mean easier lending to petroleum companies, who then might invest and increase petroleum independent of higher oil prices. I investigated this by including a measure of world GDP growth from the world bank in the regression. This did not have any effect on the estimated coeffients on price terms or on the overall fit of the model, suggesting that demand-effects are unlikely causing a bias, and therefor I excluded this variable from the final model.  I have added discussion of this on pages p. 15-16.

Other major comments:

The author leaves diagnostics and robustness checks in figures and tables in Appendix A. However, I would like a discussion in words in the text.

"I have simplified and brought forward all the analysis into the main body of the text.  I hope this will make the article easier to read and follow."

"The text is too lengthy, e.g. Fig. 3 and 4 in Section 3 are not needed, Further, above all the text in Section 5, 6 and 7 is too lengthy and repetitive."

I have eliminated some of these figures and joined together several others, and cut several sections. The manuscript is now at 25 pages plus an online appendix that details the monte-carlo experiment.

The author has to work with the language as many phrases are unfinished; a small example on p. 38: “...allowing for increased recovery percentage of the available reservoir”:

Good point - I have tried to make the exposition clearer and more concise.

1) Income tax now is 27 % (and not 28 % as is written on p. 15), while resource tax is 51 % (and not 50 % as is written on p. 15).

Thanks for the tip - I have changed the numbers in the text.

### 2015-04-09 reply_referee3.txt
wrote response to referees

--- 2015-04-09 reply_referee2.txt
@@
-Reply to Referee 2
+Reply to referee 3
 
@@
 
-Response to referee 2 comments:
+Response to comments
 
-"1) My greatest objection is that the paper lacks a discussion of why the author chooses to analyze a regression function as in e.g.
-Eq.
-(6) in the paper.
-The author has to revise this specification...To include investment and/or production costs is crucial as an increase in (marginal) costs will lead to declined production for a given oil price.
+"The different results indicate that at such micro level, the effect of price on production is sensitive to specific measures chosen for price and production.
+And if the difference in the results is mostly explained by the difference in cost of operating offshore and onshore, the micro-foundation findings in this paper would be less irrelevant for understanding the supply side of oil market on a larger scale, with offshore production contributing to about 33% of world production and deep-water to about 7%.
+As a result, contribution of this paper to the literature would be limited."
 
-I think this is good advice, and I have redone the specifications.
-Significantly, I have added a new variable to account for costs.
-From the year 2000 to the present this is the producer price index for oil and gas extraction from Statistics Norway.
-This takes into account the skyrocketing costs for rigs, steel, etc, that you note.
-Unfortunately, the series does not extend farther back, so between 1960 and 2000 I use an index of US oil and gas industry costs as a proxy, with the rationale that the industry is global and costs tend to be highly correlated across regions.
-(p.9-10)
+This is a good point worth some discussion.
+I think that you are right in that the findings are limited to off-shore production, and as you point out, this makes up a minority of world oil production.
+However, In terms of costs, many of these fields also lie on the margins, and thus are especially important in the price-formation process.
+As you also point out, the significance for countries with a high reliance on an offshore sector - like Norway, Brazil, and the UK - could be quite substantial.
+I have tried to keep a balanced discussion of this in the paper (p.1-3)
 
-"Here I will add that I do not understand the sentence on p.31/32:”..a trend of increased total extraction estimates from the Norwegian continental shelf as a whole as well as from existing fields over the last 15 years of strongly rising oil prices”."
+"The argument for adopting proper production measure is not entirely convincing.
+The paper refers to Mohn (2008)’s findings that off-shore producers concentrate investment in different production stages in response to different prices, and argues further that ”it may even be plausible that production in existing fields reacts negatively to increase in price”.
+However, investment concentrated in search of new fields doesn’t necessarily imply lower production in existing fields.
+Even though lower production in a specific field might coincide with higher prices, such negative correlation as a general observation for all fields over all time periods would be rare, especially if a larger area of fields is considered."
 
-I agree, this was poorly worded - I have rewritten the entire section.
+I see your point, and I have changed my discussion of the paper to reflect this (p.2).
 
-"Failure to include the demand side might be misleading, as price is a function of both demand and supply..."
+"It would be interesting to see how including measures of expectation could affect the results."
 
-I think you are right that I needed to do a better job considering endogeneity of demand in the paper.
-Since prices are established in equilibrium, a bias could result if demand also was correlated with changes in field production other than through the direct price-mechanism.
-For example, higher GDP could mean easier lending to petroleum companies, who then might invest and increase petroleum independent of higher oil prices.
-I investigated this by including a measure of world GDP growth from the world bank in the regression.
-This did not have any effect on the estimated coeffients on price terms or on the overall fit of the model, suggesting that demand-effects are unlikely causing a bias, and therefor I excluded this variable from the final model.
-I have added discussion of this on pages p.
-15-16.
+This is a good point, and I looked at the litterature around this question - with most finding that the current oil price is the best indicator of expected future oil prices (Mohn (2008), Alquist and Kilian (2010)).
+Another strand of literature finds that adaptive expectations plays a role (Pesaran (1990, Farzin (2001))) - where expectations are based on an average of current and past prices.
+Implicitly, I allow for this possibility by the lagged structure of my estimation.I provide some discussion of this topic on p.11.
 
-Other major comments:
+"In terms of the model estimation, there’s little information about the sample period of data.
+The concern is whether the annual real price series contains enough variation to help identify the production due to price effect and that due to physical features, especially given the relatively short production life of the fields (inferred from Figure 1 and Figure 8 in the paper)."
 
-The author leaves diagnostics and robustness checks in figures and tables in Appendix A. However, I would like a discussion in words in the text.
+I´ve tried to provide more information about the sample and data used.
+(p 9-11).
+I have also updated my data through 2014, so that I have data on field production between 1970 and 2010.
+This includes quite wide variation in the prices, so that I hope that it presents enough variation to make reasonable estimates.
+In addition, I have now excluded the smallest of the fields from the analysis, which also tend to have the shortest production life.
+Larger fields can produce for multiple decades.
 
-"I have simplified and brought forward all the analysis into the main body of the text.  I hope this will make the article easier to read and follow."
+"Also, the investment data used for estimation the investment equation lacks sufficient explanation..."
 
-"The text is too lengthy, e.g. Fig. 3 and 4 in Section 3 are not needed, Further, above all the text in Section 5, 6 and 7 is too lengthy and repetitive."
+I think you are right - and I think a good estimation of investment would require its own paper and methodology.
+For the sake of keeping within the Energy Journal page limits, and a clearer exposition of the main results, I have dropped this section from the new draft.
 
-I have eliminated some of these figures and joined together several others, and cut several sections. The manuscript is now at 25 pages plus an online appendix that details the monte-carlo experiment.
+"In addition, it’s also confusing that all fields that began oil production “prior to 2008 are excluded” to study the post-peak phase (p38).
+If only oil production after 2008 is used and all these fields took five years to build-out, the selected production data would be from the pre-peak phase.
+This is not the post-peak phase that the analysis claims to focus on."
 
-The author has to work with the language as many phrases are unfinished; a small example on p. 38: “...allowing for increased recovery percentage of the available reservoir”:
+This was a good catch - a typo on my part, I meant to say that I have excluded fields that started producing after 2008.
 
-Good point - I have tried to make the exposition clearer and more concise.
-
-1) Income tax now is 27 % (and not 28 % as is written on p. 15), while resource tax is 51 % (and not 50 % as is written on p. 15).
-
-Thanks for the tip - I have changed the numbers in the text.
+"At the end, the manuscript severely exceeds the length limit which is 25 -35 pages.
+"The paper is now 25 pages in addition to an online appendix detailing the monte-carlo experiment.

### 2015-04-09 reply_referee_1.txt
wrote response to referees

Reply to referee 1

Thank you for the careful and constructive review of my paper.  I have made a major revision, including a complete overhaul of the methodology, and a rewriting of most of the sections based on your comments, as well as those of the other referees.  I have also substantially shortened and at some places simplified the analysis in order to comply with Energy Journal length limits, as well as to focus on the main results.  I give a short summary of the major changes below, and then follow with a point-by-point response to your comments.

Methodologically, I have tried to reduce the complexity of the modeling by excluding the smallest of the fields - chosen as under 3 mill sm3 in recoverable reserves since these show significantly different production profiles than larger fields.  This should not have a major impact on the implications of the paper since they account for only .5 % of total recoverable oil reserves. I also exclude the Ekofisk field, as this has a unique double-hump profile, related to its early history (see p.9).

I was then able to in some ways simplify the methodology and in others extend it in order to deal with some of the issues you brought up. Significantly, I now use a single-variable smoothed function to control for the profile. This can be interpreted directly, and thus adds intuition to the analysis.  Perhaps more importantly, i was then able to add random effects terms - modeling the idiosyncratic variance between the fields (p. 14).  I compare these results with a model that includes a two-dimensional smoothed function (p.14-15) (but where adding a random effects term is not possible) and the main results generally agree.

In response to your suggestion that the peak of production could be endogenous to price, I now also include only a single smoothed function for the full production profile, rather than splitting the function into two pieces before and after the peak.  I have further extended the model to try to deal with the issue that you brought up of technological change biasing the results. I have also included a measure of producer prices in the norwegian oil and gas sector, to control for the fact that costs are likely highly correlated with prices.

The results are on the whole similar, though differ substantially in detail. I still find no significant evidence of a contemporary effect of oil prices on production, however now I define contemporary as the same year.  In lags 1-4, modest but statistically and economically significant effects are found, with the highest effect found after 2 years. These results are more robust and, I believe, more intuitive than those in the previous draft. In total, I believe the paper has been substantially improved due to the implimentations of your suggestions and those of the other referees.

I have chosen to limit my results to the total average effect of prices on production.  Partly this is out of purely practical consideration of keeping the exposition within Energy Journal limits. But in addition, the results for different phases of production are considerably less robust, mainly because I am only using about half the data for each analysis, but also because I now am using a single smoothed function to model the production profile. Estimations of the effect of prices on investment probably deserves a seperate paper and I have also cut this section. The analysis certainly does lose some detail with these omissions, but I believe the trade-off of better clearity, robustness, and appropriate length is worth it.

Response to comments:

"I see several problems with this assumption. One is that it ignores significant technical changes in the oil industry (changes that the paper itself discusses). Given these changes, even if oil prices had been constant throughout, fields of a given size discovered in the 1970s would surely follow a different production path than fields of the same size discovered in, say, the 2000s. If so, then the paper’s regression analysis will spuriously attribute any deviations from the average production path for fields of that size—presumably involving relatively lower production at any given point in time for the 1970s fields—to differences in oil prices over the fields’ lifetimes."

This is a good point - I have now tried to pick up the variation due to technical change by introducing a constricted smoothed function of calendar years. The function is restricted (restricting the number of knots to a maximum of 4) in order not to overfit the data - focusing on the long-term factors like technical change, rather than picking up shorter term volatility from changes in the oil-price. See p. The estimated smoothed function appears consistent with it capturing technical change - growing strongly in the early years of exploration and production, followed by a flattening out.

"A second problem, which the paper acknowledges but then to my mind erroneously dismisses as unimportant, is that field size itself—whether measured by estimates of recoverable oil or by peak production—is affected by price. At high oil prices, more oil will be viewed as recoverable. Moroever (as the paper itself argues), more investments will be made in recovering that oil, which may well affect peak production. But that implies that a field of given “true” size, as measured for example by the physical oil in place, will at different oil prices be treated as being of a different size in the paper’s regression analysis. A different baseline production path will be assigned to it and therefore different effects of oil-price changes estimated for it."

I agree that this is an important point, and was not well addressed in the previous draft.  In this version I have used estimates of total in-place oil rather than recoverable reserves as a proxy for the size of the field.  While these are still subject to uncertainty, they should not be directly correlated with movements in oil prices.

"Relatedly, given that the paper allows for only linear price effects on log production, the correlation between price and size implies that any non-linear price effects will alter the estimated baseline production paths. This in turn will bias the estimates of the linear effects."

This is a good point, and was worth checking. I attempted several alternative models with squared and other polynomial terms for the price effect, however these did not turn out to be significant, nor did they, combined lead to a better fit for the models, so I decided to stay with the linear price terms.

"A third problem is that the time at which production peaks, as measured from the start of pro- duction, is likely affected by price as well. If the oil price is high, then (again as the paper itself argues) buildout of a field is likely to be sped up, and vice versa. But that implies that a given number of years before the peak is reached, production will all else equal be lower for a field with rapid buildout induced by a high price than for a field with slower buildout induced by a low price. Since the paper’s regression analysis ignores this, it will estimate a baseline production path that averages across these two fields. As a result, it will all else equal find a spurious negative correlation between deviations of production from the estimated pre-peak baseline path and price. (More generally, estimates of that correlation will be biased downwards.)"

I have changed the model so that a total smoothed function is estimated, rather than estimating two seperate functions, split at a fixed point (the peak).  This allows for variations in the peak to be modeled endogenously in the model.

"A completely separate problem with the paper’s methodology involves the penalty for “wiggliness” in the spline estimation. Clearly, the choice of this penalty is hugely important: the more wiggly the spline estimate of the baseline production path is allowed to be, the fewer and smaller will be the deviations from that baseline, implying smaller estimated price effects. Yet apart from a cryptic reference in Appendix B to a robustness check setting an unexplained “γ parameter that controls smoothness” to 1.4 (as opposed to ?), the paper offers no discussion of how the penalty was chosen."

You are right, this detail deserves better explanation. The smooth parameter is set endogenously in the model through cross-validation. I have now noted this in the paper, provide some intuition and included a reference for the technical details (p.13)

p7 In what sense do the coefficients on the price variables estimate “the average effect of price on production over the entire production profile,” rather than just the effect on contemporaneous production (or for n-th lags, production n years in the future)?

I think my wording in the original was unclear (and after some thought, perhaps even misleading.)  In this draft, I have given a more direct interpretation, as you suggest.

p8 In what sense does Figure 2 show “the correlated production profiles of the fields” as opposed to just the pattern of field discoveries?

You are right, I think your wording is a better direct way of saying what I intended to convey - I have changed my wording (p.4). Also, in this draft I make a direct visual comparison of the bell-shaped total production curve with profiles of the largest 10 fields, showing how correlated discoveries and subsequent production coincided to create the bell-shaped total production. (fig 1)

"p17 “the methodologies used to estimate the total recoverable resource of a field are constantly evolving and it is a fair assumption that any consistent bias of the estimates are observed in older fields and corrected for in estimates for newer fields. I can then assume that existing errors are random and will not significantly bias the estimates.” I have a hard time following what is being argued here."

This is no longer included in the paper, since it is no longer relevant. (And I agree, my wording is poor.)

p19 “Yearly production is assumed to be proportional to the total size of the field.” But production is in log form, whereas size is not. In fact, I wonder how much of the bias found in the paper’s Monte Carlo analysis of what it calls the “linear model” may be due simply to the fact that size enters linearly in that model rather than in log form.

This was a good point to think about.  I tried specifications where I used the log of the in-place-oil measure.  This did not lead to any material differences in the results.  In the end, I decided to keep the absolute levels of the in-place oil estimates to be consistent with the other right-hand side variables and their interpretation.  I also included the absolute size measure in the monte-carlo experiments to check for any bias, though generally these experiments show unbiased estimators in the GAM models.

"p29 “A higher oil price means that added investments in production become attractive in order to either increase the total amount extracted from a field or to shift production forward.” Whether investment increases production or just shifts it forward in time seems very important to the general question that the paper addresses. Yet beyond this brief mention the issue is never explored."

You are right, this deserves more attention. I have added some discussion of this in the conclusion (p.21). I don´t think It is feasible to seperate these two with my current data. A combination of the efacts that the measured effects are relatively small, estimated field sizes and recoverbale oil are uncertain, and most importantly, most fields are still producing.

p29 “including several lags also allows for the possibility of adaptive expectations of future oil prices.” Whether effects of lagged prices reflect long times for investments to take effect or long times for price expectations to adjust again seems very important to the general question that the paper addresses. Yet again, beyond this brief mention the issue is never explored.

p31 “The lines in the boxplot . . . are calculated using a bayesian-inspired simulation.”

I agree, I got carried away here.  I have simplified my presenation of the results, now showing simple point estimates with 95% confidence bans calculated in the standard way (1.96*SE)

p35 How is “investment” defined and measured?

To shorten the exposition and focus the analysis on the effect of prices on production, I have cut the explicit estimation of the effects on investment.  This is a subject that probably deserves its own method and paper, and I probably was not doing it justice.

"Much of Appendix A (confusingly referred to as appendix 1 and appendix 2 at different points in the paper) is useless to the reader, because other than brief captions, no explanations are given of the figures or of the final table."

I have redone the appendix in the form of an online markdown file.  Here, I include the actual code for the Monte Carlo experiment, with comments, explanations and results imbedded directly.  I hope this improves understanding, as well as providing a direct path to replicating the results.

"Lastly, the paper is frustratingly messy, with lots of typos—including ones that a simple spell check would have caught—and confusing references to the wrong lags when discussing results."

I have paid extra attention to fixing typos and ensuring consistency.  I hope you will find this draft cleaner, and (mostly) free of errors.

### 2015-04-09 oil_prices_rr.tex
wrote response to referees

@@
 I use detailed field-level data on Norwegian off-shore oil field production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production.
-I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 3 to 5\% for a 10 dollar per barrel increase in the real price of oil.
+I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 2 to 5\% for a 10 dollar per barrel increase in the real price of oil.
 Most of this effect appears to come in the build-out or planning phase of a field's development.\\
@@
 
-\thanks{I would like to thank Klaus Mohn, Jonas Andersson, Sturla Kvamsdal, Harrison Fell, R\"ognvaldur Hannesson and Henrik Horn for valuable discussion and comments.}
+\thanks{I would like to thank three anonymous referees for helpful comments and suggestions.
+Klaus Mohn, Jonas Andersson, Sturla Kvamsdal, Harrison Fell, R\"ognvaldur Hannesson and Henrik Horn for valuable discussion and comments.
+I have recieved financial support from the Norwegian Center for Sustainable Energy Studies (CenSES)}
 % JEL Codes: Q4, L71
@@
 If this effect were to dominate, then it may even be plausible that a fall in prices leads to an \emph{increase} in production.
+Though the converse - an increase in prices leading to reduced production - is unlikely.
 
@@
 
-More so, oil field production is correlated across fields - that is, increases and decreases in production in fields are not randomly distributed across time.
-Instead, as the top panel of figure \ref{oil_decline} shows with the production profile of the 10 largest Norwegian oil fields, production tends to be correlated across fields.
+More so, major oil field discoveries, starting times and in turn production are correlated across fields as the top panel of figure \ref{oil_decline} shows with the production profile of the 10 largest Norwegian oil fields.
 The result is a total production curve that is bell-shaped over time as in the lower panel of figure \ref{oil_decline}.
@@
 
+In general I assume that oil prices are exogenous in this model.
+On the supply side, oil is a globally traded commodity.
+Since I am measuring changes in production at the field level, these are highly unlikely to have any effect on global oil prices.
+
+A more subtle possibility for endogeneity exists on the demand side.
+A bias could result if demand also was correlated with changes in field production other than through the direct price-mechanism.
+For example, higher GDP could mean easier lending to petroleum companies, who then might invest and increase petroleum independent of higher oil prices.
+I investigated this by including a measure of world GDP growth from the world bank in the regression.
+This did not have any effect on the estimated coeffients on price terms or on the overall fit of the model, suggesting that demand-side effects are unlikely to be causing a bias.
+I therefor exclude this variable from the results presented below.
+
 \section{The Effect of Oil Price on Field Production}
@@
 I then estimate the effect of price repeatedly, generating the random component before each new estimation.
-Details of the Monte Carlo experiment can be found in the on-line appendix \url{https://github.com/jmaurit/oil_prices/blob/master/mc_sim.html}.
+Details of the Monte Carlo experiment can be found in the on-line appendix \url{https://github.com/jmaurit/oil_prices/blob/master/mc_sim.pdf}.
 
@@
 
+Another important caveat, is that the changes in production could potentially either be a permanent increase in production, or a shift forward of production that may have happened later.
+Unfortunately, with the existing data and methods, I do not see a good way of distinguishing between these alternatives.
+First, there is the problem of censored data, as most of the fields in the data set are still producing, thus the information on how long the fields eventually produce are not available.
+More so, the estimates of both recoverable reserves and in-place oil are to uncertain in order to to make calculations of whether a change in production in one period led to a reduction in another - especially when the changes in production are relatively modest.
+
 The findings in this paper can provide some guidance on the effects of the recent steep fall in the oil price on Norwegian production, as well as production from other off-shore production areas.

### 2015-04-09 rr_analysis_1.r

@@ -31,7 +31,7 @@ large_fields<-prod_data[prod_data["name"]!="EKOFISK",]
 
 
 gam_mod<-gam(oil_prod_mill_sm3~s(prod_year, bs="cr") +
-	I(in_place_oil_mill_sm3/100) + I(cost_index/10) +
+	I(log(in_place_oil_mill_sm3 +.01)) + I(cost_index/10) +
     s(year, bs="cr", k=4) + I(price/10) + I(price_l1/10) + I(price_l2/10) +
     I(price_l3/10) + I(price_l4/10) + I(price_l5/10) + I(price_l6/10) + 
     I(price_l7/10) + I(price_l8/10), 
@@ -116,7 +116,7 @@ png("/Users/johannesmauritzen/research/oil_prices/figures/smooths.png",
 dev.off()
 
 
-gam_mod_2d<-gam(oil_prod_mill_sm3~s(prod_year, in_place_oil_mill_sm3) +
+gam_mod_2d<-gam(oil_prod_mill_sm3~s(prod_year, I(log(in_place_oil_mill_sm3 + .01))) +
     s(year, bs="cr", k=4) +
     I(cost_index/10) +
      I(price/10) + I(price_l1/10) + I(price_l2/10) +
@@ -145,7 +145,7 @@ labs(x="Lags", y="Estimated Coefficient on Oil Price",
 
 gam_mod_re<-gam(oil_prod_mill_sm3~s(prod_year, name, bs="re") + 
     s(year, bs="cr", k=4) +
-	I(in_place_oil_mill_sm3/100) + I(in_place_oil_mill_sm3/100) + I(cost_index/10) +
+	I(log(in_place_oil_mill_sm3 + .01)) + I(in_place_oil_mill_sm3/100) + I(cost_index/10) +
     I(price/10) + I(price_l1/10) + I(price_l2/10) +
     I(price_l3/10) + I(price_l4/10) + I(price_l5/10) + I(price_l6/10) + 
     I(price_l7/10) + I(price_l8/10), 

### 2015-04-09 rr_clean.py

@@ -20,6 +20,9 @@ brentprices["oil_price_nom"]=[float(i) for i in brentprices["oil_price_nom"]]
 brentprices["year"]=brentprices.index.year
 brent_yearly=brentprices.groupby("year").mean()
 
+# from world bank - world GDP
+world_gdp = Quandl.get("WORLDBANK/WLD_NY_GDP_MKTP_KD_ZG", authtoken="WHFoD2EKU2J8H8yGn_bS")
+
 #from bp via Quandl
 
 oil_long = Quandl.get("BP/CRUDE_OIL_PRICES", authtoken="WHFoD2EKU2J8H8yGn_bS")

### 2015-04-16 reply_referee2.txt
review of paper

@@
 Thank you for the careful and constructive review of my paper.
-I have made a major revision, including a complete overhaul of the methodology, and a rewriting of most of the sections based on your comments, as well as those of the other referees.
+I have made a major revision, including an overhaul of the methodology, and a rewriting of most of the sections based on your comments, as well as those of the other referees.
 I have also substantially shortened and at some places simplified the analysis in order to comply with Energy Journal length limits, as well as to focus on the main results.
@@
 I investigated this by including a measure of world GDP growth from the world bank in the regression.
-This did not have any effect on the estimated coeffients on price terms or on the overall fit of the model, suggesting that demand-effects are unlikely causing a bias, and therefor I excluded this variable from the final model.
+I then tested whether this significantly changed the fit of the model.
+This was not the case, suggesting that demand-effects are unlikely causing a bias, and therefor I excluded this variable from the final model.
 I have added discussion of this on pages p.
@@
 
-The author leaves diagnostics and robustness checks in figures and tables in Appendix A. However, I would like a discussion in words in the text.
+"The author leaves diagnostics and robustness checks in figures and tables in Appendix A. However, I would like a discussion in words in the text."
 
-"I have simplified and brought forward all the analysis into the main body of the text.  I hope this will make the article easier to read and follow."
+I have simplified and brought forward all the analysis into the main body of the text.  I hope this will make the article easier to read and follow.
 

### 2015-04-16 reply_referee3.txt
review of paper

@@
 Thank you for the careful and constructive review of my paper.
-I have made a major revision, including a complete overhaul of the methodology, and a rewriting of most of the sections based on your comments, as well as those of the other referees.
+I have made a major revision, including a overhaul of the methodology, and a rewriting of most of the sections based on your comments, as well as those of the other referees.
 I have also substantially shortened and at some places simplified the analysis in order to comply with Energy Journal length limits, as well as to focus on the main results.
@@
 14).
-I compare these results with a model that includes a two-dimensional smoothed function (p.14-15) (but where adding a random effects term is not possible) and the main results generally agree.
+I compare these results with a model that includes a two-dimensional smoothed function (p.14-15) -but where adding a random effects term is not possible - and the main results generally agree.
 
@@
 I think that you are right in that the findings are limited to off-shore production, and as you point out, this makes up a minority of world oil production.
-However, In terms of costs, many of these fields also lie on the margins, and thus are especially important in the price-formation process.
+However, In terms of costs, many of these fields also lie on the economic margins, and thus are especially important in the price-formation process.
 As you also point out, the significance for countries with a high reliance on an offshore sector - like Norway, Brazil, and the UK - could be quite substantial.
@@
 I have also updated my data through 2014, so that I have data on field production between 1970 and 2010.
-This includes quite wide variation in the prices, so that I hope that it presents enough variation to make reasonable estimates.
+This includes quite large variation in the prices, so that I hope that it presents enough variation to make reasonable estimates.
 In addition, I have now excluded the smallest of the fields from the analysis, which also tend to have the shortest production life.
@@
 
-This was a good catch - a typo on my part, I meant to say that I have excluded fields that started producing after 2008.
+Good catch - a typo on my part, I meant to say that I have excluded fields that started producing after 2008.
 
-"At the end, the manuscript severely exceeds the length limit which is 25 -35 pages.
-"The paper is now 25 pages in addition to an online appendix detailing the monte-carlo experiment.
+"At the end, the manuscript severely exceeds the length limit which is 25 -35 pages."
+
+The paper is now 25 pages in addition to an online appendix detailing the monte-carlo experiment.

### 2015-04-16 reply_referee_1.txt
review of paper

@@
 
-"I see several problems with this assumption.
+"...I see several problems with this assumption.
 One is that it ignores significant technical changes in the oil industry (changes that the paper itself discusses).
@@
 See p.
-The estimated smoothed function appears consistent with it capturing technical change - growing strongly in the early years of exploration and production, followed by a flattening out.
+12 The estimated smoothed function appears consistent with it capturing technical change - growing strongly in the early years of exploration and production, followed by a flattening out.
 
@@
 
-"A third problem is that the time at which production peaks, as measured from the start of pro- duction, is likely affected by price as well.
+"A third problem is that the time at which production peaks, as measured from the start of production, is likely affected by price as well.
 If the oil price is high, then (again as the paper itself argues) buildout of a field is likely to be sped up, and vice versa.
@@
 
-I think my wording in the original was unclear (and after some thought, perhaps even misleading.)  In this draft, I have given a more direct interpretation, as you suggest.
+I think my wording in the original was unclear (and after some thought, perhaps even misleading.)  In this draft, I have given a more direct interpretation, as you suggest. (p. 16-17)
 
-p8 In what sense does Figure 2 show “the correlated production profiles of the fields” as opposed to just the pattern of field discoveries?
+"In what sense does Figure 2 show “the correlated production profiles of the fields” as opposed to just the pattern of field discoveries?"
 
@@
 
-"p17 “the methodologies used to estimate the total recoverable resource of a field are constantly evolving and it is a fair assumption that any consistent bias of the estimates are observed in older fields and corrected for in estimates for newer fields.
+“including several lags also allows for the possibility of adaptive expectations of future oil prices.” Whether effects of lagged prices reflect lag times for investments to take effect or long times for price expectations to adjust again seems very important to the general question that the paper addresses.
+Yet again, beyond this brief mention the issue is never explored."
+
+You are right, this is a valid point that deserves more attention, and I have added discussion and interpretation in the conclusion (p.20).
+As you point out, the lagged effects could be interpreted either as investment lags or as lags in the updating of beliefs - likely a combination of the two.
+Disentangling of these effects would probably at a minimum require a structural model with some specific assumptions the formation of expectations, and I believe this takes me too far outside the scope of this paper.
+However, I argue that the ordering and magnitude suggest that the investment lags are dominant.
+Models of adaptive expectations generally assume that the highest weight is on the concurrent price as this contains the most information and thereafter falls.
+The estimated coefficients however show no concurrent effect, and the peak in magnitude is consistently estimated at the 2-year lag.
+
+“the methodologies used to estimate the total recoverable resource of a field are constantly evolving and it is a fair assumption that any consistent bias of the estimates are observed in older fields and corrected for in estimates for newer fields.
 I can then assume that existing errors are random and will not significantly bias the estimates.” I have a hard time following what is being argued here."
@@
 
-p19 “Yearly production is assumed to be proportional to the total size of the field.” But production is in log form, whereas size is not.
+“Yearly production is assumed to be proportional to the total size of the field.” But production is in log form, whereas size is not.
 In fact, I wonder how much of the bias found in the paper’s Monte Carlo analysis of what it calls the “linear model” may be due simply to the fact that size enters linearly in that model rather than in log form.
@@
 
-"p29 “A higher oil price means that added investments in production become attractive in order to either increase the total amount extracted from a field or to shift production forward.” Whether investment increases production or just shifts it forward in time seems very important to the general question that the paper addresses.
+“"A higher oil price means that added investments in production become attractive in order to either increase the total amount extracted from a field or to shift production forward.” Whether investment increases production or just shifts it forward in time seems very important to the general question that the paper addresses.
 Yet beyond this brief mention the issue is never explored."
@@
 I have added some discussion of this in the conclusion (p.21).
-I don´t think It is feasible to seperate these two with my current data.
-A combination of the efacts that the measured effects are relatively small, estimated field sizes and recoverbale oil are uncertain, and most importantly, most fields are still producing.
+I don´t think it is feasible to seperate these two effects with my current data because of the fact that the measured effects are relatively small, estimated field sizes and recoverbale oil are uncertain, and most importantly, most fields are still producing, thus much of my data is censored.
 
-p29 “including several lags also allows for the possibility of adaptive expectations of future oil prices.” Whether effects of lagged prices reflect long times for investments to take effect or long times for price expectations to adjust again seems very important to the general question that the paper addresses.
-Yet again, beyond this brief mention the issue is never explored.
-
-p31 “The lines in the boxplot . . . are calculated using a bayesian-inspired simulation.”
+"“The lines in the boxplot . . . are calculated using a bayesian-inspired simulation.” Meaning..."
 

### 2015-04-16 oil_prices_rr.tex
review of paper

@@
 \begin{abstract}
-I use detailed field-level data on Norwegian off-shore oil field production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production.
-I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 2 to 5\% for a 10 dollar per barrel increase in the real price of oil.
-Most of this effect appears to come in the build-out or planning phase of a field's development.\\
+I use detailed field-level data on Norwegian offshore oil field production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production.
+I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 2 to 5\% for a 10 dollar per barrel increase in the real price of oil.\\
 Keywords: Oil production,Oil Prices, Norway, Generalized Additive Model.
@@
 \thanks{I would like to thank three anonymous referees for helpful comments and suggestions.
-Klaus Mohn, Jonas Andersson, Sturla Kvamsdal, Harrison Fell, R\"ognvaldur Hannesson and Henrik Horn for valuable discussion and comments.
+I thank Klaus Mohn, Jonas Andersson, Sturla Kvamsdal, Harrison Fell, R\"ognvaldur Hannesson and Henrik Horn for valuable discussion and comments.
 I have recieved financial support from the Norwegian Center for Sustainable Energy Studies (CenSES)}
@@
 
-For most of the last century, crude oil has been the world’s single most important and valuable fuel source.\footnote{Oil is the world's largest single energy source, consisting of approximately 33\% of total energy consumption in 2013 as well as the most valuable in terms of market price per unit of energy \citep{british_petroleum_statistical_2013}} Naturally, questions of how the oil price affects the world economy as well as how oil production reacts to the oil price have been fundamental topics in economics.
+For most of the last century, crude oil has been the world's single most important and valuable fuel source.\footnote{Oil is the world's largest single energy source, consisting of approximately 33\% of total energy consumption in 2013 as well as the most valuable in terms of market price per unit of energy \citep{british_petroleum_statistical_2013}} Naturally, questions of how the oil price affects the world economy as well as how oil production reacts to the oil price have been fundamental topics in economics.
 Recent volatility in the price of oil has made this an especially relevant topic.
@@
 The studies that do use field- and well-level data, such as \citet{rao_taxation_2010}, often use data from on-shore installations.
-However an increasing amount of the world's oil production comes from hard-to-reach off-shore fields.
-Production from challenging off-shore environments is substantially different in character than from on-shore installations.
+However an increasing amount of the world's oil production comes from hard-to-reach offshore fields.
+Production from challenging offshore environments is substantially different in character than from on-shore installations.
 
 % Moreover, studies within the economics literature fail to take into account how oil prices have varying effects on the different stages of field development and production.
-The planning, build-out and depletion phase of a field are conceptually unique and how oil companies operating off-shore respond to oil prices can be expected to vary substantially.
+The planning, build-out and depletion phase of a field are conceptually unique and how oil companies operating offshore respond to oil prices can be expected to vary substantially.
 
 The effect of oil prices on producing fields is an important topic in understanding the mechanisms of how total oil supply reacts in response to price.
-The response of oil production in existing fields is especially important now as many of the major oil- producing areas, like the Norwegian Continental Shelf are mature and an increasing share of investments will be directed at capacity enhancement in producing fields rather than new field developments.
+The response of oil production in existing fields is especially important now as many of the major oil-producing areas, like the Norwegian Continental Shelf are mature and an increasing share of investments will be directed at capacity enhancement in producing fields rather than new field developments.
 How production from these fields will respond to changes in oil prices has major implications for the oil industry, the state finances of oil-producing countries and long-run oil price formation.
@@
 At a basic level a central idea of much of this theoretical work is that with a non-renewable resource, production is a decision that involves a significant opportunity cost: more production in the current period means less production in future periods.
-Within this frameworks, prices and expectations of prices become important variables in the production decision.
+Within these frameworks, prices and expectations of prices become important variables in the production decision.
 A simple Hotelling model suggests that a producer would immediately change their production in response to a change in oil price in order to dynamically optimize the total extraction.
@@
 Higher oil prices may however still lead a producer to invest in increased capacity.
-Since lag times are significant in the off-shore sector, we would then expect to see a multi-year lagged effect of prices on production.
+Since lag times are significant in the offshore sector, we would then expect to see a multi-year lagged effect of prices on production.
 
 However even with the question of lagged production and investment, some ambiguity exists.
-\citet{mohn_efforts_2008} suggests and finds evidence for the idea that in periods of high oil prices off-shore producers will invest more in risky wild-cat drilling in search of new fields, but concentrate investments in lower-risk ventures, like expanding production in existing fields, when prices are low.
+\citet{mohn_efforts_2008} suggests and finds evidence for the idea that in periods of high oil prices offshore producers will invest more in risky wild-cat drilling in search of new fields.
+But when prices are low, they tend to concentrate investments in lower-risk ventures, like expanding production in existing fields.
 If this effect were to dominate, then it may even be plausible that a fall in prices leads to an \emph{increase} in production.
@@
 The Norwegian government has been on the forefront of this movement and increasingly committed itself to transparency in the petroleum sector.
-\footnote{see http://www.regjeringen.no/en/sub/eiti---extractive-industries-tranparency/about-eiti.html?id=633586} Over the last several years, detailed data on most aspects of the country\’s oil industry has become openly available.
+\footnote{see http://www.regjeringen.no/en/sub/eiti---extractive-industries-tranparency/about-eiti.html?id=633586} Over the last several years, detailed data on most aspects of the country's oil industry has become openly available.
 
-In this article, I use historical production data from all 77 currently or formerly oil-producing fields on the Norwegian Continental Shelf in order to estimate the effect that prices have on oil production.
+In this article, I use historical production data from nearly all 77 currently or formerly oil-producing fields on the Norwegian Continental Shelf in order to estimate the effect that prices have on oil production.
 
-By looking only at the effect of price on fields that currently or previously have produced oil I am limiting the scope of this article.
+By looking only at the effect of price on fields that currently or previously have produced oil, I am limiting the scope of this article.
 The effect of oil prices on total production over an extended period of time is due not just to reactions in production in existing fields but also increased searching for new fields.
@@
 	\includegraphics[width=1\textwidth]{figures/oil_decline.png}
-	\caption{The panel shows the production from the largest 10 fields, which are temporally correlated with each other.  The total production over time is bell-shaped, as shown by the bottom panel.}
+	\caption{The top panel shows the production from the largest 10 oil fields, whose starting times are temporally correlated with each other.
+The total production over time is bell-shaped, as shown by the bottom panel.}
 	\label{oil_decline}
@@
 As a solution I use a semi-parametric model within the Generalized Additive Model frameworks of \cite{hastie_generalized_1990}.
-Here I use both a cubic spline and a two-dimensional thin-plate spline to account for the general non-parametric shape of the production profile and allowing price to enter the equation linearly while also attempting to control for the effects of varying production costs and the effects of technological change and other time-varying effects.
+Here I use both a cubic regression spline and a two-dimensional thin-plate regression spline to account for the general non-parametric shape of the production profile, while allowing price to enter the equation linearly.
+I also attempt to control for the effects of varying production costs and the effects of technological change.
 
 The results suggest that little to no concurrent effect of oil prices on production exists.
-However modest lagged effects of between 1 to 4 years exist.These results are in line with an understanding of the industry as being highly capital intensive with high operating costs.
-Maintaining spare capacity in order to adjust to changing oil prices is likely prohibitively inexpensive.
+However I estimate modest effects on the first through fourth lagged terms.
+These results are in line with an understanding of the industry as being highly capital intensive with high operating costs.
+Maintaining spare capacity in order to adjust to changing oil prices is likely prohibitively expensive.
 Instead, production is increased through investment, which may take up to several years to show an effect.
 
-\section{The effect of oil price on production: theory, simulations and empirics}
+\section{The Effect of Oil Price on Production: Theory, Simulations and Empirics}
 
@@
 \citet{black_is_1998} tests the relevance of “nesting” a structural empirical model of profit maximization that takes into account oil prices into a typical geo-engineering model of oil field production.
-They find strong evidence that taking into account profit maximization - and implicitly price - substantially improves the fit compared to a purely geo-engineering type model.
-The limitation of their methodology is that they are only able to test whether including economic factors like price affects the fit of the model but are not able to give an estimate of the effect.
+They find strong evidence that taking into account profit maximization - and implicitly price - substantially improves the fit compared to a purely geo-engineering model.
+The limitation of their methodology is that they are only able to test whether including economic factors like price affects the fit of the model but are not able to give an estimate of the magnitude of the effect.
 Methodologically, the paper also relies heavily on assumptions about the functional form of both the geo-engineering aspects of the oil producer as well as their profit-maximization.
-By taking a more flexible, semi-parametric approach to estimating the effects of oil field production profile, this paper avoids problems with overly restrictive assumptions.
+By taking a more flexible, semi-parametric approach to modeling oil field production profiles, this article avoids problems with overly restrictive assumptions.
 
-\citet{rao_taxation_2010} uses data on land-based oil wells in California to estimate the effects of tax changes and price controls.
-The author finds that short-term tax changes caused small, but significant retiming of production from oil wells.
-These findings are over-all consistent with the findings of this paper, though the author finds concurrent effects of taxation on production while I do not find any concurrent effect of prices on production.
+\citet{rao_taxation_2010} uses data on onshore oil wells in California to estimate the effects of tax changes and price controls.
+The author finds that short-term tax changes caused small, but significant re-timing of production from oil wells.
+These findings are over-all consistent with the findings of this article, though the author finds a concurrent effect of changes in taxation on production while I do not find any concurrent effect of prices on production.
 The significant differences in cost and complexity between operating offshore and onshore likely account for the different results.
-How producers react to a short-term tax change as opposed to a change in price is likely also different.
+How producers react to a short-term tax change as opposed to a change in price also likely differs.
 
@@
 
-\section{Oil production on the Norwegian Continental Shelf}
+\section{Oil Production on the Norwegian Continental Shelf}
 
-The first commercial oil well in Norwegian continental waters was discovered in December of 1969 in what became the Ekofisk oil field, the largest Norwegian oil field by estimated recoverable reserves.
+The first commercial oil well in Norwegian continental waters was discovered in December of 1969 in what became the Ekofisk oil field.
 Most of the largest fields in the North Sea were found relatively early on while more recent finds have tended to be smaller - a pattern typical of oil producing areas called creaming.
@@
 Exploration in the Norwegian Sea was opened in the early 1980’s and the first commercial field started production in 1981.
-While several mid-sized fields have been discovered, the Norwegian Sea has generally disappointed in terms of commercial oil finds and most finds have been relatively small.
+While several mid-sized fields have been discovered, the Norwegian Sea has generally disappointed in terms of commercial oil finds.
 
@@
 
-Profits from oil and gas production in Norway are subject to a resource tax of 51 \% on top of the ordinary income tax of 27 \%, thus income from petroleum production is taxed at a total marginal tax rate of 78 \%.
+Profits from oil and gas production in Norway are subject to a resource tax of 51 \% on top of the ordinary income tax of 27 \%, thus income from petroleum production is taxed at a total marginal rate of 78 \%.
 The central government also receives revenues through ownership stakes in companies, notably Statoil, where the state is the majority stakeholder.
@@
 \section{Data}
-Production data of Norwegian oil fields is obtained from the website of the Norwegian Petroleum Directorate.\footnote{http://factpages.npd.no/} The data I use includes all current or formerly oil-producing fields, from 1971 through 2014.
+Norwegian oil field production data is obtained from the website of the Norwegian Petroleum Directorate.\footnote{http://factpages.npd.no/} The data I use includes nearly all current or formerly oil-producing fields, from 1971 through 2014.
 Production data is available at a monthly frequency, though I choose to aggregate up to yearly values both to smooth over seasonality as well as short-term volatility of output due to factors such as weather or technical issues that are not relevant for this article.
@@
 	\includegraphics[width=1\textwidth]{figures/data_descriptives.png}
-	\caption{Data descriptives. The left panels show histograms of estimates of recoverable oil and in-place oil by field. The right panels shows the production cost index and real brent oil prices.}
+	\caption{Data descriptives. The left panels show histograms of estimates of recoverable oil and in-place oil by field. The right panels shows the production cost index and real Brent oil prices.}
 	\label{data_descriptives}
@@
 
-To simplify the analysis, I eliminate the smallest fields from my analysis - those with less than 3 million Standard Cubic Meters (SM3) of recoverable reserves.
+To simplify the analysis, I eliminate the smallest fields - those with less than 3 million SM3 of recoverable reserves.
 These fields tend to have substantially different production profiles than the larger fields.
-However, these fields only account for around .5 \% of total oil production so this exclusion is unlikely to have any serious practical implications.
+However, these fields only account for around 0.5 \% of total oil production so this exclusion is unlikely to have any serious practical implications.
 
@@
 It reached an initial peak in production in the late 1970's and began depleting.
-However, the introduction of water and gas injection technology led to a boost in production and a new peak in production in the early 2000's.
-All subsequent major fields used gas and water injection and have a single-hump profile.
+However, the introduction of water and gas injection technology led to a boost in production and a new peak in the early 2000's.
+All subsequent major fields used gas and water injection from the outset and have a single-hump profile.
 
@@
 
-Production costs are known to vary substantially with the price of oil.
-To control for this, I include the producer price index for Norwegian oil and gas extraction industry, obtained from Statistics Norway.
-\footnote{\url{https://www.ssb.no/en/priser-og-prisindekser/statistikker/ppi}} This series only goes back to 2000, however, so I also include an index of oil and gas well costs from the US energy information agency for the years 1960 to 2000.
+Production costs are known to covary with the price of oil.
+To control for this, I include the producer price index for the Norwegian oil and gas extraction industry, obtained from Statistics Norway.
+\footnote{\url{https://www.ssb.no/en/priser-og-prisindekser/statistikker/ppi}} Unfortunately, this series only goes back to 2000, so I also include an index of oil and gas well costs from the US energy information agency for the years 1960 to 2000.
 \footnote{\url{http://www.eia.gov/dnav/pet/pet_crd_wellcost_s1_a.htm}} While this measure is unlikely to be a perfect substitute for costs in Norway, oil and gas extraction is a global industry, where costs are highly correlated across regions.
@@
 
-An argument can be made that expectations of future oil prices can be equally if not even more important for production decisions as the current oil price.
+An argument can be made that expectations of future oil prices can be equally if not more important for production decisions as the current oil price.
 Forecasts for future oil prices are available from, among others, the International Energy Agency, but these have tended to be notoriously inaccurate and it is unlikely oil companies use these projections for their investment decisions.
@@
 \citet{mohn_investment_2008} as well as \citet{pesaran_econometric_1990} and \citet{farzin_impact_2001} find evidence for adaptive expectations, where expectations of future prices is based on a weighted average of current and past prices.
-I take account of this by including several years of price lags in my regression equations.
+This is implicitly taken into account by the price lags in my regression equations, however this also leads to some unavoidable ambiguity in interpreting the results, which I discuss in the conclusion.
 
@@
 
-\section{A generalized additive model of oil field production}
+\section{A Generalized Additive Model of Oil Field Production}
 Parametric linear models have the sizeable advantages of simplicity and interpretability and are therefore usually a good starting point for an analysis.
@@
 
-Instead of attempting to estimate the shape of the production profiles of the fields by estimating parameters on linear terms I estimate a non-parametric function for the production profile.
+Instead of attempting to estimate the shape of the production profiles of the fields by estimating parameters on linear terms, I estimate a non-parametric function for the production profile.
 At a conceptual level, this model is decomposing the variation in the data into a smoothed component representing the technical and physical forces that produce the typical bell-shaped production profile, and the remaining component, that I estimate the effect of price with.
@@
 \begin{split}
-
 	Log(production_{i,t}) & = f(production\_time_{i,t}) + f(year_t) + \beta_1 in\_place\_oil_{i,t}\\
@@
 
-$f(year_t)$ represents a function of the calendar year.  The main purpose of including this term is to control for the effects of technological change.
+$f(year_t)$ represents a restricted smooth function of the calendar time.  The main purpose of including this term is to control for the effects of technological change, which I discuss further below.
 
 Finally, $oil\_price\_lags_t$ represents the variables of interest - a vector of concurrent and lagged terms for the oil price.
-Initially I included eight lagged terms of the oil price, however in the preferred specification I drop lags 5 through 8 as they where not individually significant, nor as a group did they improve the fit of the model - as tested by an F-test.
-I also experimented with using quadratic terms for the oil price and its lags, however these could not be shown to be significant.
+I included eight lagged terms of the oil price and I also experimented with using quadratic terms, however these could not be shown to be significant.
 
@@
 Here a cubic polynomial function is used to fit the shape in sections, separated at points known as knots, but continuous up to the second derivative.
-The major advantage of this method, is that it can be represented in linear form $\boldsymbol{X \beta} $, and thus normal matrix algebra techniques can be used.
+The major advantage of this method, is that it can be represented in linear form, $\boldsymbol{X \beta} $, and thus standard, efficient matrix algebra techniques can be used.
 
-The model is then fit by minimizing,
+The smoothed component of the model is then fit by minimizing equation \ref{min_equation}.
 
 \begin{equation}
-
-\| \mathbf{y} - \mathbf{X\beta} \| ^2 + \lambda \int_{0}^{1} [f^{\prime \prime}(x)]^2dx
-
+\| \mathbf{y} - \mathbf{X\beta} \| ^2 + \lambda \int_{0}^{1} [f^{\prime \prime}(x)]^2 dx
+\label{min_equation}
 \end{equation}
 
-The latter term, an estimate of the second derivative of the function, acts a penalty for the ``wiggiliness'' of the function.
+The latter term - an estimate of the second derivative of the function - acts as a penalty for the ``wiggiliness'' of the function.
 The total smoothness can then be controlled by adjusting the $\lambda$ term.
@@
 
-The advantage of this simple model that use smoothed functions of single-variable cubic regression splines is that the smoothed terms can be interpreted directly, and thus we can do a visual check of their appropriateness.
+The advantage of this simple model that uses a single-variable cubic regression spline is that the smoothed terms can be interpreted directly, and thus we can do a visual check of the appropriateness of the smoothed function.
 The top panel of figure \ref{smooths} shows the estimated smoothed function of the field production profile, where all other co-variates are held at their average value.
@@
 
-In the second panel, the smoothed function of calendar time $year_t$ is shown.
+In the second panel, the smoothed function of calendar time, $year_t$, is shown.
 Here a cubic regression spline is also used, however the number of knots is restricted to four in order to avoid over-fitting - capturing gradual trends over time like technological change, but not the variation caused by volatile oil prices.
@@
 
-Extended differences in the oil price will likely have some long-term effects on technology and learning, which in turn may be picked up in the smoothed function here, rather than in the parametric price terms.
+Extended periods of high or low oil prices will likely have some long-term effects on technology and learning, which in turn may be picked up in the smoothed function here, rather than in the parametric price terms.
 I discuss this further in the conclusion.
@@
 
-One objection to the simple model is that a fixed effect controlling for the general production profile of oil fields does not sufficiently address idiosyncratic differences between the fields.
-To address this I can add field random effects that takes into account uncertainty based on random variation between the fields in the estimation.
+One objection to the simple model is that a fixed effect controlling for the general production profile of oil fields does not sufficiently address idiosyncratic variance between the fields.
+To address this I can add field random effects that take into account uncertainty based on random variation between the fields in the estimation.
 The smoothed function can then be written as $f(production\_time_{i,t}*Z)$.
-Where $Z\~N(0, \sigma)$ and $\sigma$ is the variance term to be estimated.
+Where $Z \sim N(0, \sigma)$ and $\sigma$ is the variance term to be estimated.
 
@@
 In particular, production profiles differ substantially by the size of the field.
-A solution is to allow the shape to vary by production.
+A solution is to allow the shape to vary by the size of the field - here I use an estimate of in-place oil in the field rather than recoverable reserves, as recoverable reserves may be correlated with oil prices.
 Thus the smoothed can be written as a function of both production time and in place oil: $f(production\_time_{i,t}, in\_place\_oil_{i,t})$.
@@
 Estimating a smoothed function of two variables requires a somewhat more involved method.
-In particular, a Thin-plate regression spline \citep{wood_thin_2003} is used.
+In particular, a thin-plate regression spline \citep{wood_thin_2003} is used.
 Consider equation \ref{thin_plate_1}.
@@
 Following the notation of \citet{wood_generalized_2006}, $g$ is the function of $x_1$ and $x_2$ that is to be estimated by $f$, which in turn is estimated by minimizing equation \ref{thin_plate_2}.
-Here $\boldsymbol{y}$ represents a vector of $y_i$’s and $\boldsymbol{f} = (f(\boldsymbol{x_1}),f(\boldsymbol{x_2}))^t$.
+Here $\boldsymbol{y}$ represents a vector of $y_i$'s and $\boldsymbol{f} = (f(\boldsymbol{x_1}),f(\boldsymbol{x_2}))^t$.
 
@@
 The actual implementation is somewhat more involved in order to increase the computational efficiency of the estimation.
-For further details I again refer to \citet{wood_thin_2003}.
+For further technical details I refer to \citet{wood_thin_2003}.
 
@@
 A bias could result if demand also was correlated with changes in field production other than through the direct price-mechanism.
-For example, higher GDP could mean easier lending to petroleum companies, who then might invest and increase petroleum independent of higher oil prices.
-I investigated this by including a measure of world GDP growth from the world bank in the regression.
-This did not have any effect on the estimated coeffients on price terms or on the overall fit of the model, suggesting that demand-side effects are unlikely to be causing a bias.
+For example, higher GDP could mean easier lending to petroleum companies, who then might invest and increase petroleum production independent of higher oil prices.
+I investigated this by including a measure of world GDP growth obtained from the World Bank in the regression.
+Using an F-test, adding this variable could not be shown to significantly increase the fit of the model, suggesting that demand-side effects are unlikely to be causing a bias.
 I therefor exclude this variable from the results presented below.
@@
 The first column shows results for the simple model with a single-dimensional smoothed function of the production profile.
-The second column shows results with field-level random effects, while the third columns shows results when a two-dimensional smoothed function is used to control for production profiles across field size.
+The second column shows results with field-level random effects, while the third columns shows results when a two-dimensional smoothed function is used to control for production profiles across field sizes.
 The latter two models are my preferred models, providing a substantially better fit, as measured by both deviance explained and $R^2$.
-Monte Carlo Experiments, which I describe further below, also indicates that the latter two models provide consistently less-biased parametric estimates of the price terms.
+Monte Carlo experiments, which I describe further below, also indicate that the latter two models provide consistently less-biased parametric estimates of the price terms.
+This is especially the case for the model with a 2-dimensional smoothed term.
 
@@
                                              & $(0.005)$      & $(0.014)$      &                  \\
-I(cost\_index/10)                            & $-0.018^{***}$ & $-0.030^{***}$ & $-0.014^{***}$   \\
+cost\_index/10                            & $-0.018^{***}$ & $-0.030^{***}$ & $-0.014^{***}$   \\
                                              & $(0.003)$      & $(0.003)$      & $(0.002)$        \\
@@
 EDF: s(prod\_year)                           & $6.381^{***}$  &                &                  \\
-                                             & $(7.249)$      &                &                  \\
 EDF: s(year)                                 & $2.768^{***}$  & $3.000^{***}$  & $3.000^{***}$    \\
-                                             & $(2.956)$      & $(3.000)$      & $(3.000)$        \\
+
 EDF: s(prod\_year,re\_field)                      &                & $33.002^{***}$ &                  \\
-                                             &                & $(72.000)$     &                  \\
+
 EDF: s(prod\_year,in\_place\_oil\_mill\_sm3) &                &                & $27.445^{***}$   \\
-                                             &                &                & $(27.895)$       \\
+
 \hline
@@
 \end{tabular}
-\caption{Model results. Standard errors in parenthesis.}
+\caption{Model results.
+Estimated standard coefficients are shown for the linear terms.
+Standard errors are in parenthesis.
+For the smoothed terms, estimated degrees of freedom are displayed, where the p-values are from F-tests of whether the terms significantly improve the fit of the model.}
 \label{GAM_model_table}
@@
 In general the results from the two models agree with each other, indicating little to no effect in the concurrent price term, but with significant coefficients estimated on the 1st to 4th lags, with a peak at the 2nd lag.
-F-tests with comparisons to restricted models without the concurrent and 5-8th lag confirm that these terms do not significantly improve the fit of the models.
+F-tests with comparisons to restricted models without the concurrent and 5-8th lags indicate that these terms do not significantly improve the fit of the models.
 
@@
 For computationally intensive Generalized Additive Models, few if any analytic results are available to confirm that the estimates are unbiased.
-Instead, I rely on Monte Carlo experiments to test for biasedness.
+Instead, I rely on Monte Carlo experiments to test for bias.
 Here, I create artificial field data using a cumulative logistic function with a linear price term and random variation.
@@
 The top panel shows one run of the artificially generated data.
-I compare estimations of a concurrent price term, $beta_hat$ made with a Generalized Linear Model and Generalized Additive Model with a 2-dimensional smooth function.
+I compare estimations of a concurrent price term, $beta\_hat$ made with a Generalized Linear Model and Generalized Additive Model with a 2-dimensional smooth function.
 I set the true parameter $beta$ to both $0$ and $0.05$, which are shown as vertical lines in the figures.
@@
 
-The main results of this research is to show that production in existing Norwegian offshore fields has no significant concurrent reaction to higher oil prices while a slight effect is estimated  with a lag of between 1 and 4 years.
+The main result of this research is to show that production in existing Norwegian offshore fields has no significant concurrent reaction to higher oil prices while a slight effect is estimated  with a lag of between 1 and 4 years.
 Oil producers do not appear to be behaving strategically in relation to short-term production - increasing or reducing production in response to changes in oil price.
@@
 In particular, price-induced technological change will likely be a smooth process with significant lag times.
-Referring back to figure \ref{smooths}Perhaps some evidence of this can be seen in the estimated smooth function of the calendar year variable, which shows a slight (and statistically insignificant) upwards slope after approximately 2005, corresponding to a preceding period of rising oil prices.
+Referring back to figure \ref{smooths}, some evidence of this can perhaps be seen in the estimated smooth function of the calendar year variable, which shows a slight (and statistically insignificant) upwards slope after approximately 2005, corresponding to a preceding period of rising oil prices.
 
-Another important caveat, is that the changes in production could potentially either be a permanent increase in production, or a shift forward of production that may have happened later.
+Another important caveat, is that the changes in production could potentially either be a permanent increase in production, or a shift forward of production that may otherwise have happened later.
 Unfortunately, with the existing data and methods, I do not see a good way of distinguishing between these alternatives.
 First, there is the problem of censored data, as most of the fields in the data set are still producing, thus the information on how long the fields eventually produce are not available.
-More so, the estimates of both recoverable reserves and in-place oil are to uncertain in order to to make calculations of whether a change in production in one period led to a reduction in another - especially when the changes in production are relatively modest.
+More so, the estimates of both recoverable reserves and in-place oil are too uncertain in order to to make calculations of whether a change in production in one period led to a reduction in another - especially when the changes in production are relatively modest.
 
-The findings in this paper can provide some guidance on the effects of the recent steep fall in the oil price on Norwegian production, as well as production from other off-shore production areas.
+Interpreting the lagged effects also involves some subtlety.
+As noted, evidence has been found for adaptive expectations of future oil prices.
+Thus, the lagged effects could be interpreted as both the result of investment lags as well as lags involved in updating of expectations.
+Completely disentangling these effects would probably at a minimum require a structural model with some specific assumptions about the formation of expectations, which is beyond the scope of this article.
+
+However, the magnitude and the ordering of the estimated price effects indicates that investment lags are likely dominant.
+Since the current price always contains the most up-to-date information, then a model of adoptive expectations would show the largest effect at the concurrent term, thereafter receding.
+The estimations in this article however can show no significant concurrent effect, with the largest magnitude effect consistently being estimated at a 2-year lag.
+
+The findings in this paper can provide some guidance on the effects of the recent steep fall in the oil price on Norwegian production, as well as production from other offshore production areas.
 The fall in total oil production that has continued since the peak in production in 2000 has recently been arrested as several new finds began production after years of increased searching and investment.
 As mentioned, a further significant boost is expected in 2019 when production from the large Sverdrup field begins production.
-The results of this paper indicate that the recent fall in prices is unlikely to have major ramifications for production in existing fields.
+The results of this article indicate that the recent fall in prices is unlikely to have major ramifications for production in existing fields.
 Anecdotal evidence however suggests that searching and exploration are being curtailed.

### 2015-04-16 rr_analysis_1.r

@@ -31,7 +31,7 @@ large_fields<-prod_data[prod_data["name"]!="EKOFISK",]
 
 
 gam_mod<-gam(oil_prod_mill_sm3~s(prod_year, bs="cr") +
-	I(log(in_place_oil_mill_sm3 +.01)) + I(cost_index/10) +
+	I(log(in_place_oil_mill_sm3 +.01)) + I(cost_index/10) + world_gdp +
     s(year, bs="cr", k=4) + I(price/10) + I(price_l1/10) + I(price_l2/10) +
     I(price_l3/10) + I(price_l4/10) + I(price_l5/10) + I(price_l6/10) + 
     I(price_l7/10) + I(price_l8/10), 
@@ -118,8 +118,8 @@ dev.off()
 
 gam_mod_2d<-gam(oil_prod_mill_sm3~s(prod_year, I(log(in_place_oil_mill_sm3 + .01))) +
     s(year, bs="cr", k=4) +
-    I(cost_index/10) +
-     I(price/10) + I(price_l1/10) + I(price_l2/10) +
+    I(cost_index/10) + world_gdp +
+    I(price/10) + I(price_l1/10) + I(price_l2/10) +
     I(price_l3/10) + I(price_l4/10) + I(price_l5/10) + I(price_l6/10) + 
     I(price_l7/10) + I(price_l8/10), 
     family=gaussian(link=log), weights=I(in_place_oil_mill_sm3/100), data=large_fields)
@@ -144,7 +144,7 @@ labs(x="Lags", y="Estimated Coefficient on Oil Price",
 #With random effects
 
 gam_mod_re<-gam(oil_prod_mill_sm3~s(prod_year, name, bs="re") + 
-    s(year, bs="cr", k=4) +
+    s(year, bs="cr", k=4) + world_gdp +
 	I(log(in_place_oil_mill_sm3 + .01)) + I(in_place_oil_mill_sm3/100) + I(cost_index/10) +
     I(price/10) + I(price_l1/10) + I(price_l2/10) +
     I(price_l3/10) + I(price_l4/10) + I(price_l5/10) + I(price_l6/10) + 

### 2015-04-16 rr_clean.py

@@ -20,8 +20,6 @@ brentprices["oil_price_nom"]=[float(i) for i in brentprices["oil_price_nom"]]
 brentprices["year"]=brentprices.index.year
 brent_yearly=brentprices.groupby("year").mean()
 
-# from world bank - world GDP
-world_gdp = Quandl.get("WORLDBANK/WLD_NY_GDP_MKTP_KD_ZG", authtoken="WHFoD2EKU2J8H8yGn_bS")
 
 #from bp via Quandl
 
@@ -176,7 +174,20 @@ prod_data["build_out"]=build_out["build_out"]
 
 prod_data.to_csv('/Users/johannesmauritzen/research/oil_prices/data/prod_data.csv', index=False)
 
+# from world bank - world GDP
+prod_data = pd.read_csv('/Users/johannesmauritzen/research/oil_prices/data/prod_data.csv')
 
+world_gdp = Quandl.get("WORLDBANK/WLD_NY_GDP_MKTP_KD_ZG", authtoken="WHFoD2EKU2J8H8yGn_bS")
+
+world_gdp['year'] = world_gdp.index.year
+
+world_gdp.reset_index(inplace=True)
+
+del world_gdp["Date"]
+world_gdp.columns = ["world_gdp", "year"]
+prod_data = prod_data.merge(world_gdp, how='left', on="year")
+
+prod_data.to_csv('/Users/johannesmauritzen/research/oil_prices/data/prod_data.csv', index=False)
 
 
 

### 2015-04-20 oil_prices_rr.tex
last update before resubmission

@@
 \begin{abstract}
-I use detailed field-level data on Norwegian offshore oil field production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production.
+I use detailed field-level data on Norwegian offshore oil production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production.
 I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 2 to 5\% for a 10 dollar per barrel increase in the real price of oil.\\
@@
 
-\thanks{I would like to thank three anonymous referees for helpful comments and suggestions.
-I thank Klaus Mohn, Jonas Andersson, Sturla Kvamsdal, Harrison Fell, R\"ognvaldur Hannesson and Henrik Horn for valuable discussion and comments.
-I have recieved financial support from the Norwegian Center for Sustainable Energy Studies (CenSES)}
+\thanks{I would like to thank three anonymous referees for comments and suggestions that substantially improved this article.
+I thank Klaus Mohn, Jonas Andersson, Sturla Kvamsdal, Harrison Fell, R\"ognvaldur Hannesson and Henrik Horn for valuable discussion.
+I have received financial support from the Norwegian Center for Sustainable Energy Studies (CenSES)}
 % JEL Codes: Q4, L71
@@
 Therefor, as a proxy for the size of the fields, I use estimates of in-place oil from the Norwegian Petroleum Directorate, which is unlikely to be correlated with prices.
-The left panels of figure \ref{data_descriptives} show the histogram of field sizes in my data set.
+The left panels of figure \ref{data_descriptives} show the histograms of field sizes as measured by estimates of recoverable oil and in-place oil.
 
@@
 These fields tend to have substantially different production profiles than the larger fields.
-However, these fields only account for around 0.5 \% of total oil production so this exclusion is unlikely to have any serious practical implications.
+However, these fields only account for around 0.5 \% of total recoverable reserves so this exclusion is unlikely to have any serious practical implications.
 
@@
 I use yearly data from the US Energy Information Agency on the real price of Brent-traded oil in 2010 dollars.
-The Brent benchmark oil price is likely the best oil price measure for Norwegian production as it is based on light sweet crude oil sourced from the North Sea.
+The Brent benchmark oil price is the best oil price measure for Norwegian production as it is based on light sweet crude oil sourced from the North Sea.
 
@@
 The left-hand-side variable, $Log(production_{i,t})$ is the log of yearly oil production for field $i$ at production time $t$ in millions of standard cubic meters (SM3).
-On the right hand side is a smoothed function of the production profile over time $f(production\_time_{i,t})$; $in\_place\_oil_{i,t}$ is the estimated in place oil of the field, and $cost\_index_{t}$ represents the producer price index for oil and gas extraction.
+On the right hand side is a smoothed function of the production profile over time $f(production\_time_{i,t})$; $in\_place\_oil_{i,t}$ is the estimated in-place oil of the field, and $cost\_index_{t}$ represents the producer price index for oil and gas extraction.
 
@@
 
-Finally, $oil\_price\_lags_t$ represents the variables of interest - a vector of concurrent and lagged terms for the oil price.
-I included eight lagged terms of the oil price and I also experimented with using quadratic terms, however these could not be shown to be significant.
+Finally, $\mathbf{oil\_price\_lags_t}$ represents the variables of interest - a vector of concurrent and lagged terms for the oil price.
+I include eight lagged terms of the oil price and I also experimented with using quadratic terms, however these could not be shown to be significant so I leave them out.
 
@@
 However, instead of setting this arbitrarily, cross-validation is used.
-Intuitively, each data point is left out and the smooth term that provides the best predicted fit for each data point is chosen.
+Intuitively, each data point is, one-by-one, left out and the smooth term that provides the best average predicted fit over all the data is chosen.
 For further details on the cubic regression spline, I refer to \citet{wood_generalized_2006}.
@@
 
-One objection to the simple model is that a fixed effect controlling for the general production profile of oil fields does not sufficiently address idiosyncratic variance between the fields.
+One objection to the simple model is that a fixed effect controlling for the general production profile of oil fields does not sufficiently take into account the idiosyncratic variance between the fields.
 To address this I can add field random effects that take into account uncertainty based on random variation between the fields in the estimation.
@@
 A solution is to allow the shape to vary by the size of the field - here I use an estimate of in-place oil in the field rather than recoverable reserves, as recoverable reserves may be correlated with oil prices.
-Thus the smoothed can be written as a function of both production time and in place oil: $f(production\_time_{i,t}, in\_place\_oil_{i,t})$.
+Thus the smoothed term can be written as a function of both production time and in place oil: $f(production\_time_{i,t}, in\_place\_oil_{i,t})$.
 
@@
 
-$J_{22}$ represents the penalty function for the smoothness of the function which can be written as in \ref{thin_plate_3}.
+$J_{22}$ represents the penalty term for the smoothness of the function, which can be written as in \ref{thin_plate_3}.
 The $22$ represents the fact that it is a penalty function of two variables with smoothness measured by the second derivative.
@@
 	\begin{equation}
-	J_{22}{f}= \diffp[2]{f}{x_1}^2 + \diffp[2]{f}{{x_1}{x_2}} + \diffp[2]{f}{x_2}^2dx_1 dx_2
+	J_{22}(f)= \Big(  \diffp[2]{f}{x_1} \Big)^2 +
+	 \Big(\frac{\partial^2 f}{\partial x_1 \partial x_2}\Big)^2 + 
+	\Big(\diffp[2]{f}{x_2} \Big)^2dx_1 dx_2
 \label{thin_plate_3}
@@
 
-Table \ref{GAM_model_table} shows the parametric coefficient terms and other summary statistics of the regression results.
+Table \ref{GAM_model_table} shows the estimated parametric coefficient terms and other summary statistics of the regression results.
 The first column shows results for the simple model with a single-dimensional smoothed function of the production profile.
@@
 In general the results from the two models agree with each other, indicating little to no effect in the concurrent price term, but with significant coefficients estimated on the 1st to 4th lags, with a peak at the 2nd lag.
-F-tests with comparisons to restricted models without the concurrent and 5-8th lags indicate that these terms do not significantly improve the fit of the models.
+A negative coefficient for the eight lag is estimated in the 2-d model, however this is not significant in the model with random effects.
+Furthermore, F-tests with comparisons to restricted models without the concurrent and 5-8th lags indicate that these terms do not as a group significantly improve the fit of the model.
 
@@
 The points represent the point estimates, while the lines represent approximate 95\% confidence intervals.
-Significant positive coefficients are estimated on lags of between 1-4.
+Significant positive coefficients are estimated on lags of between 1-4 years.
 However, little to no effect is measured on the concurrent price term.}
@@
 However, the magnitude and the ordering of the estimated price effects indicates that investment lags are likely dominant.
-Since the current price always contains the most up-to-date information, then a model of adoptive expectations would show the largest effect at the concurrent term, thereafter receding.
+Since the current price always contains the most up-to-date information, then a model of adaptive expectations would show the largest effect at the concurrent term, thereafter receding.
 The estimations in this article however can show no significant concurrent effect, with the largest magnitude effect consistently being estimated at a 2-year lag.
 
-The findings in this paper can provide some guidance on the effects of the recent steep fall in the oil price on Norwegian production, as well as production from other offshore production areas.
-The fall in total oil production that has continued since the peak in production in 2000 has recently been arrested as several new finds began production after years of increased searching and investment.
-As mentioned, a further significant boost is expected in 2019 when production from the large Sverdrup field begins production.
-The results of this article indicate that the recent fall in prices is unlikely to have major ramifications for production in existing fields.
-Anecdotal evidence however suggests that searching and exploration are being curtailed.
-Anecdotal evidence also suggests that lower oil prices may be leading to earlier shut-down decisions of depleting fields - a hypothesis that could be explored with future research.
+The findings in this paper can provide some guidance on predicting the effects of the recent steep fall in the oil price on Norwegian production, as well as production from other offshore production areas.
+The results suggest that lower oil prices are likely to have only modest ramifications for production in existing fields.
+Anecdotal evidence however suggests that searching and exploration are being substantially curtailed.
+Future oil production could then be significantly affected by way of fewer new field developments.
 

### 2015-07-24 oil_prices_rr.tex
new submissiont o Energy

@@
 \author{Johannes Mauritzen\\
+		Center for Sustainable Energy Studies (CenSES)\\
 		Department of Business and Management Science\\
@@
 
-\thanks{I would like to thank three anonymous referees for comments and suggestions that substantially improved this article.
-I thank Klaus Mohn, Jonas Andersson, Sturla Kvamsdal, Harrison Fell, R\"ognvaldur Hannesson and Henrik Horn for valuable discussion.
-I have received financial support from the Norwegian Center for Sustainable Energy Studies (CenSES)}
+\thanks{}
 % JEL Codes: Q4, L71
@@
 
-For most of the last century, crude oil has been the world's single most important and valuable fuel source.\footnote{Oil is the world's largest single energy source, consisting of approximately 33\% of total energy consumption in 2013 as well as the most valuable in terms of market price per unit of energy \citep{british_petroleum_statistical_2013}} Naturally, questions of how the oil price affects the world economy as well as how oil production reacts to the oil price have been fundamental topics in economics.
+For most of the last century, crude oil has been the world's single most important and valuable fuel source.\footnote{Oil is the world's largest single energy source, consisting of approximately 33\% of total energy consumption in 2013 as well as the most valuable in terms of market price per unit of energy \citep{british_petroleum_statistical_2013}} Naturally, questions of how the oil price affects the world economy as well as how oil production reacts to the oil price have been fundamental topics.
 Recent volatility in the price of oil has made this an especially relevant topic.
@@
 The use of total recoverable reserves is complicated since it is dependent on the price of oil and thus endogenous in the analysis.
-Therefor, as a proxy for the size of the fields, I use estimates of in-place oil from the Norwegian Petroleum Directorate, which is unlikely to be correlated with prices.
+Therefore, as a proxy for the size of the fields, I use estimates of in-place oil from the Norwegian Petroleum Directorate, which is unlikely to be correlated with prices.
 The left panels of figure \ref{data_descriptives} show the histograms of field sizes as measured by estimates of recoverable oil and in-place oil.
@@
 
-The latter term - an estimate of the second derivative of the function - acts as a penalty for the ``wiggiliness'' of the function.
+The latter term - an estimate of the second derivative of the function - acts as a penalty for the ``wiggliness'' of the function.
 The total smoothness can then be controlled by adjusting the $\lambda$ term.
@@
 Unfortunately, estimating a random effects component is not possible with a two-dimensional smoothed function for computational reasons.
-In the presentation of results, I therefor present models side-by-side that use both one-dimensional smooth functions with random effects of the field production profile and two-dimensional smooth functions.
+In the presentation of results, I therefore present models side-by-side that use both one-dimensional smooth functions with random effects of the field production profile and two-dimensional smooth functions.
 
@@
 Using an F-test, adding this variable could not be shown to significantly increase the fit of the model, suggesting that demand-side effects are unlikely to be causing a bias.
-I therefor exclude this variable from the results presented below.
+I therefore exclude this variable from the results presented below.
 
@@
 
+\section{Acknowledgments}
+I thank Klaus Mohn, Jonas Andersson, Sturla Kvamsdal, Harrison Fell, R\"ognvaldur Hannesson and Henrik Horn for valuable discussion.
+I have received financial support from the Norwegian Center for Sustainable Energy Studies (CenSES), Norwegian Research Council grant number 209697
+
 \bibliographystyle{plainnat}

### 2016-02-21 oil_prices_rr_blind.tex
pre-OBES commit

--- 2015-07-24 oil_prices_rr.tex
@@
-\documentclass[11pt]{article}
+%  oil_prices_rr_blind.tex
+\documentclass[12pt]{article}
 \usepackage[margin=1in]{geometry}
@@
 \title{The Effect of Oil Price on Field Production: Evidence from the Norwegian Continental Shelf}
-\author{Johannes Mauritzen\\
-		Center for Sustainable Energy Studies (CenSES)\\
-		Department of Business and Management Science\\
-        NHH Norwegian School of Economics\\
-        Helleveien 30, 5045\\
-        Bergen, Norway\\
-        johannes.mauritzen@nhh.no\\
-        \url{jmaurit.github.io}\\
-		}
+% \author{Johannes Mauritzen\\
+% 		Department of Industrial Economics and Technology Management\\
+%         Norwegian University of Science and Technology\\
+%         7491 Trondheim
+%         Norway\\
+%         johannes.mauritzen@iot.ntnu.no\\
+%         \url{jmaurit.github.io}\\
+% 		}
 \date{\today}
@@
 I use detailed field-level data on Norwegian offshore oil production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production.
-I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 2 to 5\% for a 10 dollar per barrel increase in the real price of oil.\\
-Keywords: Oil production,Oil Prices, Norway, Generalized Additive Model.
+I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 1 to 4\% for a 10 dollar per barrel increase in the real price of oil.\\
+\textbf{Keywords}: Oil production,Oil Prices, Norway, Generalized Additive Model.\\
+\textbf{Words}: 5970
 \end{abstract}
@@
 
-For most of the last century, crude oil has been the world's single most important and valuable fuel source.\footnote{Oil is the world's largest single energy source, consisting of approximately 33\% of total energy consumption in 2013 as well as the most valuable in terms of market price per unit of energy \citep{british_petroleum_statistical_2013}} Naturally, questions of how the oil price affects the world economy as well as how oil production reacts to the oil price have been fundamental topics.
+For most of the last century, crude oil has been the world's single most important and valuable fuel source.\footnote{Oil is the world's largest single energy source, consisting of approximately 33\% of total energy consumption in 2013 as well as the most valuable in terms of market price per unit of energy \citep{british_petroleum_statistical_2013}} Naturally, questions of how the oil price affects the world economy as well as how oil production reacts to the oil price have been fundamental topics in Economics.
 Recent volatility in the price of oil has made this an especially relevant topic.
@@
 
-\section{Acknowledgments}
-I thank Klaus Mohn, Jonas Andersson, Sturla Kvamsdal, Harrison Fell, R\"ognvaldur Hannesson and Henrik Horn for valuable discussion.
-I have received financial support from the Norwegian Center for Sustainable Energy Studies (CenSES), Norwegian Research Council grant number 209697
-
 \bibliographystyle{plainnat}

### 2016-02-21 oil_prices_title.tex
pre-OBES commit

--- 2016-02-21 oil_prices_rr_blind.tex
@@
-%  oil_prices_rr_blind.tex
+% oil_prices_title.tex
+
 \documentclass[12pt]{article}
@@
 
-\doublespacing % from package setspacs
-
 \title{The Effect of Oil Price on Field Production: Evidence from the Norwegian Continental Shelf}
-% \author{Johannes Mauritzen\\
-% 		Department of Industrial Economics and Technology Management\\
-%         Norwegian University of Science and Technology\\
-%         7491 Trondheim
-%         Norway\\
-%         johannes.mauritzen@iot.ntnu.no\\
-%         \url{jmaurit.github.io}\\
-% 		}
+\author{Johannes Mauritzen\\
+		Department of Industrial Economics and Technology Management\\
+        Norwegian University of Science and Technology\\
+        7491 Trondheim
+        Norway\\
+        johannes.mauritzen@iot.ntnu.no\\
+        \url{jmaurit.github.io}\\
+		}
 \date{\today}
@@
 \begin{document}
- \begin{spacing}{1} %sets spacing to single for title page
 	\maketitle
@@
 
-\thanks{}
-% JEL Codes: Q4, L71
- \end{spacing}
-
-\section{Introduction}
-
-For most of the last century, crude oil has been the world's single most important and valuable fuel source.\footnote{Oil is the world's largest single energy source, consisting of approximately 33\% of total energy consumption in 2013 as well as the most valuable in terms of market price per unit of energy \citep{british_petroleum_statistical_2013}} Naturally, questions of how the oil price affects the world economy as well as how oil production reacts to the oil price have been fundamental topics in Economics.
-Recent volatility in the price of oil has made this an especially relevant topic.
-
-However a significant gap exists in the literature.
-While numerous studies have taken up the issue of how the price of oil affects searching for new fields as well as total oil production at both the regional and global level, few studies exist of the effects of oil prices on oil production at the field level.
-The studies that do use field- and well-level data, such as \citet{rao_taxation_2010}, often use data from on-shore installations.
-However an increasing amount of the world's oil production comes from hard-to-reach offshore fields.
-Production from challenging offshore environments is substantially different in character than from on-shore installations.
-
-% Moreover, studies within the economics literature fail to take into account how oil prices have varying effects on the different stages of field development and production.
-The planning, build-out and depletion phase of a field are conceptually unique and how oil companies operating offshore respond to oil prices can be expected to vary substantially.
-
-The effect of oil prices on producing fields is an important topic in understanding the mechanisms of how total oil supply reacts in response to price.
-The response of oil production in existing fields is especially important now as many of the major oil-producing areas, like the Norwegian Continental Shelf are mature and an increasing share of investments will be directed at capacity enhancement in producing fields rather than new field developments.
-How production from these fields will respond to changes in oil prices has major implications for the oil industry, the state finances of oil-producing countries and long-run oil price formation.
-
-The question of the effect of oil prices on production and the more general question of optimal oil extraction has spawned a large theoretical literature dating back to the seminal work of \citet{hotelling_economics_1931}.
-\citet{krautkraemer_nonrenewable_1998} provides a good overview.
-At a basic level a central idea of much of this theoretical work is that with a non-renewable resource, production is a decision that involves a significant opportunity cost: more production in the current period means less production in future periods.
-Within these frameworks, prices and expectations of prices become important variables in the production decision.
-A simple Hotelling model suggests that a producer would immediately change their production in response to a change in oil price in order to dynamically optimize the total extraction.
-
-But in practice, the question is not as clear cut.
-Production in the Norwegian Continental Shelf - as well as most other offshore production areas - has extremely high fixed and operating costs.
-Keeping spare capacity available in order to adjust to changing oil prices might simply be too expensive.
-Instead, producers may find it more beneficial to use storage and financial instruments in order to hedge short-term price movements.
-Higher oil prices may however still lead a producer to invest in increased capacity.
-Since lag times are significant in the offshore sector, we would then expect to see a multi-year lagged effect of prices on production.
-
-However even with the question of lagged production and investment, some ambiguity exists.
-\citet{mohn_efforts_2008} suggests and finds evidence for the idea that in periods of high oil prices offshore producers will invest more in risky wild-cat drilling in search of new fields.
-But when prices are low, they tend to concentrate investments in lower-risk ventures, like expanding production in existing fields.
-If this effect were to dominate, then it may even be plausible that a fall in prices leads to an \emph{increase} in production.
-Though the converse - an increase in prices leading to reduced production - is unlikely.
-
-%Implications for theories and explanations for understanding of the price mechanism and %the role of speculation.  Hamilton block quote.
-
-For such a prominent subject, the lack of research on the role of price in oil field production is striking.
-Two main factors likely contribute to the limited literature - the availability of data and the non-linear time profile of field production.
-Large private oil companies, notably the “super majors” and state-owned oil companies have historically accounted for the vast majority of oil production and reserves.
-\footnote{See for example the economist article titled Supermajordämmerung from August 3rd, 2013: \url{http://www.economist.com/news/briefing/21582522-day-huge-integrated-international-oil-company-drawing}} These entities tend to consider field-level data as either company or state secrets.
-
-However, over the last few years, a movement towards making the petroleum and other extractive industries more transparent has taken form.
-The Norwegian government has been on the forefront of this movement and increasingly committed itself to transparency in the petroleum sector.
-\footnote{see http://www.regjeringen.no/en/sub/eiti---extractive-industries-tranparency/about-eiti.html?id=633586} Over the last several years, detailed data on most aspects of the country's oil industry has become openly available.
-
-In this article, I use historical production data from nearly all 77 currently or formerly oil-producing fields on the Norwegian Continental Shelf in order to estimate the effect that prices have on oil production.
-
-By looking only at the effect of price on fields that currently or previously have produced oil, I am limiting the scope of this article.
-The effect of oil prices on total production over an extended period of time is due not just to reactions in production in existing fields but also increased searching for new fields.
-In fact, an implication of this work is that much of the total production response from higher oil prices is likely from increased searching as well as production from previously un-economic fields.
-
-The main finding in this article is that oil production at the field level has no significant reaction to concurrent changes in the oil price.
-A slight effect can be detected at a lag of between 1 and 4 years, with a magnitude of about a 2 to 5\% increase in yearly production for a 10 dollar increase in the price of oil.
-
-% Price appears to have the most significant affect during the planning and build-out stage.  In the depletion phase of production, price is found to have little to no significant effect.
-
-The main methodological problem is the non-linear production profile of oil fields.
-Once full-scale extraction is started in an oil field, pressure in wells will start declining.
-Compensating measures such as gas and water injection have been successful, but their impact is temporary.
-In turn production rates will therefore inevitably decline.
-
-More so, major oil field discoveries, starting times and in turn production are correlated across fields as the top panel of figure \ref{oil_decline} shows with the production profile of the 10 largest Norwegian oil fields.
-The result is a total production curve that is bell-shaped over time as in the lower panel of figure \ref{oil_decline}.
-Since the oil price series is autocorrelated and non-stationary, failure to properly account for the production profile will lead to spurious estimation of the price terms in a regression.
-
-\begin{figure}
-	\includegraphics[width=1\textwidth]{figures/oil_decline.png}
-	\caption{The top panel shows the production from the largest 10 oil fields, whose starting times are temporally correlated with each other.
-The total production over time is bell-shaped, as shown by the bottom panel.}
-	\label{oil_decline}
-\end{figure}
-
-The direction of this bias can be gleaned in figure \ref{oil_decline}.
-High oil prices were present at periods of relatively low production in the late 1970s and early 1980s as well as over the last 10 years, however real prices reached some of their lowest levels at the same time as the top of production around the year 2000.
-These oil price dynamics almost certainly affected investment decisions in the industry and in turn total production (\citep{osmundsen_is_2007}, \citep{aune_financial_2010}).
-However the inverse contemporaneous relationship at the field level is entirely coincidental, but will heavily bias the estimation of the effect of price on production if field production profiles are not properly accounted for.
-
-As a solution I use a semi-parametric model within the Generalized Additive Model frameworks of \cite{hastie_generalized_1990}.
-Here I use both a cubic regression spline and a two-dimensional thin-plate regression spline to account for the general non-parametric shape of the production profile, while allowing price to enter the equation linearly.
-I also attempt to control for the effects of varying production costs and the effects of technological change.
-
-The results suggest that little to no concurrent effect of oil prices on production exists.
-However I estimate modest effects on the first through fourth lagged terms.
-These results are in line with an understanding of the industry as being highly capital intensive with high operating costs.
-Maintaining spare capacity in order to adjust to changing oil prices is likely prohibitively expensive.
-Instead, production is increased through investment, which may take up to several years to show an effect.
-
-\section{The Effect of Oil Price on Production: Theory, Simulations and Empirics}
-
-For modeling aggregate oil production, shape-fitting models, notably \citet{hubbert_energy_1962} and more recently advocated by \citet{deffeyes_hubberts_2001}, have had some success in estimating the timing of peak production at the regional and national level, but they tend to seriously underestimate the total recoverable resource of oil-producing regions and the models can be shown to be fundamentally misspecified \citep{boyce_prediction_2013}.
-Simulation type studies where aggregate oil production is modeled through an often complex combination of physical and economic processes also exist in the geo-engineering literature, but their usefulness tends to be weighed down by their complexity as they require quite detailed data and specific assumptions about functional form that can be difficult to justify \citep{brandt_review_2010}.
-
-More important to this paper are empirical estimates of the effect of price on production.
-Several econometric papers seek to answer the question of how aggregated oil supply is affected by oil prices.
-\citet{farzin_impact_2001} attempts to estimate an elasticity for the effect on added reserves of increased oil prices and finds a small though statistically significant effect.
-\citet{ramcharran_oil_2002} estimates a supply function for the total supply of oil from several OPEC countries based on data from 1973 to 1997.
-The author finds a negative price elasticity for several of the countries, and interprets this as evidence of producers targeting revenue.
-However since the author does not take into account the production profile of oil fields and the spurious correlation that can arise with autocorrelated prices, these estimates come under considerable doubt.
-
-the effect of oil-price uncertainty on drilling and exploration has also been explored.
-\citet{kellogg_effect_2014} finds that oil exploration firms in Texas do approximately respond as real-option theory would predict when it comes to the timing of drilling.
-A model and test using data from North Sea producers on the UK continental shelf by \citet{hurn_geology_1994}, on the other hand, fails to find evidence that the variance in the oil price affects the timing of oil field development.
-
-I do not attempt to directly model uncertainty, however given that the investments needed to increase oil production in an existing field are to a certain extent irreversible and that oil prices are highly volatile, the results can and probably should be interpreted with the real options framework in mind.
-
-Only a few papers utilize field-level data.
-\citet{black_is_1998} tests the relevance of “nesting” a structural empirical model of profit maximization that takes into account oil prices into a typical geo-engineering model of oil field production.
-They find strong evidence that taking into account profit maximization - and implicitly price - substantially improves the fit compared to a purely geo-engineering model.
-The limitation of their methodology is that they are only able to test whether including economic factors like price affects the fit of the model but are not able to give an estimate of the magnitude of the effect.
-Methodologically, the paper also relies heavily on assumptions about the functional form of both the geo-engineering aspects of the oil producer as well as their profit-maximization.
-By taking a more flexible, semi-parametric approach to modeling oil field production profiles, this article avoids problems with overly restrictive assumptions.
-
-\citet{rao_taxation_2010} uses data on onshore oil wells in California to estimate the effects of tax changes and price controls.
-The author finds that short-term tax changes caused small, but significant re-timing of production from oil wells.
-These findings are over-all consistent with the findings of this article, though the author finds a concurrent effect of changes in taxation on production while I do not find any concurrent effect of prices on production.
-The significant differences in cost and complexity between operating offshore and onshore likely account for the different results.
-How producers react to a short-term tax change as opposed to a change in price also likely differs.
-
-Studies using detailed Norwegian data on offshore activity also exist, though the focus has mainly been on exploration and drilling.
-\citet{mohn_exploration_2008} finds that long-term changes in the oil-price has a strong effect on exploratory drilling though little effect is measured from short-term changes in the oil price.
-\citet{osmundsen_exploration_2010} analyses drilling productivity over time on the Norwegian Continental Shelf while \citet{mohn_efforts_2008} finds that higher oil prices leads to higher reserves as well as that oil prices affect producer risk-preferences - with higher prices leading to lower success rates but larger discovery size.
-
-\section{Oil Production on the Norwegian Continental Shelf}
-
-The first commercial oil well in Norwegian continental waters was discovered in December of 1969 in what became the Ekofisk oil field.
-Most of the largest fields in the North Sea were found relatively early on while more recent finds have tended to be smaller - a pattern typical of oil producing areas called creaming.
-A major exception to this trend was the recent find of the Johan Sverdrup field which is estimated to have approximately 300 million Standard Cubic Meters (SM3) of recoverable oil.
-\footnote{The Johan Sverdrup field is estimated to begin producing oil in 2019 and so is not present in the data set used for the analysis.}
-
-Exploration in the Norwegian Sea was opened in the early 1980’s and the first commercial field started production in 1981.
-While several mid-sized fields have been discovered, the Norwegian Sea has generally disappointed in terms of commercial oil finds.
-
-Norwegian waters in the Barents Sea have also been open to exploration since the 1980s, however up until recently only a few, small finds were made and none came into commercial production.
-Recently several significant oil and gas finds have been made in the Barents Sea - notably the oil field Goliat and the large gas find Sn\o hvit, which are both currently under development but not yet producing.
-The agreement between Norway and Russia in June 2011 settling a long-running dispute over the maritime delimitation has also given a boost to new exploration in the region.
-
-Profits from oil and gas production in Norway are subject to a resource tax of 51 \% on top of the ordinary income tax of 27 \%, thus income from petroleum production is taxed at a total marginal rate of 78 \%.
-The central government also receives revenues through ownership stakes in companies, notably Statoil, where the state is the majority stakeholder.
-
-The over-all tax rate has been fairly constant through the history of Norwegian oil production, however several important changes related to the tax code have occurred.
-In 1991 a C02 tax was introduced, and in the year 2012 the tax was doubled.
-But this was mainly levied on the use and import of petroleum products.
-In the offshore sector it was levied on the burning of oil and gas and thus the main effect was on the practice of flaring natural gas that could not be transported and sold commercially.
-
-More important to the offshore sector were accounting changes that were implemented in 2002 and 2005, which were meant to encourage new entrants by allowing losses to be carried forward for tax purposes and by introducing a rebate on the tax-value of losses associated with searching and drilling.
-In general though, these rules mainly affected searching and discoveries of new fields rather than production from existing fields and I do not control for the tax changes in my model.
-More information on taxation and revenues from the offshore sector can be found at the website of the Norwegian Ministry of Finance.
-\footnote{\url{http://www.regjeringen.no/en/dep/fin/Selected-topics/taxes-and-duties/bedriftsbeskatning/Taxation-of-petroleum-activities.html?id=417318}}
-
-Rights to explore and eventually produce on the Norwegian Continental Shelf are based on a system where the government announces geographic blocks that will be opened to oil exploration and production subject to production licenses.
-Production licenses are initially granted for between 4-6 years subject to requirements that firms are actively searching in their awarded blocks.
-If oil or gas deposits are proven then the production license can be extended for up to 30 years.
-In general, the frameworks are fair, predictable and stable for companies who find commercially extractable oil deposits and regulatory interference is unlikely to be the cause of any observed changes in production from existing oil fields.
-For more information see the website of the Norwegian Petroleum Directorate.
-\footnote{\url{http://www.npd.no/en/Topics/Production-licences/Theme-articles/Production-licence--licence-to-explore-discover-and-produce-/}}
-
-\section{Data}
-Norwegian oil field production data is obtained from the website of the Norwegian Petroleum Directorate.\footnote{http://factpages.npd.no/} The data I use includes nearly all current or formerly oil-producing fields, from 1971 through 2014.
-Production data is available at a monthly frequency, though I choose to aggregate up to yearly values both to smooth over seasonality as well as short-term volatility of output due to factors such as weather or technical issues that are not relevant for this article.
-
-In addition to data on field-level production, I also have access to data on estimates of total recoverable reserves as well as estimates of total in-place oil for each field.
-The use of total recoverable reserves is complicated since it is dependent on the price of oil and thus endogenous in the analysis.
-Therefore, as a proxy for the size of the fields, I use estimates of in-place oil from the Norwegian Petroleum Directorate, which is unlikely to be correlated with prices.
-The left panels of figure \ref{data_descriptives} show the histograms of field sizes as measured by estimates of recoverable oil and in-place oil.
-
-\begin{figure}
-	\includegraphics[width=1\textwidth]{figures/data_descriptives.png}
-	\caption{Data descriptives. The left panels show histograms of estimates of recoverable oil and in-place oil by field. The right panels shows the production cost index and real Brent oil prices.}
-	\label{data_descriptives}
-\end{figure}
-
-To simplify the analysis, I eliminate the smallest fields - those with less than 3 million SM3 of recoverable reserves.
-These fields tend to have substantially different production profiles than the larger fields.
-However, these fields only account for around 0.5 \% of total recoverable reserves so this exclusion is unlikely to have any serious practical implications.
-
-I also eliminate the large Ekofisk field from my analysis.
-I do this because it has a unique double-humped production profile.
-Ekofisk was the first field to be found and the first to start production on the Norwegian Continental Shelf.
-It reached an initial peak in production in the late 1970's and began depleting.
-However, the introduction of water and gas injection technology led to a boost in production and a new peak in the early 2000's.
-All subsequent major fields used gas and water injection from the outset and have a single-hump profile.
-
-I use yearly data from the US Energy Information Agency on the real price of Brent-traded oil in 2010 dollars.
-The Brent benchmark oil price is the best oil price measure for Norwegian production as it is based on light sweet crude oil sourced from the North Sea.
-
-Production costs are known to covary with the price of oil.
-To control for this, I include the producer price index for the Norwegian oil and gas extraction industry, obtained from Statistics Norway.
-\footnote{\url{https://www.ssb.no/en/priser-og-prisindekser/statistikker/ppi}} Unfortunately, this series only goes back to 2000, so I also include an index of oil and gas well costs from the US energy information agency for the years 1960 to 2000.
-\footnote{\url{http://www.eia.gov/dnav/pet/pet_crd_wellcost_s1_a.htm}} While this measure is unlikely to be a perfect substitute for costs in Norway, oil and gas extraction is a global industry, where costs are highly correlated across regions.
-The combined production cost index and the real Brent oil price series are shown in the right panels in figure \ref{data_descriptives}.
-
-An argument can be made that expectations of future oil prices can be equally if not more important for production decisions as the current oil price.
-Forecasts for future oil prices are available from, among others, the International Energy Agency, but these have tended to be notoriously inaccurate and it is unlikely oil companies use these projections for their investment decisions.
-
-On the other hand, given the size and liquidity of oil spot markets, it is a fair assumption that the current oil prices do a good job of incorporating much of the available information about crude oil markets and that future price movements are generally difficult to predict \citep{hamilton_understanding_2008}.
-
-An active futures market does exist, but several studies have found that current oil prices are in general better than prices on futures contracts at predicting future oil prices \citep{alquist_what_2010, chinn_predictive_2005}.
-\citet{mohn_investment_2008} as well as \citet{pesaran_econometric_1990} and \citet{farzin_impact_2001} find evidence for adaptive expectations, where expectations of future prices is based on a weighted average of current and past prices.
-This is implicitly taken into account by the price lags in my regression equations, however this also leads to some unavoidable ambiguity in interpreting the results, which I discuss in the conclusion.
-
-A cleaned data set and the full code for the analysis are available at \url{jmaurit.github.io/#oil_prices}.
-
-\section{A Generalized Additive Model of Oil Field Production}
-Parametric linear models have the sizeable advantages of simplicity and interpretability and are therefore usually a good starting point for an analysis.
-However, when attempting to model the effect of price on oil field production, a standard linear model is unable to sufficiently control for the production profile and therefore heavily biases the estimate of the effect of price.
-
-Instead of attempting to estimate the shape of the production profiles of the fields by estimating parameters on linear terms, I estimate a non-parametric function for the production profile.
-At a conceptual level, this model is decomposing the variation in the data into a smoothed component representing the technical and physical forces that produce the typical bell-shaped production profile, and the remaining component, that I estimate the effect of price with.
-
-I do not, however, want to estimate smoothed curves individually for each field.
-While this would provide a good overall fit to the full data set, not enough variation in the data would be left to estimate the effect of price.
-Instead I want to estimate a general shape of the production profile for all fields and then use the remaining variation in the data to estimate the effect of price.
-
-A simple model can be written as in equation \ref{gam_price_eqn}.
-
-\begin{equation}
-\begin{split}
-	Log(production_{i,t}) & = f(production\_time_{i,t}) + f(year_t) + \beta_1 in\_place\_oil_{i,t}\\
-	 \quad & + \beta_3 cost\_index_{t} + \mathbf{\beta_2 oil\_price\_lags_t} +  
-	  \epsilon_{i,t}
-\label{gam_price_eqn}
-\end{split}
-\end{equation}
-
-In this model I am estimating the parameters and functions from all fields $i$.
-The left-hand-side variable, $Log(production_{i,t})$ is the log of yearly oil production for field $i$ at production time $t$ in millions of standard cubic meters (SM3).
-On the right hand side is a smoothed function of the production profile over time $f(production\_time_{i,t})$; $in\_place\_oil_{i,t}$ is the estimated in-place oil of the field, and $cost\_index_{t}$ represents the producer price index for oil and gas extraction.
-
-$f(year_t)$ represents a restricted smooth function of the calendar time.  The main purpose of including this term is to control for the effects of technological change, which I discuss further below.
-
-Finally, $\mathbf{oil\_price\_lags_t}$ represents the variables of interest - a vector of concurrent and lagged terms for the oil price.
-I include eight lagged terms of the oil price and I also experimented with using quadratic terms, however these could not be shown to be significant so I leave them out.
-
-The smoothed function is estimated using a cubic regression spline.
-Here a cubic polynomial function is used to fit the shape in sections, separated at points known as knots, but continuous up to the second derivative.
-The major advantage of this method, is that it can be represented in linear form, $\boldsymbol{X \beta} $, and thus standard, efficient matrix algebra techniques can be used.
-
-The smoothed component of the model is then fit by minimizing equation \ref{min_equation}.
-
-\begin{equation}
-\| \mathbf{y} - \mathbf{X\beta} \| ^2 + \lambda \int_{0}^{1} [f^{\prime \prime}(x)]^2 dx
-\label{min_equation}
-\end{equation}
-
-The latter term - an estimate of the second derivative of the function - acts as a penalty for the ``wiggliness'' of the function.
-The total smoothness can then be controlled by adjusting the $\lambda$ term.
-However, instead of setting this arbitrarily, cross-validation is used.
-Intuitively, each data point is, one-by-one, left out and the smooth term that provides the best average predicted fit over all the data is chosen.
-For further details on the cubic regression spline, I refer to \citet{wood_generalized_2006}.
-
-The advantage of this simple model that uses a single-variable cubic regression spline is that the smoothed terms can be interpreted directly, and thus we can do a visual check of the appropriateness of the smoothed function.
-The top panel of figure \ref{smooths} shows the estimated smoothed function of the field production profile, where all other co-variates are held at their average value.
-The function appears reasonable, showing the expected pattern of an initial build-out period followed by an extended depletion phase.
-
-In the second panel, the smoothed function of calendar time, $year_t$, is shown.
-Here a cubic regression spline is also used, however the number of knots is restricted to four in order to avoid over-fitting - capturing gradual trends over time like technological change, but not the variation caused by volatile oil prices.
-The function increases strongly in the early period, likely reflecting the effects of learning and technological change in a new industry in Norwegian waters, while flattening out in the latter years.
-
-Extended periods of high or low oil prices will likely have some long-term effects on technology and learning, which in turn may be picked up in the smoothed function here, rather than in the parametric price terms.
-I discuss this further in the conclusion.
-
-\begin{figure}
-	\includegraphics[width=1\textwidth]{figures/smooths.png}
-	\caption{The estimated smoothed functions over production time and calendar time.}
-	\label{smooths}
-\end{figure}
-
-One objection to the simple model is that a fixed effect controlling for the general production profile of oil fields does not sufficiently take into account the idiosyncratic variance between the fields.
-To address this I can add field random effects that take into account uncertainty based on random variation between the fields in the estimation.
-The smoothed function can then be written as $f(production\_time_{i,t}*Z)$.
-Where $Z \sim N(0, \sigma)$ and $\sigma$ is the variance term to be estimated.
-
-Another potential issue may be that the smoothed function of production time may not be sufficient in controlling for the variation in production profiles shapes across fields.
-In particular, production profiles differ substantially by the size of the field.
-A solution is to allow the shape to vary by the size of the field - here I use an estimate of in-place oil in the field rather than recoverable reserves, as recoverable reserves may be correlated with oil prices.
-Thus the smoothed term can be written as a function of both production time and in place oil: $f(production\_time_{i,t}, in\_place\_oil_{i,t})$.
-
-Estimating a smoothed function of two variables requires a somewhat more involved method.
-In particular, a thin-plate regression spline \citep{wood_thin_2003} is used.
-Consider equation \ref{thin_plate_1}.
-
-	\begin{equation}
-	y_i = g(x_1, x_2)
-	\label{thin_plate_1}
-	\end{equation}
-
-Following the notation of \citet{wood_generalized_2006}, $g$ is the function of $x_1$ and $x_2$ that is to be estimated by $f$, which in turn is estimated by minimizing equation \ref{thin_plate_2}.
-Here $\boldsymbol{y}$ represents a vector of $y_i$'s and $\boldsymbol{f} = (f(\boldsymbol{x_1}),f(\boldsymbol{x_2}))^t$.
-
-	\begin{equation}
-\min \|\boldsymbol{y-f}\|^2 + \lambda J_{22}(f)
-\label{thin_plate_2}
-	\end{equation}
-
-$J_{22}$ represents the penalty term for the smoothness of the function, which can be written as in \ref{thin_plate_3}.
-The $22$ represents the fact that it is a penalty function of two variables with smoothness measured by the second derivative.
-
-	\begin{equation}
-	J_{22}(f)= \Big(  \diffp[2]{f}{x_1} \Big)^2 +
-	 \Big(\frac{\partial^2 f}{\partial x_1 \partial x_2}\Big)^2 + 
-	\Big(\diffp[2]{f}{x_2} \Big)^2dx_1 dx_2
-\label{thin_plate_3}
-	\end{equation}
-
-In short, a function of $x_1$ and $x_2$ is found that is minimizing errors in the sense of minimizing Euclidean distance subject to a penalty function of “wiggiliness”.
-The actual implementation is somewhat more involved in order to increase the computational efficiency of the estimation.
-For further technical details I refer to \citet{wood_thin_2003}.
-
-Unfortunately, estimating a random effects component is not possible with a two-dimensional smoothed function for computational reasons.
-In the presentation of results, I therefore present models side-by-side that use both one-dimensional smooth functions with random effects of the field production profile and two-dimensional smooth functions.
-
-In general I assume that oil prices are exogenous in this model.
-On the supply side, oil is a globally traded commodity.
-Since I am measuring changes in production at the field level, these are highly unlikely to have any effect on global oil prices.
-
-A more subtle possibility for endogeneity exists on the demand side.
-A bias could result if demand also was correlated with changes in field production other than through the direct price-mechanism.
-For example, higher GDP could mean easier lending to petroleum companies, who then might invest and increase petroleum production independent of higher oil prices.
-I investigated this by including a measure of world GDP growth obtained from the World Bank in the regression.
-Using an F-test, adding this variable could not be shown to significantly increase the fit of the model, suggesting that demand-side effects are unlikely to be causing a bias.
-I therefore exclude this variable from the results presented below.
-
-\section{The Effect of Oil Price on Field Production}
-
-Table \ref{GAM_model_table} shows the estimated parametric coefficient terms and other summary statistics of the regression results.
-The first column shows results for the simple model with a single-dimensional smoothed function of the production profile.
-The second column shows results with field-level random effects, while the third columns shows results when a two-dimensional smoothed function is used to control for production profiles across field sizes.
-The latter two models are my preferred models, providing a substantially better fit, as measured by both deviance explained and $R^2$.
-Monte Carlo experiments, which I describe further below, also indicate that the latter two models provide consistently less-biased parametric estimates of the price terms.
-This is especially the case for the model with a 2-dimensional smoothed term.
-
-The variables of interest in this article is the oil price and its lags, which I include as eight linear parametric terms in the model.
-The idea of including both a concurrent oil price term as well as eight lags is that a change in price could conceivably have two effects on oil production in a field.
-First, the field operator could be operating on the basis of some short-term extraction rule - choosing to pump out less at times of lower prices in order to pump out more at periods of high prices.
-
-Alternatively, a change in price of oil can be seen as a lifting of a production constraint.
-A higher oil price means that added investments in production become attractive in order to either increase the total amount extracted from a field or to shift production forward.
-However investments in the offshore sector can be complex and lengthy, and any production lift would be expected to happen with a lag.
-As mentioned earlier, including several lags also allows for the possibility of adaptive expectations of future oil prices.
-
-\begin{table}
-\begin{center}
-\begin{tabular}{l c c c }
-\hline
-                                             & w/out Rand. Effects & w Rand. Effects & 2-d Smooth \\
-\hline
-(Intercept)                                  & $0.335^{*}$    & $-4.017^{***}$ & $-574.287^{***}$ \\
-                                             & $(0.146)$      & $(0.183)$      & $(51.125)$       \\
-in\_place\_oil\_mill\_100sm3             & $0.337^{***}$  & $-0.014$       &                  \\
-                                             & $(0.005)$      & $(0.014)$      &                  \\
-cost\_index/10                            & $-0.018^{***}$ & $-0.030^{***}$ & $-0.014^{***}$   \\
-                                             & $(0.003)$      & $(0.003)$      & $(0.002)$        \\
-price/10                                  & $-0.006$       & $0.014$        & $0.007$          \\
-                                             & $(0.009)$      & $(0.010)$      & $(0.006)$        \\
-price\_l1/10                              & $0.004$        & $0.021^{*}$    & $0.015^{*}$      \\
-                                             & $(0.010)$      & $(0.010)$      & $(0.006)$        \\
-price\_l2/10                              & $0.037^{***}$  & $0.047^{***}$  & $0.048^{***}$    \\
-                                             & $(0.009)$      & $(0.010$      & $(0.006)$        \\
-price\_l3/10                              & $0.030^{***}$  & $0.025^{**}$   & $0.033^{***}$    \\
-                                             & $(0.009)$      & $(0.009)$      & $(0.006)$        \\
-price\_l4/10                              & $0.003$        & $0.043^{***}$  & $0.028^{***}$    \\
-                                             & $(0.008)$      & $(0.008)$      & $(0.005)$        \\
-price\_l5/10                              & $-0.020^{**}$  & $0.012$        & $-0.001$         \\
-                                             & $(0.007)$      & $(0.007)$      & $(0.005)$        \\
-price\_l6/10                              & $-0.010$       & $0.008$        & $-0.001$         \\
-                                             & $(0.007)$      & $(0.007)$      & $(0.004)$        \\
-price\_l7/10                              & $-0.008$       & $0.007$        & $0.004$          \\
-                                             & $(0.007)$      & $(0.007)$      & $(0.004)$        \\
-price\_l8/10                              & $-0.044^{***}$ & $-0.011$       & $-0.027^{***}$   \\
-                                             & $(0.006)$      & $(0.007)$      & $(0.004)$        \\
-EDF: s(prod\_year)                           & $6.381^{***}$  &                &                  \\
-EDF: s(year)                                 & $2.768^{***}$  & $3.000^{***}$  & $3.000^{***}$    \\
-
-EDF: s(prod\_year,re\_field)                      &                & $33.002^{***}$ &                  \\
-
-EDF: s(prod\_year,in\_place\_oil\_mill\_sm3) &                &                & $27.445^{***}$   \\
-
-\hline
-Deviance                                     & 23062.678      & 19765.833      & 9057.204         \\
-Deviance explained                           & 0.848          & 0.870          & 0.940            \\
-Dispersion                                   & 5.410          & 5.965          & 2.325            \\
-R$^2$                                        & 0.841          & 0.865          & 0.938            \\
-GCV score                                    & 18.145         & 16.218         & 7.355            \\
-Num. obs.                                    & 1088           & 1088           & 1088             \\
-Num. smooth terms                            & 2              & 2              & 2                \\
-\hline
-\multicolumn{4}{l}{\scriptsize{$^{***}p<0.001$, $^{**}p<0.01$, $^*p<0.05$}}
-\end{tabular}
-\caption{Model results.
-Estimated standard coefficients are shown for the linear terms.
-Standard errors are in parenthesis.
-For the smoothed terms, estimated degrees of freedom are displayed, where the p-values are from F-tests of whether the terms significantly improve the fit of the model.}
-\label{GAM_model_table}
-\end{center}
-\end{table}
-
-For convenience the estimated coefficients on the price terms from the two preferred models are shown in figure \ref{price_coefficients}.
-In general the results from the two models agree with each other, indicating little to no effect in the concurrent price term, but with significant coefficients estimated on the 1st to 4th lags, with a peak at the 2nd lag.
-A negative coefficient for the eight lag is estimated in the 2-d model, however this is not significant in the model with random effects.
-Furthermore, F-tests with comparisons to restricted models without the concurrent and 5-8th lags indicate that these terms do not as a group significantly improve the fit of the model.
-
-\begin{figure}
-	\includegraphics[width=1\textwidth]{figures/price_coefficents.png}
-	\caption{Estimated coefficients from two GAM models: a model with random field effects and single smoothed function of production time, and another with a two dimensional smoothed function of production time and field size.
-The points represent the point estimates, while the lines represent approximate 95\% confidence intervals.
-Significant positive coefficients are estimated on lags of between 1-4 years.
-However, little to no effect is measured on the concurrent price term.}
-	\label{price_coefficients}
-\end{figure}
-
-For computationally intensive Generalized Additive Models, few if any analytic results are available to confirm that the estimates are unbiased.
-Instead, I rely on Monte Carlo experiments to test for bias.
-Here, I create artificial field data using a cumulative logistic function with a linear price term and random variation.
-I then estimate the effect of price repeatedly, generating the random component before each new estimation.
-Details of the Monte Carlo experiment can be found in the on-line appendix \url{https://github.com/jmaurit/oil_prices/blob/master/mc_sim.pdf}.
-
-The main results are shown in figure \ref{mc_results}.
-The top panel shows one run of the artificially generated data.
-I compare estimations of a concurrent price term, $beta\_hat$ made with a Generalized Linear Model and Generalized Additive Model with a 2-dimensional smooth function.
-I set the true parameter $beta$ to both $0$ and $0.05$, which are shown as vertical lines in the figures.
-The figure shows that while a bias exists in the estimation of the GLM model, especially for when the true value is $0.05$, the Generalized Additive Model generally provides unbiased results.
-
-\begin{figure}
-	\includegraphics[width=1\textwidth]{figures/mc_plot.png}
-	\caption{Results from a Monte-Carlo experiment.
-I generate data mimicking oil price production from a set of fields, as displayed in the top frame.
-Using GLM and GAM models, and regenerating the error component in the data, I replicate estimates of the effect of price on production, which is set at $\beta = 0$ and $\beta=0.05$.
-The density of the results for the GLM model are shown in the lower left panel, where a clear and large bias is shown relative to the true $\beta$ which are represented by the vertical lines.
-A GAM model however provides estimates close to the true $\beta$}
-	\label{mc_results}
-\end{figure}
-
-\section{Conclusion and Discussion}
-
-The main result of this research is to show that production in existing Norwegian offshore fields has no significant concurrent reaction to higher oil prices while a slight effect is estimated  with a lag of between 1 and 4 years.
-Oil producers do not appear to be behaving strategically in relation to short-term production - increasing or reducing production in response to changes in oil price.
-Instead they are likely using storage or financial instruments to hedge short-term price movements.
-Changes in oil prices can rather be seen as a relaxing of a production constraint, justifying increased investment that leads to either a higher total extraction rate or an inter-temporal shifting of production.
-
-One important caveat is that I do not claim to capture all possible mechanisms for price to influence production.
-Mechanisms that are likely gradual and smooth in nature will be absorbed by the smooth functions in the regression.
-In particular, price-induced technological change will likely be a smooth process with significant lag times.
-Referring back to figure \ref{smooths}, some evidence of this can perhaps be seen in the estimated smooth function of the calendar year variable, which shows a slight (and statistically insignificant) upwards slope after approximately 2005, corresponding to a preceding period of rising oil prices.
-
-Another important caveat, is that the changes in production could potentially either be a permanent increase in production, or a shift forward of production that may otherwise have happened later.
-Unfortunately, with the existing data and methods, I do not see a good way of distinguishing between these alternatives.
-First, there is the problem of censored data, as most of the fields in the data set are still producing, thus the information on how long the fields eventually produce are not available.
-More so, the estimates of both recoverable reserves and in-place oil are too uncertain in order to to make calculations of whether a change in production in one period led to a reduction in another - especially when the changes in production are relatively modest.
-
-Interpreting the lagged effects also involves some subtlety.
-As noted, evidence has been found for adaptive expectations of future oil prices.
-Thus, the lagged effects could be interpreted as both the result of investment lags as well as lags involved in updating of expectations.
-Completely disentangling these effects would probably at a minimum require a structural model with some specific assumptions about the formation of expectations, which is beyond the scope of this article.
-
-However, the magnitude and the ordering of the estimated price effects indicates that investment lags are likely dominant.
-Since the current price always contains the most up-to-date information, then a model of adaptive expectations would show the largest effect at the concurrent term, thereafter receding.
-The estimations in this article however can show no significant concurrent effect, with the largest magnitude effect consistently being estimated at a 2-year lag.
-
-The findings in this paper can provide some guidance on predicting the effects of the recent steep fall in the oil price on Norwegian production, as well as production from other offshore production areas.
-The results suggest that lower oil prices are likely to have only modest ramifications for production in existing fields.
-Anecdotal evidence however suggests that searching and exploration are being substantially curtailed.
-Future oil production could then be significantly affected by way of fewer new field developments.
-
-Finally, relating back to the literature on total oil supply, the modest estimated effect of prices on production adds weight to the argument of \citet{hamilton_oil_2012} that most of the increased supply of oil that comes from higher prices is from expanding the geographic and technological boundaries of oil production.
-For example exploration of deep-water oil deposits off the coast of Brazil and extraction of oil sands in western Canada.
-
-\FloatBarrier
-\section{Software}
-I use the R statistical programming package for all the analysis in this article \citep{r_core_team_r:_2013}.
-I use the R package ggplot2 for plotting \citep{wickham_ggplot2:_2009}, plyr for data manipulation and cleaning \citep{wickham_split-apply-combine_2011}, texreg for table formatting \citep{leifeld_texreg:_2013} and mgcv for implementation of the Generalized Additive Models \citep{wood_fast_2011}.
-
-\bibliographystyle{plainnat}
-\bibliography{oil_prices}
-
 \end{document}

### 2016-02-21 rr_clean.py

@@ -17,8 +17,7 @@ brentprices=brentprices[2:]
 brentprices["date"]=pd.to_datetime(brentprices["date"])
 brentprices=brentprices.set_index("date")
 brentprices["oil_price_nom"]=[float(i) for i in brentprices["oil_price_nom"]]
-brentprices["year"]=brentprices.index.year
-brent_yearly=brentprices.groupby("year").mean()
+oil
 
 
 #from bp via Quandl

### 2016-02-26 oil_prices_obes.tex
added build-out interaction effects to the model

--- 2016-02-21 oil_prices_rr_blind.tex
@@
-%  oil_prices_rr_blind.tex
 \documentclass[12pt]{article}
@@
 \title{The Effect of Oil Price on Field Production: Evidence from the Norwegian Continental Shelf}
-% \author{Johannes Mauritzen\\
-% 		Department of Industrial Economics and Technology Management\\
-%         Norwegian University of Science and Technology\\
-%         7491 Trondheim
-%         Norway\\
-%         johannes.mauritzen@iot.ntnu.no\\
-%         \url{jmaurit.github.io}\\
-% 		}
+\author{Johannes Mauritzen\\
+		Department of Economics\\
+        BI Norwegian Business School\\
+        Trondheim Campus\\
+		Trondheim, Norway\\
+        johannes.mauritzen@bi.no\\
+        \url{jmaurit.github.io}\\
+		}
 \date{\today}
@@
 I use detailed field-level data on Norwegian offshore oil production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production.
-I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 1 to 4\% for a 10 dollar per barrel increase in the real price of oil.\\
-\textbf{Keywords}: Oil production,Oil Prices, Norway, Generalized Additive Model.\\
-\textbf{Words}: 5970
+I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 2 to 7\% for a 10 dollar per barrel increase in the real price of oil.\\
+Keywords: Oil production,Oil Prices, Norway, Generalized Additive Model.
+Words: 5970
 \end{abstract}
@@
 \begin{table}
+\scriptsize
 \begin{center}
-\begin{tabular}{l c c c }
+\begin{tabular}{l c c c c }
 \hline
-                                             & w/out Rand. Effects & w Rand. Effects & 2-d Smooth \\
+ & w/out Rand. Effects & w Rand. Effects & 2-d Smooth & w Build-out \\
 \hline
-(Intercept)                                  & $0.335^{*}$    & $-4.017^{***}$ & $-574.287^{***}$ \\
-                                             & $(0.146)$      & $(0.183)$      & $(51.125)$       \\
-in\_place\_oil\_mill\_100sm3             & $0.337^{***}$  & $-0.014$       &                  \\
-                                             & $(0.005)$      & $(0.014)$      &                  \\
-cost\_index/10                            & $-0.018^{***}$ & $-0.030^{***}$ & $-0.014^{***}$   \\
-                                             & $(0.003)$      & $(0.003)$      & $(0.002)$        \\
-price/10                                  & $-0.006$       & $0.014$        & $0.007$          \\
-                                             & $(0.009)$      & $(0.010)$      & $(0.006)$        \\
-price\_l1/10                              & $0.004$        & $0.021^{*}$    & $0.015^{*}$      \\
-                                             & $(0.010)$      & $(0.010)$      & $(0.006)$        \\
-price\_l2/10                              & $0.037^{***}$  & $0.047^{***}$  & $0.048^{***}$    \\
-                                             & $(0.009)$      & $(0.010$      & $(0.006)$        \\
-price\_l3/10                              & $0.030^{***}$  & $0.025^{**}$   & $0.033^{***}$    \\
-                                             & $(0.009)$      & $(0.009)$      & $(0.006)$        \\
-price\_l4/10                              & $0.003$        & $0.043^{***}$  & $0.028^{***}$    \\
-                                             & $(0.008)$      & $(0.008)$      & $(0.005)$        \\
-price\_l5/10                              & $-0.020^{**}$  & $0.012$        & $-0.001$         \\
-                                             & $(0.007)$      & $(0.007)$      & $(0.005)$        \\
-price\_l6/10                              & $-0.010$       & $0.008$        & $-0.001$         \\
-                                             & $(0.007)$      & $(0.007)$      & $(0.004)$        \\
-price\_l7/10                              & $-0.008$       & $0.007$        & $0.004$          \\
-                                             & $(0.007)$      & $(0.007)$      & $(0.004)$        \\
-price\_l8/10                              & $-0.044^{***}$ & $-0.011$       & $-0.027^{***}$   \\
-                                             & $(0.006)$      & $(0.007)$      & $(0.004)$        \\
-EDF: s(prod\_year)                           & $6.381^{***}$  &                &                  \\
-EDF: s(year)                                 & $2.768^{***}$  & $3.000^{***}$  & $3.000^{***}$    \\
-
-EDF: s(prod\_year,re\_field)                      &                & $33.002^{***}$ &                  \\
-
-EDF: s(prod\_year,in\_place\_oil\_mill\_sm3) &                &                & $27.445^{***}$   \\
-
+Intercept                                  & $0.335$        & $-4.016^{***}$ & $-574.287^{***}$ & $-3.858^{***}$ \\
+                                             & $(0.265)$      & $(0.297)$      & $(89.476)$       & $(0.239)$      \\
+in\_place\_oil             & $0.337^{***}$  & $-0.014$       &                  & $-0.015$       \\
+                                             & $(0.009)$      & $(0.022)$      &                  & $(0.024)$      \\
+cost\_index                            & $-0.018^{**}$  & $-0.030^{***}$ & $-0.014^{***}$   & $-0.034^{***}$ \\
+                                             & $(0.006)$      & $(0.006)$      & $(0.004)$        & $(0.006)$      \\
+price                                  & $-0.006$       & $0.014$        & $0.007$          & $0.005$        \\
+                                             & $(0.017)$      & $(0.016)$      & $(0.011)$        & $(0.015)$      \\
+price\_l1                              & $0.004$        & $0.021$        & $0.015$          & $-0.003$       \\
+                                             & $(0.017)$      & $(0.017)$      & $(0.011)$        & $(0.020)$      \\
+price\_l2                              & $0.037^{*}$    & $0.047^{**}$   & $0.048^{***}$    & $0.053^{*}$    \\
+                                             & $(0.017)$      & $(0.016)$      & $(0.011)$        & $(0.023)$      \\
+price\_l3                              & $0.030$        & $0.025$        & $0.033^{***}$    & $0.017$        \\
+                                             & $(0.016)$      & $(0.014)$      & $(0.010)$        & $(0.024)$      \\
+price\_l4                              & $0.003$        & $0.043^{**}$   & $0.028^{**}$     & $0.076^{***}$  \\
+                                             & $(0.015)$      & $(0.014)$      & $(0.009)$        & $(0.020)$      \\
+price\_l5                              & $-0.020$       & $0.012$        & $-0.001$         &                \\
+                                             & $(0.013)$      & $(0.012)$      & $(0.008)$        &                \\
+price\_l6                              & $-0.010$       & $0.008$        & $-0.001$         &                \\
+                                             & $(0.012)$      & $(0.011)$      & $(0.008)$        &                \\
+price\_l7                              & $-0.008$       & $0.007$        & $0.004$          &                \\
+                                             & $(0.012)$      & $(0.012)$      & $(0.008)$        &                \\
+price\_l8                              & $-0.044^{***}$ & $-0.011$       & $-0.027^{***}$   &                \\
+                                             & $(0.011)$      & $(0.011)$      & $(0.007)$        &                \\
+EDF: s(prod\_year)                           & $6.381^{***}$  &                &                  &                \\
+                                             & $(7.249)$      &                &                  &                \\
+EDF: s(year)                                 & $2.768^{***}$  & $3.000^{***}$  & $3.000^{***}$    & $3.000^{***}$  \\
+                                             & $(2.956)$      & $(3.000)$      & $(3.000)$        & $(3.000)$      \\
+EDF: s(prod\_year,name)                      &                & $33.002^{***}$ &                  & $32.962^{***}$ \\
+                                             &                & $(72.000)$     &                  & $(72.000)$     \\
+EDF: s(prod\_year,in\_place\_oil) &                &                & $27.445^{***}$   &                \\
+                                             &                &                & $(27.895)$       &                \\
+price\_l1:build\_out                   &                &                &                  & $0.038$        \\
+                                             &                &                &                  & $(0.024)$      \\
+price\_l2:build\_out                   &                &                &                  & $-0.002$       \\
+                                             &                &                &                  & $(0.029)$      \\
+price\_l3:build\_out                   &                &                &                  & $0.003$        \\
+                                             &                &                &                  & $(0.029)$      \\
+price\_l4:build\_out                   &                &                &                  & $-0.027$       \\
+                                             &                &                &                  & $(0.024)$      \\
 \hline
-Deviance                                     & 23062.678      & 19765.833      & 9057.204         \\
-Deviance explained                           & 0.848          & 0.870          & 0.940            \\
-Dispersion                                   & 5.410          & 5.965          & 2.325            \\
-R$^2$                                        & 0.841          & 0.865          & 0.938            \\
-GCV score                                    & 18.145         & 16.218         & 7.355            \\
-Num. obs.                                    & 1088           & 1088           & 1088             \\
-Num. smooth terms                            & 2              & 2              & 2                \\
+Deviance                                     & 23062.678      & 19765.852      & 9057.204         & 19786.567      \\
+Deviance explained                           & 0.848          & 0.870          & 0.940            & 0.870          \\
+Dispersion                                   & 17.852         & 15.625         & 7.123            & 15.641         \\
+R$^2$                                        & 0.841          & 0.865          & 0.938            & 0.865          \\
+GCV score                                    & 18.145         & 16.218         & 7.355            & 16.234         \\
+Num. obs.                                    & 1088           & 1088           & 1088             & 1088           \\
+Num. smooth terms                            & 2              & 2              & 2                & 2              \\
 \hline
-\multicolumn{4}{l}{\scriptsize{$^{***}p<0.001$, $^{**}p<0.01$, $^*p<0.05$}}
+\multicolumn{5}{l}{\scriptsize{$^{***}p<0.001$, $^{**}p<0.01$, $^*p<0.05$}}
 \end{tabular}
@@
 \end{figure}
+
+One potential weakness of the estimation so far is that I do not take into account the different stages of field production, where the effect of price could be expected to differ. 
+In particular, in the build-out phase of a field, where production is started and then ramped up to full capacity, the field operator has an incentive to build out the field as quickly as possible as large amounts of capital have at that point been allocated and it is important to get oil production and in turn income flowing.
+At this stage, it is ex-ante questionable that prices would have an effect on production, as a decision to increase capacity would require a new submission to the regulators and a potentially lengthy and expensive delay.
+On the other hand, the draw down phase is characterized by overcapacity and in turn greater flexibility in increasing production by way of, for example, infill drilling that can both accelerate recovery and increase total recovery in a field.
+\footnote{I thank an anonymous referee for making this point.}
+
+To attempt to capture this heterogeneity, I introduce an indicator variable that represents the build-out phase defined as the period between the start of production up to peak production.
+I add interaction effects between this indicator variable and the block of significant lagged price effects $price\_l1$ through $price\_l4$.
+
+The main problem with introducing this heterogeneity in the model is that most field build-outs happen quickly, usually within one or two years in most small to medium sized fields.
+The sample to estimate the build out price terms is then relatively small - less than one third of the total observations.
+The small sub-sample leads to estimation problems when using the model with a 2-dimensional smooth term, so I instead extend the random effects model.
+I also remove the block of 5th through 8th year lagged price terms that could not be shown to increase the fit of the model.
+
+The estimation results for this model are shown in the fourth column of table \ref{GAM_model_table}.
+The interaction effects, labeled $price lt:build$, are not estimated to be significantly different from zero.
+Given the small sample used to estimate these terms, this is not surprising.
+On the other hand, if we compare the main effects of the lagged price terms to the standard random effects model, we notice that the point estimates on the 2nd, 3rd and 4th lags are now subsantially larger in magnitude.
+In this model, these main effect terms can be interpreted as the effect of price on production in the draw-down phase.
+This provides some evidence that the effect of price on production is indeed strongest in this latter stage of field development.
 
@@
 
-The findings in this paper can provide some guidance on predicting the effects of the recent steep fall in the oil price on Norwegian production, as well as production from other offshore production areas.
+The findings in this article can provide some guidance on predicting the effects of the recent steep fall in the oil price on Norwegian production, as well as production from other offshore production areas.
 The results suggest that lower oil prices are likely to have only modest ramifications for production in existing fields.
@@
 Future oil production could then be significantly affected by way of fewer new field developments.
+
+Policy implications arise for both energy industry regulators and financial policy makers.
+Since the evidence points to a stronger lagged effect of price on production in declining fields, this would also imply that policy measures meant to stimulate increased production would likely also be more effective in mature oil producing areas compared to newer areas.
+This is particularily important in the Norwegian Continental Shelf, as several new fields have recently come online following a period of high oil prices.
+\footnote{I thank an anonymous reviewer for pointing out this implication.}
 
@@
 
+\section{Acknowledgments}
+I thank Klaus Mohn, Jonas Andersson, Sturla Kvamsdal, Harrison Fell, R\"ognvaldur Hannesson and Henrik Horn for valuable discussion.
+I have received financial support from the Norwegian Center for Sustainable Energy Studies (CenSES), Norwegian Research Council grant number 209697
+
+\begin{spacing}{1}
 \bibliographystyle{plainnat}
 \bibliography{oil_prices}
+\end{spacing}
 

### 2016-02-26 oil_prices_title.tex
added build-out interaction effects to the model

@@
 \author{Johannes Mauritzen\\
-		Department of Industrial Economics and Technology Management\\
-        Norwegian University of Science and Technology\\
-        7491 Trondheim
-        Norway\\
+		Department of Economics\\
+        BI Norwegian Business School\\
+        Trondheim, Norway\\
         johannes.mauritzen@iot.ntnu.no\\

### 2016-02-26 rr_analysis_obes.r

@@ -28,10 +28,11 @@ prod_data<-read.csv('/Users/johannesmauritzen/research/oil_prices/data/prod_data
 
 large_fields<-prod_data[prod_data["recoverable_oil"]>5,]
 large_fields<-prod_data[prod_data["name"]!="EKOFISK",]
+large_fields<-large_fields[!is.na(large_fields$oil_prod_mill_sm3),]
 
 
 gam_mod<-gam(oil_prod_mill_sm3~s(prod_year, bs="cr") +
-	I(log(in_place_oil_mill_sm3 +.01)) + I(cost_index/10) + world_gdp +
+	I(in_place_oil_mill_sm3/100) + I(cost_index/10) + 
     s(year, bs="cr", k=4) + I(price/10) + I(price_l1/10) + I(price_l2/10) +
     I(price_l3/10) + I(price_l4/10) + I(price_l5/10) + I(price_l6/10) + 
     I(price_l7/10) + I(price_l8/10), 
@@ -119,14 +120,52 @@ cal_smooth
 # dev.off()
 
 
-gam_mod_2d<-gam(oil_prod_mill_sm3~s(prod_year, I(log(in_place_oil_mill_sm3+.0001)) +
+gam_mod_2d<-gam(oil_prod_mill_sm3~s(prod_year, in_place_oil_mill_sm3) +
     s(year, bs="cr", k=4) +
-    I(cost_index/10) + world_gdp +
+    I(cost_index/10) + 
     I(price/10) + I(price_l1/10) + I(price_l2/10) +
     I(price_l3/10) + I(price_l4/10) + I(price_l5/10) + I(price_l6/10) + 
     I(price_l7/10) + I(price_l8/10), 
     family=gaussian(link=log), weights=I(in_place_oil_mill_sm3/100), data=large_fields)
+
 summary(gam_mod_2d)
+
+#With random effects
+
+gam_mod_re<-gam(oil_prod_mill_sm3~s(prod_year, name, bs="re") + 
+    s(year, bs="cr", k=4) +
+    I(in_place_oil_mill_sm3/100) + I(cost_index/10) +
+    I(price/10) + I(price_l1/10) + I(price_l2/10) +
+    I(price_l3/10) + I(price_l4/10) + I(price_l5/10) + I(price_l6/10) + 
+    I(price_l7/10) + I(price_l8/10), 
+     family=gaussian(link=log), weights=I(in_place_oil_mill_sm3/100), data=large_fields)
+summary(gam_mod_re)
+
[…]

### 2016-03-02 oil_prices_obes.tex
updated writing

@@
 
-However a significant gap exists in the literature.
+Oil and gas supply, especially in difficult and expensive offshore areas, is the production of an often lengthy and complex chain of conditional investment decisions.
+Starting with the exploration phase, where a potential oil-producing region or "play" is explored, leases are acquired, prospects are identified and exploration wells are drilled.
+In the case of a significant hydrocarbon find, a field is appraised to judge size, difficulty and in turn profitability.
+For fields seen as sufficiently profitable, a development stage commences, where a detailed drilling and production plan is drawn up and submitted to obtain permits and regulatory approval.
+Operating agreements are signed and subcontractors acquired for offshore construction and well development.
+Finally, at the end of this multi-year process, a build-out can commence, where full production capacity is built out as fast as possible, followed by an extended depletion phase.
+All of this consists of what is referred to as upstream activity, with steps such as shipping, refining and trading, among other, making up an equally complex down-stream process.
+\footnote{\ref{leffler_deepwater} provides an accessible guide to complex offshore petroleum exploration and production industry.}
+
+A large body of litterature exists on the many aspects of the complex chain of investment and production decisions made in the oil and gas industry.
+This article looks at the limited, but important topic of producing offshore fields, where a significant gap exists in the literature.
+
 While numerous studies have taken up the issue of how the price of oil affects searching for new fields as well as total oil production at both the regional and global level, few studies exist of the effects of oil prices on oil production at the field level.
@@
 
-% Moreover, studies within the economics literature fail to take into account how oil prices have varying effects on the different stages of field development and production.
-The planning, build-out and depletion phase of a field are conceptually unique and how oil companies operating offshore respond to oil prices can be expected to vary substantially.
+Moreover, studies within the economics literature fail to take into account how oil prices have varying effects on the different stages of field development and production.
+the build-out and depletion phases of a field are conceptually unique and how oil companies operating offshore respond to oil prices can be expected to vary substantially.
 
@@
 The main finding in this article is that oil production at the field level has no significant reaction to concurrent changes in the oil price.
-A slight effect can be detected at a lag of between 1 and 4 years, with a magnitude of about a 2 to 5\% increase in yearly production for a 10 dollar increase in the price of oil.
+A slight effect can be detected at a lag of between 1 and 4 years, with a magnitude of about a 2 to 7\% increase in yearly production for a 10 dollar increase in the price of oil.
 
@@
 
-\section{The Effect of Oil Price on Production: Theory, Simulations and Empirics}
+\section{Econometric Litterature on The Effect of Oil Price on Production}
 
-For modeling aggregate oil production, shape-fitting models, notably \citet{hubbert_energy_1962} and more recently advocated by \citet{deffeyes_hubberts_2001}, have had some success in estimating the timing of peak production at the regional and national level, but they tend to seriously underestimate the total recoverable resource of oil-producing regions and the models can be shown to be fundamentally misspecified \citep{boyce_prediction_2013}.
+%For modeling aggregate oil production, shape-fitting models, notably \citet{hubbert_energy_1962} and more recently advocated by \citet{deffeyes_hubberts_2001}, have had some success in estimating the timing of peak production at the regional and national level, but they tend to seriously underestimate the total recoverable resource of oil-producing regions and the models can be shown to be fundamentally misspecified \citep{boyce_prediction_2013}.
 Simulation type studies where aggregate oil production is modeled through an often complex combination of physical and economic processes also exist in the geo-engineering literature, but their usefulness tends to be weighed down by their complexity as they require quite detailed data and specific assumptions about functional form that can be difficult to justify \citep{brandt_review_2010}.
 
-More important to this paper are empirical estimates of the effect of price on production.
-Several econometric papers seek to answer the question of how aggregated oil supply is affected by oil prices.
-\citet{farzin_impact_2001} attempts to estimate an elasticity for the effect on added reserves of increased oil prices and finds a small though statistically significant effect.
+Early econometric attempts, like \citet{adelman_fea_1975} often made the assumption that the price contained all relevant information, ignoring the role of geophysical limits at the regional level which allowed oil supply to move in opposite direction of prices.
+However \citet{hamilton_oil_1983} showed that macroeconomic events could not explain movements in oil prices, and in turn rejecting the notion that prices contained all relevant information to oil supply.
+\citet{kaufmann_oil_1991} presents a relatively early attempt at combining both economic and geophysical aspects in emperically modeling oil production at the national level.
+The author used a hubbert-like \citep{hubbert_energy_1962} shape-fitting model to capture regional geophysical dynamics, in turn using the residuals to model the effects of price on production in the lower 48 United States.
+The author predicted that depletion and reduced oil supply from the lower 48 states would continue - a prediction that proved accurate up until the technological revolution of shale oil extraction.
+
+A similar analysis combining econometric models with shape-fitting models by \citet{cleveland_forecasting_1991} also led to pessimistic forecasts of future US production.
+Further refinements of these were done by, among others, \citet{pesaran_forecasting_1995}  and \citet{moroney_integrated_1999}.
+\citet{kaufman_oil_2001} present an extension and generalisation using a Vector Error Correction Model (VECM) that relaxes the assumption that firms produce their fields in order of decreasing quality and also includes institutional factors in US oil production.
+They claim that they are able to account for most of the price variation between 1938 and 1991.
+
+\citet{adelman_mineral_1990} and \citet{adelman_modeling_1993} take an opposing view, using cointegration and causality testing to argue that scarcity, and in turn geophysical forces had little to do with the eras oil price movements.
+Instead, Adelman argued that cartel behavior of low cost producers likely stood behind the increasing oil prices.
+\citet{lynch_forecasting_2002} provides a fuller overview of these competing strands of the litterature.
+
+The economics litterature on petroleum production tends to be overweighted towards studies of the US.
+\citet{watkins_world_1998} attempts to estimate supply functions for 41 countries by using methods from estimating the relationship between the price of undeveloped reserves and reserve additions over time.
+Based on the aggregated estimates, they found that fears of increasing global scarcity at the time were overblown.
+On the issue of endogeneity of price and supply, they make the somewhat dubious argument that for non-OPEC countries, price could be taken as exogenous since individual oil firms hedged (or "managed") price swings.
+
+\citet{farzin_impact_2001} attempts to estimate an elasticity for the effect on added global reserves of increased oil prices and finds a small though statistically significant effect.
 \citet{ramcharran_oil_2002} estimates a supply function for the total supply of oil from several OPEC countries based on data from 1973 to 1997.
@@
 However since the author does not take into account the production profile of oil fields and the spurious correlation that can arise with autocorrelated prices, these estimates come under considerable doubt.
+\citet{ringlund_does_2008} estimates how rig activity reacts to changes in oil prices in non-Opec countries, and finds substantial heterogeneity based on the structure of a countries oil industry.
+Though in the long run average, they find an elasticity of close to one.
 
-the effect of oil-price uncertainty on drilling and exploration has also been explored.
-\citet{kellogg_effect_2014} finds that oil exploration firms in Texas do approximately respond as real-option theory would predict when it comes to the timing of drilling.
+On the surface, this article fits well into the previous econometrics litterature on petroleum supply in that there is a focus on establishing and decomposing the effects of geophysical and economic forces.
+However, most of the above articles focus on total long-term regional oil supply - which, as I mention above, is the result of a multiple chain of decisions from surveying of a potential play, exploration drilling, and production.
+This article, on the other hand, is considerably narrower in scope - focusing exclusively on production in active fields.
+While ambiguity exists if geophysical forces and scarcity is a significant factor that needs to be controlled for in the long term regional perspective, it is without question a dominant factor at the field level.
+
+An expanding litterature looking at more microeconomic issues of oil supply exists.
+\citet{kellogg_effect_2014} studies the effect of oil-price uncertainty on drilling and exploration.
+The author finds that oil exploration firms in Texas do approximately respond as real-option theory would predict when it comes to the timing of drilling.
 A model and test using data from North Sea producers on the UK continental shelf by \citet{hurn_geology_1994}, on the other hand, fails to find evidence that the variance in the oil price affects the timing of oil field development.
@@
 
-Only a few papers utilize field-level data.
+Studies using detailed Norwegian data on offshore activity also exist, though the focus has mainly been on exploration and drilling.
+\citet{mohn_exploration_2008} finds that long-term changes in the oil-price has a strong effect on exploratory drilling though little effect is measured from short-term changes in the oil price.
+\citet{osmundsen_exploration_2010} analyses drilling productivity over time on the Norwegian Continental Shelf while \citet{mohn_efforts_2008} finds that higher oil prices leads to higher reserves as well as that oil prices affect producer risk-preferences - with higher prices leading to lower success rates but larger discovery size.
+\citet{mohn_elastic_2010} presents a thorough econometric analysis of the role of economic and policy factors in oil exploration and reserve additions, as well as a useful summary of exploration and production on the Norwegian Continental Shelf.
+
+Most of the microeconometric articles analyzing the petroleum industry, like the ones mentioned above, focus on issues of exploration, drilling, and rig activity.
+Microeconomic studies of production from existing, operational fields are more sparse and make up a significant gap in the litterature where this article contributes.
+
+Only a few papers study production decisions utilizing field-level data.
 \citet{black_is_1998} tests the relevance of “nesting” a structural empirical model of profit maximization that takes into account oil prices into a typical geo-engineering model of oil field production.
@@
 How producers react to a short-term tax change as opposed to a change in price also likely differs.
-
-Studies using detailed Norwegian data on offshore activity also exist, though the focus has mainly been on exploration and drilling.
-\citet{mohn_exploration_2008} finds that long-term changes in the oil-price has a strong effect on exploratory drilling though little effect is measured from short-term changes in the oil price.
-\citet{osmundsen_exploration_2010} analyses drilling productivity over time on the Norwegian Continental Shelf while \citet{mohn_efforts_2008} finds that higher oil prices leads to higher reserves as well as that oil prices affect producer risk-preferences - with higher prices leading to lower success rates but larger discovery size.
 
@@
 Profits from oil and gas production in Norway are subject to a resource tax of 51 \% on top of the ordinary income tax of 27 \%, thus income from petroleum production is taxed at a total marginal rate of 78 \%.
-The central government also receives revenues through ownership stakes in companies, notably Statoil, where the state is the majority stakeholder.
+The central government also receives revenues through its majority ownership stake in Statoil.
+In addition, a wholly state-owned company, Petero, manages the development licenses that the state retains - referred to as the State's Direct Financial Interest (SDFI).
+
+The tax system is also designed to be neutral, that is a field that is profitable before tax should also be profitable after tax.
+This is accomplished by only taxing the net-profits of oil companies, allowing companies to deduct and carry forward losses, and deducting investment costs.\footnote{The Norwegian Oil and Gas industry group Norsk Petroleum provides a fuller overview: \url{http://www.norskpetroleum.no/en/framework/petroleum-tax/}}
 

### 2016-03-03 oil_prices_obes.tex
update

@@
 The response of oil production in existing fields is especially important now as many of the major oil-producing areas, like the Norwegian Continental Shelf are mature and an increasing share of investments will be directed at capacity enhancement in producing fields rather than new field developments.
+Figure \ref{investment} shows that investments in producing fields makes up a large share of total investments, nearly dwarfing investments in searching, exploration and new build-outs.
 How production from these fields will respond to changes in oil prices has major implications for the oil industry, the state finances of oil-producing countries and long-run oil price formation.
+
+\begin{figure}
+	\includegraphics[width=1\textwidth]{figures/investment.png}
+	\caption{The figure shows investments (millions NOK) in the oil and gas extraction industry on the Norwegian Continental Shelf.
+A large and growing share of investment goes into fields that are already producing as opposed to investments in search, exploration and build-out of new fields.}
+	\label{investment}
+\end{figure}
 
@@
 \citet{kellogg_effect_2014} studies the effect of oil-price uncertainty on drilling and exploration.
-The author finds that oil exploration firms in Texas do approximately respond as real-option theory would predict when it comes to the timing of drilling.
+The author finds that oil exploration firms in Texas approximately respond as real-option theory would predict when it comes to the timing of drilling.
 A model and test using data from North Sea producers on the UK continental shelf by \citet{hurn_geology_1994}, on the other hand, fails to find evidence that the variance in the oil price affects the timing of oil field development.

### 2016-03-03 rr_descriptives.r

@@ -9,6 +9,7 @@ library(lubridate)
 library(grid)
 library(arm)
 library(texreg)
+library(zoo)
 
 field_data<-read.csv('/Users/johannesmauritzen/research/oil_prices/data/field_data.csv')
 prod_data<-read.csv('/Users/johannesmauritzen/research/oil_prices/data/prod_data.csv')
@@ -19,6 +20,53 @@ small_rec_oil<-sum(field_data["recoverable_oil"][field_data["recoverable_oil"]<=
 28.3/4779.2
 #.5% of total recoverable oil.  
 
+
+#plot of investing
+#from ssb
+invest <- read.csv("https://data.ssb.no/api/v0/dataset/166334.csv?lang=no", sep=";")
+colnames(invest) <- c("type", "period", "var", "investment")
+invest <- reshape(invest, 
+  timevar = "type",
+  idvar = c("period", "var"),
+  direction = "wide")
+
+quarter <- as.Date(as.yearqtr(invest$period, format = "%YK%q"))
+invest$period <- quarter
+invest <- invest[c("period", "investment.0 Investeringer i alt", 
+	"investment.10.3.3 Produksjonsboring, felt i drift",
+	"investment.10.3 Felt i drift",
+	"investment.10.2 Feltutbygging", 
+	"investment.10.1 Leting og konseptstudier",
+	"investment.10.1.1 Unders\xf8kelsesboring")]
+
+
+invest <- invest[invest$period<as.Date("2015-01-01"),]
+
+colnames(invest) <-c("period", "Total", 
+	"In Production, Drilling", "In Production, Total", "Build-out", "Search", "Expl. Drilling")
+
+invest_long <- melt(invest, id.vars="period")
+
+investment<-ggplot(invest_long, aes(x=period, y=value, color=variable)) +
+#facet_wrap(nrow=2, ~variable, scales="free_y") +
+geom_line() +
+scale_color_grey() +
+theme_bw() +
+#geom_text() +
+annotate("text", x = as.Date("2016-01-01"), y = 50000, label = "Total") +
+annotate("text", x = as.Date("2016-06-01"), y = 21000, label = "In Production, Total") +
+annotate("text", x = as.Date("2016-01-01"), y = 19000, label = "Build-out") +
+annotate("text", x = as.Date("2016-06-01"), y = 11000, label = "In Production Drilling") +
+annotate("text", x = as.Date("2016-01-01"), y = 8000, label = "Search") +
[…]

### 2016-03-04 oil_prices_obes.tex
update draft

@@
 \usepackage{comment} %for excluding figures
+\usepackage{booktabs}
 % \excludecomment{figure} % for printing without figures
@@
 Finally, at the end of this multi-year process, a build-out can commence, where full production capacity is built out as fast as possible, followed by an extended depletion phase.
-All of this consists of what is referred to as upstream activity, with steps such as shipping, refining and trading, among other, making up an equally complex down-stream process.
-\footnote{\ref{leffler_deepwater} provides an accessible guide to complex offshore petroleum exploration and production industry.}
+All of this consists of what is referred to as upstream activity, with steps such as shipping, refining and trading, among others, making up an equally complex down-stream process.
+\footnote{\citet{leffler_deepwater_2011} provides an accessible guide to the offshore petroleum exploration and production industry.}
 
@@
 \begin{figure}
-	\includegraphics[width=1\textwidth]{figures/investment.png}
+	\center
+	\includegraphics[width=.9\textwidth]{figures/investment.png}
 	\caption{The figure shows investments (millions NOK) in the oil and gas extraction industry on the Norwegian Continental Shelf.
@@
 Further refinements of these were done by, among others, \citet{pesaran_forecasting_1995}  and \citet{moroney_integrated_1999}.
-\citet{kaufman_oil_2001} present an extension and generalisation using a Vector Error Correction Model (VECM) that relaxes the assumption that firms produce their fields in order of decreasing quality and also includes institutional factors in US oil production.
+\citet{kaufmann_oil_2001} present an extension and generalisation using a Vector Error Correction Model (VECM) that relaxes the assumption that firms produce their fields in order of decreasing quality and also includes institutional factors in US oil production.
 They claim that they are able to account for most of the price variation between 1938 and 1991.
 
-\citet{adelman_mineral_1990} and \citet{adelman_modeling_1993} take an opposing view, using cointegration and causality testing to argue that scarcity, and in turn geophysical forces had little to do with the eras oil price movements.
+\citet{adelman_mineral_1990} and \citet{adelman_modelling_1993} take an opposing view, using cointegration and causality testing to argue that scarcity, and in turn geophysical forces had little to do with the eras oil price movements.
 Instead, Adelman argued that cartel behavior of low cost producers likely stood behind the increasing oil prices.
@@
 The use of total recoverable reserves is complicated since it is dependent on the price of oil and thus endogenous in the analysis.
-Therefore, as a proxy for the size of the fields, I use estimates of in-place oil from the Norwegian Petroleum Directorate, which is unlikely to be correlated with prices.
+Therefore, as a proxy for the size of the fields, I use estimates of geological in-place oil which is not a function of economic factors.
 The left panels of figure \ref{data_descriptives} show the histograms of field sizes as measured by estimates of recoverable oil and in-place oil.
@@
 All subsequent major fields used gas and water injection from the outset and have a single-hump profile.
+
+Table \ref{field_summary} provides summary statistics over the distribution of field-level variables.
+Notably, the median field in the data set had been producing for 16 years as of the beginning of 2015 - an indication of the mature status of the Norwegian Continental Shelf.
+
+\begin{table}
+\begin{tabular}{lrrrrr}
+\toprule
+Percentiles &    5\% &    25\% &    50\% &     75\% &     95\% \\
+                      &       &        &        &         &         \\
+\midrule
+In-place Oil, Mill SM3         &  5.90 &  34.00 &  75.80 &  157.00 &  636.70 \\
+Est. Recoverable Oil           &  6.20 &  10.50 &  25.70 &   70.30 &  374.50 \\
+Production Years, to 2015      &  1.79 &   9.17 &  16.04 &   22.43 &   35.63 \\
+Tot. Cumulative Prod., to 2015, Mill SM3 &  3.61 &   8.94 &  22.94 &   57.46 &  359.04 \\
+\bottomrule
+\end{tabular}
+\caption{Summary distribution statistics of field-level data. Columns represent percentiles of data. Excluds the Ekofisk field and fields with less than 5 million SM3 of recoverable oil.}
+\label{field_summary}
+\end{table}
 
@@
 \begin{split}
-	Log(production_{i,t}) & = f(production\_time_{i,t}) + f(year_t) + \beta_1 in\_place\_oil_{i,t}\\
+	Log(production_{i,t}) & = f(production\_time_{i,t}) + f(year_t) + \beta_1 in\_place\_oil_{i}\\
 	 \quad & + \beta_3 cost\_index_{t} + \mathbf{\beta_2 oil\_price\_lags_t} +  
@@
 The left-hand-side variable, $Log(production_{i,t})$ is the log of yearly oil production for field $i$ at production time $t$ in millions of standard cubic meters (SM3).
-On the right hand side is a smoothed function of the production profile over time $f(production\_time_{i,t})$; $in\_place\_oil_{i,t}$ is the estimated in-place oil of the field, and $cost\_index_{t}$ represents the producer price index for oil and gas extraction.
+On the right hand side is a smoothed function of the production profile over time $f(production\_time_{i,t})$; $in\_place\_oil_{i}$ is the estimated in-place oil of the field, and $cost\_index_{t}$ represents the producer price index for oil and gas extraction.
 

### 2016-03-04 oxfordbes_submission/summ_stat.py

@@ -1,0 +1,53 @@
+#summ_stat.py
+
+import pandas as pd
+import tabulate as tab
+import numpy as np
+import matplotlib as mpt
+import matplotlib.pyplot as plt
+import matplotlib.patches as mpatches
+import seaborn as sns
+import statsmodels.formula.api as smf
+import statsmodels.api as sm
+import sys
+import pystan
+import scipy
+import pickle
+import json
+
+pd.options.display.max_rows = 2000
+pd.options.display.max_columns = 100
+
+#statsmodels model
+field_data = pd.read_csv('/Users/johannesmauritzen/research/oil_prices/data/field_data.csv')
+prod_data = pd.read_csv('/Users/johannesmauritzen/research/oil_prices/data/prod_data.csv')
+
+large_prod=prod_data[prod_data["recoverable_oil"]>5]
+tot_prod = large_prod.groupby("name")["oil_prod_mill_sm3"].sum()
+
+large_field = field_data[field_data.recoverable_oil>5]
+large_field = large_field[large_field.producing_date.notnull()]
+large_field["producing_date"] = pd.to_datetime(large_field.producing_date, format="%d.%m.%Y")
+large_field["prod_time"] = pd.to_datetime("2015-01-01") - large_field.producing_date
+large_field.reset_index(inplace=True)
+
+prod_years = [float(d.days)/365 for d in large_field.prod_time]
+large_field["prod_years"] = prod_years
+
+summ_data = np.empty(shape=(4,5))
+
+def get_percentiles(data):
+	return(np.percentile(data, [5, 25, 50, 75, 95]).round(2))
+
+summ_data[0 ,:] = get_percentiles(large_field.in_place_oil_mill_sm3)
+summ_data[1 ,:] = get_percentiles(large_field.recoverable_oil)
+summ_data[2 ,:] = get_percentiles(large_field.prod_years)
+summ_data[3, :] = get_percentiles(tot_prod)
+
+table = pd.DataFrame(summ_data, columns =["5%", "25%", "50%", "75%", "95%"])
+table["variable"] = ["In Place Oil, Mill SM3", "Est. Recoverable Oil", "Production Years, to 2015", "Tot. Cumulative Prod., to 2015"]
+table.set_index("variable", inplace=True)
+table.to_latex(buf="oxfordbes_submission/summ_stat.tex")
+ #rps_laws[['Total RPS %', 'Year','voluntary']].to_latex(buf="temp_tables.tex")
+
+

### 2016-03-06 oil_prices_obes.tex
corrected draft

@@
 I use detailed field-level data on Norwegian offshore oil production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production.
-I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 2 to 7\% for a 10 dollar per barrel increase in the real price of oil.\\
+I find no significant evidence of a concurrent reaction of field production to oil prices, though a modest lagged effect is found of the magnitude of approximately 2 to 7\% for a 10 dollar per barrel increase in the real price of oil.\\
 Keywords: Oil production,Oil Prices, Norway, Generalized Additive Model.
-Words: 5970
+Words: 7000
 \end{abstract}
@@
 
-Oil and gas supply, especially in difficult and expensive offshore areas, is the production of an often lengthy and complex chain of conditional investment decisions.
+Oil and gas supply, especially in difficult and expensive offshore areas, is the product of an often lengthy and complex chain of conditional investment decisions.
 Starting with the exploration phase, where a potential oil-producing region or "play" is explored, leases are acquired, prospects are identified and exploration wells are drilled.
@@
 All of this consists of what is referred to as upstream activity, with steps such as shipping, refining and trading, among others, making up an equally complex down-stream process.
-\footnote{\citet{leffler_deepwater_2011} provides an accessible guide to the offshore petroleum exploration and production industry.}
+\footnote{\citet{lefler_deepwater_2011} provides an accessible guide to the offshore petroleum exploration and production industry.}
 
-A large body of litterature exists on the many aspects of the complex chain of investment and production decisions made in the oil and gas industry.
-This article looks at the limited, but important topic of producing offshore fields, where a significant gap exists in the literature.
+A large body of economic litterature exists on the complex chain of investment and production decisions made in the upstream oil and gas industry.
+This article looks at the limited, but important topic of the effect of price on production from active offshore fields, where a significant gap exists in the literature.
 
@@
 
-Moreover, studies within the economics literature fail to take into account how oil prices have varying effects on the different stages of field development and production.
-the build-out and depletion phases of a field are conceptually unique and how oil companies operating offshore respond to oil prices can be expected to vary substantially.
-
 The effect of oil prices on producing fields is an important topic in understanding the mechanisms of how total oil supply reacts in response to price.
-The response of oil production in existing fields is especially important now as many of the major oil-producing areas, like the Norwegian Continental Shelf are mature and an increasing share of investments will be directed at capacity enhancement in producing fields rather than new field developments.
-Figure \ref{investment} shows that investments in producing fields makes up a large share of total investments, nearly dwarfing investments in searching, exploration and new build-outs.
-How production from these fields will respond to changes in oil prices has major implications for the oil industry, the state finances of oil-producing countries and long-run oil price formation.
+The response of oil production in existing fields is especially important now as many of the major oil-producing areas, like the Norwegian Continental Shelf are mature and an increasing share of total investment will be directed at capacity enhancement in producing fields rather than new field developments.
+Figure \ref{investment} shows that investments in producing fields makes up a large and growing share of total investment, nearly dwarfing investments in searching, exploration and new build-outs.
+How production from mature fields will respond to changes in oil prices has major implications for the oil industry, the state finances of oil-producing countries and long-run oil price formation.
 
@@
 	\includegraphics[width=.9\textwidth]{figures/investment.png}
-	\caption{The figure shows investments (millions NOK) in the oil and gas extraction industry on the Norwegian Continental Shelf.
-A large and growing share of investment goes into fields that are already producing as opposed to investments in search, exploration and build-out of new fields.}
+	\caption{The figure shows investment (millions NOK) in the oil and gas extraction industry on the Norwegian Continental Shelf.
+A large and growing share of total investment goes into fields that are already producing as opposed to investment in search, exploration and build-out of new fields.}
 	\label{investment}
@@
 But in practice, the question is not as clear cut.
-Production in the Norwegian Continental Shelf - as well as most other offshore production areas - has extremely high fixed and operating costs.
+Production on the Norwegian Continental Shelf - as well as most other offshore production areas - has extremely high fixed and operating costs.
 Keeping spare capacity available in order to adjust to changing oil prices might simply be too expensive.
 Instead, producers may find it more beneficial to use storage and financial instruments in order to hedge short-term price movements.
-Higher oil prices may however still lead a producer to invest in increased capacity.
+Higher oil prices may still lead a producer to invest in increased capacity.
 Since lag times are significant in the offshore sector, we would then expect to see a multi-year lagged effect of prices on production.
@@
 The Norwegian government has been on the forefront of this movement and increasingly committed itself to transparency in the petroleum sector.
-\footnote{see http://www.regjeringen.no/en/sub/eiti---extractive-industries-tranparency/about-eiti.html?id=633586} Over the last several years, detailed data on most aspects of the country's oil industry has become openly available.
+\footnote{See \url{http://www.regjeringen.no/en/sub/eiti---extractive-industries-tranparency/about-eiti.html?id=633586}} Over the last several years, detailed data on most aspects of the country's oil industry has become openly available.
 
@@
 The main finding in this article is that oil production at the field level has no significant reaction to concurrent changes in the oil price.
-A slight effect can be detected at a lag of between 1 and 4 years, with a magnitude of about a 2 to 7\% increase in yearly production for a 10 dollar increase in the price of oil.
+A modest effect can be detected at a lag of between 2 and 4 years of between 2 to 7\% increase in yearly production for a 10 dollar increase in the price of oil.
 
@@
 	\includegraphics[width=1\textwidth]{figures/oil_decline.png}
-	\caption{The top panel shows the production from the largest 10 oil fields, whose starting times are temporally correlated with each other.
+	\caption{The top panel shows the production from the largest 10 oil fields on the Norwegian Continental Shelf, whose starting times are temporally correlated with each other.
 The total production over time is bell-shaped, as shown by the bottom panel.}
@@
 High oil prices were present at periods of relatively low production in the late 1970s and early 1980s as well as over the last 10 years, however real prices reached some of their lowest levels at the same time as the top of production around the year 2000.
-These oil price dynamics almost certainly affected investment decisions in the industry and in turn total production (\citep{osmundsen_is_2007}, \citep{aune_financial_2010}).
+These oil price dynamics almost certainly affected investment decisions in the industry and in turn total production \citep{osmundsen_is_2007}, \citep{aune_financial_2010}.
 However the inverse contemporaneous relationship at the field level is entirely coincidental, but will heavily bias the estimation of the effect of price on production if field production profiles are not properly accounted for.
@@
 
-Early econometric attempts, like \citet{adelman_fea_1975} often made the assumption that the price contained all relevant information, ignoring the role of geophysical limits at the regional level which allowed oil supply to move in opposite direction of prices.
-However \citet{hamilton_oil_1983} showed that macroeconomic events could not explain movements in oil prices, and in turn rejecting the notion that prices contained all relevant information to oil supply.
-\citet{kaufmann_oil_1991} presents a relatively early attempt at combining both economic and geophysical aspects in emperically modeling oil production at the national level.
-The author used a hubbert-like \citep{hubbert_energy_1962} shape-fitting model to capture regional geophysical dynamics, in turn using the residuals to model the effects of price on production in the lower 48 United States.
-The author predicted that depletion and reduced oil supply from the lower 48 states would continue - a prediction that proved accurate up until the technological revolution of shale oil extraction.
+An extensive litterature exists that attempts to estimate the effect of oil prices on regional or global oil supply.
+Early econometric attempts, like \citet{adelman_fea_1975} often made the assumption that the price contained all relevant information, ignoring the role of geophysical limits at the regional level.
+\citet{hamilton_oil_1983} showed that macroeconomic events could not fully explain movements in oil prices, rejecting the notion that prices contained all relevant information to oil supply.
+\citet{kaufmann_oil_1991} presents a relatively early attempt at combining both economic and geophysical aspects in an emperical model of oil production in the US.
+The author used a Hbbert-like \citep{hubbert_energy_1962} shape-fitting model to capture regional geophysical dynamics, using the residuals to model the effects of price on production.
+The author predicted that depletion and reduced oil supply from the lower 48 states would continue - a prediction that proved mostly accurate up until the technological revolution of fracking and shale oil extraction.
 
@@
 
-\citet{adelman_mineral_1990} and \citet{adelman_modelling_1993} take an opposing view, using cointegration and causality testing to argue that scarcity, and in turn geophysical forces had little to do with the eras oil price movements.
-Instead, Adelman argued that cartel behavior of low cost producers likely stood behind the increasing oil prices.
-\citet{lynch_forecasting_2002} provides a fuller overview of these competing strands of the litterature.
+\citet{adelman_mineral_1990} and \citet{adelman_modelling_1993} takes an opposing view, using cointegration and causality testing to argue that scarcity, and in turn geophysical forces had little to do with the oil price movements of the 1970s and 1980s.
+Instead Adelman argues that cartel behavior of low cost producers likely stood behind increasing oil prices.
+\citet{lynch_forecasting_2002} provides a fuller overview of these competing strands of litterature.
 
-The economics litterature on petroleum production tends to be overweighted towards studies of the US.
-\citet{watkins_world_1998} attempts to estimate supply functions for 41 countries by using methods from estimating the relationship between the price of undeveloped reserves and reserve additions over time.
-Based on the aggregated estimates, they found that fears of increasing global scarcity at the time were overblown.
-On the issue of endogeneity of price and supply, they make the somewhat dubious argument that for non-OPEC countries, price could be taken as exogenous since individual oil firms hedged (or "managed") price swings.
+The economics litterature on petroleum production tends to be overweighted towards studies of the US, but some global studies have also been attempted.
+\citet{watkins_world_1998} attempt to estimate supply functions for 41 countries by estimating the relationship between the price of undeveloped reserves and reserve additions over time.
+Based on the aggregated estimates, they argued that fears of increasing global scarcity at the time were overblown.
 
-\citet{farzin_impact_2001} attempts to estimate an elasticity for the effect on added global reserves of increased oil prices and finds a small though statistically significant effect.
-\citet{ramcharran_oil_2002} estimates a supply function for the total supply of oil from several OPEC countries based on data from 1973 to 1997.
+\citet{farzin_impact_2001} attempts to estimate the price elasticity of added global reserves and finds a small positive elasticity.
+\citet{ramcharran_oil_2002} estimates a total supply function of oil from several OPEC countries based on data from 1973 to 1997.
 The author finds a negative price elasticity for several of the countries, and interprets this as evidence of producers targeting revenue.
 However since the author does not take into account the production profile of oil fields and the spurious correlation that can arise with autocorrelated prices, these estimates come under considerable doubt.
-\citet{ringlund_does_2008} estimates how rig activity reacts to changes in oil prices in non-Opec countries, and finds substantial heterogeneity based on the structure of a countries oil industry.
-Though in the long run average, they find an elasticity of close to one.
+\citet{ringlund_does_2008} estimates how rig activity reacts to changes in oil prices in non-Opec countries, and finds substantial heterogeneity based on the structure of a country's oil industry.
+Though in the long run, they find an average elasticity of close to one.
 
 On the surface, this article fits well into the previous econometrics litterature on petroleum supply in that there is a focus on establishing and decomposing the effects of geophysical and economic forces.
-However, most of the above articles focus on total long-term regional oil supply - which, as I mention above, is the result of a multiple chain of decisions from surveying of a potential play, exploration drilling, and production.
+However, most of the above articles focus on total long-term regional oil supply, which as I mention above, is the result of an extensive chain of decisions from surveying of a potential play, exploration drilling, to production.
 This article, on the other hand, is considerably narrower in scope - focusing exclusively on production in active fields.
-While ambiguity exists if geophysical forces and scarcity is a significant factor that needs to be controlled for in the long term regional perspective, it is without question a dominant factor at the field level.
+While ambiguity exists on whether geophysical forces and scarcity are significant factors that need to be controlled for in the long term regional perspective, they are without question dominant factors at the field level.
 
-An expanding litterature looking at more microeconomic issues of oil supply exists.
+Microeconomic aspects of oil supply have more recently come to the fore.
 \citet{kellogg_effect_2014} studies the effect of oil-price uncertainty on drilling and exploration.
@@
 Most of the microeconometric articles analyzing the petroleum industry, like the ones mentioned above, focus on issues of exploration, drilling, and rig activity.
-Microeconomic studies of production from existing, operational fields are more sparse and make up a significant gap in the litterature where this article contributes.
-
-Only a few papers study production decisions utilizing field-level data.
+Microeconomic studies of production from operational fields are more sparse.
 \citet{black_is_1998} tests the relevance of “nesting” a structural empirical model of profit maximization that takes into account oil prices into a typical geo-engineering model of oil field production.
 They find strong evidence that taking into account profit maximization - and implicitly price - substantially improves the fit compared to a purely geo-engineering model.
-The limitation of their methodology is that they are only able to test whether including economic factors like price affects the fit of the model but are not able to give an estimate of the magnitude of the effect.
-Methodologically, the paper also relies heavily on assumptions about the functional form of both the geo-engineering aspects of the oil producer as well as their profit-maximization.
+The major limitation of their methodology is that they are only able to test whether including economic factors like price affects the fit of the model but are not able to give an estimate of the magnitude of the effect.
+Methodologically, the paper also relies heavily on strict assumptions about the functional form of both the geo-engineering and economic aspects of their model.
 By taking a more flexible, semi-parametric approach to modeling oil field production profiles, this article avoids problems with overly restrictive assumptions.
@@
 The author finds that short-term tax changes caused small, but significant re-timing of production from oil wells.
-These findings are over-all consistent with the findings of this article, though the author finds a concurrent effect of changes in taxation on production while I do not find any concurrent effect of prices on production.
+These findings are generally consistent with the findings of this article, though the author finds a concurrent effect of changes in taxation on production while I do not find any concurrent effect of prices on production.
 The significant differences in cost and complexity between operating offshore and onshore likely account for the different results.
@@
 A major exception to this trend was the recent find of the Johan Sverdrup field which is estimated to have approximately 300 million Standard Cubic Meters (SM3) of recoverable oil.
-\footnote{The Johan Sverdrup field is estimated to begin producing oil in 2019 and so is not present in the data set used for the analysis.}
+\footnote{The Johan Sverdrup field is estimated to begin producing oil in 2019 and is not present in the data set used for the analysis.}
 
@@
 The tax system is also designed to be neutral, that is a field that is profitable before tax should also be profitable after tax.
-This is accomplished by only taxing the net-profits of oil companies, allowing companies to deduct and carry forward losses, and deducting investment costs.\footnote{The Norwegian Oil and Gas industry group Norsk Petroleum provides a fuller overview: \url{http://www.norskpetroleum.no/en/framework/petroleum-tax/}}
+This is accomplished by only taxing the net-profits of oil companies, allowing companies to deduct investment costs and deduct and carry forward losses.\footnote{The Norwegian Oil and Gas industry group Norsk Petroleum provides a fuller overview: \url{http://www.norskpetroleum.no/en/framework/petroleum-tax/}}
 
-The over-all tax rate has been fairly constant through the history of Norwegian oil production, however several important changes related to the tax code have occurred.
+The over-all tax rate has been fairly constant through the history of Norwegian petroleum production, however several important changes related to the tax code have occurred.
 In 1991 a C02 tax was introduced, and in the year 2012 the tax was doubled.
@@
 
-To simplify the analysis, I eliminate the smallest fields - those with less than 3 million SM3 of recoverable reserves.
+To simplify the analysis, I eliminate the smallest fields - those with less than 5 million SM3 of recoverable reserves.
 These fields tend to have substantially different production profiles than the larger fields.
-However, these fields only account for around 0.5 \% of total recoverable reserves so this exclusion is unlikely to have any serious practical implications.
+However, these fields only account for around 0.5 \% of total recoverable reserves.
 
@@
 \end{tabular}
-\caption{Summary distribution statistics of field-level data. Columns represent percentiles of data. Excluds the Ekofisk field and fields with less than 5 million SM3 of recoverable oil.}
+\caption{Summary distribution statistics of field-level data. Columns represent percentiles of data. Excludes the Ekofisk field and fields with less than 5 million SM3 of recoverable oil.}
 \label{field_summary}
@@
 
-I do not, however, want to estimate smoothed curves individually for each field.
+I do not want to estimate smoothed curves individually for each field.
 While this would provide a good overall fit to the full data set, not enough variation in the data would be left to estimate the effect of price.
-Instead I want to estimate a general shape of the production profile for all fields and then use the remaining variation in the data to estimate the effect of price.
-
+Instead I want to estimate a general shape of the production profile for all fields.
 A simple model can be written as in equation \ref{gam_price_eqn}.
@@
 	Log(production_{i,t}) & = f(production\_time_{i,t}) + f(year_t) + \beta_1 in\_place\_oil_{i}\\
-	 \quad & + \beta_3 cost\_index_{t} + \mathbf{\beta_2 oil\_price\_lags_t} +  
+	 \quad & + \mathbf{\beta_2 oil\_price\_lags_t} + 
+	 \beta_3 cost\_index_{t}  +  
 	  \epsilon_{i,t}
@@
 A solution is to allow the shape to vary by the size of the field - here I use an estimate of in-place oil in the field rather than recoverable reserves, as recoverable reserves may be correlated with oil prices.
-Thus the smoothed term can be written as a function of both production time and in place oil: $f(production\_time_{i,t}, in\_place\_oil_{i,t})$.
+Thus the smoothed term can be written as a function of both production time and in place oil: $f(production\_time_{i,t}, in\_place\_oil_{i})$.
 
@@
 The variables of interest in this article is the oil price and its lags, which I include as eight linear parametric terms in the model.
-The idea of including both a concurrent oil price term as well as eight lags is that a change in price could conceivably have two effects on oil production in a field.
+The reason for including both a concurrent oil price term as well as eight lags is that a change in price could conceivably have two effects on oil production in a field.
 First, the field operator could be operating on the basis of some short-term extraction rule - choosing to pump out less at times of lower prices in order to pump out more at periods of high prices.
 
-Alternatively, a change in price of oil can be seen as a lifting of a production constraint.
-A higher oil price means that added investments in production become attractive in order to either increase the total amount extracted from a field or to shift production forward.
+Alternatively, a change in the price of oil can be seen as a lifting of a production constraint.
+A higher oil price means that added investments in production becomes attractive in order to either increase the total amount extracted from a field or to shift production forward.
 However investments in the offshore sector can be complex and lengthy, and any production lift would be expected to happen with a lag.
@@
 \caption{Model results.
-Estimated standard coefficients are shown for the linear terms.
+Estimated coefficients are shown for the linear terms.
 Standard errors are in parenthesis.
-For the smoothed terms, estimated degrees of freedom are displayed, where the p-values are from F-tests of whether the terms significantly improve the fit of the model.}
+For the smoothed terms, estimated degrees of freedom are displayed, where the p-values are from F-tests of whether the smooth terms significantly improve the fit of the model.}
 \label{GAM_model_table}
@@
 For convenience the estimated coefficients on the price terms from the two preferred models are shown in figure \ref{price_coefficients}.
-In general the results from the two models agree with each other, indicating little to no effect in the concurrent price term, but with significant coefficients estimated on the 1st to 4th lags, with a peak at the 2nd lag.
-A negative coefficient for the eight lag is estimated in the 2-d model, however this is not significant in the model with random effects.
+In general the results from the two models agree with each other, indicating little to no effect on the concurrent price term, but with positive coefficients estimated on the 2st to 4th lags, with a peak at the 2nd lag.
+A negative coefficient for the eighth lag is estimated in the 2-d model, however this is not significant in the model with random effects.
 Furthermore, F-tests with comparisons to restricted models without the concurrent and 5-8th lags indicate that these terms do not as a group significantly improve the fit of the model.
+On the other hand, the block of terms for the 1st through 4th lags do significantly add to the fit of the model.
 
@@
 The points represent the point estimates, while the lines represent approximate 95\% confidence intervals.
-Significant positive coefficients are estimated on lags of between 1-4 years.
+Significant positive coefficients are estimated on lags of between 2-4 years.
 However, little to no effect is measured on the concurrent price term.}
@@
 The estimation results for this model are shown in the fourth column of table \ref{GAM_model_table}.
-The interaction effects, labeled $price lt:build$, are not estimated to be significantly different from zero.
+The interaction effects, labeled $price\_lt:build$, are not estimated to be significantly different from zero.
 Given the small sample used to estimate these terms, this is not surprising.
@@
 
-The main result of this research is to show that production in existing Norwegian offshore fields has no significant concurrent reaction to higher oil prices while a slight effect is estimated  with a lag of between 1 and 4 years.
+The main result of this research is to show that production in existing Norwegian offshore fields has no significant concurrent reaction to higher oil prices while a modest effect is estimated  with a lag of between 2 and 4 years.
 Oil producers do not appear to be behaving strategically in relation to short-term production - increasing or reducing production in response to changes in oil price.
@@
 
-Another important caveat, is that the changes in production could potentially either be a permanent increase in production, or a shift forward of production that may otherwise have happened later.
+Another important caveat is that the changes in production could potentially either be a permanent increase in production, or a shift forward of production that may otherwise have happened later.
 Unfortunately, with the existing data and methods, I do not see a good way of distinguishing between these alternatives.
@@
 Since the evidence points to a stronger lagged effect of price on production in declining fields, this would also imply that policy measures meant to stimulate increased production would likely also be more effective in mature oil producing areas compared to newer areas.
-This is particularily important in the Norwegian Continental Shelf, as several new fields have recently come online following a period of high oil prices.
+This is particularily important on the Norwegian Continental Shelf, as several new fields have recently come online following a period of high oil prices.
 \footnote{I thank an anonymous reviewer for pointing out this implication.}

### 2016-03-06 oil_prices_title.tex
corrected draft

@@
 I use detailed field-level data on Norwegian offshore oil production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production.
-I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 1 to 4\% for a 10 dollar per barrel increase in the real price of oil.\\
+I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 2 to 5\% for a 10 dollar per barrel increase in the real price of oil.\\
 \textbf{Keywords}: Oil production,Oil Prices, Norway, Generalized Additive Model.\\

### 2016-03-06 rr_analysis_obes.r

@@ -149,7 +149,7 @@ gam_mod_re_inter<-gam(oil_prod_mill_sm3~s(prod_year, name, bs="re") +
     I(price_l2/10) + 
     I(price_l3/10) + 
     I(price_l4/10) + 
-    I(price_l1/10):build_out +
+    I(price_l1/19):build_out +
     I(price_l2/10):build_out + 
     I(price_l3/10):build_out +
     I(price_l4/10):build_out,
@@ -169,8 +169,8 @@ texreg(results,
 #, weights=I(in_place_oil_mill_sm3/100)
 #year + I(year^2) +
 
-coef<-gam_mod_2d$coefficients[5:13]
-se<-summary(gam_mod_2d)$p.table[5:13,2]
+coef<-gam_mod_2d$coefficients[3:11]
+se<-summary(gam_mod_2d)$p.table[3:11,2]
 ymin<-coef-2*se
 ymax<-coef+2*se
 gam_2d_coef<-data.frame(coef=coef, se=se, lag=0:8, ymin=ymin, ymax=ymax, type="2d Smooth Model")
@@ -205,15 +205,17 @@ labs(x="Lags", y="Estimated Coefficient on Oil Price",
 
 coef_plot<-ggplot(model_coef) +
 geom_pointrange(aes(x=lag, y=coef, ymin = ymin, ymax = ymax)) +
-geom_hline(aes(y=0)) +
+geom_hline(aes(yintercept=0)) +
 facet_wrap(~type, nrow=2) +
 theme_bw() +
 labs(x="Lags", y="Estimated Coefficient on Oil Price")
 
-# png("/Users/johannesmauritzen/research/oil_prices/figures/price_coefficents.png",
-# width = 25, height =15, units = "cm", res=150, pointsize=12) 
-# print(coef_plot)
-# dev.off()
+coef_plot 
+
+png("/Users/johannesmauritzen/research/oil_prices/figures/price_coefficents.png",
+width = 25, height =15, units = "cm", res=150, pointsize=12) 
+print(coef_plot)
+dev.off()
 
 
 #Now create counterfactuals - 

### 2016-03-07 oil_prices_obes.tex
minor improvements)

@@
 Keywords: Oil production,Oil Prices, Norway, Generalized Additive Model.
-Words: 7000
+Words: 7100
 \end{abstract}
@@
 
-A large body of economic litterature exists on the complex chain of investment and production decisions made in the upstream oil and gas industry.
+A large body of economic literature exists on the complex chain of investment and production decisions made in the upstream oil and gas industry.
 This article looks at the limited, but important topic of the effect of price on production from active offshore fields, where a significant gap exists in the literature.
@@
 
-By looking only at the effect of price on fields that currently or previously have produced oil, I am limiting the scope of this article.
-The effect of oil prices on total production over an extended period of time is due not just to reactions in production in existing fields but also increased searching for new fields.
-In fact, an implication of this work is that much of the total production response from higher oil prices is likely from increased searching as well as production from previously un-economic fields.
+By looking only at the effect of price on production in fields that currently or previously have produced oil, I am limiting the scope of this article.
+The effect of oil prices on total production over the long term is due to both the reaction of production in existing fields as well as searching for new fields.
+This article is limited to the decisions oil companies can make to increase production in existing fields like making use of spare capacity or increasing production drilling.
+In fact, an implication of this work is that much of the total production response from higher oil prices is likely from increased searching as well as development of previously un-economic fields.
 
@@
 
-\section{Econometric Litterature on The Effect of Oil Price on Production}
+\section{Econometric Literature on The Effect of Oil Price on Production}
 
@@
 
-An extensive litterature exists that attempts to estimate the effect of oil prices on regional or global oil supply.
+An extensive literature exists that attempts to estimate the effect of oil prices on regional or global oil supply.
 Early econometric attempts, like \citet{adelman_fea_1975} often made the assumption that the price contained all relevant information, ignoring the role of geophysical limits at the regional level.
 \citet{hamilton_oil_1983} showed that macroeconomic events could not fully explain movements in oil prices, rejecting the notion that prices contained all relevant information to oil supply.
-\citet{kaufmann_oil_1991} presents a relatively early attempt at combining both economic and geophysical aspects in an emperical model of oil production in the US.
+\citet{kaufmann_oil_1991} presents a relatively early attempt at combining both economic and geophysical aspects in an empirical model of oil production in the US.
 The author used a Hbbert-like \citep{hubbert_energy_1962} shape-fitting model to capture regional geophysical dynamics, using the residuals to model the effects of price on production.
@@
 Instead Adelman argues that cartel behavior of low cost producers likely stood behind increasing oil prices.
-\citet{lynch_forecasting_2002} provides a fuller overview of these competing strands of litterature.
+\citet{lynch_forecasting_2002} provides a fuller overview of these competing strands of literature.
 
-The economics litterature on petroleum production tends to be overweighted towards studies of the US, but some global studies have also been attempted.
+The economics literature on petroleum production tends to be overweighted towards studies of the US, but some global studies have also been attempted.
 \citet{watkins_world_1998} attempt to estimate supply functions for 41 countries by estimating the relationship between the price of undeveloped reserves and reserve additions over time.
@@
 
-On the surface, this article fits well into the previous econometrics litterature on petroleum supply in that there is a focus on establishing and decomposing the effects of geophysical and economic forces.
+On the surface, this article fits well into the previous econometrics literature on petroleum supply in that there is a focus on establishing and decomposing the effects of geophysical and economic forces.
 However, most of the above articles focus on total long-term regional oil supply, which as I mention above, is the result of an extensive chain of decisions from surveying of a potential play, exploration drilling, to production.
@@
 
-Norwegian waters in the Barents Sea have also been open to exploration since the 1980s, however up until recently only a few, small finds were made and none came into commercial production.
+Norwegian waters in the Barents Sea have also been open to exploration since the 1980's, however up until recently only a few, small finds were made and none came into commercial production.
 Recently several significant oil and gas finds have been made in the Barents Sea - notably the oil field Goliat and the large gas find Sn\o hvit, which are both currently under development but not yet producing.
@@
 The central government also receives revenues through its majority ownership stake in Statoil.
-In addition, a wholly state-owned company, Petero, manages the development licenses that the state retains - referred to as the State's Direct Financial Interest (SDFI).
+In addition, a wholly state-owned company, Petoro, manages the development licenses that the state retains - referred to as the State's Direct Financial Interest (SDFI).
 
@@
 \section{A Generalized Additive Model of Oil Field Production}
-Parametric linear models have the sizeable advantages of simplicity and interpretability and are therefore usually a good starting point for an analysis.
+Parametric linear models have the sizable advantages of simplicity and interpretability and are therefore usually a good starting point for an analysis.
 However, when attempting to model the effect of price on oil field production, a standard linear model is unable to sufficiently control for the production profile and therefore heavily biases the estimate of the effect of price.
@@
 For convenience the estimated coefficients on the price terms from the two preferred models are shown in figure \ref{price_coefficients}.
-In general the results from the two models agree with each other, indicating little to no effect on the concurrent price term, but with positive coefficients estimated on the 2st to 4th lags, with a peak at the 2nd lag.
+In general the results from the two models agree with each other, indicating little to no effect on the concurrent price term, but with positive coefficients estimated on the 2nd to 4th lags, with a peak at the 2nd lag.
 A negative coefficient for the eighth lag is estimated in the 2-d model, however this is not significant in the model with random effects.
@@
 To attempt to capture this heterogeneity, I introduce an indicator variable that represents the build-out phase defined as the period between the start of production up to peak production.
-I add interaction effects between this indicator variable and the block of significant lagged price effects $price\_l1$ through $price\_l4$.
+I add interaction effects between this indicator variable and the block of lagged price effects $price\_l1$ through $price\_l4$.
 
@@
 Given the small sample used to estimate these terms, this is not surprising.
-On the other hand, if we compare the main effects of the lagged price terms to the standard random effects model, we notice that the point estimates on the 2nd, 3rd and 4th lags are now subsantially larger in magnitude.
+On the other hand, if we compare the main effects of the lagged price terms to the standard random effects model, we notice that the point estimates on the 2nd, 3rd and 4th lags are now substantially larger in magnitude.
 In this model, these main effect terms can be interpreted as the effect of price on production in the draw-down phase.
@@
 Since the evidence points to a stronger lagged effect of price on production in declining fields, this would also imply that policy measures meant to stimulate increased production would likely also be more effective in mature oil producing areas compared to newer areas.
-This is particularily important on the Norwegian Continental Shelf, as several new fields have recently come online following a period of high oil prices.
+This is particularly important on the Norwegian Continental Shelf, as several new fields have recently come on-line following a period of high oil prices.
 \footnote{I thank an anonymous reviewer for pointing out this implication.}

### 2016-03-07 rr_analysis_obes.r

@@ -31,7 +31,7 @@ large_fields<-prod_data[prod_data["name"]!="EKOFISK",]
 large_fields<-large_fields[!is.na(large_fields$oil_prod_mill_sm3),]
 
 
-gam_mod<-gam(oil_prod_mill_sm3~s(prod_year, bs="cr") +
+gam_mod<-gam(oil_prod_mill_sm3~s(prod_year) +
 	I(in_place_oil_mill_sm3/100) + I(cost_index/10) + 
     s(year, bs="cr", k=4) + I(price/10) + I(price_l1/10) + I(price_l2/10) +
     I(price_l3/10) + I(price_l4/10) + I(price_l5/10) + I(price_l6/10) + 
@@ -74,6 +74,8 @@ prod_smooth<-ggplot(smooth_pred1, aes(x=prod_year))+
     geom_line(aes(y=s_prod_year)) + 
     labs(x="Production Years", y="Smooth Function of Production Time") +
     theme_bw() 
+
+
 
 pdat2 <- with(large_fields,
              data.frame(prod_year = rep(mean(prod_year), 200),
@@ -133,7 +135,7 @@ summary(gam_mod_2d)
 #With random effects
 
 gam_mod_re<-gam(oil_prod_mill_sm3~s(prod_year, name, bs="re") + 
-    s(year, bs="cr", k=4) +
+    s(year, bs="cr", k=3) +
     I(in_place_oil_mill_sm3/100) + I(cost_index/10) +
     I(price/10) + I(price_l1/10) + I(price_l2/10) +
     I(price_l3/10) + I(price_l4/10) + I(price_l5/10) + I(price_l6/10) + 
@@ -142,7 +144,7 @@ gam_mod_re<-gam(oil_prod_mill_sm3~s(prod_year, name, bs="re") +
 summary(gam_mod_re)
 
 gam_mod_re_inter<-gam(oil_prod_mill_sm3~s(prod_year, name, bs="re") + 
- s(year, bs="cr", k=4) + I(cost_index/10) + 
+ s(year, bs="cr", k=3) + I(cost_index/10) + 
     I(in_place_oil_mill_sm3/100) +
     I(price/10) +
     I(price_l1/10) + 
@@ -183,7 +185,6 @@ labs(x="Lags", y="Estimated Coefficient on Oil Price",
     title="2-dim. Smooth Model")
 
 coef_plot_2d
-
 
 
 #weights=I(in_place_oil_mill_sm3/100) ,

### 2016-03-11 oil_prices_obes_blind.tex
final resubmission

--- 2016-03-07 oil_prices_obes.tex
@@
 \usepackage{comment} %for excluding figures
-\usepackage{booktabs}
+\usepackage{booktabs} %for pretty tables
 % \excludecomment{figure} % for printing without figures
@@
 
-\doublespacing % from package setspacs
+\doublespacing %from package setspacs
 
-\title{The Effect of Oil Price on Field Production: Evidence from the Norwegian Continental Shelf}
-\author{Johannes Mauritzen\\
-		Department of Economics\\
-        BI Norwegian Business School\\
-        Trondheim Campus\\
-		Trondheim, Norway\\
-        johannes.mauritzen@bi.no\\
-        \url{jmaurit.github.io}\\
-		}
+\title{The Effect of Oil Prices on Field Production: Evidence from the Norwegian Continental Shelf}
+% \author{Johannes Mauritzen\\
+		% Department of Economics\\
+  %       BI Norwegian Business School\\
+  %       Trondheim Campus\\
+		% Trondheim, Norway\\
+  %       johannes.mauritzen@bi.no\\
+  %       \url{jmaurit.github.io}\\
+		% }
 \date{\today}
@@
 I use detailed field-level data on Norwegian offshore oil production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production.
-I find no significant evidence of a concurrent reaction of field production to oil prices, though a modest lagged effect is found of the magnitude of approximately 2 to 7\% for a 10 dollar per barrel increase in the real price of oil.\\
-Keywords: Oil production,Oil Prices, Norway, Generalized Additive Model.
+I find no significant evidence of a concurrent reaction of field production to oil prices, though a modest lagged effect is found of the magnitude of approximately 2\% to 7\% for a \$10 per barrel increase in the real price of oil.\\
+Keywords: Oil production, Oil Prices, Norway, Generalized Additive Model.
 Words: 7100
@@
 
-For most of the last century, crude oil has been the world's single most important and valuable fuel source.\footnote{Oil is the world's largest single energy source, consisting of approximately 33\% of total energy consumption in 2013 as well as the most valuable in terms of market price per unit of energy \citep{british_petroleum_statistical_2013}} Naturally, questions of how the oil price affects the world economy as well as how oil production reacts to the oil price have been fundamental topics in Economics.
+For most of the past century, crude oil has been the world's single most important and valuable fuel source.\footnote{Oil is the world's largest single energy source, consisting of approximately 33\% of total energy consumption in 2013.
+Oil is also the most valuable in terms of market price per unit of energy \citep{british_petroleum_statistical_2013}.} Naturally, how the oil price affects the world economy as well as how oil production reacts to the oil price have been fundamental topics in economics.
 Recent volatility in the price of oil has made this an especially relevant topic.
 
-Oil and gas supply, especially in difficult and expensive offshore areas, is the product of an often lengthy and complex chain of conditional investment decisions.
-Starting with the exploration phase, where a potential oil-producing region or "play" is explored, leases are acquired, prospects are identified and exploration wells are drilled.
-In the case of a significant hydrocarbon find, a field is appraised to judge size, difficulty and in turn profitability.
+Oil and gas supply is the product of an often lengthy and complex chain of conditional investment decisions.
+Starting with the exploration phase, where a potential oil-producing region or ``play" is explored, leases are acquired, prospects are identified and exploration wells are drilled.
+In the case of a significant hydrocarbon find, a field is appraised to judge its size, difficulty of extraction and in turn profitability.
 For fields seen as sufficiently profitable, a development stage commences, where a detailed drilling and production plan is drawn up and submitted to obtain permits and regulatory approval.
@@
 Finally, at the end of this multi-year process, a build-out can commence, where full production capacity is built out as fast as possible, followed by an extended depletion phase.
-All of this consists of what is referred to as upstream activity, with steps such as shipping, refining and trading, among others, making up an equally complex down-stream process.
-\footnote{\citet{lefler_deepwater_2011} provides an accessible guide to the offshore petroleum exploration and production industry.}
+All of this consists of what is referred to as ``upstream" activity, with steps such as shipping, refining and trading, among others, making up an equally complex ``downstream" process.\footnote{\citet{lefler_deepwater_2011} provides an accessible guide to the offshore petroleum exploration and production industry.}
 
 A large body of economic literature exists on the complex chain of investment and production decisions made in the upstream oil and gas industry.
-This article looks at the limited, but important topic of the effect of price on production from active offshore fields, where a significant gap exists in the literature.
+This article looks at the important topic of the effect of oil prices on production from active offshore fields, where a significant gap exists in the literature.
 
-While numerous studies have taken up the issue of how the price of oil affects searching for new fields as well as total oil production at both the regional and global level, few studies exist of the effects of oil prices on oil production at the field level.
-The studies that do use field- and well-level data, such as \citet{rao_taxation_2010}, often use data from on-shore installations.
-However an increasing amount of the world's oil production comes from hard-to-reach offshore fields.
-Production from challenging offshore environments is substantially different in character than from on-shore installations.
+While numerous studies have taken up the issue of how the price of oil affects searching for new fields as well as total oil production at both the regional and global level, few studies exist on the effects of oil prices on oil production at the field level.
+The studies that do use field- and well-level data, such as \citet{rao_taxation_2010}, often use data from onshore installations.
+However, an increasing amount of the world's oil production comes from hard-to-reach offshore fields.
+Production from challenging offshore environments is substantially different in character than from onshore installations.
 
-The effect of oil prices on producing fields is an important topic in understanding the mechanisms of how total oil supply reacts in response to price.
-The response of oil production in existing fields is especially important now as many of the major oil-producing areas, like the Norwegian Continental Shelf are mature and an increasing share of total investment will be directed at capacity enhancement in producing fields rather than new field developments.
-Figure \ref{investment} shows that investments in producing fields makes up a large and growing share of total investment, nearly dwarfing investments in searching, exploration and new build-outs.
+The effect of oil prices on producing fields is an important topic in understanding the mechanisms of how total oil supply reacts to price.
+The response of oil production in existing fields is especially important now as many of the major oil-producing areas, like the Norwegian Continental Shelf, are mature and an increasing share of total investment will be directed at capacity enhancements in producing fields rather than new field developments.
+Figure \ref{investment} shows that investments in producing fields make up a large and growing share of total investment.
 How production from mature fields will respond to changes in oil prices has major implications for the oil industry, the state finances of oil-producing countries and long-run oil price formation.
@@
 	\includegraphics[width=.9\textwidth]{figures/investment.png}
-	\caption{The figure shows investment (millions NOK) in the oil and gas extraction industry on the Norwegian Continental Shelf.
+	\caption{Investment (millions of NOK) in the oil and gas extraction industry on the Norwegian Continental Shelf.
 A large and growing share of total investment goes into fields that are already producing as opposed to investment in search, exploration and build-out of new fields.}
@@
 \citet{krautkraemer_nonrenewable_1998} provides a good overview.
-At a basic level a central idea of much of this theoretical work is that with a non-renewable resource, production is a decision that involves a significant opportunity cost: more production in the current period means less production in future periods.
+At a basic level a central idea of much of this theoretical work is that with a non-renewable resource, production involves a significant opportunity cost: more production in the current period means less production in future periods.
 Within these frameworks, prices and expectations of prices become important variables in the production decision.
-A simple Hotelling model suggests that a producer would immediately change their production in response to a change in oil price in order to dynamically optimize the total extraction.
+A simple Hotelling model suggests that a producer would immediately change its production in response to a change in oil prices in order to dynamically optimize the total extraction.
 
 But in practice, the question is not as clear cut.
-Production on the Norwegian Continental Shelf - as well as most other offshore production areas - has extremely high fixed and operating costs.
-Keeping spare capacity available in order to adjust to changing oil prices might simply be too expensive.
-Instead, producers may find it more beneficial to use storage and financial instruments in order to hedge short-term price movements.
-Higher oil prices may still lead a producer to invest in increased capacity.
-Since lag times are significant in the offshore sector, we would then expect to see a multi-year lagged effect of prices on production.
+Production on the Norwegian Continental Shelf -- as in most other offshore production areas -- has extremely high fixed and operating costs.
+Keeping spare capacity available in order to adjust to changing oil prices may simply be too expensive.
+Producers may instead use storage and financial instruments in order to hedge short-term price movements.
+Higher oil prices may still lead a producer to invest in increased capacity and production drilling.
+However, since lag times are significant in the offshore sector, we would then expect to see a multi-year lagged effect of prices on production.
 
-However even with the question of lagged production and investment, some ambiguity exists.
-\citet{mohn_efforts_2008} suggests and finds evidence for the idea that in periods of high oil prices offshore producers will invest more in risky wild-cat drilling in search of new fields.
+However, even with the question of lagged production and investment, some ambiguity exists.
+\citet{mohn_efforts_2008} finds evidence for the idea that in periods of high oil prices offshore producers will invest more in risky wild-cat drilling in search of new fields.
 But when prices are low, they tend to concentrate investments in lower-risk ventures, like expanding production in existing fields.
 If this effect were to dominate, then it may even be plausible that a fall in prices leads to an \emph{increase} in production.
-Though the converse - an increase in prices leading to reduced production - is unlikely.
+Though the converse, an increase in prices leading to reduced production, is unlikely.
 
@@
 
-For such a prominent subject, the lack of research on the role of price in oil field production is striking.
-Two main factors likely contribute to the limited literature - the availability of data and the non-linear time profile of field production.
-Large private oil companies, notably the “super majors” and state-owned oil companies have historically accounted for the vast majority of oil production and reserves.
-\footnote{See for example the economist article titled Supermajordämmerung from August 3rd, 2013: \url{http://www.economist.com/news/briefing/21582522-day-huge-integrated-international-oil-company-drawing}} These entities tend to consider field-level data as either company or state secrets.
+For such a prominent subject, the lack of research on the role of prices in oil field production is striking.
+Two main factors likely contribute to the limited literature: The availability of data and the non-linear time profile of field production.
+Large private oil companies, notably the ``super majors" and state-owned oil companies, have historically accounted for the vast majority of oil production and reserves.\footnote{See for example the economist article titled Supermajordämmerung from August 3rd, 2013: \url{http://www.economist.com/news/briefing/21582522-day-huge-integrated-international-oil-company-drawing}} These entities tend to consider field-level data as either company or state secrets.
 
 However, over the last few years, a movement towards making the petroleum and other extractive industries more transparent has taken form.
-The Norwegian government has been on the forefront of this movement and increasingly committed itself to transparency in the petroleum sector.
-\footnote{See \url{http://www.regjeringen.no/en/sub/eiti---extractive-industries-tranparency/about-eiti.html?id=633586}} Over the last several years, detailed data on most aspects of the country's oil industry has become openly available.
+The Norwegian government has been on the forefront of this movement and has increasingly committed itself to transparency in the petroleum sector.\footnote{See \url{http://www.regjeringen.no/en/sub/eiti---extractive-industries-tranparency/about-eiti.html?id=633586}} Over the last several years, detailed data on most aspects of the country's oil industry have become openly available.
 
-In this article, I use historical production data from nearly all 77 currently or formerly oil-producing fields on the Norwegian Continental Shelf in order to estimate the effect that prices have on oil production.
+In this article, I use historical production data from nearly all 77 currently or formerly oil-producing fields on the Norwegian Continental Shelf in order to estimate the effect that oil prices have on oil production.
 
-By looking only at the effect of price on production in fields that currently or previously have produced oil, I am limiting the scope of this article.
-The effect of oil prices on total production over the long term is due to both the reaction of production in existing fields as well as searching for new fields.
-This article is limited to the decisions oil companies can make to increase production in existing fields like making use of spare capacity or increasing production drilling.
-In fact, an implication of this work is that much of the total production response from higher oil prices is likely from increased searching as well as development of previously un-economic fields.
+By looking only at the effect of oil prices on production in fields that currently or previously have produced oil, I am limiting the scope of this article.
+The effect of oil prices on total production over the long term is due not only to how production reacts in existing fields, but also to changes in search activity for new fields.
+This article is limited to the decisions oil companies make to increase production in existing fields like making use of spare capacity and increasing production drilling.
+In fact, an implication of this work is that much of the total production response to higher oil prices is likely from increased searching as well as from development of previously un-economic fields.
 
 The main finding in this article is that oil production at the field level has no significant reaction to concurrent changes in the oil price.
-A modest effect can be detected at a lag of between 2 and 4 years of between 2 to 7\% increase in yearly production for a 10 dollar increase in the price of oil.
+A modest effect can be detected at a lag of between 2 and 4 years of between 2\% to 7\% increase in yearly production for a \$10 increase in the price of oil.
 
@@
 
-The main methodological problem is the non-linear production profile of oil fields.
+The main methodological problem in the estimation is the non-linear production profile of oil fields.
 Once full-scale extraction is started in an oil field, pressure in wells will start declining.
 Compensating measures such as gas and water injection have been successful, but their impact is temporary.
-In turn production rates will therefore inevitably decline.
+Production rates will therefore inevitably decline.
 
-More so, major oil field discoveries, starting times and in turn production are correlated across fields as the top panel of figure \ref{oil_decline} shows with the production profile of the 10 largest Norwegian oil fields.
-The result is a total production curve that is bell-shaped over time as in the lower panel of figure \ref{oil_decline}.
+Moreover, major oil field discoveries, starting times, and in turn production are correlated across fields.
+The top panel of figure \ref{oil_decline} shows the production profile of the 10 largest Norwegian oil fields.
+The temporal correlation between the fields results in a total production curve that is bell-shaped over time as in the lower panel of figure \ref{oil_decline}.
 Since the oil price series is autocorrelated and non-stationary, failure to properly account for the production profile will lead to spurious estimation of the price terms in a regression.
@@
 The direction of this bias can be gleaned in figure \ref{oil_decline}.
-High oil prices were present at periods of relatively low production in the late 1970s and early 1980s as well as over the last 10 years, however real prices reached some of their lowest levels at the same time as the top of production around the year 2000.
-These oil price dynamics almost certainly affected investment decisions in the industry and in turn total production \citep{osmundsen_is_2007}, \citep{aune_financial_2010}.
-However the inverse contemporaneous relationship at the field level is entirely coincidental, but will heavily bias the estimation of the effect of price on production if field production profiles are not properly accounted for.
+High oil prices were present at periods of relatively low production in the late 1970s and early 1980s as well as over the last 10 years.
+However, real prices reached some of their lowest levels at the same time as the peak of production around the year 2000.
+These oil price dynamics almost certainly affected investment decisions in the industry and in turn total production \citep{osmundsen_is_2007,aune_financial_2010}.
+However, the inverse contemporaneous relationship at the field level is entirely coincidental.
+This will heavily bias the estimates of the effect of oil prices on production if field production profiles are not properly accounted for.
 
@@
 The results suggest that little to no concurrent effect of oil prices on production exists.
-However I estimate modest effects on the first through fourth lagged terms.
+However, I estimate modest effects on the second through fourth yearly lagged terms.
 These results are in line with an understanding of the industry as being highly capital intensive with high operating costs.
-Maintaining spare capacity in order to adjust to changing oil prices is likely prohibitively expensive.
-Instead, production is increased through investment, which may take up to several years to show an effect.
+Maintaining spare capacity in order to adjust to changing oil prices may be prohibitively expensive.
+Instead, production adjusts only with a lag, the result of a change in investment.
 
-\section{Econometric Literature on The Effect of Oil Price on Production}
+\section{Econometric Literature on the Effect of Oil Prices on Production}
 
@@
 An extensive literature exists that attempts to estimate the effect of oil prices on regional or global oil supply.
-Early econometric attempts, like \citet{adelman_fea_1975} often made the assumption that the price contained all relevant information, ignoring the role of geophysical limits at the regional level.
-\citet{hamilton_oil_1983} showed that macroeconomic events could not fully explain movements in oil prices, rejecting the notion that prices contained all relevant information to oil supply.
+Early econometric studies, such as \citet{adelman_fea_1975}, often made the assumption that the price contained all relevant information, ignoring the role of geophysical limits at the regional level.
+\citet{hamilton_oil_1983} showed that macroeconomic events could not fully explain movements in oil prices, rejecting the notion that prices contained all information relevant to oil supply.
 \citet{kaufmann_oil_1991} presents a relatively early attempt at combining both economic and geophysical aspects in an empirical model of oil production in the US.
-The author used a Hbbert-like \citep{hubbert_energy_1962} shape-fitting model to capture regional geophysical dynamics, using the residuals to model the effects of price on production.
-The author predicted that depletion and reduced oil supply from the lower 48 states would continue - a prediction that proved mostly accurate up until the technological revolution of fracking and shale oil extraction.
+The author used a Hubbert-like \citep{hubbert_energy_1962} shape-fitting model to capture regional geophysical dynamics, using the residuals to model the effects of price on production.
+The author predicted that depletion and reduced oil supply from the lower 48 states would continue -- a prediction that proved mostly accurate up until the technological revolution of fracking and shale oil extraction.
 
 A similar analysis combining econometric models with shape-fitting models by \citet{cleveland_forecasting_1991} also led to pessimistic forecasts of future US production.
-Further refinements of these were done by, among others, \citet{pesaran_forecasting_1995}  and \citet{moroney_integrated_1999}.
-\citet{kaufmann_oil_2001} present an extension and generalisation using a Vector Error Correction Model (VECM) that relaxes the assumption that firms produce their fields in order of decreasing quality and also includes institutional factors in US oil production.
+\citet{pesaran_forecasting_1995} and \citet{moroney_integrated_1999}, among others, made further refinements of these types of models.
+\citet{kaufmann_oil_2001} present an extension and generalization using a Vector Error Correction Model (VECM) that relaxes the assumption that firms produce their fields in order of decreasing quality and also includes institutional factors in US oil production.
 They claim that they are able to account for most of the price variation between 1938 and 1991.
 
-\citet{adelman_mineral_1990} and \citet{adelman_modelling_1993} takes an opposing view, using cointegration and causality testing to argue that scarcity, and in turn geophysical forces had little to do with the oil price movements of the 1970s and 1980s.
-Instead Adelman argues that cartel behavior of low cost producers likely stood behind increasing oil prices.
+\citet{adelman_mineral_1990, adelman_modelling_1993} takes an opposing view, using cointegration and causality testing to argue that scarcity, and in turn geophysical forces had little to do with the oil price movements of the 1970s and 1980s.
+Instead Adelman argues that the cartel behavior of low cost producers likely stood behind increasing oil prices.
 \citet{lynch_forecasting_2002} provides a fuller overview of these competing strands of literature.
@@
 The economics literature on petroleum production tends to be overweighted towards studies of the US, but some global studies have also been attempted.
-\citet{watkins_world_1998} attempt to estimate supply functions for 41 countries by estimating the relationship between the price of undeveloped reserves and reserve additions over time.
+\citet{watkins_world_1998} attempt to estimate the supply functions for 41 countries by modeling the relationship between the price of undeveloped reserves and reserve additions over time.
 Based on the aggregated estimates, they argued that fears of increasing global scarcity at the time were overblown.
-
 \citet{farzin_impact_2001} attempts to estimate the price elasticity of added global reserves and finds a small positive elasticity.
@@
 The author finds a negative price elasticity for several of the countries, and interprets this as evidence of producers targeting revenue.
-However since the author does not take into account the production profile of oil fields and the spurious correlation that can arise with autocorrelated prices, these estimates come under considerable doubt.
-\citet{ringlund_does_2008} estimates how rig activity reacts to changes in oil prices in non-Opec countries, and finds substantial heterogeneity based on the structure of a country's oil industry.
+However since the author does not take into account the production profile of oil fields and the spurious correlation that can arise with autocorrelated prices, these estimates come under some doubt.
+\citet{ringlund_does_2008} estimates how rig activity reacts to changes in oil prices in non-OPEC countries, and finds substantial heterogeneity based on the structure of a country's oil industry.
 Though in the long run, they find an average elasticity of close to one.
 
-On the surface, this article fits well into the previous econometrics literature on petroleum supply in that there is a focus on establishing and decomposing the effects of geophysical and economic forces.
-However, most of the above articles focus on total long-term regional oil supply, which as I mention above, is the result of an extensive chain of decisions from surveying of a potential play, exploration drilling, to production.
-This article, on the other hand, is considerably narrower in scope - focusing exclusively on production in active fields.
+This article fits well into the previous econometrics literature on petroleum supply in that there is a focus on establishing and decomposing the effects of geophysical and economic forces.
+However, most of the above articles focus on total long-term regional oil supply, which, as I mention above, is the result of an extensive chain of decisions from surveying of a potential play, exploration drilling, to production.
+This article, on the other hand, is considerably narrower in scope -- focusing exclusively on production in active fields.
 While ambiguity exists on whether geophysical forces and scarcity are significant factors that need to be controlled for in the long term regional perspective, they are without question dominant factors at the field level.
@@
 The author finds that oil exploration firms in Texas approximately respond as real-option theory would predict when it comes to the timing of drilling.
-A model and test using data from North Sea producers on the UK continental shelf by \citet{hurn_geology_1994}, on the other hand, fails to find evidence that the variance in the oil price affects the timing of oil field development.
-
+A model and test using data from North Sea producers on the UK continental shelf by \citet{hurn_geology_1994}, on the other hand, fails to find evidence that the variance in oil prices affects the timing of oil field development.
 I do not attempt to directly model uncertainty, however given that the investments needed to increase oil production in an existing field are to a certain extent irreversible and that oil prices are highly volatile, the results can and probably should be interpreted with the real options framework in mind.
@@
 \citet{mohn_exploration_2008} finds that long-term changes in the oil-price has a strong effect on exploratory drilling though little effect is measured from short-term changes in the oil price.
-\citet{osmundsen_exploration_2010} analyses drilling productivity over time on the Norwegian Continental Shelf while \citet{mohn_efforts_2008} finds that higher oil prices leads to higher reserves as well as that oil prices affect producer risk-preferences - with higher prices leading to lower success rates but larger discovery size.
+\citet{osmundsen_exploration_2010} analyses drilling productivity over time on the Norwegian Continental Shelf while \citet{mohn_efforts_2008} finds that higher oil prices leads to higher reserves as well as that oil prices affect producer risk-preferences -- with higher prices leading to lower success rates but larger discovery size.
 \citet{mohn_elastic_2010} presents a thorough econometric analysis of the role of economic and policy factors in oil exploration and reserve additions, as well as a useful summary of exploration and production on the Norwegian Continental Shelf.
 
-Most of the microeconometric articles analyzing the petroleum industry, like the ones mentioned above, focus on issues of exploration, drilling, and rig activity.
+Most of the microeconometric studies analyzing the petroleum industry, like the ones mentioned above, focus on issues of exploration, drilling, and rig activity.
 Microeconomic studies of production from operational fields are more sparse.
 \citet{black_is_1998} tests the relevance of “nesting” a structural empirical model of profit maximization that takes into account oil prices into a typical geo-engineering model of oil field production.
-They find strong evidence that taking into account profit maximization - and implicitly price - substantially improves the fit compared to a purely geo-engineering model.
-The major limitation of their methodology is that they are only able to test whether including economic factors like price affects the fit of the model but are not able to give an estimate of the magnitude of the effect.
+They find strong evidence that taking into account profit maximization -- and implicitly oil prices -- substantially improves the fit compared to a purely geo-engineering model.
+The major limitation of their methodology is that they are only able to test whether including economic factors like oil prices affects the fit of the model but are not able to give an estimate of the magnitude of the effect.
 Methodologically, the paper also relies heavily on strict assumptions about the functional form of both the geo-engineering and economic aspects of their model.
@@
 The significant differences in cost and complexity between operating offshore and onshore likely account for the different results.
-How producers react to a short-term tax change as opposed to a change in price also likely differs.
+How producers react to a short-term tax change as opposed to a change in the oil price also likely differs.
 
@@
 
-The first commercial oil well in Norwegian continental waters was discovered in December of 1969 in what became the Ekofisk oil field.
-Most of the largest fields in the North Sea were found relatively early on while more recent finds have tended to be smaller - a pattern typical of oil producing areas called creaming.
-A major exception to this trend was the recent find of the Johan Sverdrup field which is estimated to have approximately 300 million Standard Cubic Meters (SM3) of recoverable oil.
-\footnote{The Johan Sverdrup field is estimated to begin producing oil in 2019 and is not present in the data set used for the analysis.}
+The first commercial oil well on the Norwegian Continental Shelf was discovered in December of 1969 in what became the Ekofisk oil field.
+Most of the largest fields in the North Sea were found relatively early on while more recent finds have tended to be smaller -- a pattern typical of oil producing area called ``creaming''.
+A major exception to this trend was the recent find of the Johan Sverdrup field which is estimated to have approximately 300 million Standard Cubic Meters (SM3) of recoverable oil.\footnote{The Johan Sverdrup field is estimated to begin producing oil in 2019 and is not present in the data set used for the analysis.}
 
-Exploration in the Norwegian Sea was opened in the early 1980’s and the first commercial field started production in 1981.
+Exploration in the Norwegian Sea was opened in the early 1980s and the first commercial field started production in 1981.
 While several mid-sized fields have been discovered, the Norwegian Sea has generally disappointed in terms of commercial oil finds.
 
-Norwegian waters in the Barents Sea have also been open to exploration since the 1980's, however up until recently only a few, small finds were made and none came into commercial production.
-Recently several significant oil and gas finds have been made in the Barents Sea - notably the oil field Goliat and the large gas find Sn\o hvit, which are both currently under development but not yet producing.
+Norwegian waters in the Barents Sea have also been open to exploration since the 1980s.
+However, up until recently only a few, small finds were made and none came into commercial production.
+Recently, several significant oil and gas finds have been made in the Barents Sea -- notably the oil field Goliat and the large gas find Sn\o hvit, which are both currently under development but not yet producing.
 The agreement between Norway and Russia in June 2011 settling a long-running dispute over the maritime delimitation has also given a boost to new exploration in the region.
 
-Profits from oil and gas production in Norway are subject to a resource tax of 51 \% on top of the ordinary income tax of 27 \%, thus income from petroleum production is taxed at a total marginal rate of 78 \%.
-The central government also receives revenues through its majority ownership stake in Statoil.
-In addition, a wholly state-owned company, Petoro, manages the development licenses that the state retains - referred to as the State's Direct Financial Interest (SDFI).
+Profits from oil and gas production in Norway are subject to a resource tax of 51\% on top of the ordinary income tax of 27\%, thus income from petroleum production is taxed at a total marginal rate of 78\%.
+The State also receives revenues through its majority ownership stake in Statoil.
+In addition, a wholly state-owned company, Petoro, manages the development licenses that the State retains -- referred to as the ``State's Direct Financial Interest (SDFI).''
 
-The tax system is also designed to be neutral, that is a field that is profitable before tax should also be profitable after tax.
+The tax system is also designed to be neutral.
+That is, a field that is profitable before tax should also be profitable after tax.
 This is accomplished by only taxing the net-profits of oil companies, allowing companies to deduct investment costs and deduct and carry forward losses.\footnote{The Norwegian Oil and Gas industry group Norsk Petroleum provides a fuller overview: \url{http://www.norskpetroleum.no/en/framework/petroleum-tax/}}
@@
 In general though, these rules mainly affected searching and discoveries of new fields rather than production from existing fields and I do not control for the tax changes in my model.
-More information on taxation and revenues from the offshore sector can be found at the website of the Norwegian Ministry of Finance.
-\footnote{\url{http://www.regjeringen.no/en/dep/fin/Selected-topics/taxes-and-duties/bedriftsbeskatning/Taxation-of-petroleum-activities.html?id=417318}}
+More information on taxation and revenues from the offshore sector can be found at the website of the Norwegian Ministry of Finance.\footnote{\url{http://www.regjeringen.no/en/dep/fin/Selected-topics/taxes-and-duties/bedriftsbeskatning/Taxation-of-petroleum-activities.html?id=417318}}
 
@@
 In general, the frameworks are fair, predictable and stable for companies who find commercially extractable oil deposits and regulatory interference is unlikely to be the cause of any observed changes in production from existing oil fields.
-For more information see the website of the Norwegian Petroleum Directorate.
-\footnote{\url{http://www.npd.no/en/Topics/Production-licences/Theme-articles/Production-licence--licence-to-explore-discover-and-produce-/}}
+For more information see the website of the Norwegian Petroleum Directorate.\footnote{\url{http://www.npd.no/en/Topics/Production-licences/Theme-articles/Production-licence--licence-to-explore-discover-and-produce-/}}
 
@@
 
-To simplify the analysis, I eliminate the smallest fields - those with less than 5 million SM3 of recoverable reserves.
+To simplify the analysis, I eliminate the smallest fields -- those with less than 5 million SM3 of recoverable reserves.
 These fields tend to have substantially different production profiles than the larger fields.
-However, these fields only account for around 0.5 \% of total recoverable reserves.
+However, these fields only account for around 0.5\% of total recoverable reserves.
 
@@
 Ekofisk was the first field to be found and the first to start production on the Norwegian Continental Shelf.
-It reached an initial peak in production in the late 1970's and began depleting.
-However, the introduction of water and gas injection technology led to a boost in production and a new peak in the early 2000's.
+It reached an initial peak in production in the late 1970s and began depleting.
+However, the introduction of water and gas injection technology led to a boost in production and a new peak in the early 2000s.
 All subsequent major fields used gas and water injection from the outset and have a single-hump profile.
@@
 Table \ref{field_summary} provides summary statistics over the distribution of field-level variables.
-Notably, the median field in the data set had been producing for 16 years as of the beginning of 2015 - an indication of the mature status of the Norwegian Continental Shelf.
+Notably, the median field in the data set had been producing for 16 years as of the beginning of 2015 -- an indication of the mature status of the Norwegian Continental Shelf.
 
@@
 Production costs are known to covary with the price of oil.
-To control for this, I include the producer price index for the Norwegian oil and gas extraction industry, obtained from Statistics Norway.
-\footnote{\url{https://www.ssb.no/en/priser-og-prisindekser/statistikker/ppi}} Unfortunately, this series only goes back to 2000, so I also include an index of oil and gas well costs from the US energy information agency for the years 1960 to 2000.
+To control for this, I include the producer price index for the Norwegian oil and gas extraction industry, obtained from Statistics Norway.\footnote{\url{https://www.ssb.no/en/priser-og-prisindekser/statistikker/ppi}} Unfortunately, this series only goes back to 2000, so I also include an index of oil and gas well costs from the US energy information agency for the years 1960 to 2000.
 \footnote{\url{http://www.eia.gov/dnav/pet/pet_crd_wellcost_s1_a.htm}} While this measure is unlikely to be a perfect substitute for costs in Norway, oil and gas extraction is a global industry, where costs are highly correlated across regions.
@@
 Instead of attempting to estimate the shape of the production profiles of the fields by estimating parameters on linear terms, I estimate a non-parametric function for the production profile.
-At a conceptual level, this model is decomposing the variation in the data into a smoothed component representing the technical and physical forces that produce the typical bell-shaped production profile, and the remaining component, that I estimate the effect of price with.
+At a conceptual level, this model is decomposing the variation in the data into a smoothed component representing the technical and physical forces that produce the typical bell-shaped production profile, and the remaining component, that I use to estimate the effect of price.
 
@@
 
-Finally, $\mathbf{oil\_price\_lags_t}$ represents the variables of interest - a vector of concurrent and lagged terms for the oil price.
+Finally, $\mathbf{oil\_price\_lags_t}$ represents the variables of interest -- a vector of concurrent and lagged terms for oil prices.
 I include eight lagged terms of the oil price and I also experimented with using quadratic terms, however these could not be shown to be significant so I leave them out.
@@
 
-The latter term - an estimate of the second derivative of the function - acts as a penalty for the ``wiggliness'' of the function.
+The latter term -- an estimate of the second derivative of the function -- acts as a penalty for the ``wiggliness'' of the function.
 The total smoothness can then be controlled by adjusting the $\lambda$ term.
@@
 In the second panel, the smoothed function of calendar time, $year_t$, is shown.
-Here a cubic regression spline is also used, however the number of knots is restricted to four in order to avoid over-fitting - capturing gradual trends over time like technological change, but not the variation caused by volatile oil prices.
+Here a cubic regression spline is also used, however the number of knots is restricted to four in order to avoid over-fitting -- capturing gradual trends over time like technological change, but not the variation caused by volatile oil prices.
 The function increases strongly in the early period, likely reflecting the effects of learning and technological change in a new industry in Norwegian waters, while flattening out in the latter years.
@@
 
-Another potential issue may be that the smoothed function of production time may not be sufficient in controlling for the variation in production profiles shapes across fields.
+Another potential issue may be that the smoothed function of production time may not be sufficient for controlling for the variation in production profiles across fields.
 In particular, production profiles differ substantially by the size of the field.
-A solution is to allow the shape to vary by the size of the field - here I use an estimate of in-place oil in the field rather than recoverable reserves, as recoverable reserves may be correlated with oil prices.
+A solution is to allow the shape to vary by the size of the field -- here I use an estimate of in-place oil in the field rather than recoverable reserves, as recoverable reserves may be correlated with oil prices.
 Thus the smoothed term can be written as a function of both production time and in place oil: $f(production\_time_{i,t}, in\_place\_oil_{i})$.
@@
 
-In general I assume that oil prices are exogenous in this model.
+In general, I assume that oil prices are exogenous in this model.
 On the supply side, oil is a globally traded commodity.
@@
 
-\section{The Effect of Oil Price on Field Production}
+\section{The Effect of Oil Prices on Field Production}
 
@@
 The variables of interest in this article is the oil price and its lags, which I include as eight linear parametric terms in the model.
-The reason for including both a concurrent oil price term as well as eight lags is that a change in price could conceivably have two effects on oil production in a field.
-First, the field operator could be operating on the basis of some short-term extraction rule - choosing to pump out less at times of lower prices in order to pump out more at periods of high prices.
+The reason for including both a concurrent oil price term as well as eight lags is that a change in the price could conceivably have two effects on oil production in a field.
+First, the field operator could be operating on the basis of some short-term extraction rule -- choosing to pump out less at times of lower prices in order to pump out more at times of high prices.
 
 Alternatively, a change in the price of oil can be seen as a lifting of a production constraint.
-A higher oil price means that added investments in production becomes attractive in order to either increase the total amount extracted from a field or to shift production forward.
-However investments in the offshore sector can be complex and lengthy, and any production lift would be expected to happen with a lag.
+A higher oil price means that added investments in production become attractive in order to either increase the total amount extracted from a field or to shift production forward.
+However, investments in the offshore sector can be complex and lengthy, and any production lift would be expected to happen with a lag.
 As mentioned earlier, including several lags also allows for the possibility of adaptive expectations of future oil prices.
@@
 
-For convenience the estimated coefficients on the price terms from the two preferred models are shown in figure \ref{price_coefficients}.
+For convenience, the estimated coefficients on the price terms from the two preferred models are shown in figure \ref{price_coefficients}.
 In general the results from the two models agree with each other, indicating little to no effect on the concurrent price term, but with positive coefficients estimated on the 2nd to 4th lags, with a peak at the 2nd lag.
@@
 At this stage, it is ex-ante questionable that prices would have an effect on production, as a decision to increase capacity would require a new submission to the regulators and a potentially lengthy and expensive delay.
-On the other hand, the draw down phase is characterized by overcapacity and in turn greater flexibility in increasing production by way of, for example, infill drilling that can both accelerate recovery and increase total recovery in a field.
-\footnote{I thank an anonymous referee for making this point.}
+On the other hand, the depletion phase is characterized by overcapacity and in turn greater flexibility in increasing production by way of, for example, infill drilling that can both accelerate recovery and increase total recovery in a field.\footnote{I thank an anonymous referee for making this point.}
 
@@
 The main problem with introducing this heterogeneity in the model is that most field build-outs happen quickly, usually within one or two years in most small to medium sized fields.
-The sample to estimate the build out price terms is then relatively small - less than one third of the total observations.
+The sample to estimate the build out price terms is then relatively small -- less than one third of the total observations.
 The small sub-sample leads to estimation problems when using the model with a 2-dimensional smooth term, so I instead extend the random effects model.
@@
 	\caption{Results from a Monte-Carlo experiment.
-I generate data mimicking oil price production from a set of fields, as displayed in the top frame.
-Using GLM and GAM models, and regenerating the error component in the data, I replicate estimates of the effect of price on production, which is set at $\beta = 0$ and $\beta=0.05$.
+I generate data mimicking production from a set of fields, as displayed in the top frame.
+Using GLM and GAM models, and regenerating the error component in the data, I replicate estimates of the effect of prices on production, which is set at $\beta = 0$ and $\beta=0.05$.
 The density of the results for the GLM model are shown in the lower left panel, where a clear and large bias is shown relative to the true $\beta$ which are represented by the vertical lines.
@@
 The main result of this research is to show that production in existing Norwegian offshore fields has no significant concurrent reaction to higher oil prices while a modest effect is estimated  with a lag of between 2 and 4 years.
-Oil producers do not appear to be behaving strategically in relation to short-term production - increasing or reducing production in response to changes in oil price.
+Oil producers do not appear to be behaving strategically in relation to short-term production -- increasing or reducing production in response to changes in the oil price.
 Instead they are likely using storage or financial instruments to hedge short-term price movements.
-Changes in oil prices can rather be seen as a relaxing of a production constraint, justifying increased investment that leads to either a higher total extraction rate or an inter-temporal shifting of production.
+Changes in oil prices can rather be seen as a relaxation of a production constraint, justifying increased investment that leads to either a higher total extraction rate or an inter-temporal shifting of production.
 
@@
 Unfortunately, with the existing data and methods, I do not see a good way of distinguishing between these alternatives.
-First, there is the problem of censored data, as most of the fields in the data set are still producing, thus the information on how long the fields eventually produce are not available.
-More so, the estimates of both recoverable reserves and in-place oil are too uncertain in order to to make calculations of whether a change in production in one period led to a reduction in another - especially when the changes in production are relatively modest.
+First, there is the problem of truncated data, as most of the fields in the data set are still producing, thus the information on how long the fields eventually produce are not available.
+More so, the estimates of both recoverable reserves and in-place oil are too uncertain in order to make calculations of whether a change in production in one period led to a reduction in another -- especially when the changes in production are relatively modest.
 
@@
 As noted, evidence has been found for adaptive expectations of future oil prices.
-Thus, the lagged effects could be interpreted as both the result of investment lags as well as lags involved in updating of expectations.
+Thus, the lagged effects could be interpreted as both the result of investment lags as well as lags involved in the updating of expectations.
 Completely disentangling these effects would probably at a minimum require a structural model with some specific assumptions about the formation of expectations, which is beyond the scope of this article.
@@
 Since the current price always contains the most up-to-date information, then a model of adaptive expectations would show the largest effect at the concurrent term, thereafter receding.
-The estimations in this article however can show no significant concurrent effect, with the largest magnitude effect consistently being estimated at a 2-year lag.
+The estimations in this article however can show no significant concurrent effect, with the largest-in-magnitude effect being estimated at a 2-year lag.
 
-The findings in this article can provide some guidance on predicting the effects of the recent steep fall in the oil price on Norwegian production, as well as production from other offshore production areas.
+The findings in this article can provide some guidance on predicting the effects of the recent steep fall in oil prices on Norwegian production, as well as production from other offshore production areas.
 The results suggest that lower oil prices are likely to have only modest ramifications for production in existing fields.
@@
 Since the evidence points to a stronger lagged effect of price on production in declining fields, this would also imply that policy measures meant to stimulate increased production would likely also be more effective in mature oil producing areas compared to newer areas.
-This is particularly important on the Norwegian Continental Shelf, as several new fields have recently come on-line following a period of high oil prices.
-\footnote{I thank an anonymous reviewer for pointing out this implication.}
+This is particularly important on the Norwegian Continental Shelf, as several new fields have recently come on-line following a period of high oil prices.\footnote{I thank an anonymous reviewer for pointing out this implication.}
 
@@
 
-\section{Acknowledgments}
-I thank Klaus Mohn, Jonas Andersson, Sturla Kvamsdal, Harrison Fell, R\"ognvaldur Hannesson and Henrik Horn for valuable discussion.
-I have received financial support from the Norwegian Center for Sustainable Energy Studies (CenSES), Norwegian Research Council grant number 209697
+% \section{Acknowledgments}
+% I thank participants at the University of Stavanger Business School seminar and University of Aberdeen Business School Workshop on Petroleum Economics for helpful comments and suggestions.
+Jonas Andersson provided valuable discussion and suggestions on the Monte-Carlo study.
+Nicholas Sheard provided valuable proof reading help.
+I have received financial support from the Norwegian Center for Sustainable Energy Studies (CenSES), Norwegian Research Council grant number 209697.
 

### 2016-03-11 oil_prices_title.tex
final resubmission

@@
 
-\title{The Effect of Oil Price on Field Production: Evidence from the Norwegian Continental Shelf}
+\title{The Effect of Oil Prices on Field Production: Evidence from the Norwegian Continental Shelf}
 \author{Johannes Mauritzen\\
+        BI Norwegian Business School\\
         Department of Economics\\
-        BI Norwegian Business School\\
+        Trondheim Campus \\
         Trondheim, Norway\\
-        johannes.mauritzen@iot.ntnu.no\\
+        johannes.mauritzen@bi.no\\
         \url{jmaurit.github.io}\\
@@
 I use detailed field-level data on Norwegian offshore oil production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production.
-I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 2 to 5\% for a 10 dollar per barrel increase in the real price of oil.\\
-\textbf{Keywords}: Oil production,Oil Prices, Norway, Generalized Additive Model.\\
-\textbf{Words}: 5970
+I find no significant evidence of a concurrent reaction of field production to oil prices, though a modest lagged effect is found of the magnitude of approximately 2\% to 7\% for a \$10 dollar per barrel increase in the real price of oil.\\
+Keywords: Oil production, Oil Prices, Norway, Generalized Additive Model.
+Words: 7100
 \end{abstract}

### 2016-03-11 rr_analysis_obes.r

@@ -26,10 +26,9 @@ small_rec_oil<-sum(field_data["recoverable_oil"][field_data["recoverable_oil"]<=
 
 prod_data<-read.csv('/Users/johannesmauritzen/research/oil_prices/data/prod_data.csv')
 
-large_fields<-prod_data[prod_data["recoverable_oil"]>5,]
+large_fields<-prod_data[prod_data["recoverable_oil"]>3,]
 large_fields<-prod_data[prod_data["name"]!="EKOFISK",]
 large_fields<-large_fields[!is.na(large_fields$oil_prod_mill_sm3),]
-
 
 gam_mod<-gam(oil_prod_mill_sm3~s(prod_year) +
 	I(in_place_oil_mill_sm3/100) + I(cost_index/10) + 

### 2016-04-05 oil_prices_obes.tex
added word document version

@@
 
-\doublespacing % from package setspacs
+\doublespacing %from package setspacs
 
-\title{The Effect of Oil Price on Field Production: Evidence from the Norwegian Continental Shelf}
+\title{The Effect of Oil Prices on Field Production: Evidence from the Norwegian Continental Shelf}
 \author{Johannes Mauritzen\\
@@
 
-In short, a function of $x_1$ and $x_2$ is found that is minimizing errors in the sense of minimizing Euclidean distance subject to a penalty function of “wiggiliness”.
+In short, a function of $x_1$ and $x_2$ is found that is minimizing errors in the sense of minimizing Euclidean distance subject to a penalty function of ``wiggiliness''.
 The actual implementation is somewhat more involved in order to increase the computational efficiency of the estimation.
@@
 
-A more subtle possibility for endogeneity exists on the demand side.
+A subtler possibility for endogeneity exists on the demand side.
 A bias could result if demand also was correlated with changes in field production other than through the direct price-mechanism.
@@
 One potential weakness of the estimation so far is that I do not take into account the different stages of field production, where the effect of price could be expected to differ.
+
 In particular, in the build-out phase of a field, where production is started and then ramped up to full capacity, the field operator has an incentive to build out the field as quickly as possible as large amounts of capital have at that point been allocated and it is important to get oil production and in turn income flowing.
 At this stage, it is ex-ante questionable that prices would have an effect on production, as a decision to increase capacity would require a new submission to the regulators and a potentially lengthy and expensive delay.
-On the other hand, the draw down phase is characterized by overcapacity and in turn greater flexibility in increasing production by way of, for example, infill drilling that can both accelerate recovery and increase total recovery in a field.
-\footnote{I thank an anonymous referee for making this point.}
+On the other hand, the depletion phase is characterized by overcapacity and in turn greater flexibility in increasing production by way of, for example, infill drilling that can both accelerate recovery and increase total recovery in a field.\footnote{I thank an anonymous referee for making this point.}
 
@@
 \section{Acknowledgments}
-I thank Klaus Mohn, Jonas Andersson, Sturla Kvamsdal, Harrison Fell, R\"ognvaldur Hannesson and Henrik Horn for valuable discussion.
-I have received financial support from the Norwegian Center for Sustainable Energy Studies (CenSES), Norwegian Research Council grant number 209697
+I thank participants at the University of Stavanger Business School seminar and University of Aberdeen Business School Workshop on Petroleum Economics for helpful comments and suggestions.
+Jonas Andersson provided valuable discussion and suggestions on the Monte-Carlo study.
+Nicholas Sheard provided valuable proof reading help.
+I have received financial support from the Norwegian Center for Sustainable Energy Studies (CenSES), Norwegian Research Council grant number 209697.
 
